---
title: "Assumptions and validation - wheat RNA/protein kinetics"
author: "Kristina Gagalova"
format:
  gfm:
    toc: true
    code-tools: true
    fig-dir: "figs"
execute:
  echo: true
editor: visual
bibliography: references.bib
---

## Purpose

`de_proteomics_wheat.qmd` states a set of assumptions (A1--A3, B1--B8) but only tests three of them (A3, B6, B7). This companion notebook **tests every remaining assumption that is testable**, states which ones are violated and by how much, and grounds each modelling choice in the published literature.

Each section follows the same structure:

1.  **The assumption** --- stated precisely.
2.  **Literature** --- who else makes it, and what they found.
3.  **The test** --- runnable code.
4.  **Verdict** --- holds / violated / untestable, with the number.

A summary table of all verdicts is in §11.

## Executive summary --- what the tests found

> **This notebook now runs on the validated RNA DE.** `prepare_variety()` reads
> the RNA treatment contrasts from the project's standalone limma workflow
> rather than re-deriving them (see `R/load_validated_de.R`). That resolves what
> §2b previously recorded as an unresolved 2--6× disagreement. **Every number
> below that was computed under the superseded DE is marked ⚠ and must be
> re-checked against this render** --- the prose was written before the change
> and has not been recomputed by hand.

Five assumptions are violated and three hold; the reconciliation that was
previously unresolved is now settled. In order of how much they should change
what gets claimed:

1.  **B4 --- RNA treated as known (§7). Severely violated, and this is the headline.** `se_rna` is accepted by the model and never used. Propagating it by parametric bootstrap flips the regulated/not-regulated call for ⚠ **36% of Cadenza genes and 49% of Norin genes**. That uncertainty is entirely absent from the reported credible intervals. *Per-gene regulation calls from this model should not be published without propagating RNA error or validating orthogonally.* Aggregate statements remain sound.

    ⚠ The two percentages are pre-change. The direction of this finding is
    structural and will not reverse --- the standalone workflow's SEs are
    recovered as \|logFC/t\| and are just as absent from the fitter --- but the
    magnitudes will move, because the SEs themselves differ. Re-read §7's table.

2.  **A1 --- first-order kinetics (§3). Violated.** Residuals are systematically structured by timepoint, in *opposite directions* in the two varieties (Cadenza's response peaks then falls; Norin's is delayed). Some genes will be flagged "regulated" because the trajectory *shape* is wrong, not because the amplitude departs from 1.

3.  **Identifiability (§5). Worse than previously stated.** The "23--27% resolvable" figure counted fits pinned at the 12 h grid floor as identifiable. A fit at the floor means "≤ 12 h" --- it is censored. Cross-tabulating the `identifiable` flag against where each fit actually landed (from the main notebook's exported results, under the validated DE):

    | | Cadenza (n = 5,580) | Norin (n = 5,159) |
    |:--|--:|--:|
    | Not identifiable | 2,732 (49.0%) | 2,651 (51.4%) |
    | Identifiable, pinned at the 12 h floor | 1,304 (23.4%) | 908 (17.6%) |
    | **Identifiable and inside the band** | **392 (7.0%)** | **643 (12.5%)** |
    | Above the 72 h bound | 2,376 (42.6%) | 1,952 (37.8%) |

    Only **7% of Cadenza and 13% of Norin** genes have a half-life this design resolves. The fitted `t_half` distribution is a property of the 0/24/48/72 h sampling window, not of the wheat proteome.

4.  **B5 --- gene independence (§8). Violated.** Residual PC1 absorbs ⚠ 62--69% of variance against 25% expected. The horseshoe treats correlated genes as independent evidence, so the model is more confident than it should be.

5.  **A0 --- baseline equivalence (§2). Severely asymmetric --- far worse than previously reported.** Under the validated DE, Cadenza is clean (**0.72%** DE at 0 hpi, 432/59,931) but **Norin is at 17.6%** (10,499/59,502) --- a **24×** difference at a timepoint where infection cannot yet have acted. The previously reported figures were 0.4% and 5.8%. This also undercuts the justification for t0-centring in Norin: `Rl - Rl[,1]` is defended as removing a batch artefact, but at 17.6% the Norin baseline contrast carries substantial real signal.

6.  **Reconciliation (§2b). RESOLVED.** The RNA treatment contrasts are now read from the standalone workflow rather than re-derived, so the 2--6× disagreement no longer exists: there is one RNA DE result in the project. §2b is retained below as a record of the investigation, not as an open problem.

Holding up: abundance-dependent missingness (§10, Spearman −0.74), the horseshoe prior scale (§B6), and MCMC convergence after the adaptive-step fix (§B7).

Newly flagged, not previously listed: **the LRT null is not calibrated for this fitting procedure** --- the χ² approximation ignores grid-based half-life selection, estimated RNA, correlated timepoint contrasts, and estimated protein variances. A parametric bootstrap repeating the full procedure is needed before any p-value here is quoted as calibrated. And **abundance-dependent missingness is not the same claim as "every completely-missing cell is MNAR"**; §10 supports the former, not the latter.

> **The overall picture.** The *aggregate* findings --- the distribution of amplitudes, the existence of post-transcriptional regulation, comparisons between large gene sets --- are supported. **Individual per-gene calls are not**, for the reasons in points 1--3. Any target list coming out of this analysis should be treated as a hypothesis-generating shortlist requiring orthogonal validation, not a result.

------------------------------------------------------------------------

## 1. Setup

```{r}
#| label: setup
#| message: false

if (!requireNamespace("here", quietly = TRUE))
  stop("package 'here' is required")

## renv activation: see the note in de_proteomics_wheat.qmd -- RENV_PROJECT
## must be set before sourcing activate.R, or renv treats THIS folder as the
## project and bootstraps an empty library.
if (file.exists(here::here("renv", "activate.R"))) {
  Sys.setenv(RENV_PROJECT = here::here())
  source(here::here("renv", "activate.R"))
}

source(here::here("R", "utils.R"))
source(here::here("R", "wheat_pipeline.R"))
source(here::here("R", "07_bayesian_kinetics.R"))

need_pkgs(c("limma", "edgeR", "DESeq2", "matrixStats", "impute"))

set.seed(20260823)
DESIGN <- wheat_design()

cad <- prepare_variety("cadenza", DESIGN)
nor <- prepare_variety("norin",   DESIGN)

VARIETIES <- list(Cadenza = cad, Norin = nor)
```

```{r}
#| label: fn-fast-amplitude

#' Fast profile-likelihood amplitude fit (no MCMC).
#'
#' Identical model to M0/M1 in the main notebook -- grid over half-lives with
#' the amplitude profiled out in closed form -- but returns only the point
#' estimates. Used throughout this notebook for sensitivity analyses that
#' need dozens of refits, where running the full sampler each time would be
#' wasteful and would confound sampler noise with the effect being tested.
fit_amplitude <- function(Rl, Pl, Ps, tps, n_grid = 60) {

  grid <- seq(0, max(tps), by = 0.25)
  gi   <- match(tps, grid)

  r_lin <- 2^t(apply(Rl, 1, function(v) approx(tps, v, grid, rule = 2)$y))

  th_grid <- exp(seq(log(0.5 * diff(tps)[1]), log(6 * max(tps)),
                     length.out = n_grid))
  kd_grid <- log(2) / th_grid

  W   <- 1 / pmax(Ps, 0.05)^2
  SST <- rowSums(W * Pl^2)
  n   <- nrow(Pl)

  ssr0 <- ssr1 <- ahat <- matrix(NA_real_, n, n_grid)
  best_pred <- vector("list", n_grid)

  for (j in seq_len(n_grid)) {
    kd <- kd_grid[j]
    pl <- integrate_protein_mat(grid, r_lin, ks = kd, kd = kd, P0 = rep(1, n))
    pl <- log2(pmax(pl[, gi, drop = FALSE], 1e-9))
    best_pred[[j]] <- pl

    ssr0[, j] <- rowSums(W * (pl - Pl)^2)
    num <- rowSums(W * pl * Pl)
    den <- rowSums(W * pl^2)
    ahat[, j] <- ifelse(den > 1e-12, num / den, NA_real_)
    ssr1[, j] <- SST - ifelse(den > 1e-12, num^2 / den, 0)
  }

  j0 <- max.col(-ssr0, "first")
  j1 <- max.col(-ssr1, "first")
  ix <- function(j) cbind(seq_len(n), j)

  lrt <- pmax(ssr0[ix(j0)] - ssr1[ix(j1)], 0)
  fdr <- p.adjust(pchisq(lrt, 1, lower.tail = FALSE), "BH")

  pred <- t(vapply(seq_len(n), function(i) best_pred[[j0[i]]][i, ],
                   numeric(length(tps))))

  list(
    a         = ahat[ix(j1)],
    t_half_M0 = th_grid[j0],
    t_half_M1 = th_grid[j1],
    lrt_fdr   = fdr,
    pred      = pred,
    resid     = Pl - pred,
    at_bound  = th_grid[j0] >= max(tps)
  )
}

#' The RNA-responsive gate used throughout (matches the main notebook).
rna_gate <- function(v, min_abs_lfc = 0.5, fdr = 0.05) {
  (v$de_rna$any_fdr[rownames(v$Rl)] < fdr) &
    (apply(abs(v$Rl[, -1, drop = FALSE]), 1, max) > min_abs_lfc)
}
```

------------------------------------------------------------------------

## 2. A0 --- Baseline equivalence at 0 hpi

### The assumption

Everything downstream is **centred on t0** (`Rl - Rl[,1]`), which treats any Infected-vs-Control difference at 0 hpi as a nuisance offset to be removed. That is only legitimate if the two arms are genuinely equivalent at baseline: at 0 hpi the inoculation has only just happened and no transcriptional response is biologically possible yet.

If the arms differ substantially at t0, centring does not fix the problem --- it *hides* it, because the difference is then subtracted from every later timepoint.

### Literature

Time-course infection studies in wheat routinely sample a 0 hpi / mock baseline and treat it as the reference against which later timepoints are contrasted ([Fusarium time-course eQTL atlas](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5633761/); [Zymoseptoria infection cycle](https://academic.oup.com/plphys/article/167/3/1158/6113592)). The validity of that reference is an assumption those designs share and rarely test explicitly.

### The test

```{r}
#| label: a0-baseline

baseline_check <- function(v, label, fdr_cut = 0.05, lfc_cut = 1) {

  # de_*$lfc is UNCENTRED, so column t0 is the genuine 0 hpi contrast
  t0col <- paste0("t", DESIGN$timepoints[1])
  lastc <- paste0("t", DESIGN$timepoints[length(DESIGN$timepoints)])

  n_de <- function(de, cc)
    sum(de$fdr[, cc] < fdr_cut & abs(de$lfc[, cc]) > lfc_cut, na.rm = TRUE)

  data.frame(
    variety           = label,
    layer             = c("RNA", "protein"),
    n_tested          = c(nrow(v$de_rna$lfc), nrow(v$de_prot$lfc)),
    DE_at_0hpi        = c(n_de(v$de_rna, t0col),  n_de(v$de_prot, t0col)),
    DE_at_72hpi       = c(n_de(v$de_rna, lastc),  n_de(v$de_prot, lastc)),
    median_abs_lfc_0hpi = round(c(median(abs(v$de_rna$lfc[, t0col])),
                                  median(abs(v$de_prot$lfc[, t0col]))), 3),
    row.names = NULL
  )
}

bl <- do.call(rbind, Map(baseline_check, VARIETIES, names(VARIETIES)))
bl$pct_DE_at_0hpi <- round(100 * bl$DE_at_0hpi / bl$n_tested, 1)

knitr::kable(bl, caption = "A0: differential expression at 0 hpi, when biologically there should be almost none.")
```

```{r}
#| label: fig-a0-baseline
#| fig-cap: "Infected vs Control effect size at each timepoint. At 0 hpi the distribution should be centred and narrow; a wide 0 hpi distribution indicates a baseline confound."
#| fig-width: 10
#| fig-height: 5

op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

for (nm in names(VARIETIES)) {
  v  <- VARIETIES[[nm]]
  m  <- v$de_rna$lfc
  bp <- boxplot(as.data.frame(m), outline = FALSE, las = 1,
                col = c("#D1495B", rep("#2C6FBB", ncol(m) - 1)),
                xlab = "hours post inoculation", ylab = "RNA log2FC (Infected - Control)",
                main = sprintf("%s: RNA", nm))
  abline(h = 0, lty = 2, col = "grey40")
  legend("topleft", bty = "n", cex = 0.75, fill = c("#D1495B", "#2C6FBB"),
         legend = c("0 hpi (should be ~0)", "later timepoints"))
}
par(op)
```

### Verdict: severely asymmetric --- and Norin now crosses into double digits

Read the `pct_DE_at_0hpi` column. A few percent is unremarkable; a double-digit percentage at 0 hpi means the arms were **not** equivalent at baseline, and t0-centring is subtracting a real (non-biological) difference from every subsequent timepoint. Where that happens, treat that variety's kinetic results as provisional and investigate sample handling / batch / labelling before interpreting them.

**Under the validated DE, Norin does cross that line.** Computed directly from the standalone workflow's 0 dpi contrasts:

| Variety | DE at 0 hpi (FDR \< 0.05) | Rate | Previously reported |
|:--------|--------------------------:|-----:|--------------------:|
| Cadenza | 432 / 59,931 | **0.72%** | 0.4% |
| Norin   | 10,499 / 59,502 | **17.6%** | 5.8% |

A **24×** asymmetry, not the 14× previously stated. Two consequences to carry forward:

-   **Norin's kinetic results are provisional in the sense this verdict defines**, not merely "the weaker of the two datasets".
-   **t0-centring is doing something different in each variety.** In Cadenza it removes a small baseline offset, as intended. In Norin it subtracts a contrast in which 17.6% of genes are individually significant --- that is not plausibly all artefact, so centring is removing real signal along with any batch effect. The step is still the right default (the kinetic model requires `p(0) = r(0) = 1`), but the interpretation "we removed a batch artefact" does not hold for Norin and should not be written that way.

------------------------------------------------------------------------

## 2b. RESOLVED --- reconciliation with the standalone limma workflow

> **Status: closed.** This section previously recorded an unresolved 2--6×
> disagreement. It is retained as the record of the investigation and of *why*
> the pipeline now reads the standalone workflow's contrasts instead of
> re-deriving them --- the test below still runs and still shows the
> disagreement, because it deliberately re-derives DE locally in order to
> compare. **The resolution was not to find the bug but to stop having two
> answers**: `prepare_variety()` now takes the RNA treatment contrasts from the
> standalone workflow (`R/load_validated_de.R`), so the project holds one RNA DE
> result. The local fit is still used for the protein layer and for `ctrl_time`.
>
> The underlying cause was never identified. Anyone wanting to close that loop
> should diff the two `voom()` objects' `E` and `weights` matrices directly; the
> uniform deflation across every coefficient points at the variance estimate
> rather than at the data or the contrast definition.

### The assumption (as originally stated)

`R/wheat_pipeline.R`'s `run_de()` is assumed to reproduce the project's existing standalone limma DE workflow (`DE-varieties-limma`), since both fit `~treatment*time` by voom + `eBayes(trend = TRUE)` to the same counts.

### The test

```{r}
#| label: reconciliation

## Counts reported by the standalone workflow's decideTests(p.value = 0.01,
## lfc = 1) for the `treatmentInfected` coefficient -- Infected vs Control at
## 0 hpi -- and for the control-arm time main effects. Hard-coded here so the
## comparison is explicit and versioned; update if that workflow is re-run.
reported <- data.frame(
  variety            = c("Cadenza", "Norin"),
  genes_kept_reported = c(59931, 59502),
  t1_0hpi_reported    = c(127 + 305, 4754 + 5745),
  time1dpi_reported   = c(9878 + 9775, 12336 + 11214)
)

reconcile <- function(prefix, label) {
  v <- load_variety(prefix, DESIGN)
  meta <- v$meta
  meta$treatment <- relevel(factor(meta$treatment, levels = DESIGN$treatments),
                            ref = DESIGN$reference)
  meta$time_f <- relevel(factor(meta$time_num, levels = DESIGN$timepoints),
                         ref = as.character(DESIGN$timepoints[1]))

  dge  <- edgeR::DGEList(v$rna)
  keep <- rowSums(dge$counts >= 10) >= DESIGN$n_reps
  dge  <- edgeR::calcNormFactors(dge[keep, , keep.lib.sizes = FALSE])

  dm <- model.matrix(~ treatment * time_f, meta)
  colnames(dm) <- make.names(colnames(dm))

  f  <- limma::eBayes(limma::lmFit(limma::voom(dge, dm, plot = FALSE), dm),
                      trend = TRUE)
  dt <- limma::decideTests(f, p.value = 0.01, lfc = 1)

  data.frame(
    variety          = label,
    genes_kept_here  = nrow(dge),
    t1_0hpi_here     = sum(dt[, "treatmentT1"] != 0),
    time1dpi_here    = sum(dt[, "time_f24"] != 0)
  )
}

rec <- merge(
  rbind(reconcile("cadenza", "Cadenza"), reconcile("norin", "Norin")),
  reported, by = "variety"
)
rec <- rec[, c("variety", "genes_kept_here", "genes_kept_reported",
               "t1_0hpi_here", "t1_0hpi_reported",
               "time1dpi_here", "time1dpi_reported")]

knitr::kable(rec, caption = "Does this pipeline reproduce the standalone limma workflow on the same counts?")
```

### Verdict: discrepancy real, cause unidentified, superseded by adopting the standalone result

The gene counts match **exactly** in both varieties, which is strong evidence that the counts matrices and the filter are identical --- the filter depends on the count values, so an exact match is not a coincidence of annotation size. The design formula, reference levels and `eBayes(trend = TRUE)` call are also the same.

Despite that, **every coefficient is 2--6× smaller here**, uniformly, in both varieties and for both the treatment and the time main effects. A uniform deflation across all coefficients points at the variance estimate rather than at the data or the contrast definition.

Things ruled out already:

-   `robust = TRUE` --- makes almost no difference (Norin 3,425 vs 3,464 at FDR \< 0.05).
-   Different counts --- gene totals and post-filter counts match exactly.
-   Different contrast --- `treatmentT1` here is the same coefficient as `treatmentInfected` there.

The most likely remaining candidates are a difference in sample-to-column alignment between metadata and counts in one of the two workflows, or the standalone workflow having been run against a different revision of the counts files. Anyone reproducing this should diff the two `voom()` objects' `E` and `weights` matrices directly.

**How this was settled.** Not by finding the cause. The pipeline now reads the standalone workflow's published per-timepoint contrasts (`R/load_validated_de.R`), so the two answers no longer coexist and there is nothing left to reconcile. That is a defensible resolution because the standalone workflow is the one whose numbers appear elsewhere in the project, but it is worth being explicit that **the deflation was never explained** --- if the standalone workflow itself turns out to be the erroneous one, every downstream number moves.

Note that the qualitative conclusion in §2 is unaffected either way: **both** workflows show Norin with far more 0 hpi signal than Cadenza (here 1,766 vs 51, a 35× ratio; in the standalone results now adopted, 10,499 vs 432, a 24× ratio).

------------------------------------------------------------------------

## 3. A1 --- First-order kinetics

### The assumption

`dP/dt = ks·R(t) - kd·P(t)`: synthesis proportional to transcript level, degradation proportional to current protein level.

### Literature

This is the standard model in the field. @schwanhausser2011 fitted exactly this ODE genome-wide in mammalian cells, combining pulsed SILAC with mRNA measurements to estimate per-gene synthesis and degradation rates, and it remains the reference formulation ([Nature 473:337](https://www.nature.com/articles/nature10098)). Subsequent decomposition using the same framework attributes roughly 38% of protein-level variance to transcription, 18% to RNA degradation, 30% to translation and 14% to protein degradation, i.e. **the majority of control is post-transcriptional** --- which is the premise this whole analysis rests on.

The model excludes translational saturation, cooperative or zero-order degradation, stoichiometry-driven degradation of unassembled complex subunits, and any transcription-to-translation delay. Those are real phenomena; the defence for ignoring them is identifiability, not realism.

### The test

A misspecified functional form leaves **structured** residuals --- systematically positive at some timepoints and negative at others, consistently across genes. Independent noise does not.

```{r}
#| label: a1-residual-structure

residual_structure <- function(v, label) {

  keep <- rna_gate(v)
  f <- fit_amplitude(v$Rl[keep, , drop = FALSE],
                     v$Pl[keep, , drop = FALSE],
                     v$Ps[keep, , drop = FALSE],
                     DESIGN$timepoints)

  # standardise residuals by each gene's own SE so timepoints are comparable
  z <- f$resid / pmax(v$Ps[keep, , drop = FALSE], 0.05)

  data.frame(
    variety        = label,
    timepoint      = paste0("t", DESIGN$timepoints),
    mean_z_resid   = round(colMeans(z), 3),
    pct_positive   = round(100 * colMeans(z > 0), 1),
    # under a correct model each timepoint should sit near 0 and 50%
    row.names      = NULL
  )
}

rs <- do.call(rbind, Map(residual_structure, VARIETIES, names(VARIETIES)))
knitr::kable(rs, caption = "A1: standardised residuals by timepoint. A correct functional form gives mean ~0 and ~50% positive at every timepoint.")
```

```{r}
#| label: fig-a1-residuals
#| fig-cap: "Standardised residuals from the first-order model, by timepoint. Systematic departure from zero indicates the functional form is wrong, not merely noisy."
#| fig-width: 10
#| fig-height: 5

op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

for (nm in names(VARIETIES)) {
  v    <- VARIETIES[[nm]]
  keep <- rna_gate(v)
  f    <- fit_amplitude(v$Rl[keep, , drop = FALSE], v$Pl[keep, , drop = FALSE],
                        v$Ps[keep, , drop = FALSE], DESIGN$timepoints)
  z    <- f$resid / pmax(v$Ps[keep, , drop = FALSE], 0.05)

  boxplot(as.data.frame(z), outline = FALSE, las = 1,
          col = "#7FB2E5", names = paste0("t", DESIGN$timepoints),
          xlab = "hours post inoculation", ylab = "standardised residual",
          main = sprintf("%s: model residuals", nm))
  abline(h = 0, lwd = 2, col = "#D1495B")
}
par(op)
```

### Verdict: VIOLATED --- the residuals are strongly structured

Under a correct functional form, `mean_z_resid` should sit near 0 and `pct_positive` near 50% at every timepoint. Instead both varieties show large, systematic, timepoint-specific departures:

| Variety | t24             | t48             | t72                 |
|:--------|:----------------|:----------------|:--------------------|
| Cadenza | +0.55 (64% pos) | +0.75 (71% pos) | **−0.42 (35% pos)** |
| Norin   | −0.11 (41% pos) | +0.62 (67% pos) | **+0.78 (70% pos)** |

(t0 is identically zero by construction --- it is the centring anchor, not evidence.)

These are not small. A residual of +0.75 standardised units means the model **under-predicts** the protein response at 48 h by three quarters of a standard error, for the majority of genes simultaneously. And the two varieties are misspecified in *different directions*:

-   **Cadenza:** under-predicts at 24 and 48 h, over-predicts at 72 h --- the real protein response rises faster and then falls back, while first-order kinetics predicts monotonic accumulation toward the mRNA level.
-   **Norin:** under-predicts increasingly at 48 and 72 h --- consistent with a **delayed** response that the model, which reacts instantaneously to mRNA, cannot represent.

Both patterns are exactly what the omitted terms would produce: a transcription-to-translation delay, or saturation/turnover behaviour that is not first-order. This is a real limitation of the model, not noise.

**What follows from it.** The amplitude parameter `a` absorbs whatever misfit it can, so systematic shape error partly leaks into `a`. That means some fraction of genes called "regulated" may be flagged because the *shape* is wrong rather than because the amplitude genuinely departs from 1. The direction of that bias differs between varieties, which is a further reason not to compare `a` across Cadenza and Norin without care.

**What would fix it:** adding an explicit delay term `dP/dt = ks·R(t − τ) − kd·P` (one extra parameter, and with three informative timepoints almost certainly unidentifiable alongside `kd` and `a`), or more timepoints. Neither is available in this design, so the honest course is to report the misspecification rather than to claim first-order kinetics fits.

------------------------------------------------------------------------

## 4. A2 --- Treatment does not change the rate constants

### The assumption

`ks` and `kd` are the same in both arms.

### Verdict: untestable by construction, and that is the point

This assumption **is** the null hypothesis. The entire M0-vs-M1 test asks whether the data reject it. Testing it separately would be circular. What matters is that rejection is reported as *"the protein departs from its RNA-predicted trajectory"* rather than as a specific mechanism --- since a single amplitude parameter cannot say whether translation or degradation changed. See §5 for why that separation is impossible here.

------------------------------------------------------------------------

## 5. Identifiability --- which half-lives can this design see at all?

### The assumption

The half-life grid runs from `0.5 × Δt = 12 h` to `6 × 72 = 432 h`, and `t_half > 72 h` is flagged `identifiable = FALSE`.

### Literature

This matters more in plants than the notebook currently admits. 15N-labelling studies in *Arabidopsis* report protein half-lives spanning **several hours to several months** ([Li et al., Plant Cell 29:207](https://academic.oup.com/plcell/article/29/2/207/6099059)), and turnover work in developing wheat grain reports a 28 h lag for label incorporation alone ([Cao et al., New Phytologist](https://nph.onlinelibrary.wiley.com/doi/full/10.1111/nph.17756)). A 72 h course sampled every 24 h therefore sits inside a much wider biological range: fast-turnover proteins are below the resolution floor, and a large fraction of the plant proteome turns over far too slowly to be distinguished within the window.

### The test

Rather than assert the bound, show it: how different are the predicted trajectories for different half-lives, at the three timepoints actually sampled?

```{r}
#| label: fig-identifiability
#| fig-cap: "Predicted protein trajectory for a step change in mRNA, across half-lives. Curves that overlap at the sampled timepoints (dashed lines) cannot be told apart by this design."
#| fig-width: 10
#| fig-height: 5

tps  <- DESIGN$timepoints
grid <- seq(0, max(tps), by = 0.25)

# a unit step up in mRNA at t=0, the cleanest probe of kinetic resolution
r_step <- matrix(2^rep(1, length(grid)), nrow = 1)

th_show <- c(6, 12, 24, 48, 72, 144, 432)
cols    <- hcl.colors(length(th_show), "Zissou1")

op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

plot(NA, xlim = c(0, max(tps)), ylim = c(0, 1.05),
     xlab = "hours post inoculation", ylab = "protein log2FC (predicted)",
     main = "Resolvable half-lives")

for (i in seq_along(th_show)) {
  kd <- log(2) / th_show[i]
  p  <- integrate_protein_mat(grid, r_step, ks = kd, kd = kd, P0 = 1)
  lines(grid, log2(p[1, ]), col = cols[i], lwd = 2.5)
}
abline(v = tps[-1], lty = 3, col = "grey50")
legend("topleft", bty = "n", cex = 0.7, lwd = 2.5, col = cols,
       legend = sprintf("t1/2 = %g h", th_show), title = "half-life")

# how much does the trajectory actually change per doubling of half-life?
th_seq <- exp(seq(log(3), log(600), length.out = 60))
sep <- vapply(th_seq, function(th) {
  kd <- log(2) / th
  p  <- integrate_protein_mat(grid, r_step, ks = kd, kd = kd, P0 = 1)
  max(log2(p[1, match(tps[-1], grid)]))
}, numeric(1))

plot(th_seq, sep, log = "x", type = "l", lwd = 3, col = "#2C6FBB",
     xlab = "true half-life (h)", ylab = "max predicted log2FC at sampled points",
     main = "Signal available to distinguish half-lives")
abline(v = c(12, 72), lty = 2, col = "#D1495B", lwd = 2)
text(12, max(sep) * 0.95, " grid floor", col = "#D1495B", cex = 0.75, adj = 0)
text(72, max(sep) * 0.55, " identifiability\n bound", col = "#D1495B", cex = 0.75, adj = 0)
par(op)
```

```{r}
#| label: identifiability-fractions

id_frac <- do.call(rbind, lapply(names(VARIETIES), function(nm) {
  v <- VARIETIES[[nm]]; keep <- rna_gate(v)
  f <- fit_amplitude(v$Rl[keep, , drop = FALSE], v$Pl[keep, , drop = FALSE],
                     v$Ps[keep, , drop = FALSE], DESIGN$timepoints)
  data.frame(
    variety            = nm,
    n_fitted           = length(f$t_half_M0),
    pct_at_grid_floor  = round(100 * mean(f$t_half_M0 <= 12.001), 1),
    pct_unidentifiable = round(100 * mean(f$t_half_M0 > max(DESIGN$timepoints)), 1),
    pct_resolvable     = round(100 * mean(f$t_half_M0 > 12.001 &
                                          f$t_half_M0 <= max(DESIGN$timepoints)), 1)
  )
}))

knitr::kable(id_frac, caption = "Where the fitted half-lives fall relative to what the design can resolve.")
```

### Verdict: far fewer than a quarter --- the "resolvable" column was over-counting

⚠ The table below is regenerated by the chunk above and its numbers will change
on re-render; the correction it prompted does not.

| Variety | n fitted | at grid floor (≤12 h) | unidentifiable (\>72 h) | **"resolvable"** |
|:--------|:---------|:----------------------|:------------------------|:-----------------|
| Cadenza | 4228     | 24.0%                 | 49.2%                   | **26.8%**        |
| Norin   | 3487     | 24.0%                 | 53.1%                   | **22.9%**        |

**That last column is not what it appears to be.** It counts a fit pinned at the
12 h grid floor as resolved, but a floor fit means "≤ 12 h" --- it is censored,
and the design cannot distinguish a 12 h half-life from a 2 h one. Cross-tabulating
the `identifiable` flag against grid position, from the main notebook's exported
results under the validated DE:

| | Cadenza (n = 5,580) | Norin (n = 5,159) |
|:--|--:|--:|
| Not identifiable | 2,732 (49.0%) | 2,651 (51.4%) |
| Identifiable, pinned at the 12 h floor | 1,304 (23.4%) | 908 (17.6%) |
| **Identifiable and genuinely inside the band** | **392 (7.0%)** | **643 (12.5%)** |
| Above the 72 h bound | 2,376 (42.6%) | 1,952 (37.8%) |

So the honest figure is **7% (Cadenza) and 13% (Norin)**, not 23--27%. Roughly
half of all fitted genes land above the identifiability bound, and a further
fifth to a quarter sit pinned at the floor.

This matches the plant turnover literature rather than contradicting it: with *Arabidopsis* half-lives spanning hours to months [@li2017], a 72 h window sampled every 24 h was always going to see a narrow slice. The design is not broken --- it is being asked a question outside its resolution for most of the proteome.

**Consequences that must be carried into any write-up:**

-   A `t_half` at the floor means "≤ 12 h", not "12 h". A `t_half` above 72 h means "slower than we can see", not a number.
-   **The fitted `t_half` distribution must never be presented as an estimate of the wheat protein half-life distribution.** It is a distribution of what a 72 h window can resolve, truncated hard at both ends.
-   The `kinetics_limited` archetype is not a biological class; it is the \~50% of genes where attenuation and slow turnover are mathematically confounded.
-   Kinetic claims should be restricted to the resolvable quarter, and that denominator stated.

------------------------------------------------------------------------

## 6. B3 --- Protein variance treated as known

### The assumption

limma's moderated standard errors are used as fixed inverse-variance weights, ignoring uncertainty in the variance estimate itself.

### Literature

The moderated variance comes from @smyth2004's empirical Bayes procedure, which shrinks each feature's sample variance toward a fitted prior across all features --- the shrinkage that makes n = 3 designs analysable at all ([limma; Ritchie et al. 2015](https://academic.oup.com/nar/article/43/7/e47/2414268)). Treating the posterior variance as known is a standard pragmatic simplification, but it makes credible intervals mildly anti-conservative.

### The test

Refit with the SEs scaled up and down. If conclusions swing, the weights are doing more work than the data.

```{r}
#| label: b3-se-sensitivity

se_sensitivity <- function(v, label, mults = c(0.67, 1, 1.5)) {

  keep <- rna_gate(v)
  Rk <- v$Rl[keep, , drop = FALSE]
  Pk <- v$Pl[keep, , drop = FALSE]
  Sk <- v$Ps[keep, , drop = FALSE]

  # Reference fit computed FIRST, so it exists regardless of the order the
  # multipliers happen to be given in.
  ref <- fit_amplitude(Rk, Pk, Sk, DESIGN$timepoints)$a

  do.call(rbind, lapply(mults, function(m) {
    f <- fit_amplitude(Rk, Pk, Sk * m, DESIGN$timepoints)
    data.frame(variety = label, se_multiplier = m,
               n_regulated = sum(f$lrt_fdr < 0.05),
               median_a = round(median(f$a, na.rm = TRUE), 3),
               cor_a_with_ref = round(cor(f$a, ref, use = "complete.obs",
                                          method = "spearman"), 4))
  }))
}

knitr::kable(
  do.call(rbind, Map(se_sensitivity, VARIETIES, names(VARIETIES))),
  caption = "B3: sensitivity to the assumed protein measurement error."
)
```

### Verdict: the amplitude is invariant *algebraically*, but the call count is not

`cor_a_with_ref` comes out at exactly **1.000** in both varieties, and that is **not** evidence of robustness --- it is an algebraic identity. The profiled amplitude is

$$
a = \frac{\sum W \cdot pl \cdot Pl}{\sum W \cdot pl^2}
$$

so multiplying every weight by a constant cancels top and bottom exactly. A uniform SE rescaling therefore *cannot* move `a`, and this part of the test is uninformative by construction. (A non-uniform perturbation --- e.g. rescaling only low-abundance proteins --- would be informative, and is the version worth running if this matters.)

What the test does show is the second column, and it is not reassuring:

| Variety | SE × 0.67 | SE × 1.0 | SE × 1.5 |
|:--------|:----------|:---------|:---------|
| Cadenza | 1457      | 902      | 488      |
| Norin   | 1273      | 783      | 373      |

**The number of genes called "regulated" varies three-fold** across a plausible range of assumed measurement error. That is expected --- the LRT statistic scales directly with the weights --- but it means the count is a statement about the assumed noise level as much as about the biology. Report amplitudes and effect sizes; treat "N genes are regulated" as soft.

------------------------------------------------------------------------

## 7. B4 --- RNA trajectory treated as known

### The assumption

`se_rna` is accepted by `fit_hierarchical_kinetics()` **but never used**. The RNA trajectory enters the ODE as a fixed, exactly-known covariate.

### Literature

This is the classic errors-in-variables problem: a noisy predictor biases regression slopes toward zero and understates uncertainty. @schwanhausser2011 avoided it by measuring synthesis and degradation directly with pulsed labelling rather than inferring them from a noisy mRNA trajectory. With n = 3 per cell here, the RNA log2FCs are themselves moderated estimates with non-trivial standard errors, so the assumption is knowingly false --- the question is how much it matters.

### The test

Parametric bootstrap: redraw the RNA trajectory from its own sampling distribution, refit, and compare the resulting spread in `a` against the spread the model reports when RNA is treated as fixed.

```{r}
#| label: b4-rna-uncertainty

rna_uncertainty <- function(v, label, B = 40, n_genes = 1500) {

  keep <- which(rna_gate(v))
  set.seed(11)
  if (length(keep) > n_genes) keep <- sample(keep, n_genes)

  Rl <- v$Rl[keep, , drop = FALSE]; Rs <- v$Rs[keep, , drop = FALSE]
  Pl <- v$Pl[keep, , drop = FALSE]; Ps <- v$Ps[keep, , drop = FALSE]

  base <- fit_amplitude(Rl, Pl, Ps, DESIGN$timepoints)

  boot <- vapply(seq_len(B), function(b) {
    Rb <- Rl + matrix(rnorm(length(Rl), 0, Rs), nrow(Rl))
    Rb <- Rb - Rb[, 1]              # re-centre, as upstream does
    fit_amplitude(Rb, Pl, Ps, DESIGN$timepoints)$a
  }, numeric(nrow(Rl)))

  sd_from_rna <- apply(boot, 1, sd, na.rm = TRUE)

  list(
    summary = data.frame(
      variety            = label,
      n_genes            = nrow(Rl),
      median_a           = round(median(base$a, na.rm = TRUE), 3),
      median_sd_from_rna = round(median(sd_from_rna, na.rm = TRUE), 3),
      pct_sd_over_0.25   = round(100 * mean(sd_from_rna > 0.25, na.rm = TRUE), 1),
      pct_sign_unstable  = round(100 * mean(
        apply(boot, 1, function(x) any(x > 1, na.rm = TRUE) &&
                                   any(x < 1, na.rm = TRUE))), 1)
    ),
    sd = sd_from_rna, base = base$a
  )
}

b4 <- Map(rna_uncertainty, VARIETIES, names(VARIETIES))
knitr::kable(do.call(rbind, lapply(b4, `[[`, "summary")),
             caption = "B4: extra uncertainty in the amplitude that comes from RNA measurement error alone, which the model currently ignores.")
```

```{r}
#| label: fig-b4
#| fig-cap: "Amplitude uncertainty attributable to RNA measurement error. This spread is entirely absent from the reported credible intervals."
#| fig-width: 10
#| fig-height: 5

op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))
for (nm in names(b4)) {
  x <- b4[[nm]]
  hist(pmin(x$sd, 2), breaks = 60, col = "#E5A11F", border = NA,
       xlab = "SD of amplitude a induced by RNA noise",
       main = sprintf("%s (n = %d)", nm, length(x$sd)))
  abline(v = median(x$sd, na.rm = TRUE), col = "#D1495B", lwd = 2)
}
par(op)
```

### Verdict: SEVERELY VIOLATED --- this is the most consequential finding here

| Variety | median SD of `a` from RNA noise | \% genes with SD \> 0.25 | **% genes whose call flips** |
|:--------|:--------------------------------|:-------------------------|:-----------------------------|
| Cadenza | 0.83                            | 72.4%                    | **36.1%**                    |
| Norin   | 2.26                            | 83.4%                    | **49.4%**                    |

Read the last column carefully. For **36% of Cadenza genes and 49% of Norin genes**, redrawing the RNA trajectory from its own sampling distribution --- not changing the data, just acknowledging the error bars limma already reports --- moves the amplitude across `a = 1`. That is the boundary between "this protein tracks its mRNA" and "this protein is post-transcriptionally regulated". The qualitative conclusion for roughly **half the genes in Norin is not stable** under the RNA measurement error that the model currently pretends is zero.

For scale: the median extra SD contributed by RNA noise alone is 0.83 (Cadenza) and 2.26 (Norin), against amplitudes centred near 1. **None of this appears in the credible intervals reported by the hierarchical model**, because `se_rna` is accepted by the function signature and then never used.

This is the classic errors-in-variables situation, and it does not average out --- a noisy predictor biases the fitted slope and understates its uncertainty simultaneously.

**What this means for the manuscript.** Per-gene "regulated / not regulated" calls from this analysis are not reliable at the individual-gene level, particularly in Norin. Aggregate statements (the distribution of `a`, its median, comparisons of large gene sets) are far more robust than any individual call. Do not publish a per-gene regulated list from this model without either propagating RNA uncertainty or validating targets orthogonally.

**The fix**, in increasing order of effort: (1) report the bootstrap SD alongside the posterior SD, so readers see the full uncertainty; (2) inflate the likelihood weights to absorb the RNA contribution; (3) sample the latent RNA trajectory inside the model, i.e. the full `4A` formulation in `docs/planning.md`, which is what @schwanhausser2011 sidestepped entirely by measuring rates directly with pulsed labelling rather than inferring them from a noisy mRNA trajectory.

------------------------------------------------------------------------

## 8. B5 --- Genes are conditionally independent

### The assumption

Given the hyperparameters, genes contribute independently to the likelihood. The horseshoe pools information across genes on that basis.

### Literature

Independence is violated in principle by co-regulation, protein complex membership and shared MS normalisation. @leek2007 showed that unmodelled shared structure ("surrogate variables") is pervasive in high-throughput expression data and biases downstream inference if ignored ([PLoS Genetics 3:e161](https://journals.plos.org/plosgenetics/article?id=10.1371/journal.pgen.0030161)).

### The test

If genes were independent, the residual matrix would be structureless and its principal components would each explain a small, similar share of variance. A dominant PC means genes share an unmodelled effect.

```{r}
#| label: b5-residual-pca

resid_pca <- function(v, label) {
  keep <- rna_gate(v)
  f <- fit_amplitude(v$Rl[keep, , drop = FALSE], v$Pl[keep, , drop = FALSE],
                     v$Ps[keep, , drop = FALSE], DESIGN$timepoints)

  z <- f$resid / pmax(v$Ps[keep, , drop = FALSE], 0.05)
  z <- z[stats::complete.cases(z), , drop = FALSE]

  p  <- prcomp(z, center = TRUE, scale. = FALSE)
  ve <- round(100 * p$sdev^2 / sum(p$sdev^2), 1)

  data.frame(variety = label, n_genes = nrow(z),
             PC1_pct = ve[1], PC2_pct = ve[2],
             PC3_pct = if (length(ve) >= 3) ve[3] else NA_real_,
             # with T timepoints, independence implies ~100/T each
             expected_if_independent = round(100 / ncol(z), 1))
}

knitr::kable(do.call(rbind, Map(resid_pca, VARIETIES, names(VARIETIES))),
             caption = "B5: variance explained by residual principal components. Independence predicts an even split.")
```

### Verdict: VIOLATED

| Variety | PC1       | PC2   | PC3   | Expected if independent |
|:--------|:----------|:------|:------|:------------------------|
| Cadenza | **62.2%** | 23.4% | 14.4% | 25%                     |
| Norin   | **69.3%** | 17.5% | 13.3% | 25%                     |

Independence predicts roughly an even split across the four residual dimensions (\~25% each). Instead a single component absorbs **62--69%** of residual variance in both varieties. Genes are not contributing independent information --- they share a large, unmodelled effect.

The most likely sources, in order: the control-arm drift documented in the main notebook's A3 diagnostic (a shared developmental/circadian programme that the Infected-minus-Control contrast does not fully cancel), the shape misspecification identified in §3 above (a common residual pattern *is* exactly what a wrong functional form produces), and residual normalisation structure.

Note that §3 and §8 are measuring the same phenomenon from two angles: a systematic timepoint-wise residual offset shared across genes will necessarily show up as a dominant residual PC. That consistency is reassuring about the diagnosis even though the news is bad.

**Consequence.** The horseshoe pools information across genes assuming they are independent evidence. With this much shared structure the **effective number of independent genes is far below the nominal count**, so the hyperparameter posterior (and hence the amount of shrinkage applied) is over-confident. This does not bias the amplitudes in a specific direction, but it does mean the model is more certain than it should be --- compounding the same problem B4 identifies from a different source.

------------------------------------------------------------------------

## 9. Upstream --- imputation scheme

### The assumption

The `mixed` scheme (kNN for MAR, down-shifted normal for MNAR) is used throughout, and conclusions do not depend on that choice.

### Literature

The mixed strategy follows the consensus that MNAR dominates in bottom-up proteomics and needs left-censored imputation, while MAR is better served by neighbour-based methods. Benchmarks agree on the split: random forest / kNN perform best for MCAR/MAR, and QRILC or Perseus-style down-shifted imputation for left-censored MNAR ([Briefings in Bioinformatics 23:bbac138](https://academic.oup.com/bib/article/23/3/bbac138/6566001)). Reporting sensitivity across schemes is standard practice for exactly this reason.

### The test

```{r}
#| label: imputation-sensitivity

imputation_sensitivity <- function(prefix, label) {

  v  <- load_variety(prefix, DESIGN)
  qr <- run_qc_rna(v$rna, v$meta, DESIGN)
  qp <- run_qc_prot(v$prot, v$meta)

  de_rna <- run_de(qr$counts, v$meta, DESIGN, is_rna = TRUE)

  schemes <- list(
    mixed = run_missingness(qp$norm, v$meta)$mixed,

    # everything imputed as left-censored, ignoring the MAR/MNAR split
    downshift = {
      m <- qp$norm
      for (j in seq_len(ncol(m))) {
        nas <- is.na(m[, j]); if (!any(nas)) next
        mu <- mean(m[, j], na.rm = TRUE); s <- sd(m[, j], na.rm = TRUE)
        m[nas, j] <- rnorm(sum(nas), mu - 1.8 * s, 0.3 * s)
      }
      m
    },

    # no imputation at all: only fully-observed proteins
    complete_case = qp$norm[stats::complete.cases(qp$norm), , drop = FALSE]
  )

  ref_a <- NULL
  do.call(rbind, lapply(names(schemes), function(s) {
    m      <- schemes[[s]]
    common <- intersect(rownames(m), rownames(qr$counts))
    dp     <- run_de(m[common, , drop = FALSE], v$meta, DESIGN, is_rna = FALSE)

    Rl <- de_rna$lfc[common, , drop = FALSE]; Rl <- Rl - Rl[, 1]
    Pl <- dp$lfc;  Pl <- Pl - Pl[, 1]
    Rs <- de_rna$se[common, , drop = FALSE]; Ps <- dp$se

    gate <- (de_rna$any_fdr[common] < 0.05) &
            (apply(abs(Rl[, -1, drop = FALSE]), 1, max) > 0.5)

    f <- fit_amplitude(Rl[gate, , drop = FALSE], Pl[gate, , drop = FALSE],
                       Ps[gate, , drop = FALSE], DESIGN$timepoints)

    a <- setNames(f$a, rownames(Rl)[gate])
    if (s == "mixed") ref_a <<- a

    shared <- intersect(names(a), names(ref_a))
    data.frame(
      variety      = label,
      scheme       = s,
      n_proteins   = nrow(m),
      n_fitted     = length(a),
      n_regulated  = sum(f$lrt_fdr < 0.05),
      median_a     = round(median(a, na.rm = TRUE), 3),
      cor_with_mixed = round(cor(a[shared], ref_a[shared],
                                 method = "spearman", use = "complete.obs"), 4)
    )
  }))
}

knitr::kable(
  rbind(imputation_sensitivity("cadenza", "Cadenza"),
        imputation_sensitivity("norin",   "Norin")),
  caption = "Sensitivity of the amplitude estimate to imputation scheme."
)
```

### Verdict: the MAR/MNAR split matters materially

| Variety | scheme        | n proteins | n regulated | median `a` | cor with mixed |
|:--------|:--------------|:-----------|:------------|:-----------|:---------------|
| Cadenza | mixed         | 5861       | 902         | 0.994      | 1.00           |
| Cadenza | downshift     | 5861       | 627         | 1.138      | **0.70**       |
| Cadenza | complete_case | 2340       | 143         | 1.065      | 1.00           |
| Norin   | mixed         | 5641       | 783         | 0.754      | 1.00           |
| Norin   | downshift     | 5641       | 553         | 0.879      | **0.66**       |
| Norin   | complete_case | 2293       | 130         | 0.549      | 1.00           |

Two things to read here.

**Treating every missing value as left-censored changes the answer.** The `downshift` scheme correlates only **0.66--0.70** with `mixed` on the same genes, shifts the median amplitude by 0.12--0.14, and drops the regulated count by \~30%. So the MAR/MNAR classification is not a formality --- it materially determines the amplitudes. That makes §10 load-bearing rather than decorative --- and it is also why §10's result should be read carefully: it establishes abundance-DEPENDENT missingness, not that every completely-missing cell is MNAR. The imputation rests on the stronger claim, which is assumed rather than tested.

**The `complete_case` correlation of exactly 1.00 is expected, not reassuring.** A fully-observed protein has no imputed values, so its `mixed` matrix entries *are* its observed entries --- the amplitudes are identical by construction on that subset. What `complete_case` actually shows is the cost of refusing to impute: **only \~40% of proteins survive** (2340/5861 and 2293/5641), and the regulated count collapses to 143/130. That is the honest floor of what this dataset supports with no values invented at all.

------------------------------------------------------------------------

## 10. Upstream --- MNAR really is abundance-dependent

### The assumption

Zeros in the LFQ matrix are "not quantified", and missingness is dominated by left-censoring (low abundance falling below detection).

### Literature

MNAR is described as the predominant source of missingness in proteomic datasets, arising when concentrations fall below the instrument's detection limit ([review](https://academic.oup.com/bib/article/23/3/bbac138/6566001)). That is a claim about typical data, not a guarantee about *this* dataset, so it should be checked directly.

### The test

```{r}
#| label: fig-mnar
#| fig-cap: "Missingness against abundance. A strong negative relationship is the signature of left-censored (MNAR) missingness."
#| fig-width: 10
#| fig-height: 5

op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))
mnar_tab <- do.call(rbind, lapply(names(VARIETIES), function(nm) {
  v  <- load_variety(tolower(nm), DESIGN)
  qp <- run_qc_prot(v$prot, v$meta)
  m  <- qp$norm

  mean_int  <- rowMeans(m, na.rm = TRUE)
  miss_rate <- rowMeans(is.na(m))
  ok <- is.finite(mean_int)

  plot(mean_int[ok], miss_rate[ok], pch = 16, cex = 0.3, col = "#00000033",
       xlab = "mean log2 intensity", ylab = "missing rate",
       main = sprintf("%s", nm))
  lines(lowess(mean_int[ok], miss_rate[ok], f = 0.4), col = "#D1495B", lwd = 3)

  data.frame(variety = nm,
             spearman_abundance_vs_missing =
               round(cor(mean_int[ok], miss_rate[ok], method = "spearman"), 3),
             overall_missing_pct = round(100 * mean(is.na(m)), 1))
}))
par(op)

knitr::kable(mnar_tab, caption = "Is missingness abundance-dependent (MNAR) in this dataset?")
```

### Verdict: HOLDS as abundance-dependence --- which is a weaker claim than "MNAR-dominant"

> **Wording caution.** A strong negative abundance-vs-missingness correlation establishes that missingness is *abundance-dependent*: low-abundance proteins go missing more often. It does **not** establish that each completely-missing design cell is MNAR rather than MAR --- that is an additional, untested inferential step, and the imputation scheme rests on it (§9 shows the MAR/MNAR split materially changes the amplitudes). Write "abundance-dependent missingness"; reserve "MNAR" for the modelling assumption, and cite it as an assumption. A dropout model that avoids imputation entirely (e.g. proDA) is the sensitivity analysis that would settle this.

| Variety | Spearman(abundance, missing rate) | overall missing |
|:--------|:----------------------------------|:----------------|
| Cadenza | **−0.741**                        | 22.3%           |
| Norin   | **−0.754**                        | 21.0%           |

A Spearman of −0.74 is a strong, unambiguous abundance-dependence: low-abundance proteins are missing far more often, which is the definition of left-censoring. This matches the literature consensus that MNAR predominates in bottom-up proteomics [@lazar2016], and it justifies both the down-shifted normal component and the decision not to impute everything with kNN.

Given §9 showed the MAR/MNAR split materially changes the amplitudes, this is one of the load-bearing assumptions --- and it is the one that most clearly holds.

------------------------------------------------------------------------

## 11. Verdict summary

```{r}
#| label: verdict-summary
#| echo: false

knitr::kable(data.frame(
  Assumption = c("A0 baseline equivalence at 0 hpi",
                 "A1 first-order kinetics",
                 "A2 treatment leaves ks/kd unchanged",
                 "A3 control arm at steady state",
                 "Reconciliation with standalone limma",
                 "B2 kd fixed, not jointly sampled",
                 "B3 protein variance known",
                 "B4 RNA trajectory known exactly",
                 "B5 genes conditionally independent",
                 "B6 horseshoe sparsity scale",
                 "B7 MCMC converged",
                 "B8 delta_bound inert",
                 "Identifiability band",
                 "Imputation scheme",
                 "Abundance-dependent missingness"),
  Tested_in = c("§2", "§3", "-- (is the null)", "main nb §8b", "§2b",
                "-- (by construction)", "§6", "§7", "§8",
                "main nb §8c", "main nb §8c", "main nb §8c",
                "§5", "§9", "§10"),
  Verdict = c(
    "SEVERELY ASYMMETRIC: Cadenza 0.72% vs Norin 17.6% DE at 0 hpi",
    "VIOLATED: structured residuals, opposite direction per variety",
    "Untestable by design (it is the null being tested)",
    "VIOLATED: median control drift ~0.67 log2FC",
    "RESOLVED: standalone workflow's contrasts adopted; see 2b",
    "Known limitation, deliberate (no joint kd/a uncertainty)",
    "Amplitude invariant algebraically; sensitivity re-run due",
    "SEVERELY VIOLATED: 36% (Cad) / 49% (Nor) of calls flip [pre-change]",
    "VIOLATED: residual PC1 = 62-69% vs 25% expected",
    "HOLDS: insensitive across p0_frac 0.05-0.40",
    "HOLDS after the adaptive step-size fix (0% R-hat > 1.05)",
    "~4% of genes pinned at the bound",
    "Only 7% (Cad) / 13% (Nor) genuinely in the resolvable band",
    "MAR/MNAR split matters (downshift cor 0.66-0.70)",
    "HOLDS as abundance-DEPENDENT: Spearman -0.74 / -0.75. NOT the same claim as every missing cell being MNAR")
), caption = "Every stated assumption, where it is tested, and the verdict on this dataset.")
```

## 12. References

Beyond the inline links above, the core methodological citations are:

-   @schwanhausser2011 --- the ODE formulation this analysis adapts, and the source of the "most control is post-transcriptional" premise. [Nature 473:337](https://www.nature.com/articles/nature10098)
-   @smyth2004 **/ Ritchie et al. 2015** --- moderated variance estimation, without which n = 3 per cell would be unanalysable. [limma, NAR 43:e47](https://academic.oup.com/nar/article/43/7/e47/2414268)
-   @law2014 --- voom, the mean-variance modelling used for the RNA layer.
-   @carvalho2010 **/ @piironen2017** --- the horseshoe prior and the `p0`-based global scale used for `p0_frac`. [arXiv:1707.01694](https://arxiv.org/abs/1707.01694)
-   @makalic2016 --- the conjugate Gibbs sampler for the horseshoe scale mixture implemented in `R/07_bayesian_kinetics.R`.
-   @gelman1996 **/ @roberts2009** --- the 0.44 optimal acceptance rate for one-dimensional random-walk Metropolis, which the adaptive step-size tuning targets.
-   @gelman1992 --- the R-hat convergence statistic used in the main notebook's §8c.
-   **Proteomics imputation benchmark** --- QRILC/Perseus for MNAR, kNN/RF for MAR. [Brief Bioinform 23:bbac138](https://academic.oup.com/bib/article/23/3/bbac138/6566001)
-   **Plant protein turnover** --- half-lives from hours to months, setting the identifiability context. [Plant Cell 29:207](https://academic.oup.com/plcell/article/29/2/207/6099059)

> **One honest note on the prior.** `R/07_bayesian_kinetics.R` implements the *original* horseshoe [@carvalho2010] with the global scale rescaled by the `p0` heuristic of @piironen2017 --- **not** the full *regularized* horseshoe, which adds a slab that bounds the largest coefficients. The `delta_bound` clip serves that purpose crudely instead. This is worth stating precisely rather than citing the regularized horseshoe for a plain-horseshoe implementation.
