---
title: "DE proteomics - wheat"
author: "Kristina Gagalova"
format:
  gfm:
    toc: true
    code-tools: true
    fig-dir: "figs"
execute:
  echo: true
editor: visual
---

## Overview

Integrated RNA-seq / quantitative proteomics analysis of a wheat time-course treatment experiment, in two varieties (**Cadenza**, **Norin**), each with an independent **2 treatments x 4 timepoints x 3 replicates = 24 samples per omics layer** design.

**RNA and protein are measured on different biological samples within each variety** (unpaired / "Case B" design, per `../../CLAUDE.md` and `../../docs/PIPELINE.md`) --- not split aliquots of the same 24 samples. That rules out sample-level factor models (MOFA+/MEFISTO/DIABLO/O2PLS) and shapes every method choice below; see `../../docs/PIPELINE.md` and `../../docs/planning.md` for the full reasoning.

This notebook runs the complete pipeline end to end on the real data in `../../data/real/`:

1.  Data audit and design check
2.  QC & normalisation (RNA: `filterByExpr` + VST/voom; protein: validity filter + normalisation, adapted for this dataset -- see the note in §2)
3.  ID mapping (trivial here: `protein_id == gene_id` within each variety)
4.  Missingness diagnosis (MNAR/MAR) and imputation
5.  Temporal differential expression (`limma`, both layers)
6.  Kinetic-null / amplitude classification (the profile-likelihood model from `05_concordance_archetypes.R`)
7.  **Hierarchical Bayesian kinetics** (`07_bayesian_kinetics.R`) --- the new piece this notebook exists to run: a horseshoe-shrinkage model that rescues genes the profile-likelihood method above cannot classify on its own (the `kinetics_limited` problem this directory is named for)
8.  Case B design-cell integration (leave-one-cell-out Q\^2 + permutation null)
9.  Cross-variety validation using reciprocal-best-hit orthologs --- the one piece of genuine external validation available for real (non-simulated) data in this project

Steps 2--8 are defined once as reusable functions and applied to each variety; step 9 compares the two varieties' independent results against each other.

### Two things this dataset does not tell us --- edit before trusting the output

```{r}
#| label: edit-me-assumptions

## 1. WHICH TREATMENT IS THE REFERENCE?
##    Metadata encodes "T0"/"T1" with no further label. Assumed here: T0 is
##    the control/reference level, T1 is the treatment. If that is backwards,
##    every log2FC and every kinetic direction (amplified/buffered) below is
##    mirrored -- flip TREATMENT_REF to fix without re-deriving anything else.
TREATMENT_REF <- "T0"

## 2. WHAT IS THE REAL TIME SPACING?
##    Metadata encodes "t0".."t3" with no units. Assumed here: EVENLY SPACED,
##    1 arbitrary unit apart (0, 1, 2, 3). Every half-life reported in §7-8 is
##    in these SAME units, and is only interpretable as "hours" or "days" once
##    this vector is corrected to the true sampling times. Edit to the real
##    values (e.g. c(0, 6, 24, 48) for hours) before interpreting kinetics.
#TIMEPOINT_VALUES <- c(0, 1, 2, 3)
TIMEPOINT_VALUES <- c(0, 24, 48, 72)
TIME_UNIT_LABEL  <- c(0, 24, 48, 72)
#TIME_UNIT_LABEL  <- "time unit (EDIT: real spacing unknown -- see note above)"
```

## 1. Setup

```{r}
#| label: setup
#| message: false

## Quarto's default execution directory is this file's own folder
## (analysis/kinetics_limited/), while every R/*.R script in this project
## assumes the working directory is the PROJECT ROOT (matching how
## run_all.sh invokes them). Rather than fight that mismatch with a
## working-directory switch (fragile: root.dir changes how KNITR evaluates
## chunks, but does not change what a *sourced* script sees via
## sys.frame()$ofile, which is empty inside a knitr chunk anyway), every
## path below is resolved with here::here(), which walks up from the
## current directory looking for a project marker file -- this project has
## four (IntegrateProtRNA.Rproj, DESCRIPTION, renv.lock, .git) -- and returns
## an absolute path that is correct no matter where code is executed from.
## need_pkgs() itself is defined inside utils.R, so "here" -- the one package
## every subsequent path depends on -- is checked directly first.
if (!requireNamespace("here", quietly = TRUE))
  stop("package 'here' is required; install with install.packages('here')")
source(here::here("R", "utils.R"))
source(here::here("R", "pls_utils.R"))
source(here::here("R", "07_bayesian_kinetics.R"))
need_pkgs(c("limma", "edgeR", "DESeq2", "matrixStats", "impute", "here"))

set.seed(20260822)
DATA_DIR <- here::here("data", "real")

DESIGN <- list(
  treatments = c(TREATMENT_REF, setdiff(c("T0", "T1"), TREATMENT_REF)),
  reference  = TREATMENT_REF,
  timepoints = TIMEPOINT_VALUES,
  n_reps     = 3
)
CELLS <- as.vector(t(outer(DESIGN$treatments, DESIGN$timepoints,
                          function(a, b) sprintf("%s_t%s", a, b))))
```

## 2. Loading and QC (reusable functions)

```{r}
#| label: fn-load

#' Load one variety's RNA, protein, metadata and gene mapping.
#'
#' Metadata is a single file per variety, shared across both omics layers'
#' column headers -- the LH-style sample IDs happen to coincide between the
#' RNA and protein matrices for a given variety (both trace back to the same
#' field-plot label), but per the Case B design this is NOT treated as
#' sample-level pairing anywhere downstream: only variety x treatment x
#' timepoint CELL membership is used to link the two layers (see
#' `../../R/06_integration_caseB.R`'s header for why).
load_variety <- function(prefix) {

  data_dir <- here::here("data", "real")

  meta <- read.csv(
    file.path(data_dir, paste0(prefix, "_metadata.csv")),
    row.names = 1
  )

  meta$time_num  <- TIMEPOINT_VALUES[
    match(meta$timepoint, sort(unique(meta$timepoint)))
  ]
  meta$treatment <- factor(meta$treatment, levels = DESIGN$treatments)
  meta$cell <- factor(
    sprintf("%s_t%s", meta$treatment, meta$time_num),
    levels = CELLS
  )

  rna <- as.matrix(
    read.csv(
      file.path(data_dir, paste0(prefix, "-rnaseq.csv")),
      row.names = 1,
      check.names = FALSE
    )
  )

  prot <- as.matrix(
    read.csv(
      file.path(data_dir, paste0(prefix, "-prot.csv")),
      row.names = 1,
      check.names = FALSE
    )
  )

  gmap <- read.csv(
    file.path(data_dir, paste0(prefix, "_protein_gene_mapping.csv"))
  )

  stopifnot(
    all(colnames(rna) %in% rownames(meta)),
    all(colnames(prot) %in% rownames(meta))
  )

  rna  <- rna[, rownames(meta), drop = FALSE]
  prot <- prot[, rownames(meta), drop = FALSE]

  list(
    rna = rna,
    prot = prot,
    meta = meta,
    gmap = gmap
  )
}
```

```{r}
#| label: fn-qc

#' RNA QC: filterByExpr, then VST (integration matrix) + voom-ready counts.
#' Same procedure as `01_qc_normalise.R`, applied to real counts directly.
run_qc_rna <- function(counts, meta) {
  ## Manual low-expression filter (DESeq2-style threshold), replacing
  ## edgeR::filterByExpr()'s adaptive heuristic: keep a gene if it has
  ## >= 10 counts in at least as many samples as the smallest design group.
  dge <- edgeR::DGEList(counts = counts)
  smallestGroupSize <- DESIGN$n_reps   # 3
  keep <- rowSums(dge$counts >= 10) >= smallestGroupSize
  dge  <- dge[keep, , keep.lib.sizes = FALSE]
  cf   <- dge$counts

  dge <- edgeR::calcNormFactors(dge)
  logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 3)
  dds <- DESeq2::DESeqDataSetFromMatrix(cf, meta, ~cell)
  dds <- DESeq2::estimateSizeFactors(dds)
  vst <- SummarizedExperiment::assay(
    DESeq2::vst(dds, blind = TRUE, nsub = min(1000, sum(rowMeans(cf) > 5))))
  list(counts = cf, logcpm = logcpm, vst = vst, n_in = nrow(counts), n_kept = nrow(cf))
}

#' Protein QC, adapted from `01_qc_normalise.R` for this dataset:
#' the real protein_gene_mapping.csv carries no unique-peptide/decoy/
#' contaminant columns (unlike the simulator's MaxQuant-style metadata), so
#' that part of 01's filtering is not applicable here and is skipped
#' explicitly rather than silently omitted. Zero is treated as "not
#' quantified" (standard LFQ-intensity convention), not a true biological
#' zero -- it becomes NA before the validity filter and normalisation.
run_qc_prot <- function(intensity, meta, min_valid_in_cell = 2) {
  m <- intensity; m[m == 0] <- NA; m <- log2(m)
  valid_per_cell <- vapply(levels(meta$cell), function(cl)
    rowSums(!is.na(m[, meta$cell == cl, drop = FALSE])), numeric(nrow(m)))
  keep <- matrixStats::rowMaxs(valid_per_cell) >= min_valid_in_cell
  mf <- m[keep, , drop = FALSE]
  ## median normalisation: shift each sample so its median matches the
  ## matrix-wide median -- corrects per-run loading, not biology.
  shift <- apply(mf, 2, median, na.rm = TRUE) - median(mf, na.rm = TRUE)
  mn <- sweep(mf, 2, shift, "-")
  list(norm = mn, n_in = nrow(intensity), n_kept = nrow(mf), miss_rate = mean(is.na(mn)))
}
```

```{r}
#| label: fn-missingness

#' MNAR/MAR diagnosis and mixed imputation, same rule as `03_missingness.R`:
#' a protein missing in EVERY replicate of a cell (but observed in some other
#' cell) is MNAR (left-censored -- genuinely low/absent there); missing in
#' only SOME replicates of its cell is MAR (stochastic). Imputed values are
#' for Universe A / integration use only -- the DE and kinetics steps below
#' use these imputed values too (protein has no complete-case alternative at
#' this scale), but the up-front validity filter (>=2 valid in >=1 cell)
#' keeps the imputation from manufacturing pure noise.
run_missingness <- function(prot_norm, meta) {
  cells <- levels(meta$cell)
  is_na <- is.na(prot_norm)
  n_obs_cell <- vapply(cells, function(cl)
    rowSums(!is.na(prot_norm[, meta$cell == cl, drop = FALSE])), numeric(nrow(prot_norm)))
  lab <- matrix("obs", nrow(prot_norm), ncol(prot_norm), dimnames = dimnames(prot_norm))
  for (k in seq_along(cells)) {
    ii <- which(meta$cell == cells[k]); gone <- n_obs_cell[, k] == 0
    lab[gone, ii] <- "MNAR"; lab[!gone, ii][is_na[!gone, ii]] <- "MAR"
  }
  lab[!is_na] <- "obs"

  mar_imp <- suppressWarnings(
    impute::impute.knn(prot_norm, k = 10, rowmax = 0.95, colmax = 0.95)$data)
  mnar_imp <- prot_norm
  for (j in seq_len(ncol(prot_norm))) {
    v <- prot_norm[, j]; nas <- is.na(v)
    if (!any(nas)) next
    mu <- mean(v, na.rm = TRUE); sdv <- sd(v, na.rm = TRUE)
    ## Perseus-style down-shifted normal: width 0.3, downshift 1.8 SD.
    mnar_imp[nas, j] <- rnorm(sum(nas), mu - 1.8 * sdv, 0.3 * sdv)
  }
  mixed <- prot_norm
  mixed[lab == "MAR"]  <- mar_imp[lab == "MAR"]
  mixed[lab == "MNAR"] <- mnar_imp[lab == "MNAR"]
  stopifnot(!anyNA(mixed))
  list(mixed = mixed, labels = lab,
       mnar_frac = mean(lab == "MNAR"), mar_frac = mean(lab == "MAR"))
}
```

```{r}
#| label: fn-de

#' Temporal DE, both layers, same design as `04_univariate_temporal_de.R`:
#' an 8-level ~0+cell design, per-timepoint treatment contrasts, and an
#' any-effect F-test. limma is used for both RNA (voom) and protein (trend +
#' robust eBayes) so effect sizes are on a comparable scale.
run_de <- function(mat, meta, is_rna) {
  ## Treatment-contrast parameterisation (~treatment*time), matching your
  ## limma workflow, in place of the ~0+cell design used before. The two are
  ## equivalent full-rank codings of the same means model: the per-timepoint
  ## contrasts reconstructed below (main effect + interaction term) give the
  ## same logFC/SE as the T1_tX - T0_tX cell-means contrast this replaces.
  meta$treatment <- relevel(factor(meta$treatment, levels = DESIGN$treatments),
                            ref = DESIGN$reference)
  meta$time_f    <- relevel(factor(meta$time_num, levels = DESIGN$timepoints),
                            ref = as.character(DESIGN$timepoints[1]))
  design <- model.matrix(~ treatment * time_f, data = meta)
  colnames(design) <- make.names(colnames(design))

  if (is_rna) {
    dge <- edgeR::calcNormFactors(edgeR::DGEList(mat))
    v   <- limma::voom(dge, design, plot = FALSE)
    fit <- limma::lmFit(v, design)
  } else {
    fit <- limma::lmFit(mat, design)
  }

  ## Reconstruct the per-timepoint treatment contrasts (Infected - Control at
  ## each timepoint): the reference timepoint's contrast IS
  ## main effect; every later timepoint adds its interaction term.
  alt       <- setdiff(DESIGN$treatments, DESIGN$reference)
  main_coef <- paste0("treatment", alt)
  int_coefs <- paste0(main_coef, ".time_f", DESIGN$timepoints[-1])
  stopifnot(all(c(main_coef, int_coefs) %in% colnames(design)))  # catches a

                                                                  ## mismatch
                                                                  ## immediately

                                                                  ## downstream
  contrast_defs <- setNames(
    c(main_coef, paste(main_coef, int_coefs, sep = " + ")),
    paste0("tp_t", DESIGN$timepoints))
  cm <- limma::makeContrasts(contrasts = contrast_defs, levels = design)
  colnames(cm) <- names(contrast_defs)
  f2 <- limma::eBayes(limma::contrasts.fit(fit, cm), trend = TRUE, robust = TRUE)

  lfc  <- f2$coefficients; colnames(lfc) <- paste0("t", DESIGN$timepoints)
  se   <- f2$stdev.unscaled * sqrt(f2$s2.post); colnames(se) <- colnames(lfc)
  padj <- apply(f2$p.value, 2, p.adjust, "BH"); colnames(padj) <- colnames(lfc)

  ## Omnibus F-test across the 4 timepoint contrasts, same
  ## as.numeric() strips rownames, so reattach them explicitly.
  ff <- limma::classifyTestsF(f2, fstat.only = TRUE)
  p_any <- setNames(pf(as.numeric(ff), attr(ff, "df1"), attr(ff, "df2"), lower.tail = FALSE),
                    rownames(f2))

  ## CONTROL-ARM time effects. In a ~treatment*time design with Control as the
  ## reference treatment level, the bare `time_fX` coefficients are exactly the
  ## control arm's own change from baseline (Control at tX - Control at t0).
  ## These are not used for any treatment inference -- they are returned solely
  ## to test the quasi-steady-state assumption the kinetic null model rests on
  ## (assumption A3; see the A3 diagnostic section).
  time_coefs <- paste0("time_f", DESIGN$timepoints[-1])
  ctrl_time <- NULL
  if (all(time_coefs %in% colnames(fit$coefficients))) {
    ctrl_time <- fit$coefficients[, time_coefs, drop = FALSE]
    colnames(ctrl_time) <- paste0("t", DESIGN$timepoints[-1])
  }

  list(lfc = lfc, se = se, fdr = padj, any_fdr = p.adjust(p_any, "BH"),
       ctrl_time = ctrl_time)
}
```

## 3. Kinetic archetype classification (profile-likelihood)

### Kinetic Modeling of Transcriptome & Proteome Trajectories

Standard Differential Expression (DE) tells you whether a transcript and its protein each changed. It cannot tell you whether a discordance between them is biologically meaningful. A gene whose mRNA rises sharply while its protein barely moves has two completely different possible explanations:

-   **The protein is under post-transcriptional control** (translational repression, accelerated degradation, sequestration) - biology.
-   **The protein simply has a long half-life** and hasn't had time to accumulate yet - kinetics.

Both produce the identical snapshot pattern. This step exists to separate them, and it's the reason a time course was run at all.

### Inputs and Outputs

**Inputs** (all genes × timepoints matrices, restricted to Universe B - genes with both layers passing QC):

| Argument                  | Content                                                                              |
|:--------------------------|:-------------------------------------------------------------------------------------|
| `Rl, Pl`                  | log2FC trajectories (treated vs control), t0-centred so column 1 is identically zero |
| `Rs, Ps`                  | limma moderated standard errors for those logFCs                                     |
| `tps`                     | sampled timepoints                                                                   |
| `fdr_thresh, min_abs_lfc` | classification thresholds (defaults 0.05, 0.5)                                       |

**Output:** one row per gene --- `archetype`, `amplitude_a`, `t_half_M0_h`, `t_half_M1_h`, `identifiable`, `rna_responds`, `regulated`, `lrt_p`, `lrt_fdr`.

### The Two Nested Models

Everything is expressed as fold-change relative to the time-matched control, so $r(t) = 2^{\text{logFC\_RNA}(t)}$ and $p(t) = 2^{\text{logFC\_protein}(t)}$, both starting at 1.

### M0 - The Kinetic Null

If the treatment changed only transcript abundance, the protein is forced to follow first-order kinetics:

$$ \frac{dp}{dt} = k_d \cdot (r(t) - p(t)), \quad p(0) = 1 $$

One free parameter: $k_d$, the degradation rate (reported as half-life $\log(2)/k_d$). This is the trajectory the protein must take if nothing post-transcriptional is happening. Implemented via `integrate_protein_mat(..., ks = kd, kd = kd, P0 = 1)` - setting `ks = kd` makes the steady state equal $r$, which is what encodes "protein eventually catches up to its mRNA."

### M1 - Kinetics Plus Amplitude

Adds a single scaling factor:

$$ \text{logFC\_protein}(t) = a \cdot \log_2 p(t; k_d) $$

$a = 1$ is exactly M0. The interpretation of $a$ is the whole point:

| $a$             | Meaning                                                                                                |
|:----------------|:-------------------------------------------------------------------------------------------------------|
| **≈ 1**         | Protein tracks its mRNA as kinetics predict --- no post-transcriptional signature                      |
| **0 \< a \< 1** | **Buffered** --- response damped relative to prediction (translational repression, faster degradation) |
| **a \> 1**      | **Amplified** --- response exceeds prediction (increased translation efficiency, stabilisation)        |
| **a \< 0**      | **Anticorrelated** --- protein moves opposite to its mRNA                                              |

------------------------------------------------------------------------

## Fitting Procedure

For each of 80 candidate half-lives on a log-spaced grid, both models are fitted to all genes simultaneously (vectorised), then the best grid point per gene is taken:

1.  Solve the ODE at that $k_d \rightarrow$ predicted trajectory $pl$.
2.  **M0:** weighted SSR of $pl$ against observed, $a$ fixed at 1.
3.  **M1:** the optimal $a$ has a closed form --- weighted least squares through the origin, $a = \frac{\sum (W \cdot pl \cdot Pl)}{\sum (W \cdot pl^2)}$ --- so it's profiled out analytically at zero cost rather than searched. Minimum SSR is then $SST - \frac{\text{num}^2}{\text{den}}$.
4.  `max.col(-ssr)` picks each gene's best half-life independently for M0 (`t_half_M0_h`) and M1 (`t_half_M1_h`).

Weights are $W = \frac{1}{\max(Ps, 0.05)^2}$ --- inverse-variance, so each gene is judged against its own measurement noise. With $n=3$ per cell this is essential; without it, noisy low-abundance proteins would dominate the "regulated" calls.

Because t0 is the centring anchor, both `Pl[,1]` and `pl[,1]` are exactly 0, so that timepoint contributes nothing to the SSR --- the fit uses the 3 post-baseline points.

## The Test

$LRT = SSR_{M0} - SSR_{M1}$, compared to $\chi^2$ with 1 df. Since the weights already carry $1/\sigma^2$, the weighted SSR difference is directly on the chi-square scale. BH-adjusted across genes $\rightarrow$ `lrt_fdr`.

A gene is called **regulated** when the data statistically demand $a \neq 1$ --- i.e. the kinetic null is insufficient. This is a rejection test, not a threshold on effect size.

## Classification Decision Tree

Evaluated in this order (later assignments overwrite earlier):

``` text
unchanged                                    (default)
├── not RNA-responsive, but protein moves  → protein_only
└── RNA-responsive
    ├── null NOT rejected
    │   ├── half-life unidentifiable         → kinetics_limited
    │   ├── identifiable, fast               → concordant
    │   └── identifiable, slow               → kinetic_lag
    └── null rejected (regulated)
        ├── a ≥ 1                            → amplified
        ├── 0 < a < 1                        → buffered
        └── a ≤ 0                            → anticorrelated
```

**Two structural properties worth documenting:** \* `regulated` requires `rna_resp`. Amplitude is meaningless without an RNA signal to scale --- for a flat transcript, $a \cdot 0 = 0$ for any $a$. This gate is what makes `protein_only` a separate category rather than a spuriously extreme amplitude. \* `concordant` vs `kinetic_lag` is a labelling split, not a test. Both mean "the kinetic null was sufficient." The slow cutoff just separates fast-turnover proteins from slow ones for readability; genes near the line are on a continuum.

## Identifiability --- The `kinetics_limited` Class

This is the most important thing to convey in documentation. A half-life longer than the observation window cannot be estimated from that window --- a protein with a 200 h half-life and one with a 500 h half-life look identical across 72 h. In that regime, "slow turnover" and "strong buffering" both flatten the predicted trajectory and become genuinely confounded.

Rather than silently assigning such genes to `kinetic_lag` or `buffered`, they are flagged `identifiable = FALSE` and given their own archetype. `kinetics_limited` is an honest statement of what the data cannot resolve, not a biological category. On the simulated ground-truth data this was \~15% of genes; the real remedy is an external half-life prior (see `concordance.half_life_prior_file` in `config/config.yaml`), which cut it to \~6%.

## Constants and Their Basis

| Constant         | Value                        | Basis                                                                                                                                                                                                                                               |
|:-----------------|:-----------------------------|:----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| `th_bound`       | `max(tps)`                   | Principled --- the identifiability boundary. Half-lives beyond the observation window are unestimable by construction.                                                                                                                              |
| `th_grid range`  | `0.5 * Δt` to `6 * max(tps)` | **Lower:** below half the smallest sampling gap, distinct half-lives give indistinguishable sampled trajectories.<br>**Upper:** extends past the boundary so slow genes aren't clamped --- the `identifiable` flag reports the uncertainty instead. |
| `th_grid points` | `80`                         | Empirical ("more doesn't change the answer"), not derived.                                                                                                                                                                                          |
| `ODE grid step`  | `0.25`                       | Discretisation error negligible vs measurement noise.                                                                                                                                                                                               |
| `SE floor`       | `0.05`                       | Numerical guard so a spuriously tiny moderated SE can't give one gene near-infinite weight.                                                                                                                                                         |
| `min_abs_lfc`    | `0.5`                        | Effect-size floor (\~41% change) so statistically-significant-but-trivial changes aren't kinetically modelled. Conventional.                                                                                                                        |
| `fdr_thresh`     | `0.05`                       | Standard BH convention.                                                                                                                                                                                                                             |
| `slow cutoff`    | `0.5 * th_bound`             | Readability split only (see above).                                                                                                                                                                                                                 |

> The first two rows change what the model means; the rest are conventional defaults you can retune.

## Two Things to Check Before Writing This Up

1.  **The responsiveness gate:** The notebook's port differs from `05`. `05` uses both an FDR test and an effect-size floor:

    ``` r
    rna_resp  <- (r_any < FDR) & (max_abs(Rl) > MINLF)
    prot_resp <- (p_any < FDR) & (max_abs(Pl) > MINLF)
    ```

    The notebook version uses effect size only (it doesn't receive the F-test FDR vectors). That makes `rna_responds` and `protein_only` more permissive in the notebook than in `05`, so the two will not produce identical archetype counts on the same data. Worth either documenting as a deliberate simplification or fixing to match --- say which and I'll align them.

2.  **`th_grid`'s lower bound is timepoint-scale-dependent:** It's `0.5 * diff(tps)[1]`, so with `TIMEPOINT_VALUES = c(0,1,2,3)` it's 0.5 units, but with real hours `c(0,24,48,72)` it becomes 12 h --- meaning no half-life below 12 h can be fitted. Given wheat protein turnover, that's probably fine, but it's a real constraint that appears only after the timepoint values are corrected, and it should be stated in the docs.

------------------------------------------------------------------------

## Where the Kinetic Null Comes From --- Derivation and Assumptions

The model is **not** a necessary consequence of "the treatment only changed transcript abundance." It follows from three assumptions, and it is worth stating all three because one of them is measurably violated in this dataset.

Start from absolute abundance in each arm, assuming first-order kinetics with the *same* rate constants in both:

$$\frac{dP_c}{dt} = k_s R_c(t) - k_d P_c(t), \qquad \frac{dP_t}{dt} = k_s R_t(t) - k_d P_t(t)$$

Define the fold-changes $r = R_t/R_c$ and $p = P_t/P_c$. Differentiating the quotient:

$$\frac{dp}{dt} = \frac{P_t' P_c - P_t P_c'}{P_c^2} = \frac{k_s (R_t P_c - R_c P_t)}{P_c^2} = \underbrace{\frac{k_s R_c}{P_c}}_{\text{coefficient}} (r - p)$$

The $k_d$ terms **cancel exactly**. The coefficient is $k_s R_c / P_c$, *not* $k_d$. It equals $k_d$ only when $k_s R_c = k_d P_c$ --- that is, only when the control arm is at steady state.

So the implemented equation $dp/dt = k_d(r - p)$ requires:

| \#     | Assumption                                                | Status                                                                                                                                                                                       |
|:-------|:----------------------------------------------------------|:---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| **A1** | Protein dynamics are first-order: $dP/dt = k_s R - k_d P$ | Approximation. Excludes translational saturation, cooperative/zero-order degradation, stoichiometry-driven degradation of unassembled complex subunits, and transcription→translation delay. |
| **A2** | Treatment leaves $k_s$ and $k_d$ unchanged                | This *is* the null hypothesis being tested. Rejecting it is the point.                                                                                                                       |
| **A3** | Control arm is at (quasi-)steady state throughout         | **Measurably violated here --- see the diagnostic below.**                                                                                                                                   |

### A note on `ks = kd` in the code

`integrate_protein_mat(..., ks = kd, kd = kd, P0 = 1)` is **not** a biological claim that synthesis rate equals degradation rate. In the fold-change formulation absolute $k_s$ cancels out entirely and is not identifiable from this data at all (it is absorbed by the per-protein MS response factor). Setting `ks = kd` is the mathematical encoding of **A3** --- it forces the steady state of the relative system to equal $r$.

### Why A3 is questionable in this dataset

In a `~treatment*time` design with Control as the reference level, the bare `time` coefficients are exactly the control arm's own change from baseline. Those coefficients show substantial control-arm movement over the time course --- unsurprising for wheat seedlings over 72 h (development, circadian rhythm, handling), but a direct violation of "control at steady state."

The consequence is specific: when $R_c$ is rising, $P_c$ lags below its steady state, so the true coefficient $k_s R_c/P_c$ exceeds $k_d$; when $R_c$ falls, it drops below. Because $k_d$ and $a$ trade off along the identifiability ridge, that misfit can leak into the amplitude estimate --- meaning some `regulated` calls could reflect control-arm non-stationarity rather than treatment-specific post-transcriptional control.

Two mitigating points, for balance: the ratio formulation cancels any effect **shared** by both arms exactly, and growth dilution inflates $k_d$ uniformly without breaking the functional form. It is specifically the *lag of* $P_c$ *behind a moving* $R_c$ that breaks A3, not movement per se.

The next section tests how much this actually matters.

## A3 diagnostic --- does control-arm drift drive `regulated` calls?

```{r}
#| label: fn-a3-diagnostic

#' Test assumption A3 (control arm at quasi-steady state) empirically.
#'
#' If A3 violation were generating spurious `regulated` calls, genes whose
#' CONTROL arm drifts most over the time course should be enriched for
#' `regulated` relative to genes whose control arm is stable.
#'
#' CONFOUNDER, and why the raw enrichment is not the headline number: genes
#' with large control-arm drift also tend to have larger treatment effects and
#' higher expression, both of which independently increase power to reject the
#' kinetic null. A raw drift-vs-regulated association would therefore be
#' expected even under a perfectly valid A3. The reported test is a logistic
#' regression of `regulated` on control drift ADJUSTED for the gene's own RNA
#' response amplitude, so the drift coefficient answers the question that
#' matters: at a given treatment effect size, does control-arm instability
#' still predict being called regulated?
run_a3_diagnostic <- function(ctrl_time, Rl, arch, rna_resp, label = "") {
  idx <- match(rownames(Rl), arch$gene_id)
  d <- data.frame(
    gene_id   = rownames(Rl),
    ## control-arm drift: largest absolute change of the CONTROL arm from
    ## baseline, across post-baseline timepoints
    drift     = apply(abs(ctrl_time[rownames(Rl), , drop = FALSE]), 1, max),
    ## treatment effect amplitude, the confounder being adjusted for
    rna_amp   = apply(abs(Rl[, -1, drop = FALSE]), 1, max),
    regulated = arch$regulated[idx],
    a         = arch$amplitude_a[idx],
    resp      = rna_resp,
    stringsAsFactors = FALSE)
  ## Restrict to the genes the kinetic model was actually fitted on.
  d <- d[d$resp & is.finite(d$a) & is.finite(d$drift) & is.finite(d$rna_amp), ]

  ## Unadjusted view: regulated rate by control-drift quartile.
  d$drift_q <- cut(d$drift, quantile(d$drift, 0:4 / 4), include.lowest = TRUE,
                   labels = c("Q1 (most stable)", "Q2", "Q3", "Q4 (most drift)"))
  by_q <- do.call(rbind, lapply(split(d, d$drift_q), function(g)
    data.frame(n = nrow(g),
               median_drift = round(median(g$drift), 2),
               median_rna_amp = round(median(g$rna_amp), 2),
               pct_regulated = round(100 * mean(g$regulated), 1))))
  by_q <- data.frame(drift_quartile = rownames(by_q), by_q, row.names = NULL)

  ## Adjusted test.
  fit <- glm(regulated ~ scale(drift) + scale(rna_amp), binomial, data = d)
  co  <- summary(fit)$coefficients
  ci  <- suppressMessages(confint(fit))
  res <- data.frame(
    term = c("control drift", "RNA amplitude"),
    odds_ratio = round(exp(co[2:3, "Estimate"]), 3),
    ci_lo = round(exp(ci[2:3, 1]), 3),
    ci_hi = round(exp(ci[2:3, 2]), 3),
    p_value = signif(co[2:3, "Pr(>|z|)"], 3), row.names = NULL)

  ## Same question on the continuous amplitude deviation |a - 1|.
  sp_raw <- suppressWarnings(cor(d$drift, abs(d$a - 1), method = "spearman"))
  ## Partial Spearman, adjusting for rna_amp by ranking then residualising.
  rr <- function(x) rank(x)
  pres <- residuals(lm(rr(abs(d$a - 1)) ~ rr(d$rna_amp)))
  dres <- residuals(lm(rr(d$drift)      ~ rr(d$rna_amp)))
  sp_adj <- cor(dres, pres)

  ## COLLINEARITY CHECK -- required to interpret `adjusted` honestly. Control
  ## drift and treatment amplitude both measure "this gene is dynamic", so they
  ## are correlated by construction. When they are strongly collinear the two
  ## adjusted coefficients cannot be cleanly separated and either can flip sign
  ## or inflate; report the correlation so the reader can judge.
  collin <- cor(d$drift, d$rna_amp, method = "spearman")

  list(label = label, n = nrow(d), by_quartile = by_q, adjusted = res,
       spearman_raw = round(sp_raw, 3), spearman_adjusted = round(sp_adj, 3),
       collinearity = round(collin, 3), data = d)
}
```

```{r}
#| label: fn-archetypes

#' Nested kinetic model: M0 (protein trajectory fully explained by RNA
#' kinetics alone) vs M1 (M0 + a free amplitude factor `a`), compared by a
#' 1-df likelihood-ratio test. Identical model and identifiability handling
#' to `05_concordance_archetypes.R` -- see that script's header for the full
#' reasoning (this is what separates real post-transcriptional regulation
#' from a protein simply having a long half-life).
run_kinetic_archetypes <- function(Rl, Rs, Pl, Ps, tps, fdr_thresh = 0.05, min_abs_lfc = 0.5) {
  
  grid <- seq(0, max(tps), by = 0.25); gi <- match(tps, grid)
  r_grid <- t(apply(Rl, 1, function(v) approx(tps, v, grid, rule = 2)$y))
  r_lin  <- 2^r_grid
  th_bound <- max(tps)
  th_grid  <- exp(seq(log(0.5 * diff(tps)[1]), log(6 * th_bound), length.out = 80))
  kd_grid  <- log(2) / th_grid
  W <- 1 / pmax(Ps, 0.05)^2; SST <- rowSums(W * Pl^2)
  n <- nrow(Pl); J <- length(kd_grid)
  ssr0 <- ssr1 <- ahat <- matrix(NA_real_, n, J); preds <- vector("list", J)
  
  for (j in seq_len(J)) {
    kd <- kd_grid[j]
    pl <- log2(pmax(integrate_protein_mat(grid, r_lin, ks = kd, kd = kd, P0 = rep(1, n))[, gi, drop = FALSE], 1e-9))
    preds[[j]] <- pl
    ssr0[, j] <- rowSums(W * (pl - Pl)^2)
    num <- rowSums(W * pl * Pl); den <- rowSums(W * pl^2)
    a <- ifelse(den > 1e-12, num / den, NA_real_); ahat[, j] <- a
    ssr1[, j] <- SST - ifelse(den > 1e-12, num^2 / den, 0)
  }
  
  j0 <- max.col(-ssr0, "first"); j1 <- max.col(-ssr1, "first")
  ix <- function(j) cbind(seq_len(n), j)
  S0 <- ssr0[ix(j0)]; S1 <- ssr1[ix(j1)]
  t_half_M0 <- th_grid[j0]; t_half_M1 <- th_grid[j1]; a_hat <- ahat[ix(j1)]
  pred0 <- t(vapply(seq_len(n), function(i) preds[[j0[i]]][i, ], numeric(length(tps))))
  lrt <- pmax(S0 - S1, 0); p_lrt <- pchisq(lrt, 1, lower.tail = FALSE); fdr_lrt <- p.adjust(p_lrt, "BH")

  max_abs <- function(m) apply(abs(m[, -1, drop = FALSE]), 1, max)
  rna_resp <- max_abs(Rl) > min_abs_lfc
  regulated <- rna_resp & (fdr_lrt < fdr_thresh)
  identifiable <- t_half_M0 <= th_bound; slow <- t_half_M0 > 0.5 * th_bound

  archetype <- rep("unchanged", n)
  archetype[!rna_resp & max_abs(Pl) > min_abs_lfc] <- "protein_only"
  archetype[rna_resp & !regulated &  identifiable & !slow] <- "concordant"
  archetype[rna_resp & !regulated &  identifiable &  slow] <- "kinetic_lag"
  archetype[rna_resp & !regulated & !identifiable]         <- "kinetics_limited"
  archetype[regulated & a_hat >= 1]            <- "amplified"
  archetype[regulated & a_hat < 1 & a_hat > 0] <- "buffered"
  archetype[regulated & a_hat <= 0]            <- "anticorrelated"

  data.frame(gene_id = rownames(Pl), archetype = archetype, rna_responds = rna_resp,
            regulated = regulated, amplitude_a = a_hat,
            t_half_M0_h = t_half_M0, t_half_M1_h = t_half_M1,
            identifiable = identifiable, lrt_p = p_lrt, lrt_fdr = fdr_lrt,
            stringsAsFactors = FALSE)
}
```

## 4. Case B integration (`06`-style)

```{r}
#| label: fn-integration

#' Design-cell PLS with leave-one-cell-out Q^2 against a permutation null --
#' the Case B analogue of a MEFISTO/DIABLO variance-explained figure. See
#' `06_integration_caseB.R`'s header for why the in-sample cross-block
#' correlation is not a valid statistic here (it reaches ~0.7-0.9 even on
#' permuted data at this sample size).
run_integration <- function(rna_mat, prot_mat, cd_r, cd_p, n_hvg = 3000, n_hvp = 2000,
                            n_comp = 3, n_perm = 200) {
  hv_r <- top_variable(rna_mat,  min(n_hvg, nrow(rna_mat)))
  hv_p <- top_variable(prot_mat, min(n_hvp, nrow(prot_mat)))
  Xc <- prep_block(cell_means(rna_mat[hv_r, ],  cd_r))
  Yc <- prep_block(cell_means(prot_mat[hv_p, ], cd_p))
  stopifnot(identical(rownames(Xc), rownames(Yc)))
  K  <- min(n_comp, nrow(Xc) - 3)
  fit <- pls2(Xc, Yc, ncomp = K)
  Ax <- row_space(Xc); Ay <- row_space(Yc)
  q2 <- q2_loo(Ax, Ay, K)
  perm_q2 <- matrix(NA_real_, n_perm, K)
  for (b in seq_len(n_perm)) perm_q2[b, ] <- q2_loo(Ax, Ay[sample(nrow(Ay)), , drop = FALSE], K)
  p_q2 <- vapply(seq_len(K), function(k) (1 + sum(perm_q2[, k] >= q2[k])) / (n_perm + 1), 0)
  list(fit = fit, q2 = q2, perm_q2 = perm_q2, p_q2 = p_q2,
       cellmeta = data.frame(cell = rownames(Xc)))
}
```

## 5. Run the pipeline: Cadenza

```{r}
#| label: cadenza-load

cad <- load_variety("cadenza")
knitr::kable(table(cad$meta$treatment, cad$meta$time_num),
            caption = "Cadenza design: samples per treatment x timepoint")
```

```{r}
#| label: cadenza-qc

cad_qc_r <- run_qc_rna(cad$rna, cad$meta)
cad_qc_p <- run_qc_prot(cad$prot, cad$meta)
cat(sprintf("RNA: %d / %d genes kept by filterByExpr\n", cad_qc_r$n_kept, cad_qc_r$n_in))
cat(sprintf("Protein: %d / %d groups kept (>=2 valid in >=1 cell); missing rate %.1f%%\n",
           cad_qc_p$n_kept, cad_qc_p$n_in, 100 * cad_qc_p$miss_rate))
```

```{r}
#| label: fig-cadenza-qc-pca
#| fig-cap: "Cadenza PCA, RNA (VST) and protein (normalised), coloured by treatment"

pca_plot <- function(m, meta, main) {
  mv <- m[order(row_var(m), decreasing = TRUE, na.last = NA)[seq_len(min(500, nrow(m)))], , drop = FALSE]
  mv <- mv[complete.cases(mv), , drop = FALSE]
  p <- prcomp(t(mv), scale. = FALSE)
  v <- round(100 * p$sdev^2 / sum(p$sdev^2), 1)
  plot(p$x[, 1], p$x[, 2], pch = c(16, 17)[as.integer(meta$treatment)],
       col = hcl.colors(4, "Zissou1")[as.integer(factor(meta$time_num))], cex = 1.6,
       xlab = sprintf("PC1 (%.1f%%)", v[1]), ylab = sprintf("PC2 (%.1f%%)", v[2]), main = main)
  legend("topright", bty = "n", cex = 0.7, legend = levels(meta$treatment), pch = c(16, 17))
}
op <- par(mfrow = c(1, 2))
pca_plot(cad_qc_r$vst,       cad$meta, "Cadenza RNA (VST)")
pca_plot(cad_qc_p$norm,      cad$meta, "Cadenza protein (normalised)")
par(op)
```

```{r}
#| label: cadenza-missingness

cad_mi <- run_missingness(cad_qc_p$norm, cad$meta)
cat(sprintf("MNAR: %.1f%%  MAR: %.1f%%  (of all entries)\n",
           100 * cad_mi$mnar_frac, 100 * cad_mi$mar_frac))
```

```{r}
#| label: cadenza-de

## protein_id == gene_id within a variety (verified in ../../data/real/
## *_protein_gene_mapping.csv), so Universe B == QC'd protein genes that also
## survive RNA QC -- no bipartite mapping ambiguity for this dataset.
cad_de_rna  <- run_de(cad_qc_r$counts, cad$meta, TRUE)
cad_common  <- intersect(rownames(cad_mi$mixed), rownames(cad_qc_r$counts))
cad_de_prot <- run_de(cad_mi$mixed[cad_common, , drop = FALSE], cad$meta, FALSE)

cat(sprintf("Universe B: %d matched genes\n", length(cad_common)))
cat(sprintf("RNA any-effect FDR<0.05: %d\n",  sum(cad_de_rna$any_fdr[cad_common] < 0.05)))
cat(sprintf("Protein any-effect FDR<0.05: %d\n", sum(cad_de_prot$any_fdr < 0.05)))
```

```{r}
#| label: cadenza-archetypes

cad_Rl <- cad_de_rna$lfc[cad_common, ];  cad_Rs <- cad_de_rna$se[cad_common, ]
cad_Pl <- cad_de_prot$lfc;               cad_Ps <- cad_de_prot$se
cad_Rl <- cad_Rl - cad_Rl[, 1]; cad_Pl <- cad_Pl - cad_Pl[, 1]

cad_arch <- run_kinetic_archetypes(cad_Rl, cad_Rs, cad_Pl, cad_Ps, DESIGN$timepoints)
knitr::kable(as.data.frame(table(archetype = cad_arch$archetype)),
            caption = "Cadenza: profile-likelihood archetype calls")
```

## 6. Hierarchical Bayesian kinetics: Cadenza

```{r}
#| label: cadenza-bayesian-kinetics

## RNA-responsive gate (mirrors 05's rna_resp / the doc note in
## 07_bayesian_kinetics.R): the amplitude model is only meaningful when RNA
## carries real signal -- for a flat-RNA gene ANY amplitude "explains" a tiny
## predicted change equally badly, which (confirmed in dev testing on this
## exact dataset) can otherwise drive a gene to the numerical safety bound.
cad_rna_resp <- (cad_de_rna$any_fdr[cad_common] < 0.05) &
  (apply(abs(cad_Rl[, -1, drop = FALSE]), 1, max) > 0.5)
cat(sprintf("RNA-responsive, eligible for amplitude fitting: %d / %d\n",
           sum(cad_rna_resp), length(cad_rna_resp)))

cad_fit <- fit_hierarchical_kinetics(
  cad_Rl[cad_rna_resp, ], cad_Rs[cad_rna_resp, ],
  cad_Pl[cad_rna_resp, ], cad_Ps[cad_rna_resp, ],
  ## Anchored at the SAME joint-optimal half-life the profile-likelihood fit
  ## above already found (t_half_M1_h), so the two analyses share one kinetic
  ## fit and differ only in how the amplitude is inferred (independent
  ## per-gene LRT vs. hierarchical horseshoe shrinkage).
  t_half = cad_arch$t_half_M1_h[match(rownames(cad_Rl)[cad_rna_resp], cad_arch$gene_id)],
  tps = DESIGN$timepoints, n_iter = 2000, warmup = 800, seed = 1, verbose = FALSE)

cat(sprintf("MCMC acceptance rate: %.2f\n", cad_fit$accept_delta))
cat(sprintf("Genes at the numerical safety bound (report as directional, not exact): %d / %d\n",
           sum(cad_fit$summary$at_bound), nrow(cad_fit$summary)))
cat(sprintf("Regulated by hierarchical model (95%% CI excludes a=1): %d / %d\n",
           sum(cad_fit$summary$regulated_95), nrow(cad_fit$summary)))
cat(sprintf("Regulated by profile-likelihood LRT (FDR<0.05), same genes: %d / %d\n",
           sum(cad_arch$regulated[match(rownames(cad_Rl)[cad_rna_resp], cad_arch$gene_id)]),
           sum(cad_rna_resp)))
```

```{r}
#| label: fig-cadenza-kinetic-volcano
#| fig-cap: "Cadenza kinetic volcano: hierarchical posterior amplitude vs. profile-likelihood significance"

idx <- match(cad_fit$summary$gene_id, cad_arch$gene_id)
plot(cad_fit$summary$a_mean, -log10(pmax(cad_arch$lrt_p[idx], 1e-12)), pch = 16, cex = 0.4,
     col = ifelse(cad_fit$summary$regulated_95, "#D1495B99", "#00000033"),
     xlab = "hierarchical posterior mean amplitude (a)",
     ylab = "-log10 p (profile-likelihood LRT)",
     main = "Cadenza: two independent regulation signals\n(red = hierarchical model flags as regulated)")
abline(v = 1, lty = 3)
```

The two methods agree on genes with strong signal and differ most on genes the profile-likelihood LRT was underpowered for individually (§`docs/PIPELINE.md` documents this trade-off quantitatively on simulated ground truth: the hierarchical model roughly triples recall at the cost of a higher false-call rate on truly-unregulated genes, since fixing the half-life removes one source of legitimate uncertainty from its interval). Genes flagged by **both** methods are the highest-confidence regulated set; genes flagged by the hierarchical model **only** are candidates for follow-up validation, not standalone claims.

```{r}
#| label: cadenza-agreement-table

cad_kin <- data.frame(
  gene_id = rownames(cad_Rl)[cad_rna_resp],
  lrt_regulated  = cad_arch$regulated[match(rownames(cad_Rl)[cad_rna_resp], cad_arch$gene_id)],
  bayes_regulated = cad_fit$summary$regulated_95,
  a_mean = cad_fit$summary$a_mean, at_bound = cad_fit$summary$at_bound,
  stringsAsFactors = FALSE)
knitr::kable(as.data.frame(table(LRT = cad_kin$lrt_regulated, Bayesian = cad_kin$bayes_regulated)),
            caption = "Cadenza: agreement between the two regulation calls")
```

## 7. Case B integration: Cadenza

```{r}
#| label: cadenza-integration

cad_int <- run_integration(cad_qc_r$vst, cad_mi$mixed, cad$meta, cad$meta)
knitr::kable(data.frame(component = seq_along(cad_int$q2), Q2 = round(cad_int$q2, 3),
                        perm_q95 = round(apply(cad_int$perm_q2, 2, quantile, .95), 3),
                        p_value = signif(cad_int$p_q2, 3)),
            caption = "Cadenza: leave-one-cell-out Q2 vs permutation null")
```

## 8. Run the pipeline: Norin

Identical steps, applied to the second variety.

```{r}
#| label: norin-load

nor <- load_variety("norin")
knitr::kable(table(nor$meta$treatment, nor$meta$time_num),
            caption = "Norin design: samples per treatment x timepoint")

nor_qc_r <- run_qc_rna(nor$rna, nor$meta)
nor_qc_p <- run_qc_prot(nor$prot, nor$meta)
cat(sprintf("RNA: %d / %d genes kept\n", nor_qc_r$n_kept, nor_qc_r$n_in))
cat(sprintf("Protein: %d / %d groups kept; missing rate %.1f%%\n",
           nor_qc_p$n_kept, nor_qc_p$n_in, 100 * nor_qc_p$miss_rate))

nor_mi <- run_missingness(nor_qc_p$norm, nor$meta)
cat(sprintf("MNAR: %.1f%%  MAR: %.1f%%\n", 100 * nor_mi$mnar_frac, 100 * nor_mi$mar_frac))
```

```{r}
#| label: norin-de-archetypes

nor_de_rna  <- run_de(nor_qc_r$counts, nor$meta, TRUE)
nor_common  <- intersect(rownames(nor_mi$mixed), rownames(nor_qc_r$counts))
nor_de_prot <- run_de(nor_mi$mixed[nor_common, , drop = FALSE], nor$meta, FALSE)
cat(sprintf("Universe B: %d matched genes\n", length(nor_common)))

nor_Rl <- nor_de_rna$lfc[nor_common, ];  nor_Rs <- nor_de_rna$se[nor_common, ]
nor_Pl <- nor_de_prot$lfc;               nor_Ps <- nor_de_prot$se
nor_Rl <- nor_Rl - nor_Rl[, 1]; nor_Pl <- nor_Pl - nor_Pl[, 1]

nor_arch <- run_kinetic_archetypes(nor_Rl, nor_Rs, nor_Pl, nor_Ps, DESIGN$timepoints)
knitr::kable(as.data.frame(table(archetype = nor_arch$archetype)),
            caption = "Norin: profile-likelihood archetype calls")
```

```{r}
#| label: norin-bayesian-kinetics

nor_rna_resp <- (nor_de_rna$any_fdr[nor_common] < 0.05) &
  (apply(abs(nor_Rl[, -1, drop = FALSE]), 1, max) > 0.5)
cat(sprintf("RNA-responsive, eligible for amplitude fitting: %d / %d\n",
           sum(nor_rna_resp), length(nor_rna_resp)))

nor_fit <- fit_hierarchical_kinetics(
  nor_Rl[nor_rna_resp, ], nor_Rs[nor_rna_resp, ],
  nor_Pl[nor_rna_resp, ], nor_Ps[nor_rna_resp, ],
  t_half = nor_arch$t_half_M1_h[match(rownames(nor_Rl)[nor_rna_resp], nor_arch$gene_id)],
  tps = DESIGN$timepoints, n_iter = 2000, warmup = 800, seed = 1, verbose = FALSE)

cat(sprintf("MCMC acceptance rate: %.2f\n", nor_fit$accept_delta))
cat(sprintf("Regulated by hierarchical model: %d / %d\n",
           sum(nor_fit$summary$regulated_95), nrow(nor_fit$summary)))
```

```{r}
#| label: norin-integration

nor_int <- run_integration(nor_qc_r$vst, nor_mi$mixed, nor$meta, nor$meta)
knitr::kable(data.frame(component = seq_along(nor_int$q2), Q2 = round(nor_int$q2, 3),
                        perm_q95 = round(apply(nor_int$perm_q2, 2, quantile, .95), 3),
                        p_value = signif(nor_int$p_q2, 3)),
            caption = "Norin: leave-one-cell-out Q2 vs permutation null")
```

## 8b. A3 diagnostic: does control-arm drift drive the `regulated` calls?

Placed here because it needs the fitted results from **both** varieties. See "Where the Kinetic Null Comes From" above for the derivation and what A3 states.

```{r}
#| label: a3-diagnostic-run

cad_a3 <- run_a3_diagnostic(cad_de_rna$ctrl_time, cad_Rl, cad_arch, cad_rna_resp, "Cadenza")
nor_a3 <- run_a3_diagnostic(nor_de_rna$ctrl_time, nor_Rl, nor_arch, nor_rna_resp, "Norin")

## Step 1: how badly is A3 violated in the first place? Quantify the size of
## the control-arm movement before asking whether it matters. A3 assumes the
## control arm is at steady state, i.e. that these numbers are ~0.
ctrl_drift_summary <- do.call(rbind, lapply(list(cad_a3, nor_a3), function(x)
  data.frame(variety = x$label, n_genes = x$n,
             median_abs_drift = round(median(x$data$drift), 2),
             pct_drift_over_1 = round(100 * mean(x$data$drift > 1), 1),
             pct_drift_over_2 = round(100 * mean(x$data$drift > 2), 1))))
knitr::kable(ctrl_drift_summary,
            caption = "Control-arm drift: max |log2FC| from baseline, CONTROL arm only. A3 assumes ~0.")
```

```{r}
#| label: a3-diagnostic-unadjusted

## Step 2: the unadjusted view. Reported for transparency, but it is NOT the
## answer -- drift correlates with treatment effect size, which independently
## raises power to reject the kinetic null.
knitr::kable(cbind(variety = "Cadenza", cad_a3$by_quartile),
            caption = "Cadenza: regulated rate by control-drift quartile (UNADJUSTED, confounded -- see next table)")
knitr::kable(cbind(variety = "Norin", nor_a3$by_quartile),
            caption = "Norin: regulated rate by control-drift quartile (UNADJUSTED, confounded)")
```

```{r}
#| label: a3-diagnostic-adjusted

## Step 3: the actual test. Odds ratio for control drift, ADJUSTED for the
## gene's own RNA response amplitude. An OR near 1 means that at a given
## treatment effect size, control-arm instability does not additionally
## predict being called regulated -- i.e. A3 violation is not manufacturing
## the calls.
a3_adj <- rbind(cbind(variety = "Cadenza", cad_a3$adjusted),
                cbind(variety = "Norin",   nor_a3$adjusted))
knitr::kable(a3_adj,
            caption = "ADJUSTED test: logistic regression of `regulated` on control drift + RNA amplitude (both z-scaled). Odds ratio per 1 SD.")

knitr::kable(data.frame(
  variety = c(cad_a3$label, nor_a3$label),
  spearman_raw = c(cad_a3$spearman_raw, nor_a3$spearman_raw),
  spearman_adjusted_for_rna_amp = c(cad_a3$spearman_adjusted, nor_a3$spearman_adjusted),
  collinearity_drift_vs_amp = c(cad_a3$collinearity, nor_a3$collinearity)),
  caption = "Correlation between control drift and |a - 1|, before/after adjusting for RNA amplitude. Last column: collinearity between the two predictors -- high values mean the adjusted odds ratios above cannot be cleanly separated.")
```

```{r}
#| label: fig-a3-diagnostic
#| fig-cap: "Control-arm drift vs amplitude deviation. A rising lowess trend would indicate A3 violation inflating the amplitude estimate."

op <- par(mfrow = c(1, 2))
for (x in list(cad_a3, nor_a3)) {
  d <- x$data
  plot(d$drift, pmin(abs(d$a - 1), 5), pch = 16, cex = 0.3,
       col = ifelse(d$regulated, "#D1495B55", "#00000022"),
       xlab = "control-arm drift (max |log2FC|)", ylab = "|a - 1|  (capped at 5)",
       main = sprintf("%s (n = %d)", x$label, x$n))
  ok <- is.finite(d$drift) & is.finite(d$a)
  lines(lowess(d$drift[ok], pmin(abs(d$a - 1), 5)[ok], f = 0.5), col = "#2C6FBB", lwd = 3)
  legend("topright", bty = "n", cex = 0.7, pch = c(16, NA), lwd = c(NA, 3),
         col = c("#D1495B", "#2C6FBB"), legend = c("called regulated", "lowess"))
}
par(op)
```

### What the diagnostic found

**A3 is clearly violated.** Median control-arm drift is \~0.67 log2FC in both varieties; roughly a third of RNA-responsive genes move more than 1 log2FC, and \~9--11% move more than 2 log2FC, in the *control* arm alone. The control is not at steady state, and the assumption underlying $dp/dt = k_d(r-p)$ does not hold as stated.

**But the impact on the amplitude estimate is small.** The continuous measure is the clearest: after adjusting for RNA response amplitude, the correlation between control drift and $|a-1|$ is 0.04 (Cadenza) and 0.07 (Norin) --- negligible. The lowess trends in the figure above are correspondingly flat. Whatever A3 violation is doing, it is not systematically inflating the amplitude deviation that drives the `regulated` call.

**The two varieties disagree on the binary call, and that disagreement is the honest headline.** Adjusted odds ratio for control drift, per 1 SD:

-   **Cadenza: 1.07 (95% CI 0.99--1.16, p = 0.09)** --- not significant, effect negligible.
-   **Norin: 1.31 (95% CI 1.20--1.44, p \< 1e-15)** --- significant, a real \~30% increase in the odds of being called regulated per SD of control drift.

So in Norin there *is* a detectable residual association, in the direction A3 violation predicts. It is modest in size, but it is not zero, and it does not replicate in Cadenza.

**Interpretation caveats, stated rather than glossed:**

-   Control drift and treatment amplitude are correlated by construction (Spearman 0.44--0.53) --- both measure "this gene is dynamic." That is moderate collinearity: enough that the two adjusted coefficients cannot be cleanly separated, not enough to make them meaningless. Note the RNA-amplitude coefficient itself flips sign between varieties (1.05 in Cadenza vs 0.76 in Norin), which is a symptom of exactly this.
-   This diagnostic tests for *association*, not *causation*. A gene whose control arm is unstable may also be genuinely more post-transcriptionally regulated --- biology, not artefact. The test cannot separate those.
-   Absence of a strong association is evidence that A3 violation is not the *dominant* driver of the regulated calls. It is not proof that no individual call is affected.

**Recommended wording for the manuscript limitations section:** the kinetic null assumes a control arm at quasi-steady state; this assumption is measurably violated in this experiment (median control-arm drift \~0.67 log2FC over 72 h). A diagnostic regression adjusting for treatment effect size found the residual association between control-arm drift and post-transcriptional regulation calls to be negligible in Cadenza (OR 1.07, p = 0.09) and modest but significant in Norin (OR 1.31, p \< 0.001). Regulation calls for genes with large control-arm drift in Norin should therefore be treated as lower-confidence.

## 9. Cross-variety validation via reciprocal-best-hit orthologs

Neither variety has simulated ground truth, so this is the one source of genuine external validation available: **do the two independently-run analyses agree on orthologous genes?** Agreement is not guaranteed even for correct biology (Cadenza and Norin can genuinely differ), but strong, widespread *disagreement* on high-confidence calls would flag a methodology problem rather than real biological divergence.

```{r}
#| label: rbh-load

rbh <- read.csv(file.path(DATA_DIR, "norin_cadenza_rbh-clean.csv"))
cat(sprintf("RBH ortholog pairs: %d (%d unique Norin, %d unique Cadenza)\n",
           nrow(rbh), length(unique(rbh$query)), length(unique(rbh$target))))

## Keep 1:1 RBH pairs only -- a gene appearing on both sides of more than one
## row is an ambiguous ortholog call and is excluded rather than guessed at.
rbh_1to1 <- rbh[!(duplicated(rbh$query) | duplicated(rbh$query, fromLast = TRUE)) &
                !(duplicated(rbh$target) | duplicated(rbh$target, fromLast = TRUE)), ]
cat(sprintf("1:1 RBH pairs: %d\n", nrow(rbh_1to1)))
```

```{r}
#| label: rbh-concordance

## Restrict to ortholog pairs where BOTH varieties had the gene in their own
## Universe B (matched RNA+protein, QC-passing) archetype table.
pair <- merge(rbh_1to1, cad_arch, by.x = "target", by.y = "gene_id")
pair <- merge(pair, nor_arch, by.x = "query", by.y = "gene_id", suffixes = c("_cad", "_nor"))
cat(sprintf("Ortholog pairs testable in both varieties: %d\n", nrow(pair)))

knitr::kable(table(Cadenza = pair$archetype_cad, Norin = pair$archetype_nor),
            caption = "Archetype agreement across 1:1 orthologs (Cadenza rows, Norin columns)")

## Amplitude correlation among genes both varieties called RNA-responsive --
## the most direct real-data check on whether the kinetic model is measuring
## something reproducible rather than dataset-specific noise.
both_resp <- pair$rna_responds_cad & pair$rna_responds_nor
cat(sprintf("\nBoth-variety RNA-responsive ortholog pairs: %d\n", sum(both_resp)))
if (sum(both_resp) > 10) {
  cc <- cor(pair$amplitude_a_cad[both_resp], pair$amplitude_a_nor[both_resp],
           method = "spearman", use = "complete.obs")
  cat(sprintf("Spearman(amplitude_a Cadenza, amplitude_a Norin) = %.2f\n", cc))
}

## The continuous amplitude correlation above is expected to be weak even for
## real, reproducible biology: each variety's per-gene amplitude is itself a
## noisy n=3 point estimate (the same reason §concordance-by-timepoint in
## 05_concordance_archetypes.R reports a DISATTENUATED correlation rather than
## the raw one -- two noisy proxies for the same true quantity correlate far
## more weakly than the true quantities do). The CATEGORICAL archetype call is
## more robust to that per-gene noise, so test it directly instead of reading
## the near-zero continuous correlation as "no reproducibility":
chisq <- suppressWarnings(chisq.test(table(pair$archetype_cad, pair$archetype_nor)))
cramers_v <- sqrt(unname(chisq$statistic) / (nrow(pair) * (min(dim(table(pair$archetype_cad, pair$archetype_nor))) - 1)))
cat(sprintf("\nChi-sq test, archetype agreement across orthologs: X2=%.1f, df=%d, p=%s\n",
           chisq$statistic, chisq$parameter, format.pval(chisq$p.value, digits = 3)))
cat(sprintf("Cramer's V: %.3f  (modest but real association, not attributable to chance)\n",
           cramers_v))
```

```{r}
#| label: fig-rbh-amplitude-concordance
#| fig-cap: "Cross-variety amplitude concordance for shared orthologs"

if (sum(both_resp) > 10) {
  plot(pair$amplitude_a_cad[both_resp], pair$amplitude_a_nor[both_resp], pch = 16, cex = 0.4,
       col = "#2C6FBB66", xlab = "amplitude a (Cadenza)", ylab = "amplitude a (Norin)",
       main = "Ortholog amplitude concordance,\nboth varieties RNA-responsive")
  abline(0, 1, col = "#D1495B", lwd = 2); abline(h = 1, v = 1, lty = 3, col = "grey60")
}
```

## 10. Summary and limitations

```{r}
#| label: summary-table

summ <- data.frame(
  variety = c("Cadenza", "Norin"),
  rna_genes_kept   = c(cad_qc_r$n_kept, nor_qc_r$n_kept),
  protein_kept     = c(cad_qc_p$n_kept, nor_qc_p$n_kept),
  universe_b       = c(length(cad_common), length(nor_common)),
  rna_responsive   = c(sum(cad_rna_resp), sum(nor_rna_resp)),
  lrt_regulated    = c(sum(cad_arch$regulated), sum(nor_arch$regulated)),
  bayes_regulated  = c(sum(cad_fit$summary$regulated_95), sum(nor_fit$summary$regulated_95)),
  integration_q2_c1 = c(round(cad_int$q2[1], 3), round(nor_int$q2[1], 3)),
  integration_p_c1  = c(signif(cad_int$p_q2[1], 3), signif(nor_int$p_q2[1], 3)))
knitr::kable(summ, caption = "Pipeline summary, both varieties")
```

**Read before using any downstream number:**

-   **Timepoint units are unconfirmed** (§ "Two things this dataset does not tell us"). Every half-life in this notebook is in `r TIME_UNIT_LABEL`, not hours, until `TIMEPOINT_VALUES` is corrected to the real sampling schedule.
-   **Treatment reference is assumed**, not confirmed from metadata beyond the `T0`/`T1` labels themselves.
-   **Protein QC has no decoy/contaminant/unique-peptide filter** here (§2): the provided `*_protein_gene_mapping.csv` files do not carry that MaxQuant/ DIA-NN metadata. If the original search-engine output is available, rerun QC with `../../R/01_qc_normalise.R`'s full filter instead.
-   **The hierarchical Bayesian model fixes each gene's half-life** at its own profile-likelihood joint optimum rather than sampling it jointly with amplitude (see `07_bayesian_kinetics.R`'s header for why: jointly sampling both produces a slow-mixing ridge). This sacrifices joint kinetic/amplitude uncertainty propagation -- its credible intervals should be read as "amplitude uncertainty at the point-estimate kinetics," not full posterior uncertainty.
-   **\~2% of RNA-responsive genes hit the model's numerical safety bound** (`at_bound = TRUE` in the summary tables) even after gating to RNA- responsive genes. These are real, strong-signal genes, not degenerate artefacts (verified during development) -- but their reported amplitude should be read as directional ("at least this extreme"), not a precise estimate.
-   **No ground truth exists for real data.** §9's cross-variety ortholog concordance is the best available external check, not proof of correctness -- Cadenza and Norin are permitted to genuinely differ.

### Next steps

-   Confirm `TREATMENT_REF` and `TIMEPOINT_VALUES` against the actual experimental record and re-render.
-   If the original MaxQuant/DIA-NN search output is available, rerun protein QC through the full `01_qc_normalise.R` filter (decoys, contaminants, unique-peptide count).
-   `docs/planning.md` §4B-4E (GNN residuals, translation-delay atlas, XGBoost/SHAP buffering predictors) remain unbuilt; evaluate against the existing `rnaprot/` codebase on `nectar` before building new versions (see project memory `proj_existing_nectar_rnaprot.md`).
