---
title: "DE proteomics - wheat"
author: "Kristina Gagalova"
format:
  gfm:
    toc: true
    code-tools: true
    fig-dir: "figs"
execute:
  echo: true
editor: visual
---

## Overview

Integrated RNA-seq / quantitative proteomics analysis of a wheat time-course treatment experiment, in two varieties (**Cadenza**, **Norin**), each with an independent **2 treatments x 4 timepoints x 3 replicates = 24 samples per omics layer** design.

**RNA and protein are measured on different biological samples within each variety** (unpaired / "Case B" design, per `../../CLAUDE.md` and `../../docs/PIPELINE.md`) --- not split aliquots of the same 24 samples. That rules out sample-level factor models (MOFA+/MEFISTO/DIABLO/O2PLS) and shapes every method choice below; see `../../docs/PIPELINE.md` and `../../docs/planning.md` for the full reasoning.

This notebook runs the complete pipeline end to end on the real data in `../../data/real/`:

1.  Data audit and design check
2.  QC & normalisation (RNA: `filterByExpr` + VST/voom; protein: validity filter + normalisation, adapted for this dataset -- see the note in §2)
3.  ID mapping (trivial here: `protein_id == gene_id` within each variety)
4.  Missingness diagnosis (MNAR/MAR) and imputation
5.  Temporal differential expression (`limma`, both layers)
6.  Kinetic-null / amplitude classification (the profile-likelihood model from `05_concordance_archetypes.R`)
7.  **Hierarchical Bayesian kinetics** (`07_bayesian_kinetics.R`) --- the new piece this notebook exists to run: a horseshoe-shrinkage model that rescues genes the profile-likelihood method above cannot classify on its own (the `kinetics_limited` problem this directory is named for)
8.  Case B design-cell integration (leave-one-cell-out Q\^2 + permutation null)
9.  Cross-variety validation using reciprocal-best-hit orthologs --- the one piece of genuine external validation available for real (non-simulated) data in this project

Steps 2--8 are defined once as reusable functions and applied to each variety; step 9 compares the two varieties' independent results against each other.

------------------------------------------------------------------------

# Conclusions

**Read this section before the analysis.** Every assumption behind this pipeline has been tested in the companion notebook [`assumptions_validation.qmd`](assumptions_validation.qmd); five are violated and one is unresolved. The results below are real, but they support **population-level conclusions and not individual-gene claims**, and that distinction determines what can go in a manuscript.

## What this analysis establishes

**1. RNA and protein responses are coupled across the infection time course, and the coupling is statistically demonstrable despite unpaired samples.**

Leave-one-design-cell-out cross-validation, tested against a permutation null: given the transcriptome of a treatment × timepoint cell the model has never seen, it predicts that cell's proteome better than chance.

| Variety | Q² (1 component) | permutation p |
|:--------|:-----------------|:--------------|
| Cadenza | **0.269**        | 0.010         |
| Norin   | **0.084**        | 0.020         |

This is the most robust result in the project. It cannot be inflated by the in-sample correlation artefact that makes naive cross-block correlation meaningless at this sample size (a PLS on 8 pseudo-samples reaches r ≈ 0.72 on *permuted* data), because Q² is computed on held-out cells.

**2. Post-transcriptional regulation is present, and its magnitude is larger than measurement error alone explains.**

A substantial fraction of RNA-responsive genes have protein trajectories that depart from what their own transcript trajectory predicts under first-order kinetics. This holds in both varieties, analysed independently.

**3. The two varieties differ, and Cadenza is the better-behaved dataset.** Cadenza has a clean baseline (0.4% of genes differentially expressed at 0 hpi, when infection cannot yet have acted), higher predictive coupling, and no significant association between control-arm drift and regulation calls. Norin fails all three (5.8% baseline DE; Q² = 0.084; drift OR 1.31, p \< 1e-15). Where they disagree, weight Cadenza.

**4. Independent analyses of the two varieties agree on orthologous genes above chance** (χ² = 88.2, p = 3.7e-15 for archetype agreement). The effect is modest (Cramér's V = 0.14) but it cannot be an artefact of the procedure --- the varieties share no samples and were processed separately.

## What this analysis does *not* establish

**Per-gene regulation calls are not reliable.** The model treats the RNA trajectory as known exactly (`se_rna` is accepted by `fit_hierarchical_kinetics()` and never used). Propagating limma's own RNA standard errors by parametric bootstrap flips the regulated / not-regulated call for **36% of Cadenza genes and 49% of Norin genes**. None of that uncertainty appears in the reported credible intervals. Treat any named gene as a hypothesis requiring orthogonal validation (PRM, western, qPCR).

**The count of regulated genes is soft.** It moves 3× across plausible measurement-error assumptions (1457 → 902 → 488 as protein SEs are scaled 0.67× → 1× → 1.5×), \~30% across imputation scheme, and 25% from the MCMC convergence fix alone. Always quote it with its settings.

**Half-lives are mostly unresolvable.** Only **23--27%** of fitted genes fall in the band this design can estimate; \~50% sit above the 72 h identifiability bound and \~24% at the 12 h grid floor. A reported `t_half` of 12 h means "≤ 12 h"; above 72 h means "slower than we can see". **The fitted `t_half` distribution is a property of the sampling window, not of the wheat proteome.**

**Mechanism cannot be assigned.** The amplitude parameter `a` conflates changed translation with changed degradation --- absolute `k_s` cancels algebraically in fold-change space and is not identifiable from this data (derivation in "Where the Kinetic Null Comes From", below). Write "departs from its RNA-predicted trajectory", never "translationally repressed".

**Raw DE counts are pending reconciliation.** This pipeline reports 2--6× fewer differentially expressed genes than the project's standalone limma workflow on counts that match exactly (see `assumptions_validation.qmd` §2b). Until that is resolved, no DE count from either source should be published.

## Assumption status

| Assumption                 | Status                | Evidence                                   |
|:---------------------------|:----------------------|:-------------------------------------------|
| MNAR-dominant missingness  | **Holds**             | Spearman −0.74 / −0.75                     |
| Horseshoe prior scale      | **Holds**             | Insensitive, `p0_frac` 0.05--0.40          |
| MCMC convergence           | **Holds**             | 0% R-hat \> 1.05, ESS ≈ 1600               |
| Protein variance known     | Partly                | Amplitude invariant; call count varies 3×  |
| A0 baseline equivalence    | **Asymmetric**        | Cadenza 0.4%, Norin 5.8% at 0 hpi          |
| A3 control at steady state | **Violated**          | Median control drift ≈ 0.67 log2FC         |
| A1 first-order kinetics    | **Violated**          | Structured residuals, opposite per variety |
| B5 gene independence       | **Violated**          | Residual PC1 62--69% vs 25% expected       |
| B4 RNA known exactly       | **Severely violated** | 36% / 49% of calls flip                    |
| DE reconciliation          | **Unresolved**        | 2--6× disagreement, cause unknown          |

## Recommended framing for the manuscript

> Integrating unpaired transcriptome and proteome time courses at the design-cell level, we find statistically significant predictive coupling between the two layers (leave-one-cell-out Q² = 0.27, permutation p = 0.01 in Cadenza), and evidence that a substantial fraction of the proteome departs from transcript-predicted trajectories. Because the design provides three biological replicates and four timepoints over 72 h, and because RNA measurement error is not propagated into the kinetic model, these findings support population-level conclusions about the extent of post-transcriptional regulation but do not support confident per-gene classification; individual candidates require orthogonal validation.

## What would most improve this

1.  **Resolve the DE reconciliation** (`assumptions_validation.qmd` §2b) --- upstream of everything, and a bug hunt rather than a modelling question.
2.  **Propagate RNA uncertainty** into the kinetic model --- converts per-gene calls from unusable to usable; the single highest-value fix.
3.  **Investigate Norin's 0 hpi baseline** --- plausibly upstream of Norin's other anomalies.
4.  **Orthogonally validate 5--10 targets** --- the only route to a publishable gene-level claim given the above.

------------------------------------------------------------------------

### Design constants

Both of these were originally unknown from `data/real/*_metadata.csv` alone (which encodes only `T0`/`T1` and `t0`..`t3`) and have since been resolved against the full RNA-seq sample sheet:

```{r}
#| label: edit-me-assumptions

## 1. WHICH TREATMENT IS THE REFERENCE?  -- CONFIRMED
##    data/real/*_metadata.csv encodes only "T0"/"T1". Cross-referencing the
##    sample IDs against the full RNA-seq sample sheet (which carries the
##    real labels) resolves it: LH4 is Cadenza/NEG/0h there and T0/t0 here,
##    LH10 is Cadenza/PN143/0h there and T1/t0 here. So:
##        T0 = NEG   = uninfected control  -> reference level
##        T1 = PN143 = infected            -> treatment
##    Flipping this mirrors every log2FC and every kinetic direction
##    (amplified <-> buffered), so it is worth the cross-check.
TREATMENT_REF <- "T0"

## 2. WHAT IS THE REAL TIME SPACING?  -- CONFIRMED
##    The same sample sheet carries an `hour` column: t0/t1/t2/t3 are
##    0 / 24 / 48 / 72 hours (i.e. 0-3 dpi), evenly spaced at 24 h.
##    Every half-life reported downstream is therefore in HOURS.
TIMEPOINT_VALUES <- c(0, 24, 48, 72)
TIME_UNIT_LABEL  <- "hours"
```

Note the consequence for the kinetic model: the half-life search grid starts at `0.5 * 24 = 12 h`, so **no half-life below 12 h can be resolved** by this design. Proteins turning over faster than that are reported at the grid floor.

## 1. Setup

```{r}
#| label: setup
#| message: false

## Quarto's default execution directory is this file's own folder
## (analysis/kinetics_limited/), while every R/*.R script in this project
## assumes the working directory is the PROJECT ROOT (matching how
## run_all.sh invokes them). Rather than fight that mismatch with a
## working-directory switch (fragile: root.dir changes how KNITR evaluates
## chunks, but does not change what a *sourced* script sees via
## sys.frame()$ofile, which is empty inside a knitr chunk anyway), every
## path below is resolved with here::here(), which walks up from the
## current directory looking for a project marker file -- this project has
## four (IntegrateProtRNA.Rproj, DESCRIPTION, renv.lock, .git) -- and returns
## an absolute path that is correct no matter where code is executed from.
## need_pkgs() itself is defined inside utils.R, so "here" -- the one package
## every subsequent path depends on -- is checked directly first.
if (!requireNamespace("here", quietly = TRUE))
  stop("package 'here' is required; install with install.packages('here')")
source(here::here("R", "utils.R"))
source(here::here("R", "pls_utils.R"))
source(here::here("R", "07_bayesian_kinetics.R"))
need_pkgs(c("limma", "edgeR", "DESeq2", "matrixStats", "impute", "here"))

set.seed(20260822)
DATA_DIR <- here::here("data", "real")

DESIGN <- list(
  treatments = c(TREATMENT_REF, setdiff(c("T0", "T1"), TREATMENT_REF)),
  reference  = TREATMENT_REF,
  timepoints = TIMEPOINT_VALUES,
  n_reps     = 3
)
CELLS <- as.vector(t(outer(DESIGN$treatments, DESIGN$timepoints,
                          function(a, b) sprintf("%s_t%s", a, b))))
```

## 2. Loading and QC (reusable functions)

```{r}
#| label: fn-load

#' Load one variety's RNA, protein, metadata and gene mapping.
#'
#' Metadata is a single file per variety, shared across both omics layers'
#' column headers -- the LH-style sample IDs happen to coincide between the
#' RNA and protein matrices for a given variety (both trace back to the same
#' field-plot label), but per the Case B design this is NOT treated as
#' sample-level pairing anywhere downstream: only variety x treatment x
#' timepoint CELL membership is used to link the two layers (see
#' `../../R/06_integration_caseB.R`'s header for why).
load_variety <- function(prefix) {

  data_dir <- here::here("data", "real")

  meta <- read.csv(
    file.path(data_dir, paste0(prefix, "_metadata.csv")),
    row.names = 1
  )

  meta$time_num  <- TIMEPOINT_VALUES[
    match(meta$timepoint, sort(unique(meta$timepoint)))
  ]
  meta$treatment <- factor(meta$treatment, levels = DESIGN$treatments)
  meta$cell <- factor(
    sprintf("%s_t%s", meta$treatment, meta$time_num),
    levels = CELLS
  )

  rna <- as.matrix(
    read.csv(
      file.path(data_dir, paste0(prefix, "-rnaseq.csv")),
      row.names = 1,
      check.names = FALSE
    )
  )

  prot <- as.matrix(
    read.csv(
      file.path(data_dir, paste0(prefix, "-prot.csv")),
      row.names = 1,
      check.names = FALSE
    )
  )

  gmap <- read.csv(
    file.path(data_dir, paste0(prefix, "_protein_gene_mapping.csv"))
  )

  stopifnot(
    all(colnames(rna) %in% rownames(meta)),
    all(colnames(prot) %in% rownames(meta))
  )

  rna  <- rna[, rownames(meta), drop = FALSE]
  prot <- prot[, rownames(meta), drop = FALSE]

  list(
    rna = rna,
    prot = prot,
    meta = meta,
    gmap = gmap
  )
}
```

```{r}
#| label: fn-qc

#' RNA QC: filterByExpr, then VST (integration matrix) + voom-ready counts.
#' Same procedure as `01_qc_normalise.R`, applied to real counts directly.
run_qc_rna <- function(counts, meta) {
  ## Manual low-expression filter (DESeq2-style threshold), replacing
  ## edgeR::filterByExpr()'s adaptive heuristic: keep a gene if it has
  ## >= 10 counts in at least as many samples as the smallest design group.
  dge <- edgeR::DGEList(counts = counts)
  smallestGroupSize <- DESIGN$n_reps   # 3
  keep <- rowSums(dge$counts >= 10) >= smallestGroupSize
  dge  <- dge[keep, , keep.lib.sizes = FALSE]
  cf   <- dge$counts

  dge <- edgeR::calcNormFactors(dge)
  logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 3)
  dds <- DESeq2::DESeqDataSetFromMatrix(cf, meta, ~cell)
  dds <- DESeq2::estimateSizeFactors(dds)
  vst <- SummarizedExperiment::assay(
    DESeq2::vst(dds, blind = TRUE, nsub = min(1000, sum(rowMeans(cf) > 5))))
  list(counts = cf, logcpm = logcpm, vst = vst, n_in = nrow(counts), n_kept = nrow(cf))
}

#' Protein QC, adapted from `01_qc_normalise.R` for this dataset:
#' the real protein_gene_mapping.csv carries no unique-peptide/decoy/
#' contaminant columns (unlike the simulator's MaxQuant-style metadata), so
#' that part of 01's filtering is not applicable here and is skipped
#' explicitly rather than silently omitted. Zero is treated as "not
#' quantified" (standard LFQ-intensity convention), not a true biological
#' zero -- it becomes NA before the validity filter and normalisation.
run_qc_prot <- function(intensity, meta, min_valid_in_cell = 2) {

  # Zero means "not quantified", not a biological zero
  m <- intensity
  m[m == 0] <- NA
  m <- log2(m)

  # Validity filter: >= n valid values in AT LEAST ONE design cell.
  # A global ">= 50% valid" rule would delete genuine on/off biology.
  valid_per_cell <- vapply(
    levels(meta$cell),
    function(cl) rowSums(!is.na(m[, meta$cell == cl, drop = FALSE])),
    numeric(nrow(m))
  )

  keep <- matrixStats::rowMaxs(valid_per_cell) >= min_valid_in_cell
  mf   <- m[keep, , drop = FALSE]

  # Median normalisation: shift each sample so its median matches the
  # matrix-wide median. Corrects per-run loading, not biology.
  shift <- apply(mf, 2, median, na.rm = TRUE) - median(mf, na.rm = TRUE)
  mn    <- sweep(mf, 2, shift, "-")

  list(
    norm      = mn,
    n_in      = nrow(intensity),
    n_kept    = nrow(mf),
    miss_rate = mean(is.na(mn))
  )
}
```

```{r}
#| label: fn-missingness

#' MNAR/MAR diagnosis and mixed imputation, same rule as `03_missingness.R`:
#' a protein missing in EVERY replicate of a cell (but observed in some other
#' cell) is MNAR (left-censored -- genuinely low/absent there); missing in
#' only SOME replicates of its cell is MAR (stochastic). Imputed values are
#' for Universe A / integration use only -- the DE and kinetics steps below
#' use these imputed values too (protein has no complete-case alternative at
#' this scale), but the up-front validity filter (>=2 valid in >=1 cell)
#' keeps the imputation from manufacturing pure noise.
run_missingness <- function(prot_norm, meta) {
  cells <- levels(meta$cell)
  is_na <- is.na(prot_norm)
  n_obs_cell <- vapply(cells, function(cl)
    rowSums(!is.na(prot_norm[, meta$cell == cl, drop = FALSE])), numeric(nrow(prot_norm)))
  lab <- matrix("obs", nrow(prot_norm), ncol(prot_norm), dimnames = dimnames(prot_norm))
  for (k in seq_along(cells)) {
    ii <- which(meta$cell == cells[k]); gone <- n_obs_cell[, k] == 0
    lab[gone, ii] <- "MNAR"; lab[!gone, ii][is_na[!gone, ii]] <- "MAR"
  }
  lab[!is_na] <- "obs"

  mar_imp <- suppressWarnings(
    impute::impute.knn(prot_norm, k = 10, rowmax = 0.95, colmax = 0.95)$data)
  mnar_imp <- prot_norm
  for (j in seq_len(ncol(prot_norm))) {
    v <- prot_norm[, j]; nas <- is.na(v)
    if (!any(nas)) next
    mu <- mean(v, na.rm = TRUE); sdv <- sd(v, na.rm = TRUE)
    ## Perseus-style down-shifted normal: width 0.3, downshift 1.8 SD.
    mnar_imp[nas, j] <- rnorm(sum(nas), mu - 1.8 * sdv, 0.3 * sdv)
  }
  mixed <- prot_norm
  mixed[lab == "MAR"]  <- mar_imp[lab == "MAR"]
  mixed[lab == "MNAR"] <- mnar_imp[lab == "MNAR"]
  stopifnot(!anyNA(mixed))
  list(mixed = mixed, labels = lab,
       mnar_frac = mean(lab == "MNAR"), mar_frac = mean(lab == "MAR"))
}
```

```{r}
#| label: fn-de

#' Temporal DE, both layers, same design as `04_univariate_temporal_de.R`:
#' an 8-level ~0+cell design, per-timepoint treatment contrasts, and an
#' any-effect F-test. limma is used for both RNA (voom) and protein (trend +
#' robust eBayes) so effect sizes are on a comparable scale.
run_de <- function(mat, meta, is_rna) {
  ## Treatment-contrast parameterisation (~treatment*time), matching your
  ## limma workflow, in place of the ~0+cell design used before. The two are
  ## equivalent full-rank codings of the same means model: the per-timepoint
  ## contrasts reconstructed below (main effect + interaction term) give the
  ## same logFC/SE as the T1_tX - T0_tX cell-means contrast this replaces.
  meta$treatment <- relevel(factor(meta$treatment, levels = DESIGN$treatments),
                            ref = DESIGN$reference)
  meta$time_f    <- relevel(factor(meta$time_num, levels = DESIGN$timepoints),
                            ref = as.character(DESIGN$timepoints[1]))
  design <- model.matrix(~ treatment * time_f, data = meta)
  colnames(design) <- make.names(colnames(design))

  if (is_rna) {
    dge <- edgeR::calcNormFactors(edgeR::DGEList(mat))
    v   <- limma::voom(dge, design, plot = FALSE)
    fit <- limma::lmFit(v, design)
  } else {
    fit <- limma::lmFit(mat, design)
  }

  ## Reconstruct the per-timepoint treatment contrasts (Infected - Control at
  ## each timepoint): the reference timepoint's contrast IS
  ## main effect; every later timepoint adds its interaction term.
  alt       <- setdiff(DESIGN$treatments, DESIGN$reference)
  main_coef <- paste0("treatment", alt)
  int_coefs <- paste0(main_coef, ".time_f", DESIGN$timepoints[-1])
  stopifnot(all(c(main_coef, int_coefs) %in% colnames(design)))  # catches a

                                                                  ## mismatch
                                                                  ## immediately

                                                                  ## downstream
  contrast_defs <- setNames(
    c(main_coef, paste(main_coef, int_coefs, sep = " + ")),
    paste0("tp_t", DESIGN$timepoints))
  cm <- limma::makeContrasts(contrasts = contrast_defs, levels = design)
  colnames(cm) <- names(contrast_defs)
  f2 <- limma::eBayes(limma::contrasts.fit(fit, cm), trend = TRUE, robust = TRUE)

  lfc  <- f2$coefficients; colnames(lfc) <- paste0("t", DESIGN$timepoints)
  se   <- f2$stdev.unscaled * sqrt(f2$s2.post); colnames(se) <- colnames(lfc)
  padj <- apply(f2$p.value, 2, p.adjust, "BH"); colnames(padj) <- colnames(lfc)

  ## Omnibus F-test across the 4 timepoint contrasts, same
  ## as.numeric() strips rownames, so reattach them explicitly.
  ff <- limma::classifyTestsF(f2, fstat.only = TRUE)
  p_any <- setNames(pf(as.numeric(ff), attr(ff, "df1"), attr(ff, "df2"), lower.tail = FALSE),
                    rownames(f2))

  ## CONTROL-ARM time effects. In a ~treatment*time design with Control as the
  ## reference treatment level, the bare `time_fX` coefficients are exactly the
  ## control arm's own change from baseline (Control at tX - Control at t0).
  ## These are not used for any treatment inference -- they are returned solely
  ## to test the quasi-steady-state assumption the kinetic null model rests on
  ## (assumption A3; see the A3 diagnostic section).
  time_coefs <- paste0("time_f", DESIGN$timepoints[-1])
  ctrl_time <- NULL
  if (all(time_coefs %in% colnames(fit$coefficients))) {
    ctrl_time <- fit$coefficients[, time_coefs, drop = FALSE]
    colnames(ctrl_time) <- paste0("t", DESIGN$timepoints[-1])
  }

  list(lfc = lfc, se = se, fdr = padj, any_fdr = p.adjust(p_any, "BH"),
       ctrl_time = ctrl_time)
}
```

## 3. Kinetic archetype classification (profile-likelihood)

> **Reference material.** §3--§4 document the model, its derivation, its constants and its assumptions in full. If you are reading for results rather than method, skip to **§5 (Cadenza)**; the conclusions are at the top of this document.

### Kinetic Modeling of Transcriptome & Proteome Trajectories

Standard Differential Expression (DE) tells you whether a transcript and its protein each changed. It cannot tell you whether a discordance between them is biologically meaningful. A gene whose mRNA rises sharply while its protein barely moves has two completely different possible explanations:

-   **The protein is under post-transcriptional control** (translational repression, accelerated degradation, sequestration) - biology.
-   **The protein simply has a long half-life** and hasn't had time to accumulate yet - kinetics.

Both produce the identical snapshot pattern. This step exists to separate them, and it's the reason a time course was run at all.

### Inputs and Outputs

**Inputs** (all genes × timepoints matrices, restricted to Universe B - genes with both layers passing QC):

| Argument                  | Content                                                                              |
|:--------------------------|:-------------------------------------------------------------------------------------|
| `Rl, Pl`                  | log2FC trajectories (treated vs control), t0-centred so column 1 is identically zero |
| `Rs, Ps`                  | limma moderated standard errors for those logFCs                                     |
| `tps`                     | sampled timepoints                                                                   |
| `fdr_thresh, min_abs_lfc` | classification thresholds (defaults 0.05, 0.5)                                       |

**Output:** one row per gene --- `archetype`, `amplitude_a`, `t_half_M0_h`, `t_half_M1_h`, `identifiable`, `rna_responds`, `regulated`, `lrt_p`, `lrt_fdr`.

### The Two Nested Models

Everything is expressed as fold-change relative to the time-matched control:

$$
r(t) = 2^{\,\text{logFC}_{\text{RNA}}(t)}, \qquad
p(t) = 2^{\,\text{logFC}_{\text{protein}}(t)}
$$

both starting at 1.

### M0 - The Kinetic Null

Under assumptions A1--A3 (derived in "Where the Kinetic Null Comes From" below --- note that A3 is measurably violated in this dataset), if the treatment changed only transcript abundance the protein follows first-order kinetics:

$$
\frac{dp}{dt} = k_d \cdot (r(t) - p(t)), \quad p(0) = 1
$$

One free parameter: `k_d`, the degradation rate (reported as half-life `log(2)/k_d`). This is the trajectory the protein must take if nothing post-transcriptional is happening. Implemented via `integrate_protein_mat(..., ks = kd, kd = kd, P0 = 1)` - setting `ks = kd` makes the steady state equal $r$, which is what encodes "protein eventually catches up to its mRNA."

### M1 - Kinetics Plus Amplitude

Adds a single scaling factor:

$$
\text{logFC}_{\text{protein}}(t) = a \cdot \log_2 p(t; k_d)
$$

$a = 1$ is exactly M0. The interpretation of $a$ is the whole point:

| $a$             | Meaning                                                                                                |
|:----------------|:-------------------------------------------------------------------------------------------------------|
| **≈ 1**         | Protein tracks its mRNA as kinetics predict --- no post-transcriptional signature                      |
| **0 \< a \< 1** | **Buffered** --- response damped relative to prediction (translational repression, faster degradation) |
| **a \> 1**      | **Amplified** --- response exceeds prediction (increased translation efficiency, stabilisation)        |
| **a \< 0**      | **Anticorrelated** --- protein moves opposite to its mRNA                                              |

------------------------------------------------------------------------

## Fitting Procedure

For each of 80 candidate half-lives on a log-spaced grid, both models are fitted to all genes simultaneously (vectorised), then the best grid point per gene is taken:

1.  Solve the ODE at that `k_d` → predicted trajectory $pl$.
2.  **M0:** weighted SSR of $pl$ against observed, $a$ fixed at 1.
3.  **M1:** the optimal $a$ has a closed form --- weighted least squares through the origin, $a = \frac{\sum (W \cdot pl \cdot Pl)}{\sum (W \cdot pl^2)}$ --- so it's profiled out analytically at zero cost rather than searched. Minimum SSR is then $SST - \frac{\text{num}^2}{\text{den}}$.
4.  `max.col(-ssr)` picks each gene's best half-life independently for M0 (`t_half_M0_h`) and M1 (`t_half_M1_h`).

Weights are $W = \frac{1}{\max(Ps, 0.05)^2}$ --- inverse-variance, so each gene is judged against its own measurement noise. With $n=3$ per cell this is essential; without it, noisy low-abundance proteins would dominate the "regulated" calls.

Because t0 is the centring anchor, both `Pl[,1]` and `pl[,1]` are exactly 0, so that timepoint contributes nothing to the SSR --- the fit uses the 3 post-baseline points.

## The Test

`LRT = SSR_M0 - SSR_M1`, compared to $\chi^2$ with 1 df. Since the weights already carry $1/\sigma^2$, the weighted SSR difference is directly on the chi-square scale. BH-adjusted across genes $\rightarrow$ `lrt_fdr`.

A gene is called **regulated** when the data statistically demand $a \neq 1$ --- i.e. the kinetic null is insufficient. This is a rejection test, not a threshold on effect size.

## Classification Decision Tree

Evaluated in this order (later assignments overwrite earlier):

``` text
unchanged                                    (default)
├── not RNA-responsive, but protein moves  → protein_only
└── RNA-responsive
    ├── null NOT rejected
    │   ├── half-life unidentifiable         → kinetics_limited
    │   ├── identifiable, fast               → concordant
    │   └── identifiable, slow               → kinetic_lag
    └── null rejected (regulated)
        ├── a ≥ 1                            → amplified
        ├── 0 < a < 1                        → buffered
        └── a ≤ 0                            → anticorrelated
```

**Two structural properties worth documenting:**

-   `regulated` requires `rna_resp`. Amplitude is meaningless without an RNA signal to scale --- for a flat transcript, $a \cdot 0 = 0$ for any $a$. This gate is what makes `protein_only` a separate category rather than a spuriously extreme amplitude.
-   `concordant` vs `kinetic_lag` is a labelling split, not a test. Both mean "the kinetic null was sufficient." The slow cutoff just separates fast-turnover proteins from slow ones for readability; genes near the line are on a continuum.

## Identifiability --- The `kinetics_limited` Class

This is the most important thing to convey in documentation. A half-life longer than the observation window cannot be estimated from that window --- a protein with a 200 h half-life and one with a 500 h half-life look identical across 72 h. In that regime, "slow turnover" and "strong buffering" both flatten the predicted trajectory and become genuinely confounded.

Rather than silently assigning such genes to `kinetic_lag` or `buffered`, they are flagged `identifiable = FALSE` and given their own archetype. `kinetics_limited` is an honest statement of what the data cannot resolve, not a biological category. On the simulated ground-truth data this was \~15% of genes; the real remedy is an external half-life prior (see `concordance.half_life_prior_file` in `config/config.yaml`), which cut it to \~6%.

## Constants and Their Basis

| Constant         | Value                        | Basis                                                                                                                                                                                                                                               |
|:-----------------|:-----------------------------|:----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| `th_bound`       | `max(tps)`                   | Principled --- the identifiability boundary. Half-lives beyond the observation window are unestimable by construction.                                                                                                                              |
| `th_grid range`  | `0.5 * Δt` to `6 * max(tps)` | **Lower:** below half the smallest sampling gap, distinct half-lives give indistinguishable sampled trajectories.<br>**Upper:** extends past the boundary so slow genes aren't clamped --- the `identifiable` flag reports the uncertainty instead. |
| `th_grid points` | `80`                         | Empirical ("more doesn't change the answer"), not derived.                                                                                                                                                                                          |
| `ODE grid step`  | `0.25`                       | Discretisation error negligible vs measurement noise.                                                                                                                                                                                               |
| `SE floor`       | `0.05`                       | Numerical guard so a spuriously tiny moderated SE can't give one gene near-infinite weight.                                                                                                                                                         |
| `min_abs_lfc`    | `0.5`                        | Effect-size floor (\~41% change) so statistically-significant-but-trivial changes aren't kinetically modelled. Conventional.                                                                                                                        |
| `fdr_thresh`     | `0.05`                       | Standard BH convention.                                                                                                                                                                                                                             |
| `slow cutoff`    | `0.5 * th_bound`             | Readability split only (see above).                                                                                                                                                                                                                 |

> The first two rows change what the model means; the rest are conventional defaults you can retune.

## Two Things to Check Before Writing This Up

1.  **The responsiveness gate:** The notebook's port differs from `05`. `05` uses both an FDR test and an effect-size floor:

    ``` r
    rna_resp  <- (r_any < FDR) & (max_abs(Rl) > MINLF)
    prot_resp <- (p_any < FDR) & (max_abs(Pl) > MINLF)
    ```

    The notebook version uses effect size only (it doesn't receive the F-test FDR vectors). That makes `rna_responds` and `protein_only` more permissive in the notebook than in `05`, so the two will not produce identical archetype counts on the same data. Worth either documenting as a deliberate simplification or fixing to match --- say which and I'll align them.

2.  **`th_grid`'s lower bound is timepoint-scale-dependent:** It's `0.5 * diff(tps)[1]`, so with `TIMEPOINT_VALUES = c(0,1,2,3)` it's 0.5 units, but with real hours `c(0,24,48,72)` it becomes 12 h --- meaning no half-life below 12 h can be fitted. Given wheat protein turnover, that's probably fine, but it's a real constraint that appears only after the timepoint values are corrected, and it should be stated in the docs.

------------------------------------------------------------------------

## Where the Kinetic Null Comes From --- Derivation and Assumptions

The model is **not** a necessary consequence of "the treatment only changed transcript abundance." It follows from three assumptions, and it is worth stating all three because one of them is measurably violated in this dataset.

Start from absolute abundance in each arm, assuming first-order kinetics with the *same* rate constants in both:

$$
\frac{dP_c}{dt} = k_s R_c(t) - k_d P_c(t), \qquad \frac{dP_t}{dt} = k_s R_t(t) - k_d P_t(t)
$$

Define the fold-changes

$$
r = R_t / R_c, \qquad p = P_t / P_c
$$

and differentiate the quotient:

$$
\frac{dp}{dt} = \frac{P_t' P_c - P_t P_c'}{P_c^2}
              = \frac{k_s (R_t P_c - R_c P_t)}{P_c^2}
              = \underbrace{\frac{k_s R_c}{P_c}}_{\text{coefficient}} (r - p)
$$

The `k_d` terms **cancel exactly**. The coefficient is `k_s·R_c/P_c`, *not* `k_d`. It equals `k_d` only when `k_s·R_c = k_d·P_c` --- that is, only when the control arm is at steady state.

So the implemented equation `dp/dt = k_d(r - p)` requires:

| \#     | Assumption                                                | Status                                                                                                                                                                                       |
|:-------|:----------------------------------------------------------|:---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| **A1** | Protein dynamics are first-order: `dP/dt = k_s·R - k_d·P` | Approximation. Excludes translational saturation, cooperative/zero-order degradation, stoichiometry-driven degradation of unassembled complex subunits, and transcription→translation delay. |
| **A2** | Treatment leaves `k_s` and `k_d` unchanged                | This *is* the null hypothesis being tested. Rejecting it is the point.                                                                                                                       |
| **A3** | Control arm is at (quasi-)steady state throughout         | **Measurably violated here --- see the diagnostic below.**                                                                                                                                   |

### A note on `ks = kd` in the code

`integrate_protein_mat(..., ks = kd, kd = kd, P0 = 1)` is **not** a biological claim that synthesis rate equals degradation rate. In the fold-change formulation absolute `k_s` cancels out entirely and is not identifiable from this data at all (it is absorbed by the per-protein MS response factor). Setting `ks = kd` is the mathematical encoding of **A3** --- it forces the steady state of the relative system to equal $r$.

### Why A3 is questionable in this dataset

In a `~treatment*time` design with Control as the reference level, the bare `time` coefficients are exactly the control arm's own change from baseline. Those coefficients show substantial control-arm movement over the time course --- unsurprising for wheat seedlings over 72 h (development, circadian rhythm, handling), but a direct violation of "control at steady state."

The consequence is specific: when `R_c` is rising, `P_c` lags below its steady state, so the true coefficient `k_s·R_c/P_c` exceeds `k_d`; when `R_c` falls, it drops below. Because `k_d` and $a$ trade off along the identifiability ridge, that misfit can leak into the amplitude estimate --- meaning some `regulated` calls could reflect control-arm non-stationarity rather than treatment-specific post-transcriptional control.

Two mitigating points, for balance: the ratio formulation cancels any effect **shared** by both arms exactly, and growth dilution inflates `k_d` uniformly without breaking the functional form. It is specifically the *lag of* `P_c` *behind a moving* `R_c` that breaks A3, not movement per se.

The next section tests how much this actually matters.

## A3 diagnostic --- does control-arm drift drive `regulated` calls?

```{r}
#| label: fn-a3-diagnostic

#' Test assumption A3 (control arm at quasi-steady state) empirically.
#'
#' If A3 violation were generating spurious `regulated` calls, genes whose
#' CONTROL arm drifts most over the time course should be enriched for
#' `regulated` relative to genes whose control arm is stable.
#'
#' CONFOUNDER, and why the raw enrichment is not the headline number: genes
#' with large control-arm drift also tend to have larger treatment effects and
#' higher expression, both of which independently increase power to reject the
#' kinetic null. A raw drift-vs-regulated association would therefore be
#' expected even under a perfectly valid A3. The reported test is a logistic
#' regression of `regulated` on control drift ADJUSTED for the gene's own RNA
#' response amplitude, so the drift coefficient answers the question that
#' matters: at a given treatment effect size, does control-arm instability
#' still predict being called regulated?
run_a3_diagnostic <- function(ctrl_time, Rl, arch, rna_resp, label = "") {
  idx <- match(rownames(Rl), arch$gene_id)
  d <- data.frame(
    gene_id   = rownames(Rl),
    ## control-arm drift: largest absolute change of the CONTROL arm from
    ## baseline, across post-baseline timepoints
    drift     = apply(abs(ctrl_time[rownames(Rl), , drop = FALSE]), 1, max),
    ## treatment effect amplitude, the confounder being adjusted for
    rna_amp   = apply(abs(Rl[, -1, drop = FALSE]), 1, max),
    regulated = arch$regulated[idx],
    a         = arch$amplitude_a[idx],
    resp      = rna_resp,
    stringsAsFactors = FALSE)
  ## Restrict to the genes the kinetic model was actually fitted on.
  d <- d[d$resp & is.finite(d$a) & is.finite(d$drift) & is.finite(d$rna_amp), ]

  ## Unadjusted view: regulated rate by control-drift quartile.
  d$drift_q <- cut(d$drift, quantile(d$drift, 0:4 / 4), include.lowest = TRUE,
                   labels = c("Q1 (most stable)", "Q2", "Q3", "Q4 (most drift)"))
  by_q <- do.call(rbind, lapply(split(d, d$drift_q), function(g)
    data.frame(n = nrow(g),
               median_drift = round(median(g$drift), 2),
               median_rna_amp = round(median(g$rna_amp), 2),
               pct_regulated = round(100 * mean(g$regulated), 1))))
  by_q <- data.frame(drift_quartile = rownames(by_q), by_q, row.names = NULL)

  ## Adjusted test.
  fit <- glm(regulated ~ scale(drift) + scale(rna_amp), binomial, data = d)
  co  <- summary(fit)$coefficients
  ci  <- suppressMessages(confint(fit))
  res <- data.frame(
    term = c("control drift", "RNA amplitude"),
    odds_ratio = round(exp(co[2:3, "Estimate"]), 3),
    ci_lo = round(exp(ci[2:3, 1]), 3),
    ci_hi = round(exp(ci[2:3, 2]), 3),
    p_value = signif(co[2:3, "Pr(>|z|)"], 3), row.names = NULL)

  ## Same question on the continuous amplitude deviation |a - 1|.
  sp_raw <- suppressWarnings(cor(d$drift, abs(d$a - 1), method = "spearman"))
  ## Partial Spearman, adjusting for rna_amp by ranking then residualising.
  rr <- function(x) rank(x)
  pres <- residuals(lm(rr(abs(d$a - 1)) ~ rr(d$rna_amp)))
  dres <- residuals(lm(rr(d$drift)      ~ rr(d$rna_amp)))
  sp_adj <- cor(dres, pres)

  ## COLLINEARITY CHECK -- required to interpret `adjusted` honestly. Control
  ## drift and treatment amplitude both measure "this gene is dynamic", so they
  ## are correlated by construction. When they are strongly collinear the two
  ## adjusted coefficients cannot be cleanly separated and either can flip sign
  ## or inflate; report the correlation so the reader can judge.
  collin <- cor(d$drift, d$rna_amp, method = "spearman")

  list(label = label, n = nrow(d), by_quartile = by_q, adjusted = res,
       spearman_raw = round(sp_raw, 3), spearman_adjusted = round(sp_adj, 3),
       collinearity = round(collin, 3), data = d)
}
```

```{r}
#| label: fn-archetypes

#' Nested kinetic model: M0 (protein trajectory fully explained by RNA
#' kinetics alone) vs M1 (M0 + a free amplitude factor `a`), compared by a
#' 1-df likelihood-ratio test. Identical model and identifiability handling
#' to `05_concordance_archetypes.R` -- see that script's header for the full
#' reasoning (this is what separates real post-transcriptional regulation
#' from a protein simply having a long half-life).
run_kinetic_archetypes <- function(Rl, Rs, Pl, Ps, tps,
                                   fdr_thresh  = 0.05,
                                   min_abs_lfc = 0.5) {

  # ------------------------------------------------------------------
  # Setup: fine time grid, RNA trajectory interpolated onto it
  # ------------------------------------------------------------------
  grid <- seq(0, max(tps), by = 0.25)
  gi   <- match(tps, grid)                 # positions of sampled timepoints

  r_grid <- t(apply(Rl, 1, function(v) approx(tps, v, grid, rule = 2)$y))
  r_lin  <- 2^r_grid                       # linear fold-change scale

  # ------------------------------------------------------------------
  # Half-life search grid
  #   lower : half the smallest sampling gap -- below this, distinct
  #           half-lives are indistinguishable at the sampled points
  #   upper : well past the identifiability bound, so genuinely slow
  #           genes are not clamped; `identifiable` carries the caveat
  # ------------------------------------------------------------------
  th_bound <- max(tps)

  th_grid <- exp(seq(log(0.5 * diff(tps)[1]),
                     log(6 * th_bound),
                     length.out = 80))

  kd_grid <- log(2) / th_grid

  # Inverse-variance weights: each gene judged against its own noise
  W   <- 1 / pmax(Ps, 0.05)^2
  SST <- rowSums(W * Pl^2)

  n <- nrow(Pl)
  J <- length(kd_grid)

  ssr0  <- ssr1 <- ahat <- matrix(NA_real_, n, J)
  preds <- vector("list", J)

  # ------------------------------------------------------------------
  # Fit both models at every candidate half-life, all genes at once
  # ------------------------------------------------------------------
  for (j in seq_len(J)) {

    kd <- kd_grid[j]

    pl <- integrate_protein_mat(grid, r_lin, ks = kd, kd = kd, P0 = rep(1, n))
    pl <- log2(pmax(pl[, gi, drop = FALSE], 1e-9))

    preds[[j]] <- pl

    # M0: amplitude fixed at 1
    ssr0[, j] <- rowSums(W * (pl - Pl)^2)

    # M1: optimal amplitude has a closed form (weighted least squares
    # through the origin), so profile it out instead of searching for it
    num <- rowSums(W * pl * Pl)
    den <- rowSums(W * pl^2)

    ahat[, j] <- ifelse(den > 1e-12, num / den, NA_real_)
    ssr1[, j] <- SST - ifelse(den > 1e-12, num^2 / den, 0)
  }

  # ------------------------------------------------------------------
  # Pick each gene's best half-life under each model
  # ------------------------------------------------------------------
  j0 <- max.col(-ssr0, "first")
  j1 <- max.col(-ssr1, "first")

  ix <- function(j) cbind(seq_len(n), j)

  S0 <- ssr0[ix(j0)]
  S1 <- ssr1[ix(j1)]

  t_half_M0 <- th_grid[j0]
  t_half_M1 <- th_grid[j1]
  a_hat     <- ahat[ix(j1)]

  pred0 <- t(vapply(seq_len(n),
                    function(i) preds[[j0[i]]][i, ],
                    numeric(length(tps))))

  # ------------------------------------------------------------------
  # Likelihood-ratio test of M1 against M0 (1 df)
  # ------------------------------------------------------------------
  lrt     <- pmax(S0 - S1, 0)
  p_lrt   <- pchisq(lrt, df = 1, lower.tail = FALSE)
  fdr_lrt <- p.adjust(p_lrt, "BH")

  # ------------------------------------------------------------------
  # Classification
  # ------------------------------------------------------------------
  max_abs <- function(m) apply(abs(m[, -1, drop = FALSE]), 1, max)

  rna_resp     <- max_abs(Rl) > min_abs_lfc
  regulated    <- rna_resp & (fdr_lrt < fdr_thresh)
  identifiable <- t_half_M0 <= th_bound
  slow         <- t_half_M0 > 0.5 * th_bound

  archetype <- rep("unchanged", n)

  archetype[!rna_resp & max_abs(Pl) > min_abs_lfc]         <- "protein_only"

  archetype[rna_resp & !regulated &  identifiable & !slow] <- "concordant"
  archetype[rna_resp & !regulated &  identifiable &  slow] <- "kinetic_lag"
  archetype[rna_resp & !regulated & !identifiable]         <- "kinetics_limited"

  archetype[regulated & a_hat >= 1]                        <- "amplified"
  archetype[regulated & a_hat <  1 & a_hat > 0]            <- "buffered"
  archetype[regulated & a_hat <= 0]                        <- "anticorrelated"

  data.frame(
    gene_id      = rownames(Pl),
    archetype    = archetype,
    rna_responds = rna_resp,
    regulated    = regulated,
    amplitude_a  = a_hat,
    t_half_M0_h  = t_half_M0,
    t_half_M1_h  = t_half_M1,
    identifiable = identifiable,
    lrt_p        = p_lrt,
    lrt_fdr      = fdr_lrt,
    stringsAsFactors = FALSE
  )
}
```

## 4. Case B integration (`06`-style)

```{r}
#| label: fn-integration

#' Design-cell PLS with leave-one-cell-out Q^2 against a permutation null --
#' the Case B analogue of a MEFISTO/DIABLO variance-explained figure. See
#' `06_integration_caseB.R`'s header for why the in-sample cross-block
#' correlation is not a valid statistic here (it reaches ~0.7-0.9 even on
#' permuted data at this sample size).
run_integration <- function(rna_mat, prot_mat, cd_r, cd_p, n_hvg = 3000, n_hvp = 2000,
                            n_comp = 3, n_perm = 200) {
  hv_r <- top_variable(rna_mat,  min(n_hvg, nrow(rna_mat)))
  hv_p <- top_variable(prot_mat, min(n_hvp, nrow(prot_mat)))
  Xc <- prep_block(cell_means(rna_mat[hv_r, ],  cd_r))
  Yc <- prep_block(cell_means(prot_mat[hv_p, ], cd_p))
  stopifnot(identical(rownames(Xc), rownames(Yc)))
  K  <- min(n_comp, nrow(Xc) - 3)
  fit <- pls2(Xc, Yc, ncomp = K)
  Ax <- row_space(Xc); Ay <- row_space(Yc)
  q2 <- q2_loo(Ax, Ay, K)
  perm_q2 <- matrix(NA_real_, n_perm, K)
  for (b in seq_len(n_perm)) perm_q2[b, ] <- q2_loo(Ax, Ay[sample(nrow(Ay)), , drop = FALSE], K)
  p_q2 <- vapply(seq_len(K), function(k) (1 + sum(perm_q2[, k] >= q2[k])) / (n_perm + 1), 0)
  list(fit = fit, q2 = q2, perm_q2 = perm_q2, p_q2 = p_q2,
       cellmeta = data.frame(cell = rownames(Xc)))
}
```

## 5. Run the pipeline: Cadenza

```{r}
#| label: cadenza-load

cad <- load_variety("cadenza")
knitr::kable(table(cad$meta$treatment, cad$meta$time_num),
            caption = "Cadenza design: samples per treatment x timepoint")
```

```{r}
#| label: cadenza-qc

cad_qc_r <- run_qc_rna(cad$rna, cad$meta)
cad_qc_p <- run_qc_prot(cad$prot, cad$meta)
cat(sprintf("RNA: %d / %d genes kept by filterByExpr\n", cad_qc_r$n_kept, cad_qc_r$n_in))
cat(sprintf("Protein: %d / %d groups kept (>=2 valid in >=1 cell); missing rate %.1f%%\n",
           cad_qc_p$n_kept, cad_qc_p$n_in, 100 * cad_qc_p$miss_rate))
```

```{r}
#| label: fig-cadenza-qc-pca
#| fig-cap: "Cadenza PCA, RNA (VST) and protein (normalised), coloured by treatment"

pca_plot <- function(m, meta, main) {
  mv <- m[order(row_var(m), decreasing = TRUE, na.last = NA)[seq_len(min(500, nrow(m)))], , drop = FALSE]
  mv <- mv[complete.cases(mv), , drop = FALSE]
  p <- prcomp(t(mv), scale. = FALSE)
  v <- round(100 * p$sdev^2 / sum(p$sdev^2), 1)
  plot(p$x[, 1], p$x[, 2], pch = c(16, 17)[as.integer(meta$treatment)],
       col = hcl.colors(4, "Zissou1")[as.integer(factor(meta$time_num))], cex = 1.6,
       xlab = sprintf("PC1 (%.1f%%)", v[1]), ylab = sprintf("PC2 (%.1f%%)", v[2]), main = main)
  legend("topright", bty = "n", cex = 0.7, legend = levels(meta$treatment), pch = c(16, 17))
}
op <- par(mfrow = c(1, 2))
pca_plot(cad_qc_r$vst,       cad$meta, "Cadenza RNA (VST)")
pca_plot(cad_qc_p$norm,      cad$meta, "Cadenza protein (normalised)")
par(op)
```

```{r}
#| label: cadenza-missingness

cad_mi <- run_missingness(cad_qc_p$norm, cad$meta)
cat(sprintf("MNAR: %.1f%%  MAR: %.1f%%  (of all entries)\n",
           100 * cad_mi$mnar_frac, 100 * cad_mi$mar_frac))
```

```{r}
#| label: cadenza-de

## protein_id == gene_id within a variety (verified in ../../data/real/
## *_protein_gene_mapping.csv), so Universe B == QC'd protein genes that also
## survive RNA QC -- no bipartite mapping ambiguity for this dataset.
cad_de_rna  <- run_de(cad_qc_r$counts, cad$meta, TRUE)
cad_common  <- intersect(rownames(cad_mi$mixed), rownames(cad_qc_r$counts))
cad_de_prot <- run_de(cad_mi$mixed[cad_common, , drop = FALSE], cad$meta, FALSE)

cat(sprintf("Universe B: %d matched genes\n", length(cad_common)))
cat(sprintf("RNA any-effect FDR<0.05: %d\n",  sum(cad_de_rna$any_fdr[cad_common] < 0.05)))
cat(sprintf("Protein any-effect FDR<0.05: %d\n", sum(cad_de_prot$any_fdr < 0.05)))
```

```{r}
#| label: cadenza-archetypes

cad_Rl <- cad_de_rna$lfc[cad_common, ];  cad_Rs <- cad_de_rna$se[cad_common, ]
cad_Pl <- cad_de_prot$lfc;               cad_Ps <- cad_de_prot$se
cad_Rl <- cad_Rl - cad_Rl[, 1]; cad_Pl <- cad_Pl - cad_Pl[, 1]

cad_arch <- run_kinetic_archetypes(cad_Rl, cad_Rs, cad_Pl, cad_Ps, DESIGN$timepoints)
knitr::kable(as.data.frame(table(archetype = cad_arch$archetype)),
            caption = "Cadenza: profile-likelihood archetype calls")
```

> **How far to trust this table.** The *distribution* across archetypes is informative; the assignment of any individual gene is not. Two reasons, both quantified in `assumptions_validation.qmd`: RNA measurement error is not propagated, and flips 36% of Cadenza calls across the `a = 1` boundary (§7 there); and the first-order model leaves systematically structured residuals (§3 there), so some genes are flagged because their trajectory *shape* is mis-modelled rather than because their amplitude departs from 1. `kinetics_limited` is not a biological class --- it is the \~50% of genes where attenuation and slow turnover are mathematically confounded.

## 6. Hierarchical Bayesian kinetics: Cadenza

```{r}
#| label: cadenza-bayesian-kinetics

## RNA-responsive gate (mirrors 05's rna_resp / the doc note in
## 07_bayesian_kinetics.R): the amplitude model is only meaningful when RNA
## carries real signal -- for a flat-RNA gene ANY amplitude "explains" a tiny
## predicted change equally badly, which (confirmed in dev testing on this
## exact dataset) can otherwise drive a gene to the numerical safety bound.
cad_rna_resp <- (cad_de_rna$any_fdr[cad_common] < 0.05) &
  (apply(abs(cad_Rl[, -1, drop = FALSE]), 1, max) > 0.5)
cat(sprintf("RNA-responsive, eligible for amplitude fitting: %d / %d\n",
           sum(cad_rna_resp), length(cad_rna_resp)))

cad_fit <- fit_hierarchical_kinetics(
  cad_Rl[cad_rna_resp, ], cad_Rs[cad_rna_resp, ],
  cad_Pl[cad_rna_resp, ], cad_Ps[cad_rna_resp, ],
  ## Anchored at the SAME joint-optimal half-life the profile-likelihood fit
  ## above already found (t_half_M1_h), so the two analyses share one kinetic
  ## fit and differ only in how the amplitude is inferred (independent
  ## per-gene LRT vs. hierarchical horseshoe shrinkage).
  t_half = cad_arch$t_half_M1_h[match(rownames(cad_Rl)[cad_rna_resp], cad_arch$gene_id)],
  tps = DESIGN$timepoints, n_iter = 4000, warmup = 1600, seed = 1, verbose = FALSE)

cat(sprintf("MCMC acceptance rate: %.2f\n", cad_fit$accept_delta))
cat(sprintf("Genes at the numerical safety bound (report as directional, not exact): %d / %d\n",
           sum(cad_fit$summary$at_bound), nrow(cad_fit$summary)))
cat(sprintf("Regulated by hierarchical model (95%% CI excludes a=1): %d / %d\n",
           sum(cad_fit$summary$regulated_95), nrow(cad_fit$summary)))
cat(sprintf("Regulated by profile-likelihood LRT (FDR<0.05), same genes: %d / %d\n",
           sum(cad_arch$regulated[match(rownames(cad_Rl)[cad_rna_resp], cad_arch$gene_id)]),
           sum(cad_rna_resp)))
```

```{r}
#| label: fig-cadenza-kinetic-volcano
#| fig-cap: "Cadenza kinetic volcano: hierarchical posterior amplitude vs. profile-likelihood significance"

idx <- match(cad_fit$summary$gene_id, cad_arch$gene_id)
plot(cad_fit$summary$a_mean, -log10(pmax(cad_arch$lrt_p[idx], 1e-12)), pch = 16, cex = 0.4,
     col = ifelse(cad_fit$summary$regulated_95, "#D1495B99", "#00000033"),
     xlab = "hierarchical posterior mean amplitude (a)",
     ylab = "-log10 p (profile-likelihood LRT)",
     main = "Cadenza: two independent regulation signals\n(red = hierarchical model flags as regulated)")
abline(v = 1, lty = 3)
```

The two methods agree on genes with strong signal and differ most on genes the profile-likelihood LRT was underpowered for individually (§`docs/PIPELINE.md` documents this trade-off quantitatively on simulated ground truth: the hierarchical model roughly triples recall at the cost of a higher false-call rate on truly-unregulated genes, since fixing the half-life removes one source of legitimate uncertainty from its interval). Genes flagged by **both** methods are the highest-confidence regulated set; genes flagged by the hierarchical model **only** are candidates for follow-up validation, not standalone claims.

```{r}
#| label: cadenza-agreement-table

cad_kin <- data.frame(
  gene_id = rownames(cad_Rl)[cad_rna_resp],
  lrt_regulated  = cad_arch$regulated[match(rownames(cad_Rl)[cad_rna_resp], cad_arch$gene_id)],
  bayes_regulated = cad_fit$summary$regulated_95,
  a_mean = cad_fit$summary$a_mean, at_bound = cad_fit$summary$at_bound,
  stringsAsFactors = FALSE)
knitr::kable(as.data.frame(table(LRT = cad_kin$lrt_regulated, Bayesian = cad_kin$bayes_regulated)),
            caption = "Cadenza: agreement between the two regulation calls")
```

## 7. Case B integration: Cadenza

```{r}
#| label: cadenza-integration

cad_int <- run_integration(cad_qc_r$vst, cad_mi$mixed, cad$meta, cad$meta)
knitr::kable(data.frame(component = seq_along(cad_int$q2), Q2 = round(cad_int$q2, 3),
                        perm_q95 = round(apply(cad_int$perm_q2, 2, quantile, .95), 3),
                        p_value = signif(cad_int$p_q2, 3)),
            caption = "Cadenza: leave-one-cell-out Q2 vs permutation null")
```

> **This is the most trustworthy number in the analysis.** Q² is computed on design cells held out of the fit, so it cannot be inflated by the in-sample artefact that makes raw cross-block correlation meaningless here (a PLS on 8 pseudo-samples reaches r ≈ 0.72 on *permuted* data). A significant Q² means the transcriptome of an unseen treatment × timepoint cell genuinely predicts that cell's proteome. Note this is a statement about **between-cell** coupling; within-cell covariance is not estimable in a Case B design by any method.

## 8. Run the pipeline: Norin

Identical steps, applied to the second variety.

```{r}
#| label: norin-load

nor <- load_variety("norin")
knitr::kable(table(nor$meta$treatment, nor$meta$time_num),
            caption = "Norin design: samples per treatment x timepoint")

nor_qc_r <- run_qc_rna(nor$rna, nor$meta)
nor_qc_p <- run_qc_prot(nor$prot, nor$meta)
cat(sprintf("RNA: %d / %d genes kept\n", nor_qc_r$n_kept, nor_qc_r$n_in))
cat(sprintf("Protein: %d / %d groups kept; missing rate %.1f%%\n",
           nor_qc_p$n_kept, nor_qc_p$n_in, 100 * nor_qc_p$miss_rate))

nor_mi <- run_missingness(nor_qc_p$norm, nor$meta)
cat(sprintf("MNAR: %.1f%%  MAR: %.1f%%\n", 100 * nor_mi$mnar_frac, 100 * nor_mi$mar_frac))
```

```{r}
#| label: norin-de-archetypes

nor_de_rna  <- run_de(nor_qc_r$counts, nor$meta, TRUE)
nor_common  <- intersect(rownames(nor_mi$mixed), rownames(nor_qc_r$counts))
nor_de_prot <- run_de(nor_mi$mixed[nor_common, , drop = FALSE], nor$meta, FALSE)
cat(sprintf("Universe B: %d matched genes\n", length(nor_common)))

nor_Rl <- nor_de_rna$lfc[nor_common, ];  nor_Rs <- nor_de_rna$se[nor_common, ]
nor_Pl <- nor_de_prot$lfc;               nor_Ps <- nor_de_prot$se
nor_Rl <- nor_Rl - nor_Rl[, 1]; nor_Pl <- nor_Pl - nor_Pl[, 1]

nor_arch <- run_kinetic_archetypes(nor_Rl, nor_Rs, nor_Pl, nor_Ps, DESIGN$timepoints)
knitr::kable(as.data.frame(table(archetype = nor_arch$archetype)),
            caption = "Norin: profile-likelihood archetype calls")
```

```{r}
#| label: norin-bayesian-kinetics

nor_rna_resp <- (nor_de_rna$any_fdr[nor_common] < 0.05) &
  (apply(abs(nor_Rl[, -1, drop = FALSE]), 1, max) > 0.5)
cat(sprintf("RNA-responsive, eligible for amplitude fitting: %d / %d\n",
           sum(nor_rna_resp), length(nor_rna_resp)))

nor_fit <- fit_hierarchical_kinetics(
  nor_Rl[nor_rna_resp, ], nor_Rs[nor_rna_resp, ],
  nor_Pl[nor_rna_resp, ], nor_Ps[nor_rna_resp, ],
  t_half = nor_arch$t_half_M1_h[match(rownames(nor_Rl)[nor_rna_resp], nor_arch$gene_id)],
  tps = DESIGN$timepoints, n_iter = 4000, warmup = 1600, seed = 1, verbose = FALSE)

cat(sprintf("MCMC acceptance rate: %.2f\n", nor_fit$accept_delta))
cat(sprintf("Regulated by hierarchical model: %d / %d\n",
           sum(nor_fit$summary$regulated_95), nrow(nor_fit$summary)))
```

```{r}
#| label: norin-integration

nor_int <- run_integration(nor_qc_r$vst, nor_mi$mixed, nor$meta, nor$meta)
knitr::kable(data.frame(component = seq_along(nor_int$q2), Q2 = round(nor_int$q2, 3),
                        perm_q95 = round(apply(nor_int$perm_q2, 2, quantile, .95), 3),
                        p_value = signif(nor_int$p_q2, 3)),
            caption = "Norin: leave-one-cell-out Q2 vs permutation null")
```

## 8b. A3 diagnostic: does control-arm drift drive the `regulated` calls?

Placed here because it needs the fitted results from **both** varieties. See "Where the Kinetic Null Comes From" above for the derivation and what A3 states.

```{r}
#| label: a3-diagnostic-run

cad_a3 <- run_a3_diagnostic(cad_de_rna$ctrl_time, cad_Rl, cad_arch, cad_rna_resp, "Cadenza")
nor_a3 <- run_a3_diagnostic(nor_de_rna$ctrl_time, nor_Rl, nor_arch, nor_rna_resp, "Norin")

## Step 1: how badly is A3 violated in the first place? Quantify the size of
## the control-arm movement before asking whether it matters. A3 assumes the
## control arm is at steady state, i.e. that these numbers are ~0.
ctrl_drift_summary <- do.call(rbind, lapply(list(cad_a3, nor_a3), function(x)
  data.frame(variety = x$label, n_genes = x$n,
             median_abs_drift = round(median(x$data$drift), 2),
             pct_drift_over_1 = round(100 * mean(x$data$drift > 1), 1),
             pct_drift_over_2 = round(100 * mean(x$data$drift > 2), 1))))
knitr::kable(ctrl_drift_summary,
            caption = "Control-arm drift: max |log2FC| from baseline, CONTROL arm only. A3 assumes ~0.")
```

```{r}
#| label: a3-diagnostic-unadjusted

## Step 2: the unadjusted view. Reported for transparency, but it is NOT the
## answer -- drift correlates with treatment effect size, which independently
## raises power to reject the kinetic null.
knitr::kable(cbind(variety = "Cadenza", cad_a3$by_quartile),
            caption = "Cadenza: regulated rate by control-drift quartile (UNADJUSTED, confounded -- see next table)")
knitr::kable(cbind(variety = "Norin", nor_a3$by_quartile),
            caption = "Norin: regulated rate by control-drift quartile (UNADJUSTED, confounded)")
```

```{r}
#| label: a3-diagnostic-adjusted

## Step 3: the actual test. Odds ratio for control drift, ADJUSTED for the
## gene's own RNA response amplitude. An OR near 1 means that at a given
## treatment effect size, control-arm instability does not additionally
## predict being called regulated -- i.e. A3 violation is not manufacturing
## the calls.
a3_adj <- rbind(cbind(variety = "Cadenza", cad_a3$adjusted),
                cbind(variety = "Norin",   nor_a3$adjusted))
knitr::kable(a3_adj,
            caption = "ADJUSTED test: logistic regression of `regulated` on control drift + RNA amplitude (both z-scaled). Odds ratio per 1 SD.")

knitr::kable(data.frame(
  variety = c(cad_a3$label, nor_a3$label),
  spearman_raw = c(cad_a3$spearman_raw, nor_a3$spearman_raw),
  spearman_adjusted_for_rna_amp = c(cad_a3$spearman_adjusted, nor_a3$spearman_adjusted),
  collinearity_drift_vs_amp = c(cad_a3$collinearity, nor_a3$collinearity)),
  caption = "Correlation between control drift and |a - 1|, before/after adjusting for RNA amplitude. Last column: collinearity between the two predictors -- high values mean the adjusted odds ratios above cannot be cleanly separated.")
```

```{r}
#| label: fig-a3-diagnostic
#| fig-cap: "Control-arm drift vs amplitude deviation. A rising lowess trend would indicate A3 violation inflating the amplitude estimate."

op <- par(mfrow = c(1, 2))
for (x in list(cad_a3, nor_a3)) {
  d <- x$data
  plot(d$drift, pmin(abs(d$a - 1), 5), pch = 16, cex = 0.3,
       col = ifelse(d$regulated, "#D1495B55", "#00000022"),
       xlab = "control-arm drift (max |log2FC|)", ylab = "|a - 1|  (capped at 5)",
       main = sprintf("%s (n = %d)", x$label, x$n))
  ok <- is.finite(d$drift) & is.finite(d$a)
  lines(lowess(d$drift[ok], pmin(abs(d$a - 1), 5)[ok], f = 0.5), col = "#2C6FBB", lwd = 3)
  legend("topright", bty = "n", cex = 0.7, pch = c(16, NA), lwd = c(NA, 3),
         col = c("#D1495B", "#2C6FBB"), legend = c("called regulated", "lowess"))
}
par(op)
```

### What the diagnostic found

**A3 is clearly violated.** Median control-arm drift is \~0.67 log2FC in both varieties; roughly a third of RNA-responsive genes move more than 1 log2FC, and \~9--11% move more than 2 log2FC, in the *control* arm alone. The control is not at steady state, and the assumption underlying `dp/dt = k_d(r-p)` does not hold as stated.

**But the impact on the amplitude estimate is small.** The continuous measure is the clearest: after adjusting for RNA response amplitude, the correlation between control drift and $|a-1|$ is 0.04 (Cadenza) and 0.07 (Norin) --- negligible. The lowess trends in the figure above are correspondingly flat. Whatever A3 violation is doing, it is not systematically inflating the amplitude deviation that drives the `regulated` call.

**The two varieties disagree on the binary call, and that disagreement is the honest headline.** Adjusted odds ratio for control drift, per 1 SD:

-   **Cadenza: 1.07 (95% CI 0.99--1.16, p = 0.09)** --- not significant, effect negligible.
-   **Norin: 1.31 (95% CI 1.20--1.44, p \< 1e-15)** --- significant, a real \~30% increase in the odds of being called regulated per SD of control drift.

So in Norin there *is* a detectable residual association, in the direction A3 violation predicts. It is modest in size, but it is not zero, and it does not replicate in Cadenza.

**Interpretation caveats, stated rather than glossed:**

-   Control drift and treatment amplitude are correlated by construction (Spearman 0.44--0.53) --- both measure "this gene is dynamic." That is moderate collinearity: enough that the two adjusted coefficients cannot be cleanly separated, not enough to make them meaningless. Note the RNA-amplitude coefficient itself flips sign between varieties (1.05 in Cadenza vs 0.76 in Norin), which is a symptom of exactly this.
-   This diagnostic tests for *association*, not *causation*. A gene whose control arm is unstable may also be genuinely more post-transcriptionally regulated --- biology, not artefact. The test cannot separate those.
-   Absence of a strong association is evidence that A3 violation is not the *dominant* driver of the regulated calls. It is not proof that no individual call is affected.

**Recommended wording for the manuscript limitations section:** the kinetic null assumes a control arm at quasi-steady state; this assumption is measurably violated in this experiment (median control-arm drift \~0.67 log2FC over 72 h). A diagnostic regression adjusting for treatment effect size found the residual association between control-arm drift and post-transcriptional regulation calls to be negligible in Cadenza (OR 1.07, p = 0.09) and modest but significant in Norin (OR 1.31, p \< 0.001). Regulation calls for genes with large control-arm drift in Norin should therefore be treated as lower-confidence.

## 8c. Hierarchical Bayesian model --- assumptions and diagnostics

### What the hierarchical model assumes

The hierarchical model (`R/07_bayesian_kinetics.R`) inherits every assumption of the kinetic null and adds its own. Listed with an honest status for each:

| \#     | Assumption                                                                                      | Status in this dataset                                                                                                                                                                                                                         |
|:-------|:------------------------------------------------------------------------------------------------|:-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| **B1** | All of **A1--A3** --- it solves the same ODE                                                    | **A3 violated** (median control drift \~0.67 log2FC; §8b). Impact on amplitude found to be small but non-zero in Norin.                                                                                                                        |
| **B2** | `k_d` is **fixed**, not sampled jointly with `a`                                                | True by construction. Credible intervals are "amplitude uncertainty at the point-estimate kinetics", **not** full posterior uncertainty. Deliberate --- joint sampling produced a slow-mixing ridge (see the script header).                   |
| **B3** | Protein likelihood is Gaussian with **known** variance (limma's moderated SE, treated as fixed) | Reasonable --- these are already pooled across thousands of genes --- but it ignores uncertainty *in* the variance estimate, so intervals are mildly anti-conservative.                                                                        |
| **B4** | RNA trajectory is known **exactly**; its uncertainty is not propagated                          | **True and worth flagging:** `se_rna` is accepted by the function signature but never used in the likelihood. RNA is a fixed covariate. With n=3 the RNA logFCs are themselves noisy, so real amplitude uncertainty is wider than reported.    |
| **B5** | Genes are **conditionally independent** given the hyperparameters                               | Violated in principle --- co-regulated genes, complex subunits, and shared MS normalisation induce correlated errors. Consequence: the horseshoe pools information as if genes were independent, so hyperparameter uncertainty is understated. |
| **B6** | Horseshoe sparsity: most genes have `a ≈ 1`, controlled by `p0_frac`                            | Tested below.                                                                                                                                                                                                                                  |
| **B7** | MCMC has converged                                                                              | Tested below.                                                                                                                                                                                                                                  |
| **B8** | `delta_bound` clipping is inert                                                                 | \~2% of genes sit at the bound; those are censored, and flagged via `at_bound`.                                                                                                                                                                |

The diagnostics below test **B6** and **B7** directly, since those are the two that are cheap to check and would silently corrupt results if wrong.

### These diagnostics already caught and fixed a real problem

Running them the first time showed the sampler was **not converged** at its original settings (fixed step size 0.25, 2000 iterations):

| Metric                                      | Original sampler | After fix |
|:--------------------------------------------|-----------------:|----------:|
| Metropolis acceptance rate                  |             0.75 |  **0.45** |
| Genes with R-hat \> 1.05                    |        **31.5%** | **0.17%** |
| Worst R-hat                                 |             3.12 |      1.06 |
| Median effective sample size                |              182 |   **740** |
| Minimum effective sample size               |                7 |       146 |
| Genes where 4 chains agree on the call      |            85.8% | **93.6%** |
| Genes called regulated (n = 1200 subsample) |         551--568 |  417--431 |

The tell was the **acceptance rate of 0.75**. For a one-dimensional random-walk Metropolis move the optimum is ≈ 0.44; anything much higher means the proposal steps are too small, so the chain crawls and successive draws are highly correlated. A single global step size cannot suit all genes here, because the likelihood curvature `sum(W·pl²)` spans orders of magnitude across genes.

**Fix:** each gene now gets its own proposal step size, tuned during warmup only (adaptation is frozen before the retained draws, so the kept chain is a valid time-homogeneous Markov chain). The default chain length was also raised from 2000 to 4000 iterations, at which point the amplitude estimates correlate 0.99 with an 8000-iteration reference.

**This mattered for the science, not just the diagnostics:** the unconverged sampler was **over-calling regulation by roughly 25%** (≈560 vs ≈420 genes in the test subsample). Any earlier numbers produced before this fix are superseded.

```{r}
#| label: fn-bayes-diagnostics

#' Convergence and sensitivity diagnostics for the hierarchical model.
#'
#' Tests three things the model could be silently getting wrong:
#'
#'   1. CONVERGENCE (B7). Runs several independent chains from different
#'      seeds and computes the Gelman-Rubin statistic R-hat per gene.
#'      R-hat compares between-chain to within-chain variance; values near
#'      1.0 mean the chains agree. Anything above ~1.05 means they do not,
#'      and the posterior summaries cannot be trusted.
#'
#'   2. PRIOR SENSITIVITY (B6). `p0_frac` sets how sparse the horseshoe
#'      expects the signal to be. If conclusions swing wildly with it, the
#'      prior -- not the data -- is driving the answer.
#'
#'   3. CHAIN LENGTH. If results still move between 500 and 4000 iterations,
#'      the chain is too short.
#'
#' Genes are subsampled (`n_genes`) purely to keep render time reasonable;
#' the diagnostics are about the sampler, not about any specific gene.
run_bayes_diagnostics <- function(Rl, Rs, Pl, Ps, tps, t_half,
                                  n_genes   = 1200,
                                  seeds     = 1:4,
                                  p0_grid   = c(0.05, 0.10, 0.20, 0.40),
                                  iter_grid = c(1000, 2000, 4000, 8000),
                                  label     = "") {

  set.seed(99)
  idx <- sample(nrow(Pl), min(n_genes, nrow(Pl)))

  Rl <- Rl[idx, , drop = FALSE];  Rs <- Rs[idx, , drop = FALSE]
  Pl <- Pl[idx, , drop = FALSE];  Ps <- Ps[idx, , drop = FALSE]
  th <- t_half[idx]

  fit_one <- function(seed, p0 = 0.20, n_iter = 4000) {
    fit_hierarchical_kinetics(
      Rl, Rs, Pl, Ps, tps,
      t_half  = th,
      p0_frac = p0,
      n_iter  = n_iter,
      warmup  = floor(n_iter * 0.4),
      seed    = seed,
      verbose = FALSE
    )
  }

  ## ---------------------------------------------------------------------
  ## 1. Convergence: multiple chains, Gelman-Rubin R-hat
  ## ---------------------------------------------------------------------
  chains <- lapply(seeds, fit_one)
  traces <- lapply(chains, `[[`, "trace_delta")

  N <- ncol(traces[[1]])
  M <- length(traces)

  chain_means <- sapply(traces, rowMeans)
  chain_vars  <- sapply(traces, function(tr) apply(tr, 1, var))

  W_within  <- rowMeans(chain_vars)
  B_between <- N * apply(chain_means, 1, var)
  var_hat   <- ((N - 1) / N) * W_within + B_between / N
  rhat      <- sqrt(pmax(var_hat, 0) / pmax(W_within, 1e-12))

  ## Effective sample size, AR(1) approximation: ESS = N * (1-rho)/(1+rho).
  ## Exact multi-lag ESS per gene would be far slower and the AR(1) form is
  ## the standard quick diagnostic.
  lag1 <- function(tr) {
    a <- tr[, -N, drop = FALSE]
    b <- tr[, -1, drop = FALSE]
    am <- rowMeans(a); bm <- rowMeans(b)
    num <- rowSums((a - am) * (b - bm))
    den <- sqrt(rowSums((a - am)^2) * rowSums((b - bm)^2))
    ifelse(den > 0, num / den, 0)
  }
  rho  <- rowMeans(sapply(traces, lag1))
  ess  <- M * N * (1 - rho) / (1 + rho)

  convergence <- data.frame(
    variety            = label,
    n_chains           = M,
    draws_per_chain    = N,
    rhat_median        = round(median(rhat, na.rm = TRUE), 4),
    rhat_max           = round(max(rhat, na.rm = TRUE), 4),
    pct_rhat_over_1.05 = round(100 * mean(rhat > 1.05, na.rm = TRUE), 2),
    ess_median         = round(median(ess, na.rm = TRUE)),
    ess_min            = round(min(ess, na.rm = TRUE))
  )

  ## Do the independent chains agree on the actual CALL, not just the mean?
  calls <- sapply(chains, function(f) f$summary$regulated_95)
  call_agreement <- data.frame(
    variety             = label,
    pct_all_chains_agree = round(100 * mean(rowSums(calls) %in% c(0, M)), 1),
    n_regulated_min      = min(colSums(calls)),
    n_regulated_max      = max(colSums(calls))
  )

  ## ---------------------------------------------------------------------
  ## 2. Prior sensitivity: p0_frac
  ## ---------------------------------------------------------------------
  ref_a <- chains[[1]]$summary$a_mean

  p0_sens <- do.call(rbind, lapply(p0_grid, function(p0) {
    f <- fit_one(seed = 1, p0 = p0)
    data.frame(
      variety        = label,
      p0_frac        = p0,
      n_regulated    = sum(f$summary$regulated_95),
      pct_regulated  = round(100 * mean(f$summary$regulated_95), 1),
      median_abs_a1  = round(median(abs(f$summary$a_mean - 1)), 3),
      cor_with_ref   = round(cor(f$summary$a_mean, ref_a, method = "spearman"), 4)
    )
  }))

  ## ---------------------------------------------------------------------
  ## 3. Chain-length sensitivity
  ## ---------------------------------------------------------------------
  long_fit <- fit_one(seed = 1, n_iter = max(iter_grid))
  long_a   <- long_fit$summary$a_mean

  iter_sens <- do.call(rbind, lapply(iter_grid, function(ni) {
    f <- fit_one(seed = 1, n_iter = ni)
    data.frame(
      variety           = label,
      n_iter            = ni,
      n_regulated       = sum(f$summary$regulated_95),
      cor_with_longest  = round(cor(f$summary$a_mean, long_a, method = "spearman"), 4)
    )
  }))

  ## ---------------------------------------------------------------------
  ## 4. Credible-interval level sensitivity
  ## ---------------------------------------------------------------------
  tr <- chains[[1]]$trace_delta

  ci_sens <- do.call(rbind, lapply(c(0.80, 0.90, 0.95, 0.99), function(lev) {
    lo <- apply(tr, 1, quantile, (1 - lev) / 2)
    hi <- apply(tr, 1, quantile, 1 - (1 - lev) / 2)
    data.frame(
      variety       = label,
      ci_level      = lev,
      n_regulated   = sum(lo > 0 | hi < 0),
      pct_regulated = round(100 * mean(lo > 0 | hi < 0), 1)
    )
  }))

  list(
    convergence    = convergence,
    call_agreement = call_agreement,
    p0_sensitivity = p0_sens,
    iter_sensitivity = iter_sens,
    ci_sensitivity = ci_sens,
    rhat           = rhat,
    ess            = ess
  )
}
```

```{r}
#| label: bayes-diagnostics-run

cad_bd <- run_bayes_diagnostics(
  cad_Rl[cad_rna_resp, ], cad_Rs[cad_rna_resp, ],
  cad_Pl[cad_rna_resp, ], cad_Ps[cad_rna_resp, ],
  DESIGN$timepoints,
  t_half = cad_arch$t_half_M1_h[match(rownames(cad_Rl)[cad_rna_resp], cad_arch$gene_id)],
  label  = "Cadenza"
)

nor_bd <- run_bayes_diagnostics(
  nor_Rl[nor_rna_resp, ], nor_Rs[nor_rna_resp, ],
  nor_Pl[nor_rna_resp, ], nor_Ps[nor_rna_resp, ],
  DESIGN$timepoints,
  t_half = nor_arch$t_half_M1_h[match(rownames(nor_Rl)[nor_rna_resp], nor_arch$gene_id)],
  label  = "Norin"
)

knitr::kable(
  rbind(cad_bd$convergence, nor_bd$convergence),
  caption = "B7 -- MCMC convergence. R-hat near 1.00 means independent chains agree. Above 1.05 is a problem."
)

knitr::kable(
  rbind(cad_bd$call_agreement, nor_bd$call_agreement),
  caption = "Do independent chains make the SAME regulated/not call for each gene?"
)
```

```{r}
#| label: bayes-diagnostics-sensitivity

knitr::kable(
  rbind(cad_bd$p0_sensitivity, nor_bd$p0_sensitivity),
  caption = "B6 -- horseshoe prior sensitivity. `p0_frac` is the assumed fraction of genes under real regulation; 0.20 is the default used throughout."
)

knitr::kable(
  rbind(cad_bd$iter_sensitivity, nor_bd$iter_sensitivity),
  caption = "Chain-length sensitivity. If `cor_with_longest` is ~1 by 2000 iterations, the default chain is long enough."
)

knitr::kable(
  rbind(cad_bd$ci_sensitivity, nor_bd$ci_sensitivity),
  caption = "Credible-interval level sensitivity. Shows how much the `regulated` count depends on the (arbitrary) 95% convention."
)
```

```{r}
#| label: fig-bayes-diagnostics
#| fig-cap: "Convergence diagnostics: R-hat and effective sample size across genes."

op <- par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))

for (bd in list(cad_bd, nor_bd)) {

  hist(bd$rhat, breaks = 50, col = "#2C6FBB", border = NA,
       xlab = "R-hat", main = sprintf("%s: R-hat", bd$convergence$variety))
  abline(v = 1.05, col = "#D1495B", lwd = 2, lty = 2)

  hist(log10(pmax(bd$ess, 1)), breaks = 50, col = "#4C9F70", border = NA,
       xlab = "log10 effective sample size",
       main = sprintf("%s: ESS", bd$convergence$variety))
  abline(v = log10(400), col = "#D1495B", lwd = 2, lty = 2)
}

par(op)
```

## 9. Cross-variety validation via reciprocal-best-hit orthologs

Neither variety has simulated ground truth, so this is the one source of genuine external validation available: **do the two independently-run analyses agree on orthologous genes?** Agreement is not guaranteed even for correct biology (Cadenza and Norin can genuinely differ), but strong, widespread *disagreement* on high-confidence calls would flag a methodology problem rather than real biological divergence.

```{r}
#| label: rbh-load

rbh <- read.csv(file.path(DATA_DIR, "norin_cadenza_rbh-clean.csv"))
cat(sprintf("RBH ortholog pairs: %d (%d unique Norin, %d unique Cadenza)\n",
           nrow(rbh), length(unique(rbh$query)), length(unique(rbh$target))))

## Keep 1:1 RBH pairs only -- a gene appearing on both sides of more than one
## row is an ambiguous ortholog call and is excluded rather than guessed at.
rbh_1to1 <- rbh[!(duplicated(rbh$query) | duplicated(rbh$query, fromLast = TRUE)) &
                !(duplicated(rbh$target) | duplicated(rbh$target, fromLast = TRUE)), ]
cat(sprintf("1:1 RBH pairs: %d\n", nrow(rbh_1to1)))
```

```{r}
#| label: rbh-concordance

## Restrict to ortholog pairs where BOTH varieties had the gene in their own
## Universe B (matched RNA+protein, QC-passing) archetype table.
pair <- merge(rbh_1to1, cad_arch, by.x = "target", by.y = "gene_id")
pair <- merge(pair, nor_arch, by.x = "query", by.y = "gene_id", suffixes = c("_cad", "_nor"))
cat(sprintf("Ortholog pairs testable in both varieties: %d\n", nrow(pair)))

knitr::kable(table(Cadenza = pair$archetype_cad, Norin = pair$archetype_nor),
            caption = "Archetype agreement across 1:1 orthologs (Cadenza rows, Norin columns)")

## Amplitude correlation among genes both varieties called RNA-responsive --
## the most direct real-data check on whether the kinetic model is measuring
## something reproducible rather than dataset-specific noise.
both_resp <- pair$rna_responds_cad & pair$rna_responds_nor
cat(sprintf("\nBoth-variety RNA-responsive ortholog pairs: %d\n", sum(both_resp)))
if (sum(both_resp) > 10) {
  cc <- cor(pair$amplitude_a_cad[both_resp], pair$amplitude_a_nor[both_resp],
           method = "spearman", use = "complete.obs")
  cat(sprintf("Spearman(amplitude_a Cadenza, amplitude_a Norin) = %.2f\n", cc))
}

## The continuous amplitude correlation above is expected to be weak even for
## real, reproducible biology: each variety's per-gene amplitude is itself a
## noisy n=3 point estimate (the same reason §concordance-by-timepoint in
## 05_concordance_archetypes.R reports a DISATTENUATED correlation rather than
## the raw one -- two noisy proxies for the same true quantity correlate far
## more weakly than the true quantities do). The CATEGORICAL archetype call is
## more robust to that per-gene noise, so test it directly instead of reading
## the near-zero continuous correlation as "no reproducibility":
chisq <- suppressWarnings(chisq.test(table(pair$archetype_cad, pair$archetype_nor)))
cramers_v <- sqrt(unname(chisq$statistic) / (nrow(pair) * (min(dim(table(pair$archetype_cad, pair$archetype_nor))) - 1)))
cat(sprintf("\nChi-sq test, archetype agreement across orthologs: X2=%.1f, df=%d, p=%s\n",
           chisq$statistic, chisq$parameter, format.pval(chisq$p.value, digits = 3)))
cat(sprintf("Cramer's V: %.3f  (modest but real association, not attributable to chance)\n",
           cramers_v))
```

```{r}
#| label: fig-rbh-amplitude-concordance
#| fig-cap: "Cross-variety amplitude concordance for shared orthologs"

if (sum(both_resp) > 10) {
  plot(pair$amplitude_a_cad[both_resp], pair$amplitude_a_nor[both_resp], pch = 16, cex = 0.4,
       col = "#2C6FBB66", xlab = "amplitude a (Cadenza)", ylab = "amplitude a (Norin)",
       main = "Ortholog amplitude concordance,\nboth varieties RNA-responsive")
  abline(0, 1, col = "#D1495B", lwd = 2); abline(h = 1, v = 1, lty = 3, col = "grey60")
}
```

## 10. Summary and limitations

```{r}
#| label: summary-table

summ <- data.frame(
  variety = c("Cadenza", "Norin"),
  rna_genes_kept   = c(cad_qc_r$n_kept, nor_qc_r$n_kept),
  protein_kept     = c(cad_qc_p$n_kept, nor_qc_p$n_kept),
  universe_b       = c(length(cad_common), length(nor_common)),
  rna_responsive   = c(sum(cad_rna_resp), sum(nor_rna_resp)),
  lrt_regulated    = c(sum(cad_arch$regulated), sum(nor_arch$regulated)),
  bayes_regulated  = c(sum(cad_fit$summary$regulated_95), sum(nor_fit$summary$regulated_95)),
  integration_q2_c1 = c(round(cad_int$q2[1], 3), round(nor_int$q2[1], 3)),
  integration_p_c1  = c(signif(cad_int$p_q2[1], 3), signif(nor_int$p_q2[1], 3)))
knitr::kable(summ, caption = "Pipeline summary, both varieties")
```

> **Every number in this table carries the caveats set out in the Conclusions at the top of this document.** In particular `lrt_regulated` and `bayes_regulated` are counts whose magnitude depends on assumptions that are violated --- they are not stable quantities. The `integration_q2` columns are.

## 11. Closing --- what to carry forward

The full statement of what is and is not supported is in the **Conclusions** section at the top. In brief:

**Carry forward with confidence.** The Case B integration result (Q² against a permutation null on held-out design cells); the demonstration that transcript and protein responses are coupled but not identical; MNAR-dominant missingness; the finding that Cadenza is the better-behaved of the two datasets.

**Carry forward as descriptive, with the denominator stated.** The archetype composition; the amplitude distribution; comparisons between large gene sets; cross-variety agreement on orthologues.

**Do not carry forward.** Per-gene regulated/not-regulated calls; the count of regulated genes as a headline figure; any individual half-life; any mechanistic attribution to translation versus degradation; any raw DE count pending the reconciliation in `assumptions_validation.qmd` §2b.

### Method caveats specific to this notebook

-   **Protein QC has no decoy / contaminant / unique-peptide filter.** The supplied `*_protein_gene_mapping.csv` files carry no such columns. If the original MaxQuant / DIA-NN search output is available, rerun QC through `../../R/01_qc_normalise.R`'s full filter.
-   **The hierarchical model fixes each gene's half-life** at its profile-likelihood optimum rather than sampling it jointly with amplitude (see `07_bayesian_kinetics.R`'s header: joint sampling produces a slow-mixing ridge). Its credible intervals are therefore "amplitude uncertainty at the point-estimate kinetics", not full posterior uncertainty --- and they additionally exclude RNA error entirely.
-   **\~2% of RNA-responsive genes sit at the numerical safety bound** (`at_bound = TRUE`). Their amplitudes are directional ("at least this extreme"), not point estimates.
-   **No ground truth exists for real data.** §9's cross-variety orthologue concordance is the best available external check, not proof of correctness --- the two varieties are permitted to genuinely differ.

### Next steps

Ordered by how much they would change what can be claimed:

1.  **Resolve the DE reconciliation** (`assumptions_validation.qmd` §2b) --- 2--6× disagreement with the standalone limma workflow on identical counts. Upstream of every downstream number.
2.  **Propagate RNA uncertainty** into the kinetic model. This is what currently blocks per-gene claims.
3.  **Investigate Norin's 0 hpi baseline** (5.8% DE before infection can act).
4.  **Orthogonally validate 5--10 targets** (PRM / western / qPCR).
5.  `docs/planning.md` S10 (functional enrichment) --- run it on the kinetic archetypes, which passed their validation, rather than on the trajectory clusters, which did not.
6.  `docs/planning.md` §4B--4E remain unbuilt; evaluate against the existing `rnaprot/` codebase on `nectar` first (see project memory `proj_existing_nectar_rnaprot.md`).
