---
title: "DE proteomics - wheat"
author: "Kristina Gagalova"
format:
  gfm:
    toc: true
    code-tools: true
    fig-dir: "figs"
execute:
  echo: true
editor: visual
---

## Overview

Integrated RNA-seq / quantitative proteomics analysis of a wheat time-course treatment experiment, in two varieties (**Cadenza**, **Norin**), each with an independent **2 treatments x 4 timepoints x 3 replicates = 24 samples per omics layer** design.

**RNA and protein are measured on different biological samples within each variety** (unpaired / "Case B" design, per `../../CLAUDE.md` and `../../docs/PIPELINE.md`) --- not split aliquots of the same 24 samples. That rules out sample-level factor models (MOFA+/MEFISTO/DIABLO/O2PLS) and shapes every method choice below; see `../../docs/PIPELINE.md` and `../../docs/planning.md` for the full reasoning.

This notebook runs the complete pipeline end to end on the real data in `../../data/real/`:

1.  Data audit and design check
2.  QC & normalisation (RNA: `filterByExpr` + VST/voom; protein: validity filter + normalisation, adapted for this dataset -- see the note in §2)
3.  ID mapping (trivial here: `protein_id == gene_id` within each variety)
4.  Missingness diagnosis (MNAR/MAR) and imputation
5.  Temporal differential expression (`limma`, both layers)
6.  Kinetic-null / amplitude classification (the profile-likelihood model from `05_concordance_archetypes.R`)
7.  **Hierarchical Bayesian kinetics** (`07_bayesian_kinetics.R`) --- the new piece this notebook exists to run: a horseshoe-shrinkage model that rescues genes the profile-likelihood method above cannot classify on its own (the `kinetics_limited` problem this directory is named for)
8.  Case B design-cell integration (leave-one-cell-out Q\^2 + permutation null)
9.  Cross-variety validation using reciprocal-best-hit orthologs --- the one piece of genuine external validation available for real (non-simulated) data in this project

Steps 2--8 are defined once as reusable functions and applied to each variety; step 9 compares the two varieties' independent results against each other.

### Two things this dataset does not tell us --- edit before trusting the output

```{r}
#| label: edit-me-assumptions

## 1. WHICH TREATMENT IS THE REFERENCE?
##    Metadata encodes "T0"/"T1" with no further label. Assumed here: T0 is
##    the control/reference level, T1 is the treatment. If that is backwards,
##    every log2FC and every kinetic direction (amplified/buffered) below is
##    mirrored -- flip TREATMENT_REF to fix without re-deriving anything else.
TREATMENT_REF <- "T0"

## 2. WHAT IS THE REAL TIME SPACING?
##    Metadata encodes "t0".."t3" with no units. Assumed here: EVENLY SPACED,
##    1 arbitrary unit apart (0, 1, 2, 3). Every half-life reported in §7-8 is
##    in these SAME units, and is only interpretable as "hours" or "days" once
##    this vector is corrected to the true sampling times. Edit to the real
##    values (e.g. c(0, 6, 24, 48) for hours) before interpreting kinetics.
TIMEPOINT_VALUES <- c(0, 1, 2, 3)
TIME_UNIT_LABEL  <- c(0, 24, 48, 72)
#TIME_UNIT_LABEL  <- "time unit (EDIT: real spacing unknown -- see note above)"
```

## 1. Setup

```{r}
#| label: setup
#| message: false

## Quarto's default execution directory is this file's own folder
## (analysis/kinetics_limited/), while every R/*.R script in this project
## assumes the working directory is the PROJECT ROOT (matching how
## run_all.sh invokes them). Rather than fight that mismatch with a
## working-directory switch (fragile: root.dir changes how KNITR evaluates
## chunks, but does not change what a *sourced* script sees via
## sys.frame()$ofile, which is empty inside a knitr chunk anyway), every
## path below is resolved with here::here(), which walks up from the
## current directory looking for a project marker file -- this project has
## four (IntegrateProtRNA.Rproj, DESCRIPTION, renv.lock, .git) -- and returns
## an absolute path that is correct no matter where code is executed from.
## need_pkgs() itself is defined inside utils.R, so "here" -- the one package
## every subsequent path depends on -- is checked directly first.
if (!requireNamespace("here", quietly = TRUE))
  stop("package 'here' is required; install with install.packages('here')")
source(here::here("R", "utils.R"))
source(here::here("R", "pls_utils.R"))
source(here::here("R", "07_bayesian_kinetics.R"))
need_pkgs(c("limma", "edgeR", "DESeq2", "matrixStats", "impute", "here"))

set.seed(20260822)
DATA_DIR <- here::here("data", "real")

DESIGN <- list(
  treatments = c(TREATMENT_REF, setdiff(c("T0", "T1"), TREATMENT_REF)),
  reference  = TREATMENT_REF,
  timepoints = TIMEPOINT_VALUES,
  n_reps     = 3
)
CELLS <- as.vector(t(outer(DESIGN$treatments, DESIGN$timepoints,
                          function(a, b) sprintf("%s_t%s", a, b))))
```

## 2. Loading and QC (reusable functions)

```{r}
#| label: fn-load

#' Load one variety's RNA, protein, metadata and gene mapping.
#'
#' Metadata is a single file per variety, shared across both omics layers'
#' column headers -- the LH-style sample IDs happen to coincide between the
#' RNA and protein matrices for a given variety (both trace back to the same
#' field-plot label), but per the Case B design this is NOT treated as
#' sample-level pairing anywhere downstream: only variety x treatment x
#' timepoint CELL membership is used to link the two layers (see
#' `../../R/06_integration_caseB.R`'s header for why).
load_variety <- function(prefix) {

  data_dir <- here::here("data", "real")

  meta <- read.csv(
    file.path(data_dir, paste0(prefix, "_metadata.csv")),
    row.names = 1
  )

  meta$time_num  <- TIMEPOINT_VALUES[
    match(meta$timepoint, sort(unique(meta$timepoint)))
  ]
  meta$treatment <- factor(meta$treatment, levels = DESIGN$treatments)
  meta$cell <- factor(
    sprintf("%s_t%s", meta$treatment, meta$time_num),
    levels = CELLS
  )

  rna <- as.matrix(
    read.csv(
      file.path(data_dir, paste0(prefix, "-rnaseq.csv")),
      row.names = 1,
      check.names = FALSE
    )
  )

  prot <- as.matrix(
    read.csv(
      file.path(data_dir, paste0(prefix, "-prot.csv")),
      row.names = 1,
      check.names = FALSE
    )
  )

  gmap <- read.csv(
    file.path(data_dir, paste0(prefix, "_protein_gene_mapping.csv"))
  )

  stopifnot(
    all(colnames(rna) %in% rownames(meta)),
    all(colnames(prot) %in% rownames(meta))
  )

  rna  <- rna[, rownames(meta), drop = FALSE]
  prot <- prot[, rownames(meta), drop = FALSE]

  list(
    rna = rna,
    prot = prot,
    meta = meta,
    gmap = gmap
  )
}
```

```{r}
#| label: fn-qc

#' RNA QC: filterByExpr, then VST (integration matrix) + voom-ready counts.
#' Same procedure as `01_qc_normalise.R`, applied to real counts directly.
run_qc_rna <- function(counts, meta) {
  keep <- edgeR::filterByExpr(counts, group = meta$cell, min.count = 10, min.total.count = 15)
  cf <- counts[keep, , drop = FALSE]
  dge <- edgeR::calcNormFactors(edgeR::DGEList(cf))
  logcpm <- edgeR::cpm(dge, log = TRUE, prior.count = 3)
  dds <- DESeq2::DESeqDataSetFromMatrix(cf, meta, ~cell)
  dds <- DESeq2::estimateSizeFactors(dds)
  vst <- SummarizedExperiment::assay(
    DESeq2::vst(dds, blind = TRUE, nsub = min(1000, sum(rowMeans(cf) > 5))))
  list(counts = cf, logcpm = logcpm, vst = vst, n_in = nrow(counts), n_kept = nrow(cf))
}

#' Protein QC, adapted from `01_qc_normalise.R` for this dataset:
#' the real protein_gene_mapping.csv carries no unique-peptide/decoy/
#' contaminant columns (unlike the simulator's MaxQuant-style metadata), so
#' that part of 01's filtering is not applicable here and is skipped
#' explicitly rather than silently omitted. Zero is treated as "not
#' quantified" (standard LFQ-intensity convention), not a true biological
#' zero -- it becomes NA before the validity filter and normalisation.
run_qc_prot <- function(intensity, meta, min_valid_in_cell = 2) {
  m <- intensity; m[m == 0] <- NA; m <- log2(m)
  valid_per_cell <- vapply(levels(meta$cell), function(cl)
    rowSums(!is.na(m[, meta$cell == cl, drop = FALSE])), numeric(nrow(m)))
  keep <- matrixStats::rowMaxs(valid_per_cell) >= min_valid_in_cell
  mf <- m[keep, , drop = FALSE]
  ## median normalisation: shift each sample so its median matches the
  ## matrix-wide median -- corrects per-run loading, not biology.
  shift <- apply(mf, 2, median, na.rm = TRUE) - median(mf, na.rm = TRUE)
  mn <- sweep(mf, 2, shift, "-")
  list(norm = mn, n_in = nrow(intensity), n_kept = nrow(mf), miss_rate = mean(is.na(mn)))
}
```

```{r}
#| label: fn-missingness

#' MNAR/MAR diagnosis and mixed imputation, same rule as `03_missingness.R`:
#' a protein missing in EVERY replicate of a cell (but observed in some other
#' cell) is MNAR (left-censored -- genuinely low/absent there); missing in
#' only SOME replicates of its cell is MAR (stochastic). Imputed values are
#' for Universe A / integration use only -- the DE and kinetics steps below
#' use these imputed values too (protein has no complete-case alternative at
#' this scale), but the up-front validity filter (>=2 valid in >=1 cell)
#' keeps the imputation from manufacturing pure noise.
run_missingness <- function(prot_norm, meta) {
  cells <- levels(meta$cell)
  is_na <- is.na(prot_norm)
  n_obs_cell <- vapply(cells, function(cl)
    rowSums(!is.na(prot_norm[, meta$cell == cl, drop = FALSE])), numeric(nrow(prot_norm)))
  lab <- matrix("obs", nrow(prot_norm), ncol(prot_norm), dimnames = dimnames(prot_norm))
  for (k in seq_along(cells)) {
    ii <- which(meta$cell == cells[k]); gone <- n_obs_cell[, k] == 0
    lab[gone, ii] <- "MNAR"; lab[!gone, ii][is_na[!gone, ii]] <- "MAR"
  }
  lab[!is_na] <- "obs"

  mar_imp <- suppressWarnings(
    impute::impute.knn(prot_norm, k = 10, rowmax = 0.95, colmax = 0.95)$data)
  mnar_imp <- prot_norm
  for (j in seq_len(ncol(prot_norm))) {
    v <- prot_norm[, j]; nas <- is.na(v)
    if (!any(nas)) next
    mu <- mean(v, na.rm = TRUE); sdv <- sd(v, na.rm = TRUE)
    ## Perseus-style down-shifted normal: width 0.3, downshift 1.8 SD.
    mnar_imp[nas, j] <- rnorm(sum(nas), mu - 1.8 * sdv, 0.3 * sdv)
  }
  mixed <- prot_norm
  mixed[lab == "MAR"]  <- mar_imp[lab == "MAR"]
  mixed[lab == "MNAR"] <- mnar_imp[lab == "MNAR"]
  stopifnot(!anyNA(mixed))
  list(mixed = mixed, labels = lab,
       mnar_frac = mean(lab == "MNAR"), mar_frac = mean(lab == "MAR"))
}
```

```{r}
#| label: fn-de

#' Temporal DE, both layers, same design as `04_univariate_temporal_de.R`:
#' an 8-level ~0+cell design, per-timepoint treatment contrasts, and an
#' any-effect F-test. limma is used for both RNA (voom) and protein (trend +
#' robust eBayes) so effect sizes are on a comparable scale.
run_de <- function(mat, meta, is_rna) {
  design <- model.matrix(~0 + cell, meta); colnames(design) <- levels(meta$cell)
  cn <- function(tr, t) sprintf("%s_t%s", tr, t)
  alt <- setdiff(DESIGN$treatments, DESIGN$reference)
  tp <- setNames(sprintf("%s - %s", cn(alt, DESIGN$timepoints), cn(DESIGN$reference, DESIGN$timepoints)),
                paste0("tp_t", DESIGN$timepoints))
  it <- setNames(sprintf("(%s - %s) - (%s - %s)",
                         cn(alt, DESIGN$timepoints[-1]), cn(DESIGN$reference, DESIGN$timepoints[-1]),
                         cn(alt, DESIGN$timepoints[1]),  cn(DESIGN$reference, DESIGN$timepoints[1])),
                paste0("int_t", DESIGN$timepoints[-1]))
  if (is_rna) {
    dge <- edgeR::calcNormFactors(edgeR::DGEList(mat))
    v <- limma::voom(dge, design, plot = FALSE)
    fit <- limma::lmFit(v, design); trend <- FALSE
  } else {
    fit <- limma::lmFit(mat, design); trend <- TRUE
  }
  cm <- limma::makeContrasts(contrasts = c(tp, it), levels = design)
  colnames(cm) <- c(names(tp), names(it))
  f2 <- limma::eBayes(limma::contrasts.fit(fit, cm), trend = trend, robust = TRUE)

  lfc <- f2$coefficients[, names(tp), drop = FALSE]; colnames(lfc) <- paste0("t", DESIGN$timepoints)
  se  <- (f2$stdev.unscaled * sqrt(f2$s2.post))[, names(tp), drop = FALSE]; colnames(se) <- colnames(lfc)
  padj <- apply(f2$p.value[, names(tp), drop = FALSE], 2, p.adjust, "BH"); colnames(padj) <- colnames(lfc)

  ff <- limma::classifyTestsF(f2[, names(tp)], fstat.only = TRUE)
  ## as.numeric(ff) strips rownames -- reattach explicitly (caught in dev
  ## testing on real data: without this, any_fdr[gene_subset] silently
  ## returns NA for every gene instead of erroring).
  p_any <- setNames(pf(as.numeric(ff), attr(ff, "df1"), attr(ff, "df2"), lower.tail = FALSE),
                    rownames(f2))
  list(lfc = lfc, se = se, fdr = padj, any_fdr = p.adjust(p_any, "BH"))
}
```

## 3. Kinetic archetype classification (profile-likelihood, `05`-style)

```{r}
#| label: fn-archetypes

#' Nested kinetic model: M0 (protein trajectory fully explained by RNA
#' kinetics alone) vs M1 (M0 + a free amplitude factor `a`), compared by a
#' 1-df likelihood-ratio test. Identical model and identifiability handling
#' to `05_concordance_archetypes.R` -- see that script's header for the full
#' reasoning (this is what separates real post-transcriptional regulation
#' from a protein simply having a long half-life).
run_kinetic_archetypes <- function(Rl, Rs, Pl, Ps, tps, fdr_thresh = 0.05, min_abs_lfc = 0.5) {
  grid <- seq(0, max(tps), by = 0.25); gi <- match(tps, grid)
  r_grid <- t(apply(Rl, 1, function(v) approx(tps, v, grid, rule = 2)$y))
  r_lin  <- 2^r_grid
  th_bound <- max(tps)
  th_grid  <- exp(seq(log(0.5 * diff(tps)[1]), log(6 * th_bound), length.out = 80))
  kd_grid  <- log(2) / th_grid
  W <- 1 / pmax(Ps, 0.05)^2; SST <- rowSums(W * Pl^2)
  n <- nrow(Pl); J <- length(kd_grid)
  ssr0 <- ssr1 <- ahat <- matrix(NA_real_, n, J); preds <- vector("list", J)
  for (j in seq_len(J)) {
    kd <- kd_grid[j]
    pl <- log2(pmax(integrate_protein_mat(grid, r_lin, ks = kd, kd = kd, P0 = rep(1, n))[, gi, drop = FALSE], 1e-9))
    preds[[j]] <- pl
    ssr0[, j] <- rowSums(W * (pl - Pl)^2)
    num <- rowSums(W * pl * Pl); den <- rowSums(W * pl^2)
    a <- ifelse(den > 1e-12, num / den, NA_real_); ahat[, j] <- a
    ssr1[, j] <- SST - ifelse(den > 1e-12, num^2 / den, 0)
  }
  j0 <- max.col(-ssr0, "first"); j1 <- max.col(-ssr1, "first")
  ix <- function(j) cbind(seq_len(n), j)
  S0 <- ssr0[ix(j0)]; S1 <- ssr1[ix(j1)]
  t_half_M0 <- th_grid[j0]; t_half_M1 <- th_grid[j1]; a_hat <- ahat[ix(j1)]
  pred0 <- t(vapply(seq_len(n), function(i) preds[[j0[i]]][i, ], numeric(length(tps))))
  lrt <- pmax(S0 - S1, 0); p_lrt <- pchisq(lrt, 1, lower.tail = FALSE); fdr_lrt <- p.adjust(p_lrt, "BH")

  max_abs <- function(m) apply(abs(m[, -1, drop = FALSE]), 1, max)
  rna_resp <- max_abs(Rl) > min_abs_lfc
  regulated <- rna_resp & (fdr_lrt < fdr_thresh)
  identifiable <- t_half_M0 <= th_bound; slow <- t_half_M0 > 0.5 * th_bound

  archetype <- rep("unchanged", n)
  archetype[!rna_resp & max_abs(Pl) > min_abs_lfc] <- "protein_only"
  archetype[rna_resp & !regulated &  identifiable & !slow] <- "concordant"
  archetype[rna_resp & !regulated &  identifiable &  slow] <- "kinetic_lag"
  archetype[rna_resp & !regulated & !identifiable]         <- "kinetics_limited"
  archetype[regulated & a_hat >= 1]            <- "amplified"
  archetype[regulated & a_hat < 1 & a_hat > 0] <- "buffered"
  archetype[regulated & a_hat <= 0]            <- "anticorrelated"

  data.frame(gene_id = rownames(Pl), archetype = archetype, rna_responds = rna_resp,
            regulated = regulated, amplitude_a = a_hat,
            t_half_M0_h = t_half_M0, t_half_M1_h = t_half_M1,
            identifiable = identifiable, lrt_p = p_lrt, lrt_fdr = fdr_lrt,
            stringsAsFactors = FALSE)
}
```

## 4. Case B integration (`06`-style)

```{r}
#| label: fn-integration

#' Design-cell PLS with leave-one-cell-out Q^2 against a permutation null --
#' the Case B analogue of a MEFISTO/DIABLO variance-explained figure. See
#' `06_integration_caseB.R`'s header for why the in-sample cross-block
#' correlation is not a valid statistic here (it reaches ~0.7-0.9 even on
#' permuted data at this sample size).
run_integration <- function(rna_mat, prot_mat, cd_r, cd_p, n_hvg = 3000, n_hvp = 2000,
                            n_comp = 3, n_perm = 200) {
  hv_r <- top_variable(rna_mat,  min(n_hvg, nrow(rna_mat)))
  hv_p <- top_variable(prot_mat, min(n_hvp, nrow(prot_mat)))
  Xc <- prep_block(cell_means(rna_mat[hv_r, ],  cd_r))
  Yc <- prep_block(cell_means(prot_mat[hv_p, ], cd_p))
  stopifnot(identical(rownames(Xc), rownames(Yc)))
  K  <- min(n_comp, nrow(Xc) - 3)
  fit <- pls2(Xc, Yc, ncomp = K)
  Ax <- row_space(Xc); Ay <- row_space(Yc)
  q2 <- q2_loo(Ax, Ay, K)
  perm_q2 <- matrix(NA_real_, n_perm, K)
  for (b in seq_len(n_perm)) perm_q2[b, ] <- q2_loo(Ax, Ay[sample(nrow(Ay)), , drop = FALSE], K)
  p_q2 <- vapply(seq_len(K), function(k) (1 + sum(perm_q2[, k] >= q2[k])) / (n_perm + 1), 0)
  list(fit = fit, q2 = q2, perm_q2 = perm_q2, p_q2 = p_q2,
       cellmeta = data.frame(cell = rownames(Xc)))
}
```

## 5. Run the pipeline: Cadenza

```{r}
#| label: cadenza-load

cad <- load_variety("cadenza")
knitr::kable(table(cad$meta$treatment, cad$meta$time_num),
            caption = "Cadenza design: samples per treatment x timepoint")
```

```{r}
#| label: cadenza-qc

cad_qc_r <- run_qc_rna(cad$rna, cad$meta)
cad_qc_p <- run_qc_prot(cad$prot, cad$meta)
cat(sprintf("RNA: %d / %d genes kept by filterByExpr\n", cad_qc_r$n_kept, cad_qc_r$n_in))
cat(sprintf("Protein: %d / %d groups kept (>=2 valid in >=1 cell); missing rate %.1f%%\n",
           cad_qc_p$n_kept, cad_qc_p$n_in, 100 * cad_qc_p$miss_rate))
```

```{r}
#| label: fig-cadenza-qc-pca
#| fig-cap: "Cadenza PCA, RNA (VST) and protein (normalised), coloured by treatment"

pca_plot <- function(m, meta, main) {
  mv <- m[order(row_var(m), decreasing = TRUE, na.last = NA)[seq_len(min(500, nrow(m)))], , drop = FALSE]
  mv <- mv[complete.cases(mv), , drop = FALSE]
  p <- prcomp(t(mv), scale. = FALSE)
  v <- round(100 * p$sdev^2 / sum(p$sdev^2), 1)
  plot(p$x[, 1], p$x[, 2], pch = c(16, 17)[as.integer(meta$treatment)],
       col = hcl.colors(4, "Zissou1")[as.integer(factor(meta$time_num))], cex = 1.6,
       xlab = sprintf("PC1 (%.1f%%)", v[1]), ylab = sprintf("PC2 (%.1f%%)", v[2]), main = main)
  legend("topright", bty = "n", cex = 0.7, legend = levels(meta$treatment), pch = c(16, 17))
}
op <- par(mfrow = c(1, 2))
pca_plot(cad_qc_r$vst,       cad$meta, "Cadenza RNA (VST)")
pca_plot(cad_qc_p$norm,      cad$meta, "Cadenza protein (normalised)")
par(op)
```

```{r}
#| label: cadenza-missingness

cad_mi <- run_missingness(cad_qc_p$norm, cad$meta)
cat(sprintf("MNAR: %.1f%%  MAR: %.1f%%  (of all entries)\n",
           100 * cad_mi$mnar_frac, 100 * cad_mi$mar_frac))
```

```{r}
#| label: cadenza-de

## protein_id == gene_id within a variety (verified in ../../data/real/
## *_protein_gene_mapping.csv), so Universe B == QC'd protein genes that also
## survive RNA QC -- no bipartite mapping ambiguity for this dataset.
cad_de_rna  <- run_de(cad_qc_r$counts, cad$meta, TRUE)
cad_common  <- intersect(rownames(cad_mi$mixed), rownames(cad_qc_r$counts))
cad_de_prot <- run_de(cad_mi$mixed[cad_common, , drop = FALSE], cad$meta, FALSE)

cat(sprintf("Universe B: %d matched genes\n", length(cad_common)))
cat(sprintf("RNA any-effect FDR<0.05: %d\n",  sum(cad_de_rna$any_fdr[cad_common] < 0.05)))
cat(sprintf("Protein any-effect FDR<0.05: %d\n", sum(cad_de_prot$any_fdr < 0.05)))
```

```{r}
#| label: cadenza-archetypes

cad_Rl <- cad_de_rna$lfc[cad_common, ];  cad_Rs <- cad_de_rna$se[cad_common, ]
cad_Pl <- cad_de_prot$lfc;               cad_Ps <- cad_de_prot$se
cad_Rl <- cad_Rl - cad_Rl[, 1]; cad_Pl <- cad_Pl - cad_Pl[, 1]

cad_arch <- run_kinetic_archetypes(cad_Rl, cad_Rs, cad_Pl, cad_Ps, DESIGN$timepoints)
knitr::kable(as.data.frame(table(archetype = cad_arch$archetype)),
            caption = "Cadenza: profile-likelihood archetype calls")
```

## 6. Hierarchical Bayesian kinetics: Cadenza

```{r}
#| label: cadenza-bayesian-kinetics

## RNA-responsive gate (mirrors 05's rna_resp / the doc note in
## 07_bayesian_kinetics.R): the amplitude model is only meaningful when RNA
## carries real signal -- for a flat-RNA gene ANY amplitude "explains" a tiny
## predicted change equally badly, which (confirmed in dev testing on this
## exact dataset) can otherwise drive a gene to the numerical safety bound.
cad_rna_resp <- (cad_de_rna$any_fdr[cad_common] < 0.05) &
  (apply(abs(cad_Rl[, -1, drop = FALSE]), 1, max) > 0.5)
cat(sprintf("RNA-responsive, eligible for amplitude fitting: %d / %d\n",
           sum(cad_rna_resp), length(cad_rna_resp)))

cad_fit <- fit_hierarchical_kinetics(
  cad_Rl[cad_rna_resp, ], cad_Rs[cad_rna_resp, ],
  cad_Pl[cad_rna_resp, ], cad_Ps[cad_rna_resp, ],
  ## Anchored at the SAME joint-optimal half-life the profile-likelihood fit
  ## above already found (t_half_M1_h), so the two analyses share one kinetic
  ## fit and differ only in how the amplitude is inferred (independent
  ## per-gene LRT vs. hierarchical horseshoe shrinkage).
  t_half = cad_arch$t_half_M1_h[match(rownames(cad_Rl)[cad_rna_resp], cad_arch$gene_id)],
  tps = DESIGN$timepoints, n_iter = 2000, warmup = 800, seed = 1, verbose = FALSE)

cat(sprintf("MCMC acceptance rate: %.2f\n", cad_fit$accept_delta))
cat(sprintf("Genes at the numerical safety bound (report as directional, not exact): %d / %d\n",
           sum(cad_fit$summary$at_bound), nrow(cad_fit$summary)))
cat(sprintf("Regulated by hierarchical model (95%% CI excludes a=1): %d / %d\n",
           sum(cad_fit$summary$regulated_95), nrow(cad_fit$summary)))
cat(sprintf("Regulated by profile-likelihood LRT (FDR<0.05), same genes: %d / %d\n",
           sum(cad_arch$regulated[match(rownames(cad_Rl)[cad_rna_resp], cad_arch$gene_id)]),
           sum(cad_rna_resp)))
```

```{r}
#| label: fig-cadenza-kinetic-volcano
#| fig-cap: "Cadenza kinetic volcano: hierarchical posterior amplitude vs. profile-likelihood significance"

idx <- match(cad_fit$summary$gene_id, cad_arch$gene_id)
plot(cad_fit$summary$a_mean, -log10(pmax(cad_arch$lrt_p[idx], 1e-12)), pch = 16, cex = 0.4,
     col = ifelse(cad_fit$summary$regulated_95, "#D1495B99", "#00000033"),
     xlab = "hierarchical posterior mean amplitude (a)",
     ylab = "-log10 p (profile-likelihood LRT)",
     main = "Cadenza: two independent regulation signals\n(red = hierarchical model flags as regulated)")
abline(v = 1, lty = 3)
```

The two methods agree on genes with strong signal and differ most on genes the profile-likelihood LRT was underpowered for individually (§`docs/PIPELINE.md` documents this trade-off quantitatively on simulated ground truth: the hierarchical model roughly triples recall at the cost of a higher false-call rate on truly-unregulated genes, since fixing the half-life removes one source of legitimate uncertainty from its interval). Genes flagged by **both** methods are the highest-confidence regulated set; genes flagged by the hierarchical model **only** are candidates for follow-up validation, not standalone claims.

```{r}
#| label: cadenza-agreement-table

cad_kin <- data.frame(
  gene_id = rownames(cad_Rl)[cad_rna_resp],
  lrt_regulated  = cad_arch$regulated[match(rownames(cad_Rl)[cad_rna_resp], cad_arch$gene_id)],
  bayes_regulated = cad_fit$summary$regulated_95,
  a_mean = cad_fit$summary$a_mean, at_bound = cad_fit$summary$at_bound,
  stringsAsFactors = FALSE)
knitr::kable(as.data.frame(table(LRT = cad_kin$lrt_regulated, Bayesian = cad_kin$bayes_regulated)),
            caption = "Cadenza: agreement between the two regulation calls")
```

## 7. Case B integration: Cadenza

```{r}
#| label: cadenza-integration

cad_int <- run_integration(cad_qc_r$vst, cad_mi$mixed, cad$meta, cad$meta)
knitr::kable(data.frame(component = seq_along(cad_int$q2), Q2 = round(cad_int$q2, 3),
                        perm_q95 = round(apply(cad_int$perm_q2, 2, quantile, .95), 3),
                        p_value = signif(cad_int$p_q2, 3)),
            caption = "Cadenza: leave-one-cell-out Q2 vs permutation null")
```

## 8. Run the pipeline: Norin

Identical steps, applied to the second variety.

```{r}
#| label: norin-load

nor <- load_variety("norin")
knitr::kable(table(nor$meta$treatment, nor$meta$time_num),
            caption = "Norin design: samples per treatment x timepoint")

nor_qc_r <- run_qc_rna(nor$rna, nor$meta)
nor_qc_p <- run_qc_prot(nor$prot, nor$meta)
cat(sprintf("RNA: %d / %d genes kept\n", nor_qc_r$n_kept, nor_qc_r$n_in))
cat(sprintf("Protein: %d / %d groups kept; missing rate %.1f%%\n",
           nor_qc_p$n_kept, nor_qc_p$n_in, 100 * nor_qc_p$miss_rate))

nor_mi <- run_missingness(nor_qc_p$norm, nor$meta)
cat(sprintf("MNAR: %.1f%%  MAR: %.1f%%\n", 100 * nor_mi$mnar_frac, 100 * nor_mi$mar_frac))
```

```{r}
#| label: norin-de-archetypes

nor_de_rna  <- run_de(nor_qc_r$counts, nor$meta, TRUE)
nor_common  <- intersect(rownames(nor_mi$mixed), rownames(nor_qc_r$counts))
nor_de_prot <- run_de(nor_mi$mixed[nor_common, , drop = FALSE], nor$meta, FALSE)
cat(sprintf("Universe B: %d matched genes\n", length(nor_common)))

nor_Rl <- nor_de_rna$lfc[nor_common, ];  nor_Rs <- nor_de_rna$se[nor_common, ]
nor_Pl <- nor_de_prot$lfc;               nor_Ps <- nor_de_prot$se
nor_Rl <- nor_Rl - nor_Rl[, 1]; nor_Pl <- nor_Pl - nor_Pl[, 1]

nor_arch <- run_kinetic_archetypes(nor_Rl, nor_Rs, nor_Pl, nor_Ps, DESIGN$timepoints)
knitr::kable(as.data.frame(table(archetype = nor_arch$archetype)),
            caption = "Norin: profile-likelihood archetype calls")
```

```{r}
#| label: norin-bayesian-kinetics

nor_rna_resp <- (nor_de_rna$any_fdr[nor_common] < 0.05) &
  (apply(abs(nor_Rl[, -1, drop = FALSE]), 1, max) > 0.5)
cat(sprintf("RNA-responsive, eligible for amplitude fitting: %d / %d\n",
           sum(nor_rna_resp), length(nor_rna_resp)))

nor_fit <- fit_hierarchical_kinetics(
  nor_Rl[nor_rna_resp, ], nor_Rs[nor_rna_resp, ],
  nor_Pl[nor_rna_resp, ], nor_Ps[nor_rna_resp, ],
  t_half = nor_arch$t_half_M1_h[match(rownames(nor_Rl)[nor_rna_resp], nor_arch$gene_id)],
  tps = DESIGN$timepoints, n_iter = 2000, warmup = 800, seed = 1, verbose = FALSE)

cat(sprintf("MCMC acceptance rate: %.2f\n", nor_fit$accept_delta))
cat(sprintf("Regulated by hierarchical model: %d / %d\n",
           sum(nor_fit$summary$regulated_95), nrow(nor_fit$summary)))
```

```{r}
#| label: norin-integration

nor_int <- run_integration(nor_qc_r$vst, nor_mi$mixed, nor$meta, nor$meta)
knitr::kable(data.frame(component = seq_along(nor_int$q2), Q2 = round(nor_int$q2, 3),
                        perm_q95 = round(apply(nor_int$perm_q2, 2, quantile, .95), 3),
                        p_value = signif(nor_int$p_q2, 3)),
            caption = "Norin: leave-one-cell-out Q2 vs permutation null")
```

## 9. Cross-variety validation via reciprocal-best-hit orthologs

Neither variety has simulated ground truth, so this is the one source of genuine external validation available: **do the two independently-run analyses agree on orthologous genes?** Agreement is not guaranteed even for correct biology (Cadenza and Norin can genuinely differ), but strong, widespread *disagreement* on high-confidence calls would flag a methodology problem rather than real biological divergence.

```{r}
#| label: rbh-load

rbh <- read.csv(file.path(DATA_DIR, "norin_cadenza_rbh-clean.csv"))
cat(sprintf("RBH ortholog pairs: %d (%d unique Norin, %d unique Cadenza)\n",
           nrow(rbh), length(unique(rbh$query)), length(unique(rbh$target))))

## Keep 1:1 RBH pairs only -- a gene appearing on both sides of more than one
## row is an ambiguous ortholog call and is excluded rather than guessed at.
rbh_1to1 <- rbh[!(duplicated(rbh$query) | duplicated(rbh$query, fromLast = TRUE)) &
                !(duplicated(rbh$target) | duplicated(rbh$target, fromLast = TRUE)), ]
cat(sprintf("1:1 RBH pairs: %d\n", nrow(rbh_1to1)))
```

```{r}
#| label: rbh-concordance

## Restrict to ortholog pairs where BOTH varieties had the gene in their own
## Universe B (matched RNA+protein, QC-passing) archetype table.
pair <- merge(rbh_1to1, cad_arch, by.x = "target", by.y = "gene_id")
pair <- merge(pair, nor_arch, by.x = "query", by.y = "gene_id", suffixes = c("_cad", "_nor"))
cat(sprintf("Ortholog pairs testable in both varieties: %d\n", nrow(pair)))

knitr::kable(table(Cadenza = pair$archetype_cad, Norin = pair$archetype_nor),
            caption = "Archetype agreement across 1:1 orthologs (Cadenza rows, Norin columns)")

## Amplitude correlation among genes both varieties called RNA-responsive --
## the most direct real-data check on whether the kinetic model is measuring
## something reproducible rather than dataset-specific noise.
both_resp <- pair$rna_responds_cad & pair$rna_responds_nor
cat(sprintf("\nBoth-variety RNA-responsive ortholog pairs: %d\n", sum(both_resp)))
if (sum(both_resp) > 10) {
  cc <- cor(pair$amplitude_a_cad[both_resp], pair$amplitude_a_nor[both_resp],
           method = "spearman", use = "complete.obs")
  cat(sprintf("Spearman(amplitude_a Cadenza, amplitude_a Norin) = %.2f\n", cc))
}

## The continuous amplitude correlation above is expected to be weak even for
## real, reproducible biology: each variety's per-gene amplitude is itself a
## noisy n=3 point estimate (the same reason §concordance-by-timepoint in
## 05_concordance_archetypes.R reports a DISATTENUATED correlation rather than
## the raw one -- two noisy proxies for the same true quantity correlate far
## more weakly than the true quantities do). The CATEGORICAL archetype call is
## more robust to that per-gene noise, so test it directly instead of reading
## the near-zero continuous correlation as "no reproducibility":
chisq <- suppressWarnings(chisq.test(table(pair$archetype_cad, pair$archetype_nor)))
cramers_v <- sqrt(unname(chisq$statistic) / (nrow(pair) * (min(dim(table(pair$archetype_cad, pair$archetype_nor))) - 1)))
cat(sprintf("\nChi-sq test, archetype agreement across orthologs: X2=%.1f, df=%d, p=%s\n",
           chisq$statistic, chisq$parameter, format.pval(chisq$p.value, digits = 3)))
cat(sprintf("Cramer's V: %.3f  (modest but real association, not attributable to chance)\n",
           cramers_v))
```

```{r}
#| label: fig-rbh-amplitude-concordance
#| fig-cap: "Cross-variety amplitude concordance for shared orthologs"

if (sum(both_resp) > 10) {
  plot(pair$amplitude_a_cad[both_resp], pair$amplitude_a_nor[both_resp], pch = 16, cex = 0.4,
       col = "#2C6FBB66", xlab = "amplitude a (Cadenza)", ylab = "amplitude a (Norin)",
       main = "Ortholog amplitude concordance,\nboth varieties RNA-responsive")
  abline(0, 1, col = "#D1495B", lwd = 2); abline(h = 1, v = 1, lty = 3, col = "grey60")
}
```

## 10. Summary and limitations

```{r}
#| label: summary-table

summ <- data.frame(
  variety = c("Cadenza", "Norin"),
  rna_genes_kept   = c(cad_qc_r$n_kept, nor_qc_r$n_kept),
  protein_kept     = c(cad_qc_p$n_kept, nor_qc_p$n_kept),
  universe_b       = c(length(cad_common), length(nor_common)),
  rna_responsive   = c(sum(cad_rna_resp), sum(nor_rna_resp)),
  lrt_regulated    = c(sum(cad_arch$regulated), sum(nor_arch$regulated)),
  bayes_regulated  = c(sum(cad_fit$summary$regulated_95), sum(nor_fit$summary$regulated_95)),
  integration_q2_c1 = c(round(cad_int$q2[1], 3), round(nor_int$q2[1], 3)),
  integration_p_c1  = c(signif(cad_int$p_q2[1], 3), signif(nor_int$p_q2[1], 3)))
knitr::kable(summ, caption = "Pipeline summary, both varieties")
```

**Read before using any downstream number:**

-   **Timepoint units are unconfirmed** (§ "Two things this dataset does not tell us"). Every half-life in this notebook is in `r TIME_UNIT_LABEL`, not hours, until `TIMEPOINT_VALUES` is corrected to the real sampling schedule.
-   **Treatment reference is assumed**, not confirmed from metadata beyond the `T0`/`T1` labels themselves.
-   **Protein QC has no decoy/contaminant/unique-peptide filter** here (§2): the provided `*_protein_gene_mapping.csv` files do not carry that MaxQuant/ DIA-NN metadata. If the original search-engine output is available, rerun QC with `../../R/01_qc_normalise.R`'s full filter instead.
-   **The hierarchical Bayesian model fixes each gene's half-life** at its own profile-likelihood joint optimum rather than sampling it jointly with amplitude (see `07_bayesian_kinetics.R`'s header for why: jointly sampling both produces a slow-mixing ridge). This sacrifices joint kinetic/amplitude uncertainty propagation -- its credible intervals should be read as "amplitude uncertainty at the point-estimate kinetics," not full posterior uncertainty.
-   **\~2% of RNA-responsive genes hit the model's numerical safety bound** (`at_bound = TRUE` in the summary tables) even after gating to RNA- responsive genes. These are real, strong-signal genes, not degenerate artefacts (verified during development) -- but their reported amplitude should be read as directional ("at least this extreme"), not a precise estimate.
-   **No ground truth exists for real data.** §9's cross-variety ortholog concordance is the best available external check, not proof of correctness -- Cadenza and Norin are permitted to genuinely differ.

### Next steps

-   Confirm `TREATMENT_REF` and `TIMEPOINT_VALUES` against the actual experimental record and re-render.
-   If the original MaxQuant/DIA-NN search output is available, rerun protein QC through the full `01_qc_normalise.R` filter (decoys, contaminants, unique-peptide count).
-   `docs/planning.md` §4B-4E (GNN residuals, translation-delay atlas, XGBoost/SHAP buffering predictors) remain unbuilt; evaluate against the existing `rnaprot/` codebase on `nectar` before building new versions (see project memory `proj_existing_nectar_rnaprot.md`).
