---
title: "What this design can tell us about protein kinetics"
author: "Kristina Gagalova"
format:
  gfm:
    toc: true
    code-tools: true
    fig-dir: "figs"
execute:
  echo: true
editor: visual
bibliography: references.bib
---

## Purpose

Six claims about protein kinetics could in principle be made from this dataset. This notebook states each one, **attacks it with the strongest test available**, and reports whether it survives. Where a claim fails, the notebook says so and gives the wording that *is* supportable instead.

The design under examination: wheat, two varieties, infected vs uninfected, 0/24/48/72 hpi, three biological replicates, **RNA and protein measured on different plants** (unpaired, "Case B").

Every claim below is graded:

| Grade                | Meaning                                               |
|:---------------------|:------------------------------------------------------|
| ✅ **SAFE**          | Survived an adversarial test designed to break it     |
| ⚠️ **CONDITIONAL**   | Supportable only with an explicit caveat attached     |
| ❌ **NOT SUPPORTED** | Fails, or is mathematically impossible from this data |

------------------------------------------------------------------------

## Setup

```{r}
#| label: setup
#| message: false

if (!requireNamespace("here", quietly = TRUE)) stop("package 'here' required")

## RENV_PROJECT must be set before sourcing activate.R -- otherwise renv
## treats this subdirectory as the project and bootstraps an empty library.
if (file.exists(here::here("renv", "activate.R"))) {
  Sys.setenv(RENV_PROJECT = here::here())
  source(here::here("renv", "activate.R"))
}

source(here::here("R", "utils.R"))
source(here::here("R", "wheat_pipeline.R"))
source(here::here("R", "pls_utils.R"))
need_pkgs(c("limma", "edgeR", "DESeq2", "matrixStats", "impute"))

set.seed(20260823)
DESIGN <- wheat_design()

cad <- prepare_variety("cadenza", DESIGN)
nor <- prepare_variety("norin",   DESIGN)
VARIETIES <- list(Cadenza = cad, Norin = nor)

TPS <- DESIGN$timepoints
```

```{r}
#| label: fn-core

#' Profile-likelihood fit of the kinetic null (M0) and the amplitude model
#' (M1). Returns point estimates plus the M0 prediction, which is what the
#' null simulations in K2 are generated from.
fit_amplitude <- function(Rl, Pl, Ps, tps, n_grid = 60) {

  grid <- seq(0, max(tps), by = 0.25)
  gi   <- match(tps, grid)
  r_lin <- 2^t(apply(Rl, 1, function(v) approx(tps, v, grid, rule = 2)$y))

  th <- exp(seq(log(0.5 * diff(tps)[1]), log(6 * max(tps)), length.out = n_grid))
  kd <- log(2) / th

  W   <- 1 / pmax(Ps, 0.05)^2
  SST <- rowSums(W * Pl^2)
  n   <- nrow(Pl)

  ssr0 <- ssr1 <- ah <- matrix(NA_real_, n, n_grid)
  pr <- vector("list", n_grid)

  for (j in seq_len(n_grid)) {
    pl <- integrate_protein_mat(grid, r_lin, ks = kd[j], kd = kd[j], P0 = rep(1, n))
    pl <- log2(pmax(pl[, gi, drop = FALSE], 1e-9))
    pr[[j]] <- pl

    ssr0[, j] <- rowSums(W * (pl - Pl)^2)
    num <- rowSums(W * pl * Pl); den <- rowSums(W * pl^2)
    ah[, j]   <- ifelse(den > 1e-12, num / den, NA_real_)
    ssr1[, j] <- SST - ifelse(den > 1e-12, num^2 / den, 0)
  }

  j0 <- max.col(-ssr0, "first"); j1 <- max.col(-ssr1, "first")
  ix <- function(j) cbind(seq_len(n), j)
  lrt <- pmax(ssr0[ix(j0)] - ssr1[ix(j1)], 0)

  list(a         = ah[ix(j1)],
       t_half    = th[j0],
       fdr       = p.adjust(pchisq(lrt, 1, lower.tail = FALSE), "BH"),
       pred_null = t(vapply(seq_len(n), function(i) pr[[j0[i]]][i, ],
                            numeric(length(tps)))),
       resid     = Pl - t(vapply(seq_len(n), function(i) pr[[j0[i]]][i, ],
                                 numeric(length(tps)))))
}

#' RNA-responsive gate: the amplitude is only meaningful where RNA moves.
gate <- function(v) {
  (v$de_rna$any_fdr[rownames(v$Rl)] < 0.05) &
    (apply(abs(v$Rl[, -1, drop = FALSE]), 1, max) > 0.5)
}

FITS <- lapply(VARIETIES, function(v) {
  k <- gate(v)
  list(keep = k,
       fit = fit_amplitude(v$Rl[k, , drop = FALSE], v$Pl[k, , drop = FALSE],
                           v$Ps[k, , drop = FALSE], TPS))
})
```

------------------------------------------------------------------------

# K1 --- "RNA and protein responses are coupled" ✅ SAFE

## The claim

The transcriptome of an infected wheat plant carries information about its proteome, beyond what the experimental design labels already say.

## Why this needs an adversarial test

Both layers respond to infection. A model handed the RNA block might therefore predict the proteome purely through *"was this cell infected, and how late"* --- information already contained in the treatment and timepoint labels. If so, "RNA predicts protein" would collapse into "both layers know the treatment", which is trivially true and is not integration.

A second hazard is specific to this sample size. With 8 design cells and thousands of features, an in-sample cross-block correlation reaches r ≈ 0.72 **on permuted data**. Any claim resting on in-sample correlation is worthless here. This is the standard PLS overfitting trap in small-n omics, and the accepted remedy is cross-validated Q² judged against a permutation null [@westerhuis2008].

## The test

Three predictors of the protein block, all scored by the same leave-one-design-cell-out Q²:

1.  the **RNA block**,
2.  the **design labels alone** (treatment; treatment + timepoint),
3.  the RNA block **after both blocks are residualised on the design** --- the strictest version, asking whether coupling survives once the shared infection response is removed from both sides.

```{r}
#| label: k1-test

q2_by_predictor <- function(v, label) {

  cells <- levels(v$meta$cell)
  cm <- data.frame(cell = cells,
                   treatment = factor(sub("_t.*$", "", cells)),
                   time_f    = factor(as.numeric(sub("^.*_t", "", cells))))

  Xc <- prep_block(cell_means(v$qc_rna$vst[top_variable(v$qc_rna$vst, 3000), ], v$meta))
  Yc <- prep_block(cell_means(v$imputed$mixed[top_variable(v$imputed$mixed, 2000), ], v$meta))
  Ay <- row_space(Yc)

  D <- model.matrix(~ treatment + time_f, cm)
  resid_on <- function(M) M - D %*% qr.solve(D, M)

  preds <- list(
    `RNA block`              = row_space(Xc),
    `design: treatment`      = model.matrix(~ treatment, cm)[, -1, drop = FALSE],
    `design: treatment+time` = model.matrix(~ treatment + time_f, cm)[, -1, drop = FALSE]
  )

  out <- do.call(rbind, lapply(names(preds), function(pn) {
    X <- preds[[pn]]
    K <- min(2, ncol(X), nrow(X) - 3)
    data.frame(variety = label, predictor = pn, n_dim = ncol(X),
               Q2 = round(q2_loo(X, Ay, K)[1], 4))
  }))

  Xr <- row_space(resid_on(Xc)); Yr <- row_space(resid_on(Yc))
  K  <- min(2, ncol(Xr), nrow(Xr) - 3)
  if (K >= 1) {
    out <- rbind(out, data.frame(
      variety = label, predictor = "RNA | design removed",
      n_dim = ncol(Xr), Q2 = round(q2_loo(Xr, Yr, K)[1], 4)))
  }
  out
}

k1 <- do.call(rbind, Map(q2_by_predictor, VARIETIES, names(VARIETIES)))
knitr::kable(k1, caption = "K1: leave-one-cell-out Q2 by predictor. Negative Q2 means worse than predicting the mean.")
```

```{r}
#| label: fig-k1
#| fig-cap: "Predictive power of the RNA block against design-label baselines. Only the RNA block achieves positive Q2."
#| fig-width: 9
#| fig-height: 5

op <- par(mfrow = c(1, 2), mar = c(9, 4.5, 3, 1))
for (nm in names(VARIETIES)) {
  d <- k1[k1$variety == nm, ]
  cols <- ifelse(d$Q2 > 0, "#2C6FBB", "#D1495B")
  bp <- barplot(d$Q2, names.arg = d$predictor, las = 2, col = cols,
                ylab = expression(Q^2), main = nm,
                ylim = range(c(d$Q2, 0)) * 1.25)
  abline(h = 0, lwd = 2)
  text(bp, d$Q2, sprintf("%.3f", d$Q2), pos = ifelse(d$Q2 > 0, 3, 1), cex = 0.8)
}
par(op)
```

## Verdict: ✅ SAFE --- the attack failed, and strengthened the claim

The design labels alone give **negative** Q² in both varieties: knowing treatment and timepoint predicts the proteome *worse* than simply predicting its mean. The RNA block gives positive Q² in both. And when the design is residualised out of both blocks, RNA's predictive power **increases** (Cadenza 0.27 → 0.33; Norin 0.08 → 0.26).

So the coupling is not the shared infection response. It is gene-level covariation between the two layers that survives removal of the common treatment and time effects. Norin's low raw Q² (0.084) appears to be *depressed* by its design main effects --- consistent with the baseline anomaly documented in `assumptions_validation.qmd` §2 --- and recovers once they are removed.

## Safe wording

> Cross-validated on held-out design cells and tested against a permutation null, the transcriptome predicts the proteome significantly better than chance (Q² = 0.27, p = 0.010 in Cadenza; Q² = 0.08, p = 0.020 in Norin). Design labels alone yield negative Q², and the association strengthens after removing treatment and timepoint main effects from both layers, indicating gene-level cross-layer covariation rather than a shared response to infection.

**Do not** extend this to within-condition covariation: with unpaired samples, only between-design-cell covariance is estimable, by any method.

------------------------------------------------------------------------

# K2 --- "Post-transcriptional regulation is present" ✅ SAFE

## The claim

Some proteins depart from the trajectory their own transcript predicts under first-order kinetics --- i.e. regulation acts after transcription.

## Why this needs an adversarial test

The departure is declared by a likelihood-ratio test against a χ² null. But three assumptions behind that null are violated (`assumptions_validation.qmd`): the first-order model leaves structured residuals (§3), genes are not conditionally independent (§8), and RNA measurement error is ignored entirely (§7). Any of those can make a χ² null anti-conservative --- inflating the apparent number of regulated genes.

## The test

Simulate protein trajectories where **no regulation exists by construction**: generate each gene's protein trajectory from its own RNA trajectory with amplitude `a = 1` exactly, adding noise at the observed protein SE. Then run the identical pipeline and count how many genes it still calls regulated. That count is the false-positive rate. A second variant additionally perturbs the RNA trajectory within its own standard error, incorporating the ignored uncertainty.

```{r}
#| label: k2-test

null_calibration <- function(v, f, label, B = 20) {

  k  <- f$keep
  Rl <- v$Rl[k, , drop = FALSE]; Rs <- v$Rs[k, , drop = FALSE]
  Pl <- v$Pl[k, , drop = FALSE]; Ps <- v$Ps[k, , drop = FALSE]

  n_obs <- sum(f$fit$fdr < 0.05)

  fp <- fp_rna <- integer(B)
  for (b in seq_len(B)) {
    # protein generated under a = 1 exactly: no regulation present
    Pn <- f$fit$pred_null + matrix(rnorm(length(Pl), 0, Ps), nrow(Pl))
    Pn <- Pn - Pn[, 1]
    fp[b] <- sum(fit_amplitude(Rl, Pn, Ps, TPS)$fdr < 0.05)

    # same, but also acknowledging RNA measurement error
    Rb <- Rl + matrix(rnorm(length(Rl), 0, Rs), nrow(Rl))
    Rb <- Rb - Rb[, 1]
    fp_rna[b] <- sum(fit_amplitude(Rb, Pn, Ps, TPS)$fdr < 0.05)
  }

  data.frame(
    variety              = label,
    n_gated              = nrow(Rl),
    observed_regulated   = n_obs,
    observed_pct         = round(100 * n_obs / nrow(Rl), 1),
    null_protein_noise   = round(median(fp)),
    null_pct             = round(100 * median(fp) / nrow(Rl), 1),
    null_plus_rna_error  = round(median(fp_rna)),
    null_rna_pct         = round(100 * median(fp_rna) / nrow(Rl), 1),
    excess_over_null     = round(n_obs - median(fp_rna)),
    pct_of_calls_real    = round(100 * (n_obs - median(fp_rna)) / max(n_obs, 1))
  )
}

k2 <- do.call(rbind, Map(function(v, f, nm) null_calibration(v, f, nm),
                         VARIETIES, FITS, names(VARIETIES)))
knitr::kable(k2, caption = "K2: observed regulated genes vs the count obtained when NO regulation exists by construction.")
```

```{r}
#| label: fig-k2
#| fig-cap: "Observed regulated-gene counts against null simulations in which amplitude is exactly 1. The gap is the evidence for genuine post-transcriptional regulation."
#| fig-width: 9
#| fig-height: 5

op <- par(mfrow = c(1, 2), mar = c(8, 4.5, 3, 1))
for (i in seq_len(nrow(k2))) {
  d <- k2[i, ]
  vals <- c(d$observed_regulated, d$null_protein_noise, d$null_plus_rna_error)
  bp <- barplot(vals, names.arg = c("observed", "null\n(protein noise)",
                                    "null\n(+ RNA error)"),
                las = 2, col = c("#2C6FBB", "#BFBFBF", "#D1495B"),
                ylab = "genes called regulated", main = d$variety,
                ylim = c(0, max(vals) * 1.2))
  text(bp, vals, vals, pos = 3, cex = 0.85)
}
par(op)
```

## Verdict: ✅ SAFE --- the attack failed, with a caveat on magnitude

|                                   | Cadenza                | Norin                  |
|:----------------------------------|:-----------------------|:-----------------------|
| Observed "regulated"              | 902 (21.3%)            | 783 (22.5%)            |
| Null: `a = 1`, protein noise only | 169 (4.0%)             | 217 (6.2%)             |
| Null: `a = 1`, **+ RNA error**    | 222 (5.2%)             | 264 (7.6%)             |
| **Excess over null**              | **680 (75% of calls)** | **518 (66% of calls)** |

Three things follow.

**The χ² null is adequately calibrated.** Under data generated with amplitude exactly 1, the pipeline falsely calls 4.0% (Cadenza) and 6.2% (Norin) of genes regulated, against a nominal 5% FDR. Despite A1, B5 and B4 all being violated, the test statistic is not badly broken. Norin's slight excess (6.2%) is consistent with it being the noisier dataset throughout.

**RNA error inflates the null only modestly** --- 4.0% → 5.2% and 6.2% → 7.6%. This is the aggregate false-positive rate, and it is a *different quantity* from the 36--49% per-gene flip rate in K5. Both are true and there is no contradiction: genes whose amplitude sits close to 1 flip easily under perturbation, while the genes generating the excess have large departures that do not. The count is stable; the membership is not.

**The observed count is 3--4× the null.** Roughly **two thirds to three quarters** of the regulated calls are excess over what no-regulation data produces. Post-transcriptional regulation is present at a scale this design can detect.

**Quote the excess, not the raw count.** "902 regulated genes" overstates it by \~25%; "approximately 680 genes in excess of a no-regulation null (75% of 902 calls)" is the defensible form. And note the count still carries the 3× sensitivity to assumed measurement error documented in `assumptions_validation.qmd` §6, so it should always travel with its settings.

## Safe wording

> Simulating protein trajectories under a strict no-regulation null (amplitude fixed at 1, observed noise, RNA error propagated) yields 222 (Cadenza) and 264 (Norin) false positives, against 902 and 783 observed. Approximately 680 and 518 genes respectively exceed the null, indicating that post-transcriptional regulation affects a substantial minority of the RNA-responsive proteome.

This validates the **existence and approximate extent** of post-transcriptional regulation. It does **not** validate the identity of any particular gene --- see K5.

------------------------------------------------------------------------

# K3 --- "We can estimate protein half-lives" ⚠️ CONDITIONAL

## The claim

The fitted `t_half` values describe protein turnover in wheat.

## Why this needs a test

A half-life much longer than the observation window is unmeasurable from that window: trajectories for 200 h and 500 h half-lives are indistinguishable across 72 h. Plant proteins make this acute --- 15N labelling in *Arabidopsis* finds half-lives spanning **hours to months** [@li2017], so a 72 h course sampled every 24 h can only see a slice of the distribution.

## The test

Show the resolvable band directly, then report where the fitted values actually fall.

```{r}
#| label: fig-k3
#| fig-cap: "Left: predicted protein response to a step change in mRNA, by half-life; curves overlapping at the sampled timepoints (dashed) are indistinguishable. Right: where the fitted half-lives fall."
#| fig-width: 11
#| fig-height: 5

grid <- seq(0, max(TPS), by = 0.25)
r_step <- matrix(2^rep(1, length(grid)), nrow = 1)
th_show <- c(6, 12, 24, 48, 72, 144, 432)
cols <- hcl.colors(length(th_show), "Zissou1")

op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))

plot(NA, xlim = c(0, max(TPS)), ylim = c(0, 1.05),
     xlab = "hours post inoculation", ylab = "predicted protein log2FC",
     main = "Which half-lives are distinguishable?")
for (i in seq_along(th_show)) {
  kd <- log(2) / th_show[i]
  p <- integrate_protein_mat(grid, r_step, ks = kd, kd = kd, P0 = 1)
  lines(grid, log2(p[1, ]), col = cols[i], lwd = 2.5)
}
abline(v = TPS[-1], lty = 3, col = "grey50")
legend("topleft", bty = "n", cex = 0.7, lwd = 2.5, col = cols,
       legend = sprintf("%g h", th_show), title = "half-life")

k3 <- do.call(rbind, Map(function(f, nm) {
  th <- f$fit$t_half
  data.frame(variety = nm, n = length(th),
             floor_pct  = round(100 * mean(th <= 12.001), 1),
             resolv_pct = round(100 * mean(th > 12.001 & th <= max(TPS)), 1),
             above_pct  = round(100 * mean(th > max(TPS)), 1))
}, FITS, names(FITS)))

bp <- barplot(t(as.matrix(k3[, c("floor_pct", "resolv_pct", "above_pct")])),
              beside = FALSE, names.arg = k3$variety,
              col = c("#D1495B", "#4C9F70", "#BFBFBF"),
              ylab = "% of fitted genes", main = "Where fitted half-lives fall",
              ylim = c(0, 105))
legend("top", bty = "n", cex = 0.7, horiz = TRUE,
       fill = c("#D1495B", "#4C9F70", "#BFBFBF"),
       legend = c("at 12 h floor", "resolvable", "beyond 72 h"))
par(op)

knitr::kable(k3, caption = "K3: only the middle band is estimable from this design.")
```

## Verdict: ⚠️ CONDITIONAL --- valid only for the resolvable minority

Roughly **half** of fitted genes sit above the 72 h identifiability bound and about a quarter are pinned at the 12 h grid floor. Only **23--27%** have a half-life this design can estimate.

A value at the floor means "≤ 12 h", not 12 h. A value above the bound means "slower than we can observe", not a number. The `kinetics_limited` archetype is not a biological category --- it is the region where attenuation and slow turnover are mathematically confounded.

## Safe wording

> For the \~25% of RNA-responsive genes whose implied half-life falls within the resolvable window (12--72 h), turnover estimates are informative. Values at the grid boundaries are censored and are reported as bounds. The fitted half-life distribution reflects the sampling window rather than the underlying proteome and is not interpreted as an estimate of wheat protein turnover.

**Do not** publish the `t_half` histogram as a description of wheat protein half-lives.

------------------------------------------------------------------------

# K4 --- "We can say whether translation or degradation changed" ❌ NOT SUPPORTED

## The claim

A gene with `a < 1` is translationally repressed (or conversely, more rapidly degraded).

## Why this fails --- algebra, not data quality

Working in fold-change relative to the time-matched control, the absolute synthesis rate **cancels exactly**. From the derivation in `de_proteomics_wheat.qmd`:

$$
\frac{dp}{dt} = \frac{k_s R_c}{P_c}\,(r - p)
$$

Absolute `k_s` is absorbed into the per-protein MS response factor and never appears alone. No amount of extra replication or timepoints within this design recovers it: fold-change data cannot separate a change in synthesis from a change in degradation, because both act on the same single amplitude parameter.

This is precisely why @schwanhausser2011 used **pulsed SILAC** --- direct measurement of synthesis and degradation --- rather than inferring rates from mRNA trajectories. Pulse-chase and heavy-water labelling exist for the same reason.

```{r}
#| label: fig-k4
#| fig-cap: "Two mechanistically different genes producing identical fold-change trajectories. Nothing in this dataset can distinguish them."
#| fig-width: 7
#| fig-height: 5

# a gene whose synthesis rate halves vs one whose degradation rate doubles:
# in fold-change space the predicted trajectories are the same curve.
grid <- seq(0, max(TPS), by = 0.25)
r <- matrix(2^(1 - exp(-grid / 12)), nrow = 1)   # some rising mRNA response

kd1 <- log(2) / 24
p1 <- log2(integrate_protein_mat(grid, r, ks = kd1,     kd = kd1, P0 = 1)[1, ])
p2 <- log2(integrate_protein_mat(grid, r, ks = kd1 * 3, kd = kd1 * 3, P0 = 1)[1, ])

plot(grid, p1, type = "l", lwd = 6, col = "#2C6FBB",
     xlab = "hours post inoculation", ylab = "protein log2FC",
     main = "Different (ks, kd), identical observable")
lines(grid, p2, lwd = 2, lty = 2, col = "#D1495B")
legend("bottomright", bty = "n", lwd = c(6, 2), lty = c(1, 2),
       col = c("#2C6FBB", "#D1495B"),
       legend = c("ks = kd = log2/24",
                  "ks = kd = 3x higher (same ratio)"))
```

## Verdict: ❌ NOT SUPPORTED --- and not fixable within this design

## Safe wording

> Departures from the RNA-predicted trajectory are reported as changes in response amplitude. Because absolute synthesis rate is not identifiable from fold-change data, these are not attributed to translation or degradation individually; separating them would require pulse labelling.

Write **"departs from its RNA-predicted trajectory"**. Never write "translationally repressed", "stabilised" or "degraded faster".

------------------------------------------------------------------------

# K5 --- "Gene X is post-transcriptionally regulated" ❌ NOT SUPPORTED

## The claim

A named gene from the regulated list is a validated finding.

## The test

`assumptions_validation.qmd` §7 propagates the RNA standard errors that the model ignores, by parametric bootstrap, and asks how often the regulated / not-regulated call changes.

```{r}
#| label: k5-test

k5 <- do.call(rbind, Map(function(v, f, nm) {
  k <- f$keep
  set.seed(11)
  idx <- sample(which(k), min(1200, sum(k)))

  Rl <- v$Rl[idx, , drop = FALSE]; Rs <- v$Rs[idx, , drop = FALSE]
  Pl <- v$Pl[idx, , drop = FALSE]; Ps <- v$Ps[idx, , drop = FALSE]

  base <- fit_amplitude(Rl, Pl, Ps, TPS)$a
  B <- 30
  boot <- vapply(seq_len(B), function(b) {
    Rb <- Rl + matrix(rnorm(length(Rl), 0, Rs), nrow(Rl))
    Rb <- Rb - Rb[, 1]
    fit_amplitude(Rb, Pl, Ps, TPS)$a
  }, numeric(nrow(Rl)))

  flips <- mean(apply(boot, 1, function(x)
    any(x > 1, na.rm = TRUE) && any(x < 1, na.rm = TRUE)))

  data.frame(variety = nm, n_genes = nrow(Rl),
             median_a = round(median(base, na.rm = TRUE), 3),
             median_bootstrap_sd = round(median(apply(boot, 1, sd, na.rm = TRUE)), 3),
             pct_call_flips = round(100 * flips, 1))
}, VARIETIES, FITS, names(VARIETIES)))

knitr::kable(k5, caption = "K5: how often the regulated/not call changes when RNA measurement error is acknowledged.")
```

## Verdict: ❌ NOT SUPPORTED at the individual-gene level

A large fraction of calls flip across the `a = 1` boundary under nothing more than the RNA error bars limma already reports. Those genes are not distinguishable from unregulated ones by this analysis.

## Safe wording

> Genes with the largest amplitude departures are reported as candidates for post-transcriptional regulation, requiring orthogonal validation. Because RNA measurement error is not propagated into the kinetic model, individual classifications are unstable and no gene-level claim is made without independent confirmation.

------------------------------------------------------------------------

# K6 --- "The first-order kinetic model fits the data" ❌ NOT SUPPORTED

## The test

Under a correct functional form, standardised residuals sit near zero at every timepoint with about half positive. Systematic structure means the model shape is wrong.

```{r}
#| label: fig-k6
#| fig-cap: "Standardised residuals by timepoint. Systematic departure from zero indicates the functional form is misspecified, not merely noisy."
#| fig-width: 10
#| fig-height: 5

op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))
k6 <- do.call(rbind, Map(function(v, f, nm) {
  z <- f$fit$resid / pmax(v$Ps[f$keep, , drop = FALSE], 0.05)
  boxplot(as.data.frame(z), outline = FALSE, las = 1, col = "#7FB2E5",
          names = paste0("t", TPS), xlab = "hours post inoculation",
          ylab = "standardised residual", main = nm)
  abline(h = 0, lwd = 2, col = "#D1495B")
  data.frame(variety = nm, timepoint = paste0("t", TPS),
             mean_z = round(colMeans(z), 3),
             pct_positive = round(100 * colMeans(z > 0), 1))
}, VARIETIES, FITS, names(VARIETIES)))
par(op)

knitr::kable(k6, caption = "K6: a correct model gives mean ~0 and ~50% positive at every timepoint.")
```

## Verdict: ❌ NOT SUPPORTED --- the model is misspecified

Residuals are systematically structured, and in **opposite directions** in the two varieties: Cadenza's protein response rises faster and falls back sooner than first-order kinetics predicts, while Norin's is delayed. Both patterns are the signature of an omitted delay or saturation term.

The consequence for K2 and K5: the amplitude `a` absorbs whatever shape error it can, so some genes are flagged because their trajectory *shape* is mis-modelled rather than because their amplitude genuinely departs from 1.

## Safe wording

> A first-order synthesis/degradation model was used as the reference trajectory. Residual analysis indicates systematic departure from this functional form, consistent with an unmodelled translational delay; amplitude estimates should therefore be interpreted as summarising deviation from a first-order reference rather than as mechanistic rate parameters.

**Do not** describe the model as fitting well, or report R² / goodness-of-fit as support.

------------------------------------------------------------------------

# Summary --- what is safe to claim

```{r}
#| label: summary
#| echo: false

knitr::kable(data.frame(
  Claim = c("K1 RNA and protein responses are coupled",
            "K2 Post-transcriptional regulation is present",
            "K3 Protein half-lives can be estimated",
            "K4 Translation vs degradation can be separated",
            "K5 Gene X is post-transcriptionally regulated",
            "K6 The first-order model fits"),
  Grade = c("SAFE", "SAFE (quote the excess)", "CONDITIONAL", "NOT SUPPORTED",
            "NOT SUPPORTED", "NOT SUPPORTED"),
  Basis = c("Q2 vs permutation null; design-only Q2 negative; survives residualisation",
            "Observed 3-4x a strict no-regulation null (excess 680 / 518 genes)",
            "Only ~25% of genes in the resolvable 12-72 h band",
            "ks cancels algebraically in fold-change space",
            "Calls flip under RNA error the model ignores",
            "Residuals structured, opposite direction per variety")
), caption = "Grades for each kinetic claim.")
```

## The one-paragraph version

> This design supports **population-level statements about the extent of post-transcriptional regulation**, and a validated claim that transcript and protein responses are coupled beyond the shared infection response. It does **not** support per-gene regulation calls, mechanistic attribution to translation or degradation, or a description of wheat protein turnover. The limiting factors are three biological replicates, four timepoints across 72 h, unpaired omics layers, and a kinetic model whose functional form is measurably misspecified.

## References

Full entries in `references.bib`. The load-bearing ones for this notebook:

-   @schwanhausser2011 --- the ODE formulation, and the pulsed-SILAC approach that makes K4 answerable when it is answerable at all.
-   @westerhuis2008 --- cross-validated Q² with a permutation null as the correct validation for PLS in small-n omics, which is the basis of K1.
-   @li2017 --- plant protein half-lives spanning hours to months, which sets the K3 identifiability context.
-   @smyth2004 --- the moderated variances that make n = 3 analysable, and whose standard errors K5 propagates.
