---
title: "RNA-protein condition-profile distance --- can genes be ranked by discordance?"
author: "Kristina Gagalova"
format:
  gfm:
    toc: true
    code-tools: true
    fig-dir: "figs"
execute:
  echo: true
editor: visual
bibliography: ../kinetics_limited/references.bib
---

## Purpose and scope

This notebook asks a question distinct from anything in `analysis/kinetics_limited` or `analysis/dimensionality-reduction`:

> **For an individual gene, does its protein's condition-response profile look like its RNA's condition-response profile --- and can that resemblance be turned into a per-gene distance that ranks genes by RNA/protein discordance?**

`kinetics_limited` tests a *mechanistic* null (first-order kinetics) per gene. `dimensionality-reduction` asks whether RNA predicts protein *in aggregate*, across all genes at once. This notebook asks a *geometric* question per gene: place each gene's RNA profile and its own protein profile as two points in a shared coordinate system built from the population's dominant response shapes, and measure how far apart they land.

### Relationship to the existing Python work

Two notebooks in `KristinaGagalova/autoencoders-test` motivate this:

**`rna_protein_unpaired_per_variety.ipynb`** is a thin driver over `rnaprot/unpaired_reviewer.py` (a revised module -- not the `unpaired.py` referenced in `dimensionality_reduction_wheat.qmd`; the two have different loss weights, 3.0/0.75 here vs 2.0/0.5 there). It re-runs essentially the same leave-condition-out comparison (mean, design, cognate ridge, PCA+ridge, PLS, design-residualised variants, condition-aligned AE) already built in this repository's `dimensionality_reduction_wheat.qmd` §7 (D4) and §7b (D7). Two things in it are genuinely different from what is already here, and are picked up as an addendum (**D8**) in that notebook rather than duplicated wholesale in a second \~40-chunk document:

1.  Feature selection is done **inside each outer CV fold**, not once globally before the loop -- D4/D7 currently select the top-variable genes once, which is a selection-leakage risk.
2.  A permutation null (`N_PERM=200`) is applied to the PCA/PLS/design baselines (not to the autoencoder).

See `dimensionality_reduction_wheat.qmd` §7c for that check. This document does not repeat D4/D7.

**`rna_protein_umap_shared_space.ipynb`** is the actual source of the "distance between genes" idea, and *is* new work relative to everything else in this repository. It builds condition-mean profiles for RNA and for cognate proteins, fits a shared coordinate system (PCA or UMAP) on the RNA profiles only, projects the protein profiles into that same system, and measures the Euclidean distance between a gene's RNA point and its own protein point. Two things about it matter for how it is ported here:

1.  **Its own synthetic validation found that 2D UMAP hurts the ranking**: distance-to-truth-divergence correlation was Spearman $\rho = 0.923$ in the full aligned input space, but only $\rho = 0.713$ after compressing to a 2D UMAP embedding. That is carried over below as a **hypothesis to test on this project's own ground truth**, not assumed.
2.  **It was never actually run to completion on real data.** The real-data cells stop at the visualization step; no per-gene distance table or ranking exists for Cadenza or Norin. Building that is the new work in §6-§8 below.

------------------------------------------------------------------------

## Analysis outline

| Step | What                                           | Why                                                       |
|:-------------|:-------------------------|:-------------------------------|
| §1   | Setup and data (real + simulated)              | Validate on known ground truth before touching real data  |
| §2   | The method and its assumptions                 | Each with references                                      |
| §3   | Implementation                                 | Shared-space construction, from scratch                   |
| §4   | **G1** Does distance separate true archetypes? | Validation against `truth.tsv`, incl. the `lag_only` trap |
| §5   | **G2** Full space vs 2D embedding              | Replicates the source notebook's own finding, on our data |
| §6   | **G3** Permutation null on real data           | No ground truth on real data -- this substitutes for it   |
| §7   | **G4** Replicate-level stability               | Bootstrap over which replicates feed the condition means  |
| §8   | Apply to real data                             | Per-gene distance tables, Cadenza and Norin               |
| §9   | Assumption validation summary                  |                                                           |
| §10  | What is safe to claim                          |                                                           |

------------------------------------------------------------------------

## 1. Setup

```{r}
#| label: setup
#| message: false

if (!requireNamespace("here", quietly = TRUE)) stop("package 'here' required")

## RENV_PROJECT must be set before sourcing activate.R, or renv treats this
## subdirectory as the project and bootstraps an empty library.
if (file.exists(here::here("renv", "activate.R"))) {
  Sys.setenv(RENV_PROJECT = here::here())
  source(here::here("renv", "activate.R"))
}

source(here::here("R", "utils.R"))
source(here::here("R", "wheat_pipeline.R"))
need_pkgs(c("limma", "edgeR", "DESeq2", "matrixStats", "impute", "uwot", "yaml", "MASS", "ggplot2"))

library(ggplot2)
library(MASS)

set.seed(20260823)
```

```{r}
#| label: load-real

DESIGN <- wheat_design()
cad <- prepare_variety("cadenza", DESIGN)
nor <- prepare_variety("norin",   DESIGN)
REAL <- list(Cadenza = cad, Norin = nor)
```

```{r}
#| label: load-simulated
#| message: false

#' Loader for the project's own Case B ground-truth simulator
#' (`R/00_simulate_data.R`), parallel to `load_variety()` in
#' `wheat_pipeline.R` but pointed at `data/simulated` and its different file
#' layout. This gives archetype-labelled truth to validate the distance
#' metric against, before it is trusted on real data -- the same validate-
#' on-simulation-first discipline used for the rest of this pipeline.
load_simulated <- function(cfg, data_dir = here::here("data", "simulated")) {

  rna_counts <- as.matrix(read.delim(file.path(data_dir, "rna_counts.tsv"),
                                     row.names = 1, check.names = FALSE))
  prot_int   <- as.matrix(read.delim(file.path(data_dir, "protein_intensities.tsv"),
                                     row.names = 1, check.names = FALSE))
  rna_meta   <- read.delim(file.path(data_dir, "rna_coldata.tsv"),
                           row.names = "sample_id", stringsAsFactors = FALSE)
  prot_meta  <- read.delim(file.path(data_dir, "protein_coldata.tsv"),
                           row.names = "sample_id", stringsAsFactors = FALSE)
  truth <- read.delim(file.path(data_dir, "truth.tsv"), stringsAsFactors = FALSE)
  gmap  <- read.delim(file.path(data_dir, "truth_gene_group_map.tsv"),
                      stringsAsFactors = FALSE)

  lv <- cell_levels(cfg)
  rna_meta$cell  <- factor(rna_meta$cell,  levels = lv)
  prot_meta$cell <- factor(prot_meta$cell, levels = lv)

  design <- list(treatments = cfg$design$treatments,
                timepoints = cfg$design$timepoints,
                n_reps     = cfg$design$n_reps)

  list(rna  = rna_counts[, rownames(rna_meta), drop = FALSE],
       prot = prot_int[,  rownames(prot_meta), drop = FALSE],
       rna_meta = rna_meta, prot_meta = prot_meta,
       design = design, truth = truth, gmap = gmap)
}

sim_cfg <- load_config(here::here("config", "config.yaml"))
sim_raw <- load_simulated(sim_cfg)

sim_qc <- local({
  qr <- run_qc_rna(sim_raw$rna, sim_raw$rna_meta, sim_raw$design)
  qp <- run_qc_prot(sim_raw$prot, sim_raw$prot_meta)
  mi <- run_missingness(qp$norm, sim_raw$prot_meta)
  list(vst = qr$vst, prot = mi$mixed)
})

cat("Simulated genes after RNA QC:   ", nrow(sim_qc$vst),  "\n")
cat("Simulated groups after prot QC: ", nrow(sim_qc$prot), "\n")
```

------------------------------------------------------------------------

## 2. The method and its assumptions

For each gene present in both layers:

1.  Take its **condition-mean profile** in each layer separately (8 treatment x timepoint cells) -- the only aggregation licensed under Case B, per `dimensionality_reduction_wheat.qmd` §7b (D7).
2.  **Row-centre** each profile (subtract the gene's own across-condition mean). This compares response *shape*, not absolute level -- required regardless of Case B, since RNA (VST) and protein (log2 intensity) are on incommensurable physical scales to begin with.
3.  **Standardise per condition** (z-score each of the 8 condition columns across genes), fit on the RNA block only, and apply the same transform to protein. This matches the source notebook's `StandardScaler().fit(Xr)` step.
4.  Fit **PCA on the RNA profiles only** (their dominant shapes of response across the gene population), and project the protein profiles into that same space with no refitting.
5.  **Distance** = Euclidean distance between a gene's transformed RNA point and its own transformed protein point -- computed in the full standardised space, in the top-2 PCA components, and in a 2D UMAP embedding fit on RNA and transformed for protein.

### Assumptions, stated up front

| \#  | Assumption                                                                               | Status                                                                      |
|:------------|:-------------------------------|:--------------------------|
| A1  | The RNA\<-\>protein correspondence used is genuinely 1:1 for the genes scored            | Enforced by construction (see §4); ambiguous groups excluded                |
| A2  | Condition means are the right granularity for a cross-block comparison                   | Carried over from D7, not re-derived here                                   |
| A3  | Comparing row-centred *shape*, not absolute level, is the right notion of "look alike"   | Design choice, stated not tested                                            |
| A4  | Distance in the full standardised space is more informative than in a 2D embedding       | **Tested in G2**, not assumed                                               |
| A5  | A short Euclidean distance means concordant regulation, not just similar kinetics/timing | **Tested in G1** -- this is where `lag_only` genes are the adversarial case |

@mcinnes2018 and @becht2019 are the basis for A4: UMAP is explicitly designed to preserve local neighbourhood structure, not global pairwise distances, so a 2D UMAP distance between two specific, possibly-distant points is not guaranteed to track their true distance in the original space -- exactly the degradation the source notebook's own synthetic check already found ($\rho$ 0.923 to 0.713).

------------------------------------------------------------------------

## 3. Implementation

```{r}
#| label: fn-shared-space

#' Per-gene shape: subtract the gene's own across-condition mean.
row_center <- function(M) M - rowMeans(M)

#' Fit a per-column (per-condition) z-scorer on one matrix, return a function
#' that applies it to any matrix with the same columns. Fit on RNA only, per
#' the source notebook.
fit_col_scaler <- function(M) {
  mu  <- colMeans(M)
  sdv <- apply(M, 2, sd); sdv[sdv < 1e-8] <- 1
  function(X) sweep(sweep(X, 2, mu, "-"), 2, sdv, "/")
}

#' Build the shared space: row-centre, z-score on RNA's column moments, PCA
#' fit on RNA only. Returns the standardised profiles (full-space distance
#' needs nothing else, since the PCA rotation below is orthonormal and
#' therefore distance-preserving) and their PCA scores (for the 2D check).
build_shared_space <- function(R_cond, P_cond) {
  stopifnot(identical(rownames(R_cond), rownames(P_cond)))
  scaler <- fit_col_scaler(row_center(R_cond))
  Rs <- scaler(row_center(R_cond))
  Ps <- scaler(row_center(P_cond))

  sv <- svd(Rs)
  list(Rs = Rs, Ps = Ps,
       Zr = Rs %*% sv$v, Zp = Ps %*% sv$v,
       var_explained = sv$d^2 / sum(sv$d^2))
}

euclid <- function(A, B, k = NULL) {
  if (!is.null(k)) { A <- A[, seq_len(k), drop = FALSE]; B <- B[, seq_len(k), drop = FALSE] }
  sqrt(rowSums((A - B)^2))
}

#' Calculate correlation and amplitude for each gene
#' Returns: correlation (directional concordance), amplitude ratio (relative response magnitude)
gene_metrics <- function(R_centered, P_centered) {
  # Correlation: directional concordance (correlation of response profiles)
  cors <- sapply(seq_len(nrow(R_centered)), function(i) {
    cor(R_centered[i, ], P_centered[i, ], method = "pearson")
  })
  cors[is.na(cors)] <- 0  # Handle constant profiles

  # Amplitude: log2 ratio of response magnitudes
  # ||R|| = sqrt(sum of squared values across conditions)
  R_norm <- sqrt(rowSums(R_centered^2))
  P_norm <- sqrt(rowSums(P_centered^2))
  # Avoid log(0); replace with 0
  amps <- ifelse(R_norm > 1e-10 & P_norm > 1e-10,
                 log2(P_norm / R_norm),
                 0)

  data.frame(correlation = cors, amplitude = amps)
}

#' Calculate infection-response contrasts (Infected - Control) for each timepoint
#' Returns condition means for the 4 timepoint-level contrasts
infection_contrasts <- function(counts, meta) {
  # Assumes meta has columns: condition, treatment, timepoint
  # where treatment = "infected" or "control"

  timepoints <- sort(unique(meta$timepoint))
  contrasts <- data.frame()

  for (tp in timepoints) {
    idx_inf <- meta$treatment == "infected" & meta$timepoint == tp
    idx_ctl <- meta$treatment == "control"  & meta$timepoint == tp

    if (sum(idx_inf) > 0 && sum(idx_ctl) > 0) {
      mean_inf <- rowMeans(counts[, idx_inf, drop = FALSE])
      mean_ctl <- rowMeans(counts[, idx_ctl, drop = FALSE])
      contrast <- mean_inf - mean_ctl

      if (nrow(contrasts) == 0) {
        contrasts <- data.frame(contrast)
        colnames(contrasts)[1] <- paste0("dpi_", tp)
      } else {
        contrasts[[paste0("dpi_", tp)]] <- contrast
      }
    }
  }
  contrasts
}

#' UMAP fit on RNA, protein transformed into the same embedding -- matching
#' the source notebook's asymmetric fit/transform (never a joint/symmetric
#' fit, which would let protein leak into the axes it is being scored against).
umap_distance <- function(Rs, Ps, seed = 1) {
  nn <- max(2, min(15, nrow(Rs) %/% 4))
  set.seed(seed)
  fit <- uwot::umap(Rs, n_neighbors = nn, min_dist = 0.10, n_components = 2,
                    metric = "euclidean", ret_model = TRUE)
  Ep <- uwot::umap_transform(Ps, fit)
  sqrt(rowSums((fit$embedding - Ep)^2))
}
```

------------------------------------------------------------------------

## 4. G1 --- Does distance separate the true archetypes?

The simulator's `truth.tsv` labels every gene with the *mechanism* that generated it: `unchanged`, `concordant`, `buffered`, `mrna_only`, `protein_only`, `anticorrelated`, `lag_only`. None of that label is used to build the distance metric -- it is used only to grade it afterwards.

`lag_only` is the adversarial case, carried over from the kinetics notebooks: it is **mechanistically fully concordant** (protein is driven purely by RNA, first-order kinetics, `A2` holds exactly) but has a **slow protein half-life**, so its condition-profile *shape* lags the RNA shape. A method that cannot tell "concordant but slow" from "genuinely decoupled" will misclassify it -- `analysis/kinetics_limited/assumptions_validation.qmd` documents the same trap for naive quadrant classification. This is the sharpest test of A5.

```{r}
#| label: g1-simulated-distance

## Restrict to genes with an unambiguous 1:1 RNA<->protein-group
## correspondence. Isoform-split and shared-peptide groups (the simulator's
## deliberately messy 1:many / many:1 cases) are excluded from this analysis
## by design -- a distance metric between a gene and an ambiguous protein
## mixture is not answering the question this notebook asks.
gmap <- sim_raw$gmap
one_to_one <- with(gmap,
  ave(seq_along(gene_id), gene_id, FUN = length) == 1 &
  ave(seq_along(group),   group,   FUN = length) == 1 &
  frac == 1)
map1 <- gmap[one_to_one, c("gene_id", "group")]
map1 <- map1[map1$gene_id %in% rownames(sim_qc$vst) &
             map1$group  %in% rownames(sim_qc$prot), ]

cat("1:1 genes available for validation:", nrow(map1), "/", nrow(gmap), "\n")

R_cond_sim <- cell_means(sim_qc$vst[map1$gene_id, , drop = FALSE],  sim_raw$rna_meta)
P_cond_sim <- cell_means(sim_qc$prot[map1$group,  , drop = FALSE], sim_raw$prot_meta)
rownames(P_cond_sim) <- map1$gene_id

sp_sim     <- build_shared_space(R_cond_sim, P_cond_sim)
d_full_sim <- euclid(sp_sim$Rs, sp_sim$Ps)
d_pca2_sim <- euclid(sp_sim$Zr, sp_sim$Zp, k = 2)
d_umap_sim <- umap_distance(sp_sim$Rs, sp_sim$Ps)

arch <- setNames(sim_raw$truth$archetype, sim_raw$truth$gene_id)[map1$gene_id]

g1 <- data.frame(gene_id = map1$gene_id, archetype = arch,
                 d_full = d_full_sim, d_pca2 = d_pca2_sim, d_umap = d_umap_sim)
```

```{r}
#| label: fig-g1
#| fig-cap: "G1: full-space distance by true archetype (simulated data). concordant/unchanged are the null; lag_only is the adversarial case -- mechanistically concordant but kinetically slow."
#| fig-width: 9
#| fig-height: 5

ord <- c("unchanged", "concordant", "lag_only", "buffered",
        "mrna_only", "protein_only", "anticorrelated")
boxplot(d_full ~ factor(archetype, levels = ord), data = g1,
       las = 2, xlab = "", ylab = "distance (full standardised space)",
       col = ifelse(ord %in% c("unchanged", "concordant"), "#2C6FBB",
             ifelse(ord == "lag_only", "#E8A33D", "#D1495B")))
```

```{r}
#| label: g1-separation

#' Rank-based separation of a "positive" (expected-far) archetype set from a
#' "negative" (expected-close) baseline. AUC > 0.5 means the positive set
#' really does score farther, in the direction the mechanism predicts.
sep_stat <- function(d, pos, neg) {
  w <- suppressWarnings(wilcox.test(d[pos], d[neg]))
  data.frame(n_pos = sum(pos), n_neg = sum(neg),
            auc = round(unname(w$statistic) / (sum(pos) * sum(neg)), 3),
            p   = signif(w$p.value, 3))
}

baseline <- g1$archetype %in% c("unchanged", "concordant")
truly_discordant <- g1$archetype %in% c("buffered", "mrna_only",
                                        "protein_only", "anticorrelated")
lag <- g1$archetype == "lag_only"

g1_sep <- rbind(
  data.frame(comparison = "genuinely discordant vs concordant/unchanged",
            sep_stat(g1$d_full, truly_discordant, baseline)),
  data.frame(comparison = "lag_only vs concordant/unchanged",
            sep_stat(g1$d_full, lag, baseline))
)
knitr::kable(g1_sep, caption = "G1: separation of archetypes by full-space distance. AUC near 0.5 means the metric cannot tell the groups apart.")
```

------------------------------------------------------------------------

## 5. G2 --- Full space vs 2D embedding

```{r}
#| label: g2-embedding-comparison

g2 <- rbind(
  data.frame(space = "full standardised space",
            rbind(sep_stat(g1$d_full, truly_discordant, baseline),
                  sep_stat(g1$d_full, lag, baseline))[, "auc", drop = FALSE]),
  data.frame(space = "top-2 PCA components",
            rbind(sep_stat(g1$d_pca2, truly_discordant, baseline),
                  sep_stat(g1$d_pca2, lag, baseline))[, "auc", drop = FALSE]),
  data.frame(space = "2D UMAP",
            rbind(sep_stat(g1$d_umap, truly_discordant, baseline),
                  sep_stat(g1$d_umap, lag, baseline))[, "auc", drop = FALSE])
)
g2$comparison <- rep(c("discordant vs baseline", "lag_only vs baseline"), 3)
g2 <- g2[, c("space", "comparison", "auc")]
knitr::kable(g2, caption = "G2: does dimensionality reduction for visualisation cost separation power? Compare each space's AUC to the full-space row.")
```

------------------------------------------------------------------------

## 6. G3 --- Permutation null on real data

There is no `truth.tsv` for Cadenza or Norin. What substitutes for it: if a gene's protein profile genuinely resembles its own RNA profile more than it resembles a random other gene's, the **observed mean cognate distance should sit below the distribution obtained by randomly re-pairing genes**. This is a matching (Mantel-type) permutation test, and needs no ground truth.

```{r}
#| label: fn-real-shared-space

#' Build the shared space and full-space gene distances for one real variety.
#' Enforces assumption A1 (1:1 gene<->protein correspondence) by checking that
#' the protein matrix has already been collapsed to 1:1 identifiers.
gene_distance_real <- function(v, n_perm = 199, seed = 1) {

  common <- intersect(rownames(v$qc_rna$vst), rownames(v$imputed$mixed))

  # A1: Verify 1:1 correspondence
  # In the simulation, this is enforced by explicit filtering in §4.
  # For real data, we verify that row names are already clean 1:1 identifiers.
  # If the protein matrix contains ambiguous groups or isoforms, they should have
  # been collapsed/filtered upstream by prepare_variety() or equivalent.
  stopifnot(
    "Protein matrix must have unique row identifiers (assumption A1)" =
      length(unique(rownames(v$imputed$mixed))) == nrow(v$imputed$mixed)
  )

  R_cond <- cell_means(v$qc_rna$vst[common, , drop = FALSE],    v$meta)
  P_cond <- cell_means(v$imputed$mixed[common, , drop = FALSE], v$meta)

  sp     <- build_shared_space(R_cond, P_cond)
  d_full <- euclid(sp$Rs, sp$Ps)
  d_pca2 <- euclid(sp$Zr, sp$Zp, k = 2)

  # Calculate correlation and amplitude for each gene
  metrics <- gene_metrics(sp$Rs, sp$Ps)

  # UMAP fit on RNA, transform protein (asymmetric, never joint)
  nn <- max(2, min(15, nrow(sp$Rs) %/% 4))
  set.seed(seed)
  umap_fit <- uwot::umap(sp$Rs, n_neighbors = nn, min_dist = 0.10, n_components = 2,
                         metric = "euclidean", ret_model = TRUE)
  Er <- umap_fit$embedding
  Ep <- uwot::umap_transform(sp$Ps, umap_fit)
  d_umap <- sqrt(rowSums((Er - Ep)^2))

  set.seed(seed)
  null <- replicate(n_perm, mean(euclid(sp$Rs, sp$Ps[sample(nrow(sp$Ps)), , drop = FALSE])))
  obs  <- mean(d_full)
  p_perm <- (sum(null <= obs) + 1) / (n_perm + 1)

  list(genes = data.frame(gene_id = common,
                          d_full = d_full,
                          d_pca2 = d_pca2,
                          d_umap = d_umap,
                          correlation = metrics$correlation,
                          amplitude = metrics$amplitude),
       space = list(Er = Er, Ep = Ep),
       obs = obs, null = null, p_perm = p_perm, n_genes = length(common))
}

GD <- setNames(Map(function(v, s) gene_distance_real(v, seed = s), REAL, seq_along(REAL)), names(REAL))

#' Infection-response contrast analysis: compare Infected - Control at each timepoint
#' Instead of 8 condition means, use 4-point infection-response trajectories
gene_distance_infection_contrast <- function(v, n_perm = 199, seed = 1) {
  # Extract infection and control means for each timepoint
  # v$meta should have: condition, treatment (infected/control), timepoint

  common <- intersect(rownames(v$qc_rna$vst), rownames(v$imputed$mixed))

  # Verify 1:1 mapping
  stopifnot(
    "Protein matrix must have unique row identifiers (assumption A1)" =
      length(unique(rownames(v$imputed$mixed))) == nrow(v$imputed$mixed)
  )

  # Calculate contrasts for RNA and protein
  R_contrast <- infection_contrasts(v$qc_rna$vst[common, ], v$meta)
  P_contrast <- infection_contrasts(v$imputed$mixed[common, ], v$meta)

  # If contrasts are empty or don't match, return NULL (metadata may not have treatment column)
  if (nrow(R_contrast) == 0 || nrow(P_contrast) == 0 || ncol(R_contrast) != ncol(P_contrast)) {
    return(NULL)
  }

  # Build shared space on contrasts
  sp <- build_shared_space(R_contrast, P_contrast)
  d_full <- euclid(sp$Rs, sp$Ps)
  metrics <- gene_metrics(sp$Rs, sp$Ps)

  # UMAP
  nn <- max(2, min(15, nrow(sp$Rs) %/% 4))
  set.seed(seed)
  umap_fit <- uwot::umap(sp$Rs, n_neighbors = nn, min_dist = 0.10, n_components = 2,
                         metric = "euclidean", ret_model = TRUE)
  Er <- umap_fit$embedding
  Ep <- uwot::umap_transform(sp$Ps, umap_fit)
  d_umap <- sqrt(rowSums((Er - Ep)^2))

  set.seed(seed)
  null <- replicate(n_perm, mean(euclid(sp$Rs, sp$Ps[sample(nrow(sp$Ps)), , drop = FALSE])))
  obs <- mean(d_full)
  p_perm <- (sum(null <= obs) + 1) / (n_perm + 1)

  list(genes = data.frame(gene_id = common,
                          d_full = d_full,
                          d_umap = d_umap,
                          correlation = metrics$correlation,
                          amplitude = metrics$amplitude),
       space = list(Er = Er, Ep = Ep),
       obs = obs, null = null, p_perm = p_perm, n_genes = length(common))
}

# Try to compute infection-contrast version (may be NULL if metadata lacks treatment info)
GD_contrast <- tryCatch(
  setNames(Map(function(v, s) gene_distance_infection_contrast(v, seed = s), REAL, seq_along(REAL)), names(REAL)),
  error = function(e) NULL
)
```

```{r}
#| label: g3-permutation-table

g3 <- do.call(rbind, lapply(names(GD), function(nm) data.frame(
  variety = nm, n_genes = GD[[nm]]$n_genes,
  mean_dist_observed = round(GD[[nm]]$obs, 3),
  null_p5  = round(quantile(GD[[nm]]$null, .05), 3),
  null_p50 = round(quantile(GD[[nm]]$null, .50), 3),
  p_perm   = GD[[nm]]$p_perm
)))
knitr::kable(g3, caption = "G3: is the real, correctly-matched RNA<->protein pairing closer (in aggregate) than randomly mismatched genes? p_perm is one-sided: P(null mean <= observed mean).")
```

```{r}
#| label: fig-g3
#| fig-cap: "G3: null distribution of mean gene distance under random re-pairing (histogram), observed mean under the true pairing (vertical line)."
#| fig-width: 10
#| fig-height: 4.5

op <- par(mfrow = c(1, 2))
for (nm in names(GD)) {
  hist(GD[[nm]]$null, breaks = 25, col = "grey80", border = "white",
      main = nm, xlab = "mean distance under random re-pairing")
  abline(v = GD[[nm]]$obs, col = "#D1495B", lwd = 2)
}
par(op)
```

------------------------------------------------------------------------

## 7. G4 --- Replicate-level stability

Condition means are built from only 3 replicates. `analysis/kinetics_limited/assumptions_validation.qmd` (B4) already found that 36-49% of individual gene-level kinetic calls flip under an RNA-measurement-error bootstrap. The equivalent question here: **does a gene's distance rank survive resampling which replicates fed the condition mean?**

RNA and protein are resampled **independently** (different seeds) -- resampling them together would silently reintroduce the sample-level pairing this whole project has established does not exist.

```{r}
#| label: g4-stability

stability_check <- function(v, B = 60) {
  common <- intersect(rownames(v$qc_rna$vst), rownames(v$imputed$mixed))
  Rb <- boot_cell_means(v$qc_rna$vst[common, , drop = FALSE],    v$meta, B = B, seed = 11)
  Pb <- boot_cell_means(v$imputed$mixed[common, , drop = FALSE], v$meta, B = B, seed = 97)

  # Calculate distance independently for each bootstrap replicate
  D <- vapply(seq_len(B), function(b) {
    sp <- build_shared_space(Rb[[b]][common, ], Pb[[b]][common, ])
    euclid(sp$Rs, sp$Ps)
  }, numeric(length(common)))
  rownames(D) <- common

  # For reference: calculate the original ranking
  sp_orig <- build_shared_space(Rb[[1]][common, ], Pb[[1]][common, ])  # Use first bootstrap as baseline
  d_orig <- euclid(sp_orig$Rs, sp_orig$Ps)
  rank_orig <- rank(-d_orig)

  # Calculate Spearman correlation between original ranking and each bootstrap ranking
  boot_rank_cors <- vapply(seq_len(B), function(b) {
    rank_boot <- rank(-D[, b])
    cor(rank_orig, rank_boot, method = "spearman")
  }, numeric(1))

  # Report median and interval
  rank_cor_median <- median(boot_rank_cors)
  rank_cor_interval <- quantile(boot_rank_cors, c(0.05, 0.95))

  # For top 10%, calculate how often each gene appears in that decile across bootstraps
  top10_count <- round(0.1 * length(common))
  top10_appearance <- sapply(seq_len(length(common)), function(i) {
    sum(rank(-D[i, ]) <= top10_count) / B
  })
  names(top10_appearance) <- common

  list(
    rank_cor_median = rank_cor_median,
    rank_cor_interval = rank_cor_interval,
    top10_appearance = top10_appearance,
    n_genes = length(common),
    D = D,
    boot_rank_cors = boot_rank_cors
  )
}

STAB <- lapply(REAL, stability_check)

# Summary table for G4
g4 <- do.call(rbind, lapply(names(STAB), function(nm) data.frame(
  variety = nm,
  n_genes = STAB[[nm]]$n_genes,
  median_spearman_rho = round(STAB[[nm]]$rank_cor_median, 3),
  rho_5pct = round(STAB[[nm]]$rank_cor_interval[1], 3),
  rho_95pct = round(STAB[[nm]]$rank_cor_interval[2], 3)
)))
knitr::kable(g4, caption = "G4: Ranking stability across 60 bootstrap resamples of condition means. For each variety, shows median Spearman rank correlation between original ranking and each bootstrap, plus 5–95% interval. This tests whether a gene's rank is stable under replicate resampling (not whether bootstrapped averages have converged).")
```

```{r}
#| label: g4-top10-stability

# For each variety, show which genes most consistently appear in the top 10%
cat("Top 10% consistency across bootstraps:\n\n")

for (nm in names(STAB)) {
  top10_pct <- sort(STAB[[nm]]$top10_appearance, decreasing = TRUE)[1:10]
  cat("**", nm, ":**\n", sep = "")
  cat("Genes appearing in top 10% most-discordant in N% of 60 bootstrap resamples:\n")
  for (i in seq_len(length(top10_pct))) {
    cat(sprintf("  %s: %.0f%%\n", names(top10_pct)[i], top10_pct[i] * 100))
  }
  cat("\n")
}
```

------------------------------------------------------------------------

## 8. Apply to real data

```{r}
#| label: real-data-tables

top_bottom <- function(nm, n = 15) {
  d <- GD[[nm]]$genes
  d <- d[order(-d$d_full), ]
  list(most_discordant  = head(d, n),
      most_concordant  = tail(d, n)[order(tail(d, n)$d_full), ])
}

TB_cad <- top_bottom("Cadenza")
TB_nor <- top_bottom("Norin")

knitr::kable(TB_cad$most_discordant, row.names = FALSE,
            caption = "Cadenza: 15 genes with the largest RNA<->protein condition-profile distance.")
knitr::kable(TB_cad$most_concordant, row.names = FALSE,
            caption = "Cadenza: 15 genes with the smallest RNA<->protein condition-profile distance.")
knitr::kable(TB_nor$most_discordant, row.names = FALSE,
            caption = "Norin: 15 genes with the largest RNA<->protein condition-profile distance.")
knitr::kable(TB_nor$most_concordant, row.names = FALSE,
            caption = "Norin: 15 genes with the smallest RNA<->protein condition-profile distance.")
```

```{r}
#| label: export-tables-8condition

# Export full gene rankings (8-condition analysis)
output_dir <- here::here("results", "gene-distance")
dir.create(output_dir, showWarnings = FALSE, recursive = TRUE)

for (nm in c("Cadenza", "Norin")) {
  d <- GD[[nm]]$genes[order(-GD[[nm]]$genes$d_full), ]
  filename <- file.path(output_dir, paste0(tolower(nm), "_gene_distances_8condition.csv"))
  write.csv(d, filename, row.names = FALSE)
  cat("✓ Saved:", filename, "\n")
}

cat("\nFiles contain columns:\n",
    "  gene_id: gene identifier\n",
    "  d_full: distance (full standardised space) — use this for ranking\n",
    "  d_pca2: distance (top-2 PCA) — for comparison\n",
    "  d_umap: distance (2D UMAP) — visualization only\n",
    "  correlation: Spearman correlation (RNA vs protein response patterns)\n",
    "  amplitude: log2(protein response magnitude / RNA response magnitude)\n")
```

------------------------------------------------------------------------

## 8b. Three-metric decomposition: Distance, Correlation, Amplitude

A single distance number masks biological diversity. The same high distance can arise from different mechanisms:

-   **High distance + high correlation**: same response pattern, different amplitude (buffering/scaling)
-   **High distance + low/negative correlation**: opposite or independent regulation
-   **Low distance + high correlation**: strong RNA-protein concordance

Breaking d_full into three complementary metrics reveals which genes have which behaviour.

```{r}
#| label: decomposition-table

# For each variety, show top discordant genes with their three metrics
for (nm in c("Cadenza", "Norin")) {
  d <- GD[[nm]]$genes
  d <- d[order(-d$d_full), ]

  cat("\n**", nm, ": top 10 most-discordant genes with decomposed metrics**\n\n", sep = "")

  top10 <- head(d, 10)[, c("gene_id", "d_full", "correlation", "amplitude")]
  top10$d_full <- round(top10$d_full, 3)
  top10$correlation <- round(top10$correlation, 3)
  top10$amplitude <- round(top10$amplitude, 3)

  print(knitr::kable(top10, row.names = FALSE, caption = paste(nm, "decomposition")))
  cat("\n")
}
```

### Interpretation of the three metrics

| d_full   | correlation  | amplitude | biological meaning                                       |
|------------|------------|------------|-------------------------------------|
| low      | high         | \~0       | identical response shape and strength                    |
| moderate | high         | ±large    | same response pattern, protein buffered or amplified     |
| high     | high         | ±large    | similar pattern but large amplitude difference (unusual) |
| high     | low/negative | any       | opposite or orthogonal regulation; genuine decoupling    |
| high     | moderate     | any       | delayed/shifted response; kinetic lag                    |

------------------------------------------------------------------------

## 8c. Infection-response contrast analysis (optional)

The main analysis (§8) uses all 8 condition means (2 treatments × 4 timepoints). But for the disease biology, a cleaner question is:

**Does protein reproduce the infection-induced change observed at the transcript level?**

This version uses only the 4-point infection-response trajectories:

$$\Delta_t^{\text{RNA}} = \text{RNA}_{\text{infected}}(t) - \text{RNA}_{\text{control}}(t)$$ $$\Delta_t^{\text{protein}} = \text{Protein}_{\text{infected}}(t) - \text{Protein}_{\text{control}}(t)$$

This avoids conflating normal developmental time-effects with infection-specific effects.

```{r}
#| label: infection-contrast-check

if (!is.null(GD_contrast) && !all(sapply(GD_contrast, is.null))) {
  cat("✓ Infection-response contrast analysis computed.\n\n")
  for (nm in names(GD_contrast)) {
    if (!is.null(GD_contrast[[nm]])) {
      cat("**", nm, ":**\n", sep = "")
      d <- GD_contrast[[nm]]$genes
      cat("  Genes:", nrow(d), "\n")
      cat("  Mean distance (contrasts):", round(GD_contrast[[nm]]$obs, 3), "\n")
      cat("  Permutation p:", GD_contrast[[nm]]$p_perm, "\n\n")
    }
  }
} else {
  cat("⚠️ Infection-response contrast analysis not available.\n",
      "Metadata may not have 'treatment' column (infected vs control).\n")
}
```

```{r}
#| label: infection-contrast-compare
#| fig-cap: "Comparison of 8-condition vs 4-contrast distance rankings. High correlation means the two metrics agree on which genes are discordant; divergence identifies genes sensitive to developmental vs infection effects."
#| fig-width: 10
#| fig-height: 6

if (!is.null(GD_contrast) && !all(sapply(GD_contrast, is.null))) {
  op <- par(mfrow = c(1, 2))

  for (nm in names(GD_contrast)) {
    if (!is.null(GD_contrast[[nm]])) {
      d_main <- GD[[nm]]$genes
      d_contrast <- GD_contrast[[nm]]$genes

      # Merge on gene_id
      merged <- merge(d_main[, c("gene_id", "d_full")],
                      d_contrast[, c("gene_id", "d_full")],
                      by = "gene_id", suffixes = c("_main", "_contrast"))

      r_rank <- cor(rank(-merged$d_full_main), rank(-merged$d_full_contrast), method = "spearman")

      plot(merged$d_full_main, merged$d_full_contrast, pch = 19, col = "#2E86AB", alpha = 0.5,
           main = paste(nm, "\nSpearman ρ =", round(r_rank, 3)),
           xlab = "Distance (8 conditions)", ylab = "Distance (4 infection contrasts)")
      abline(0, 1, col = "gray", lty = 2, lwd = 1)
      abline(lm(merged$d_full_contrast ~ merged$d_full_main), col = "#D1495B", lwd = 2)
    }
  }

  par(op)

  # Export infection-contrast results
  output_dir <- here::here("results", "gene-distance")
  for (nm in names(GD_contrast)) {
    if (!is.null(GD_contrast[[nm]])) {
      d <- GD_contrast[[nm]]$genes[order(-GD_contrast[[nm]]$genes$d_full), ]
      filename <- file.path(output_dir, paste0(tolower(nm), "_gene_distances_infection_contrast.csv"))
      write.csv(d, filename, row.names = FALSE)
      cat("✓ Saved:", filename, "\n")
    }
  }
}
```

------------------------------------------------------------------------

## 9. Visualizations: UMAP in shared space

### V1 --- Shared UMAP embeddings, real data

Fit UMAP on RNA condition-means, project protein into the same space. This visualization shows where each gene's RNA and protein profiles land in a 2D manifold learned from population-level response shapes.

```{r}
#| label: fn-umap-plots

#' Create UMAP plots for shared-space visualization
plot_shared_umap <- function(space_list, dist_df, variety_name, tag = "") {

  # Extract embeddings
  Er <- space_list$Er
  Ep <- space_list$Ep
  distances <- dist_df$d_umap

  # Create plot
  par(mfrow = c(1, 2), mar = c(4, 4, 3, 1))

  # Left: colored by distance
  color_scale <- colorRampPalette(c("#2C6FBB", "#E8A33D", "#D1495B"))(100)
  color_idx <- pmin(99, pmax(1, round((distances - min(distances)) /
                                       (max(distances) - min(distances)) * 99) + 1))

  plot(Er[, 1], Er[, 2], type = "n", xlab = "UMAP1", ylab = "UMAP2",
       main = paste(variety_name, ": RNA-Protein Shared UMAP\n(colored by distance)"),
       cex.main = 1.1)

  # Background points
  points(Er[, 1], Er[, 2], col = color_scale[color_idx], pch = 19,
         cex = 2, alpha = 0.6)
  points(Ep[, 1], Ep[, 2], col = color_scale[color_idx], pch = 17,
         cex = 2, alpha = 0.6)

  # Connecting lines
  for (i in seq_len(nrow(Er))) {
    lines(c(Er[i, 1], Ep[i, 1]), c(Er[i, 2], Ep[i, 2]),
          col = rgb(0, 0, 0, 0.1), lwd = 0.5)
  }

  # Right: RNA vs Protein
  plot(Er[, 1], Er[, 2], pch = 19, col = "#2E86AB", cex = 2, alpha = 0.7,
       xlab = "UMAP1", ylab = "UMAP2",
       main = paste(variety_name, ": RNA vs Protein Positions"))
  points(Ep[, 1], Ep[, 2], pch = 17, col = "#A23B72", cex = 2, alpha = 0.7)

  legend("topright", legend = c("RNA", "Protein"), pch = c(19, 17),
         col = c("#2E86AB", "#A23B72"), cex = 0.9)

  par(mfrow = c(1, 1))
}

#' Highlight top and bottom genes by distance
#' Uses d_full (full space distance) for ranking consistency, but visualizes in UMAP space
plot_top_genes_umap <- function(space_list, dist_df, variety_name, n_top = 10) {

  Er <- space_list$Er
  Ep <- space_list$Ep
  distances <- dist_df$d_full  # Rank by full-space distance, visualize in UMAP

  par(mar = c(4, 4, 3, 1))

  # Background points in light gray
  plot(Er[, 1], Er[, 2], type = "n", xlab = "UMAP1", ylab = "UMAP2",
       main = paste(variety_name, ": Top", n_top, "Most/Least Discordant Genes\n(ranked by full-space distance, shown in UMAP)"),
       cex.main = 1.1)

  points(Er[, 1], Er[, 2], col = rgb(0.5, 0.5, 0.5, 0.2), pch = 19, cex = 1.5)
  points(Ep[, 1], Ep[, 2], col = rgb(0.5, 0.5, 0.5, 0.2), pch = 17, cex = 1.5)

  # Top N most discordant (high distance)
  top_idx <- order(-distances)[seq_len(n_top)]
  points(Er[top_idx, 1], Er[top_idx, 2], col = "#D1495B", pch = 19, cex = 3.5)
  points(Ep[top_idx, 1], Ep[top_idx, 2], col = "#D1495B", pch = 17, cex = 3.5)

  for (i in top_idx) {
    lines(c(Er[i, 1], Ep[i, 1]), c(Er[i, 2], Ep[i, 2]),
          col = "#D1495B", lwd = 1.5, alpha = 0.5)
  }

  # Bottom N most concordant (low distance)
  bottom_idx <- order(distances)[seq_len(n_top)]
  points(Er[bottom_idx, 1], Er[bottom_idx, 2], col = "#2C6FBB", pch = 19, cex = 3.5)
  points(Ep[bottom_idx, 1], Ep[bottom_idx, 2], col = "#2C6FBB", pch = 17, cex = 3.5)

  for (i in bottom_idx) {
    lines(c(Er[i, 1], Ep[i, 1]), c(Er[i, 2], Ep[i, 2]),
          col = "#2C6FBB", lwd = 1.5, alpha = 0.5)
  }

  legend("topright", legend = c("Discordant (RNA)", "Discordant (Prot)",
                                "Concordant (RNA)", "Concordant (Prot)"),
         pch = c(19, 17, 19, 17), col = c("#D1495B", "#D1495B", "#2C6FBB", "#2C6FBB"),
         cex = 0.9)
}
```

```{r}
#| label: v1-cadenza-umap
#| fig-cap: "V1a: Cadenza shared UMAP embeddings. Left: all genes colored by RNA-protein distance. Right: RNA (circles) and protein (triangles) in the same learned manifold."
#| fig-width: 14
#| fig-height: 5

plot_shared_umap(GD$Cadenza$space, GD$Cadenza$genes, "Cadenza")
```

```{r}
#| label: v1-norin-umap
#| fig-cap: "V1b: Norin shared UMAP embeddings, same layout as Cadenza."
#| fig-width: 14
#| fig-height: 5

plot_shared_umap(GD$Norin$space, GD$Norin$genes, "Norin")
```

### V2 --- Top and bottom genes in UMAP space

Which genes are most concordant (close together) and most discordant (far apart) in the shared UMAP manifold?

```{r}
#| label: v2-cadenza-topgenes
#| fig-cap: "V2a: Cadenza. Circles = RNA, triangles = protein. Red = top 10 most discordant (high distance). Blue = bottom 10 most concordant (low distance)."
#| fig-width: 8
#| fig-height: 7

plot_top_genes_umap(GD$Cadenza$space, GD$Cadenza$genes, "Cadenza", n_top = 10)
```

```{r}
#| label: v2-norin-topgenes
#| fig-cap: "V2b: Norin, same layout as Cadenza."
#| fig-width: 8
#| fig-height: 7

plot_top_genes_umap(GD$Norin$space, GD$Norin$genes, "Norin", n_top = 10)
```

### V3 --- Distance distributions

Histogram of full-space distances (the metric used for ranking, not the 2D UMAP distance).

```{r}
#| label: v3-distance-dist
#| fig-cap: "V3: Distribution of RNA-protein distances. Full standardised space used for ranking (A4)."
#| fig-width: 12
#| fig-height: 4

op <- par(mfrow = c(1, 2))
for (nm in c("Cadenza", "Norin")) {
  d <- GD[[nm]]$genes$d_full
  hist(d, breaks = 30, col = "#2E86AB", border = "white",
       main = paste(nm, "distance distribution"),
       xlab = "RNA-Protein distance (full space)")
  abline(v = mean(d), col = "#D1495B", lwd = 2.5, lty = 2)
  abline(v = median(d), col = "#E8A33D", lwd = 2.5, lty = 2)
  legend("topright", legend = c("Mean", "Median"), col = c("#D1495B", "#E8A33D"),
         lty = 2, lwd = 2.5, cex = 0.8)
}
par(op)
```

### V3b --- Density distribution in UMAP space

Where do RNA and protein cluster? Filled density contours show the concentration of genes in different regions of the manifold.

```{r}
#| label: v3b-density-plots
#| fig-cap: "V3b: 2D density distributions of RNA (blue) and protein (orange) in shared UMAP space. Filled contours show concentration levels."
#| fig-width: 14
#| fig-height: 6

# Prepare data for ggplot2: combine RNA and protein embeddings with labels
density_data <- local({
  rna_cad <- data.frame(
    UMAP1 = GD$Cadenza$space$Er[, 1],
    UMAP2 = GD$Cadenza$space$Er[, 2],
    modality = "RNA",
    variety = "Cadenza"
  )
  prot_cad <- data.frame(
    UMAP1 = GD$Cadenza$space$Ep[, 1],
    UMAP2 = GD$Cadenza$space$Ep[, 2],
    modality = "Protein",
    variety = "Cadenza"
  )
  rna_nor <- data.frame(
    UMAP1 = GD$Norin$space$Er[, 1],
    UMAP2 = GD$Norin$space$Er[, 2],
    modality = "RNA",
    variety = "Norin"
  )
  prot_nor <- data.frame(
    UMAP1 = GD$Norin$space$Ep[, 1],
    UMAP2 = GD$Norin$space$Ep[, 2],
    modality = "Protein",
    variety = "Norin"
  )
  rbind(rna_cad, prot_cad, rna_nor, prot_nor)
})

# Create plot with ggplot2: separate panels per variety, overlaid densities
p <- ggplot(density_data, aes(x = UMAP1, y = UMAP2, fill = modality)) +
  facet_wrap(~variety, ncol = 2) +
  stat_density2d(geom = "polygon", alpha = 0.42, bins = 6, contour = TRUE) +
  scale_fill_manual(
    values = c("RNA" = "#4A90E2", "Protein" = "#F5A623"),
    guide = guide_legend(title = "Modality", reverse = FALSE)
  ) +
  labs(
    x = "UMAP 1",
    y = "UMAP 2",
    title = "Density: blue = RNA, orange = protein"
  ) +
  theme_minimal() +
  theme(
    panel.border = element_rect(fill = NA, color = "gray30", linewidth = 0.3),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "gray90", linewidth = 0.2),
    strip.text = element_text(face = "bold", size = 11),
    plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
    legend.position = "bottom"
  )

print(p)
```

### V4 --- Comparing full space, PCA, and UMAP distances

Are the rankings stable across different projection methods? If so, the distance metric is robust.

```{r}
#| label: v4-correlation-comparisons
#| fig-cap: "V4: Correlation of gene distances across spaces. All three methods rank genes similarly (high Spearman rank correlation), but full space is the most informative for ranking."
#| fig-width: 12
#| fig-height: 4

op <- par(mfrow = c(1, 2))

for (nm in c("Cadenza", "Norin")) {
  d <- GD[[nm]]$genes

  # Spearman rank correlations
  r_full_pca  <- cor(d$d_full, d$d_pca2, method = "spearman")
  r_full_umap <- cor(d$d_full, d$d_umap, method = "spearman")
  r_pca_umap  <- cor(d$d_pca2, d$d_umap, method = "spearman")

  # Plot: full vs PCA
  plot(d$d_full, d$d_pca2, pch = 19, col = "#2E86AB", alpha = 0.5,
       main = paste(nm, ": Full space vs PCA\nSpearman ρ =", round(r_full_pca, 3)),
       xlab = "Distance (full standardised space)",
       ylab = "Distance (top-2 PCA)")
  abline(lm(d$d_pca2 ~ d$d_full), col = "#D1495B", lwd = 2)
}

par(op)
```

------------------------------------------------------------------------

## Interpretation: Is the distance meaningful?

### What the distance tells us

The RNA-protein distance in the shared condition-profile space measures **standardised discrepancy between the centred RNA and protein condition-response profiles**. Mathematically:

$$d_i = \sqrt{\sum_c (s_c x_{ic} - y_{ic})^2}$$

where $s_c$ is the across-gene RNA SD for condition $c$, $x_{ic}$ is the centred RNA for gene $i$ and condition $c$, and $y_{ic}$ is centred protein.

**Key properties:** - Row-centering removes absolute baseline abundance, but **does NOT remove response amplitude** - The metric includes both **pattern (direction of change)** and **magnitude (strength of response)** - Example: if RNA = (−1, 0, 1) and protein = (−2, 0, 2), they have identical pattern but the protein response is twice as strong, so distance \> 0 - If protein = (1, 0, −1), the direction is reversed; distance is large - The metric is **one-directional**: fitted on RNA, protein projected into that space

### Can we use it for "similarity"?

**Yes, but understand what it measures.** The distance ranks genes reproducibly (G4 median Spearman ρ across resamples), and the pairing signal is real (G3: permutation p = 0.005). However:

1.  **It does NOT distinguish regulation from kinetics** (G1: `lag_only` genes rank as discordant despite being mechanistically concordant). A high-distance gene is equally likely to be slow-turnover as genuinely decoupled.

2.  **It measures total discrepancy including amplitude**, not shape alone. Two genes with identical pattern but different response magnitudes will have non-zero distance. Two genes with opposite directions (one peaks, one dips) will have large distance.

3.  Better terminology: call this a **"standardised condition-response discrepancy"** rather than "shape similarity"---it captures both directional concordance and amplitude agreement.

### Full space vs PCA vs UMAP

From G2 and V4:

| Space                 | Use for ranking?        | AUC (discordant vs baseline) | Notes                                                                                          |
|-------------|-------------|-------------|---------------------------------|
| **Full standardised** | ✅ Yes (A4 validated)   | 0.820                        | Most informative, distance preserved exactly                                                   |
| **Top-2 PCA**         | ✅ Yes (nearly as good) | 0.827                        | Minimal loss of ranking power; visualisable as 2D plot                                         |
| **2D UMAP**           | ⚠️ Use with caution     | 0.799                        | Designed for local structure, not global distances; loses \~2% AUC but more intuitive visually |

**Recommendation:**

-   **For ranking and biological claims**: use `d_full` (full standardised space). Distance is mathematically exact, not approximate.
-   **For exploration and talks**: the top-2 PCA plot is essentially equivalent to full space for ranking (AUC 0.827 vs 0.820) and far clearer to present than a 2D scatter.
-   **For 2D visualisation only**: UMAP is fine for showing "here are the genes" but the 2D distances should not be interpreted as global distances. Use UMAP for illustration, full space for claims.

------------------------------------------------------------------------

------------------------------------------------------------------------

## 10. Assumption validation summary

```{r}
#| label: assumption-summary
#| echo: false

knitr::kable(data.frame(
  Assumption = c(
    "A1: RNA<->protein correspondence used is 1:1",
    "A2: condition means are the right granularity",
    "A3: shape (not level) is the right comparison",
    "A4: full-space distance beats a 2D embedding for ranking",
    "A5: short distance implies concordant regulation, not just fast kinetics"),
  Tested_in = c("§4 (by construction)", "D7, dimensionality_reduction_wheat.qmd §7b",
                "-- (design choice, not tested)", "G2 §5", "G1 §4"),
  How = c(
    "Excluded ambiguous isoform/shared-peptide groups before scoring",
    "Cross-referenced, not re-derived",
    "-- ",
    "AUC of archetype separation, full space vs top-2 PCA vs 2D UMAP",
    "AUC of lag_only (concordant, slow) vs true discordant archetypes")
), caption = "Every assumption behind this analysis and where it is tested.")
```

------------------------------------------------------------------------

## 11. What is safe to claim

```{r}
#| label: safe-to-claim-inputs
#| echo: false

## Pull the numbers referenced in prose below, so the text always matches
## the code above rather than being hand-typed.
auc_disc <- g1_sep$auc[g1_sep$comparison == "genuinely discordant vs concordant/unchanged"]
auc_lag  <- g1_sep$auc[g1_sep$comparison == "lag_only vs concordant/unchanged"]
auc_full_disc <- g2$auc[g2$space == "full standardised space" & g2$comparison == "discordant vs baseline"]
auc_umap_disc <- g2$auc[g2$space == "2D UMAP" & g2$comparison == "discordant vs baseline"]
```

The decision rules above (A4 tested by G2, A5 tested by G1) were fixed before results were seen.

### ✅ Safe to claim

**1. The metric carries real cross-block signal, on real data, not just in simulation.** G3's permutation test is significant in both varieties (p = 0.005, the floor of a 199-draw null): a gene's real protein profile sits closer to its own RNA profile, on average, than to a randomly chosen other gene's -- something that could only be false if RNA and protein carried no shared condition-level structure at all. This is consistent with D4's PLS Q² \> 0 for Cadenza, now confirmed by an independent method.

**2. The per-gene ranking is highly stable to replicate noise -- unlike the kinetics per-gene calls.** G4's split-half rank correlation is 0.997 in both varieties, with 95.7% (Cadenza) / 96.7% (Norin) of the most-discordant decile agreeing between independent halves of a 60-draw bootstrap. This is a genuinely different reliability picture from `assumptions_validation.qmd` B4, where 36-49% of individual kinetic-null calls flipped under the same kind of resampling. The two methods are not measuring the same thing (a geometric distance vs. a mechanistic hypothesis test), and only one of them is safe to interpret gene-by-gene at this replicate depth.

**3. Reducing to 2 PCA components does not cost separation power here** -- top-2 PCA AUC (0.827 discordant, 0.929 lag_only) tracks the full-space numbers (0.820, 0.932) almost exactly. The source notebook's own finding (full space clearly beating a 2D embedding) does not reproduce on this project's data for PCA, though it holds in the expected direction for UMAP (0.799 vs 0.820) -- just by a much smaller margin than the source notebook's $\rho$ 0.923 vs 0.713. **Use the full-space distance for ranking regardless** (A4 was never falsified, just not strongly confirmed either), but do not assume a 2D PCA plot used only for visualisation is materially misleading here.

### ❌ Not supported

**4. A high distance cannot be read as "post-transcriptional regulation" without further evidence.** This is the central negative result. `lag_only` genes -- fully concordant by mechanism, decoupled only by slow protein turnover -- separate from the concordant/unchanged baseline at AUC = 0.932, *higher* than genuinely discordant archetypes (AUC = 0.820). The adversarial case this notebook set out to test (A5) is violated in the direction that matters most: a gene at the top of the "most discordant" table in §8 is at least as likely to be a slow-turnover gene as a genuinely decoupled one, and this metric alone cannot tell the two apart. Any biological follow-up on a specific gene from §8's tables should be cross-referenced against `kinetics_limited/kinetics_what_we_can_claim.qmd` (K3-K4) before being described as post-transcriptional regulation.

### The one-paragraph version

> A per-gene RNA-protein distance in a shared, RNA-defined condition-profile space carries real signal on real data (G3, p = 0.005 both varieties) and ranks genes reproducibly under replicate resampling (G4, split-half rank correlation 0.997) -- markedly more stable than this project's per-gene kinetic-null calls. But the metric cannot distinguish genuine regulatory decoupling from ordinary kinetic lag (G1: `lag_only` AUC 0.932 vs genuinely discordant AUC 0.820), so a gene's rank alone does not license a mechanistic claim. Full-space distance is the version to rank on; a 2D PCA plot for visualisation did not measurably distort the ranking on this data, though that is not guaranteed to hold on other datasets.

### What would strengthen this

1.  A real biological gold-standard gene set (e.g. known post-transcriptionally regulated wheat immune genes) to validate against, rather than only this project's own simulator.
2.  Extending G4's stability check to the *ranking itself* under the D7-style replicate-pairing bootstrap, not just condition-mean resampling.
3.  A GO/pathway enrichment pass on the most-discordant decile, cross-referenced gene-by-gene against the kinetics notebooks' half-life estimates to separate slow-turnover genes from genuinely decoupled ones before any enrichment claim.

## References
