---
title: "RNA-protein condition-profile distance --- can genes be ranked by discordance?"
author: "Kristina Gagalova"
format:
  gfm:
    toc: true
    code-tools: true
    fig-dir: "figs"
execute:
  echo: true
editor: visual
bibliography: ../kinetics_limited/references.bib
---

## Purpose and scope

This notebook asks a question distinct from anything in `analysis/kinetics_limited` or `analysis/dimensionality-reduction`:

> **For an individual gene, does its protein's condition-response profile look like its RNA's condition-response profile --- and can that resemblance be turned into a per-gene distance that ranks genes by RNA/protein discordance?**

`kinetics_limited` tests a *mechanistic* null (first-order kinetics) per gene. `dimensionality-reduction` asks whether RNA predicts protein *in aggregate*, across all genes at once. This notebook asks a *geometric* question per gene: place each gene's RNA profile and its own protein profile as two points in a shared coordinate system built from the population's dominant response shapes, and measure how far apart they land.

### Relationship to the existing Python work

Two notebooks in `KristinaGagalova/autoencoders-test` motivate this:

**`rna_protein_unpaired_per_variety.ipynb`** is a thin driver over `rnaprot/unpaired_reviewer.py` (a revised module -- not the `unpaired.py` referenced in `dimensionality_reduction_wheat.qmd`; the two have different loss weights, 3.0/0.75 here vs 2.0/0.5 there). It re-runs essentially the same leave-condition-out comparison (mean, design, cognate ridge, PCA+ridge, PLS, design-residualised variants, condition-aligned AE) already built in this repository's `dimensionality_reduction_wheat.qmd` §7 (D4) and §7b (D7). Two things in it are genuinely different from what is already here, and are picked up as an addendum (**D8**) in that notebook rather than duplicated wholesale in a second \~40-chunk document:

1.  Feature selection is done **inside each outer CV fold**, not once globally before the loop -- D4/D7 currently select the top-variable genes once, which is a selection-leakage risk.
2.  A permutation null (`N_PERM=200`) is applied to the PCA/PLS/design baselines (not to the autoencoder).

See `dimensionality_reduction_wheat.qmd` §7c for that check. This document does not repeat D4/D7.

**`rna_protein_umap_shared_space.ipynb`** is the actual source of the "distance between genes" idea, and *is* new work relative to everything else in this repository. It builds condition-mean profiles for RNA and for cognate proteins, fits a shared coordinate system (PCA or UMAP) on the RNA profiles only, projects the protein profiles into that same system, and measures the Euclidean distance between a gene's RNA point and its own protein point. Two things about it matter for how it is ported here:

1.  **Its own synthetic validation found that 2D UMAP hurts the ranking**: distance-to-truth-divergence correlation was Spearman $\rho = 0.923$ in the full aligned input space, but only $\rho = 0.713$ after compressing to a 2D UMAP embedding. That is carried over below as a **hypothesis to test on this project's own ground truth**, not assumed.
2.  **It was never actually run to completion on real data.** The real-data cells stop at the visualization step; no per-gene distance table or ranking exists for Cadenza or Norin. Building that is the new work in §6-§8 below.

------------------------------------------------------------------------

## Analysis outline

| Step | What                                           | Why                                                       |
|:-----|:-----------------------------------------------|:----------------------------------------------------------|
| §1   | Setup and data (real + simulated)              | Validate on known ground truth before touching real data  |
| §2   | The method and its assumptions                 | Each with references                                      |
| §3   | Implementation                                 | Shared-space construction, from scratch                   |
| §4   | **G1** Does distance separate true archetypes? | Validation against `truth.tsv`, incl. the `lag_only` trap |
| §5   | **G2** Full space vs 2D embedding              | Replicates the source notebook's own finding, on our data |
| §6   | **G3** Permutation null on real data           | No ground truth on real data -- this substitutes for it   |
| §7   | **G4** Replicate-level stability               | Bootstrap over which replicates feed the condition means  |
| §8   | Apply to real data                             | Per-gene distance tables, Cadenza and Norin               |
| §9   | Assumption validation summary                  |                                                           |
| §10  | What is safe to claim                          |                                                           |

------------------------------------------------------------------------

## 1. Setup

```{r}
#| label: setup
#| message: false

if (!requireNamespace("here", quietly = TRUE)) stop("package 'here' required")

## RENV_PROJECT must be set before sourcing activate.R, or renv treats this
## subdirectory as the project and bootstraps an empty library.
if (file.exists(here::here("renv", "activate.R"))) {
  Sys.setenv(RENV_PROJECT = here::here())
  source(here::here("renv", "activate.R"))
}

source(here::here("R", "utils.R"))
source(here::here("R", "wheat_pipeline.R"))
need_pkgs(c("limma", "edgeR", "DESeq2", "matrixStats", "impute", "uwot", "yaml"))

set.seed(20260823)
```

```{r}
#| label: load-real

DESIGN <- wheat_design()
cad <- prepare_variety("cadenza", DESIGN)
nor <- prepare_variety("norin",   DESIGN)
REAL <- list(Cadenza = cad, Norin = nor)
```

```{r}
#| label: load-simulated
#| message: false

#' Loader for the project's own Case B ground-truth simulator
#' (`R/00_simulate_data.R`), parallel to `load_variety()` in
#' `wheat_pipeline.R` but pointed at `data/simulated` and its different file
#' layout. This gives archetype-labelled truth to validate the distance
#' metric against, before it is trusted on real data -- the same validate-
#' on-simulation-first discipline used for the rest of this pipeline.
load_simulated <- function(cfg, data_dir = here::here("data", "simulated")) {

  rna_counts <- as.matrix(read.delim(file.path(data_dir, "rna_counts.tsv"),
                                     row.names = 1, check.names = FALSE))
  prot_int   <- as.matrix(read.delim(file.path(data_dir, "protein_intensities.tsv"),
                                     row.names = 1, check.names = FALSE))
  rna_meta   <- read.delim(file.path(data_dir, "rna_coldata.tsv"),
                           row.names = "sample_id", stringsAsFactors = FALSE)
  prot_meta  <- read.delim(file.path(data_dir, "protein_coldata.tsv"),
                           row.names = "sample_id", stringsAsFactors = FALSE)
  truth <- read.delim(file.path(data_dir, "truth.tsv"), stringsAsFactors = FALSE)
  gmap  <- read.delim(file.path(data_dir, "truth_gene_group_map.tsv"),
                      stringsAsFactors = FALSE)

  lv <- cell_levels(cfg)
  rna_meta$cell  <- factor(rna_meta$cell,  levels = lv)
  prot_meta$cell <- factor(prot_meta$cell, levels = lv)

  design <- list(treatments = cfg$design$treatments,
                timepoints = cfg$design$timepoints,
                n_reps     = cfg$design$n_reps)

  list(rna  = rna_counts[, rownames(rna_meta), drop = FALSE],
       prot = prot_int[,  rownames(prot_meta), drop = FALSE],
       rna_meta = rna_meta, prot_meta = prot_meta,
       design = design, truth = truth, gmap = gmap)
}

sim_cfg <- load_config(here::here("config", "config.yaml"))
sim_raw <- load_simulated(sim_cfg)

sim_qc <- local({
  qr <- run_qc_rna(sim_raw$rna, sim_raw$rna_meta, sim_raw$design)
  qp <- run_qc_prot(sim_raw$prot, sim_raw$prot_meta)
  mi <- run_missingness(qp$norm, sim_raw$prot_meta)
  list(vst = qr$vst, prot = mi$mixed)
})

cat("Simulated genes after RNA QC:   ", nrow(sim_qc$vst),  "\n")
cat("Simulated groups after prot QC: ", nrow(sim_qc$prot), "\n")
```

------------------------------------------------------------------------

## 2. The method and its assumptions

For each gene present in both layers:

1.  Take its **condition-mean profile** in each layer separately (8 treatment x timepoint cells) -- the only aggregation licensed under Case B, per `dimensionality_reduction_wheat.qmd` §7b (D7).
2.  **Row-centre** each profile (subtract the gene's own across-condition mean). This compares response *shape*, not absolute level -- required regardless of Case B, since RNA (VST) and protein (log2 intensity) are on incommensurable physical scales to begin with.
3.  **Standardise per condition** (z-score each of the 8 condition columns across genes), fit on the RNA block only, and apply the same transform to protein. This matches the source notebook's `StandardScaler().fit(Xr)` step.
4.  Fit **PCA on the RNA profiles only** (their dominant shapes of response across the gene population), and project the protein profiles into that same space with no refitting.
5.  **Distance** = Euclidean distance between a gene's transformed RNA point and its own transformed protein point -- computed in the full standardised space, in the top-2 PCA components, and in a 2D UMAP embedding fit on RNA and transformed for protein.

### Assumptions, stated up front

| \#  | Assumption                                                                               | Status                                                                      |
|:----|:-----------------------------------------------------------------------------------------|:----------------------------------------------------------------------------|
| A1  | The RNA\<-\>protein correspondence used is genuinely 1:1 for the genes scored            | Enforced by construction (see §4); ambiguous groups excluded                |
| A2  | Condition means are the right granularity for a cross-block comparison                   | Carried over from D7, not re-derived here                                   |
| A3  | Comparing row-centred *shape*, not absolute level, is the right notion of "look alike"   | Design choice, stated not tested                                            |
| A4  | Distance in the full standardised space is more informative than in a 2D embedding       | **Tested in G2**, not assumed                                               |
| A5  | A short Euclidean distance means concordant regulation, not just similar kinetics/timing | **Tested in G1** -- this is where `lag_only` genes are the adversarial case |

@mcinnes2018 and @becht2019 are the basis for A4: UMAP is explicitly designed to preserve local neighbourhood structure, not global pairwise distances, so a 2D UMAP distance between two specific, possibly-distant points is not guaranteed to track their true distance in the original space -- exactly the degradation the source notebook's own synthetic check already found ($\rho$ 0.923 to 0.713).

------------------------------------------------------------------------

## 3. Implementation

```{r}
#| label: fn-shared-space

#' Per-gene shape: subtract the gene's own across-condition mean.
row_center <- function(M) M - rowMeans(M)

#' Fit a per-column (per-condition) z-scorer on one matrix, return a function
#' that applies it to any matrix with the same columns. Fit on RNA only, per
#' the source notebook.
fit_col_scaler <- function(M) {
  mu  <- colMeans(M)
  sdv <- apply(M, 2, sd); sdv[sdv < 1e-8] <- 1
  function(X) sweep(sweep(X, 2, mu, "-"), 2, sdv, "/")
}

#' Build the shared space: row-centre, z-score on RNA's column moments, PCA
#' fit on RNA only. Returns the standardised profiles (full-space distance
#' needs nothing else, since the PCA rotation below is orthonormal and
#' therefore distance-preserving) and their PCA scores (for the 2D check).
build_shared_space <- function(R_cond, P_cond) {
  stopifnot(identical(rownames(R_cond), rownames(P_cond)))
  scaler <- fit_col_scaler(row_center(R_cond))
  Rs <- scaler(row_center(R_cond))
  Ps <- scaler(row_center(P_cond))

  sv <- svd(Rs)
  list(Rs = Rs, Ps = Ps,
       Zr = Rs %*% sv$v, Zp = Ps %*% sv$v,
       var_explained = sv$d^2 / sum(sv$d^2))
}

euclid <- function(A, B, k = NULL) {
  if (!is.null(k)) { A <- A[, seq_len(k), drop = FALSE]; B <- B[, seq_len(k), drop = FALSE] }
  sqrt(rowSums((A - B)^2))
}

#' UMAP fit on RNA, protein transformed into the same embedding -- matching
#' the source notebook's asymmetric fit/transform (never a joint/symmetric
#' fit, which would let protein leak into the axes it is being scored against).
umap_distance <- function(Rs, Ps, seed = 1) {
  nn <- max(2, min(15, nrow(Rs) %/% 4))
  set.seed(seed)
  fit <- uwot::umap(Rs, n_neighbors = nn, min_dist = 0.10, n_components = 2,
                    metric = "euclidean", ret_model = TRUE)
  Ep <- uwot::umap_transform(Ps, fit)
  sqrt(rowSums((fit$embedding - Ep)^2))
}
```

------------------------------------------------------------------------

## 4. G1 --- Does distance separate the true archetypes?

The simulator's `truth.tsv` labels every gene with the *mechanism* that generated it: `unchanged`, `concordant`, `buffered`, `mrna_only`, `protein_only`, `anticorrelated`, `lag_only`. None of that label is used to build the distance metric -- it is used only to grade it afterwards.

`lag_only` is the adversarial case, carried over from the kinetics notebooks: it is **mechanistically fully concordant** (protein is driven purely by RNA, first-order kinetics, `A2` holds exactly) but has a **slow protein half-life**, so its condition-profile *shape* lags the RNA shape. A method that cannot tell "concordant but slow" from "genuinely decoupled" will misclassify it -- `analysis/kinetics_limited/assumptions_validation.qmd` documents the same trap for naive quadrant classification. This is the sharpest test of A5.

```{r}
#| label: g1-simulated-distance

## Restrict to genes with an unambiguous 1:1 RNA<->protein-group
## correspondence. Isoform-split and shared-peptide groups (the simulator's
## deliberately messy 1:many / many:1 cases) are excluded from this analysis
## by design -- a distance metric between a gene and an ambiguous protein
## mixture is not answering the question this notebook asks.
gmap <- sim_raw$gmap
one_to_one <- with(gmap,
  ave(seq_along(gene_id), gene_id, FUN = length) == 1 &
  ave(seq_along(group),   group,   FUN = length) == 1 &
  frac == 1)
map1 <- gmap[one_to_one, c("gene_id", "group")]
map1 <- map1[map1$gene_id %in% rownames(sim_qc$vst) &
             map1$group  %in% rownames(sim_qc$prot), ]

cat("1:1 genes available for validation:", nrow(map1), "/", nrow(gmap), "\n")

R_cond_sim <- cell_means(sim_qc$vst[map1$gene_id, , drop = FALSE],  sim_raw$rna_meta)
P_cond_sim <- cell_means(sim_qc$prot[map1$group,  , drop = FALSE], sim_raw$prot_meta)
rownames(P_cond_sim) <- map1$gene_id

sp_sim     <- build_shared_space(R_cond_sim, P_cond_sim)
d_full_sim <- euclid(sp_sim$Rs, sp_sim$Ps)
d_pca2_sim <- euclid(sp_sim$Zr, sp_sim$Zp, k = 2)
d_umap_sim <- umap_distance(sp_sim$Rs, sp_sim$Ps)

arch <- setNames(sim_raw$truth$archetype, sim_raw$truth$gene_id)[map1$gene_id]

g1 <- data.frame(gene_id = map1$gene_id, archetype = arch,
                 d_full = d_full_sim, d_pca2 = d_pca2_sim, d_umap = d_umap_sim)
```

```{r}
#| label: fig-g1
#| fig-cap: "G1: full-space distance by true archetype (simulated data). concordant/unchanged are the null; lag_only is the adversarial case -- mechanistically concordant but kinetically slow."
#| fig-width: 9
#| fig-height: 5

ord <- c("unchanged", "concordant", "lag_only", "buffered",
        "mrna_only", "protein_only", "anticorrelated")
boxplot(d_full ~ factor(archetype, levels = ord), data = g1,
       las = 2, xlab = "", ylab = "distance (full standardised space)",
       col = ifelse(ord %in% c("unchanged", "concordant"), "#2C6FBB",
             ifelse(ord == "lag_only", "#E8A33D", "#D1495B")))
```

```{r}
#| label: g1-separation

#' Rank-based separation of a "positive" (expected-far) archetype set from a
#' "negative" (expected-close) baseline. AUC > 0.5 means the positive set
#' really does score farther, in the direction the mechanism predicts.
sep_stat <- function(d, pos, neg) {
  w <- suppressWarnings(wilcox.test(d[pos], d[neg]))
  data.frame(n_pos = sum(pos), n_neg = sum(neg),
            auc = round(unname(w$statistic) / (sum(pos) * sum(neg)), 3),
            p   = signif(w$p.value, 3))
}

baseline <- g1$archetype %in% c("unchanged", "concordant")
truly_discordant <- g1$archetype %in% c("buffered", "mrna_only",
                                        "protein_only", "anticorrelated")
lag <- g1$archetype == "lag_only"

g1_sep <- rbind(
  data.frame(comparison = "genuinely discordant vs concordant/unchanged",
            sep_stat(g1$d_full, truly_discordant, baseline)),
  data.frame(comparison = "lag_only vs concordant/unchanged",
            sep_stat(g1$d_full, lag, baseline))
)
knitr::kable(g1_sep, caption = "G1: separation of archetypes by full-space distance. AUC near 0.5 means the metric cannot tell the groups apart.")
```

------------------------------------------------------------------------

## 5. G2 --- Full space vs 2D embedding

```{r}
#| label: g2-embedding-comparison

g2 <- rbind(
  data.frame(space = "full standardised space",
            rbind(sep_stat(g1$d_full, truly_discordant, baseline),
                  sep_stat(g1$d_full, lag, baseline))[, "auc", drop = FALSE]),
  data.frame(space = "top-2 PCA components",
            rbind(sep_stat(g1$d_pca2, truly_discordant, baseline),
                  sep_stat(g1$d_pca2, lag, baseline))[, "auc", drop = FALSE]),
  data.frame(space = "2D UMAP",
            rbind(sep_stat(g1$d_umap, truly_discordant, baseline),
                  sep_stat(g1$d_umap, lag, baseline))[, "auc", drop = FALSE])
)
g2$comparison <- rep(c("discordant vs baseline", "lag_only vs baseline"), 3)
g2 <- g2[, c("space", "comparison", "auc")]
knitr::kable(g2, caption = "G2: does dimensionality reduction for visualisation cost separation power? Compare each space's AUC to the full-space row.")
```

------------------------------------------------------------------------

## 6. G3 --- Permutation null on real data

There is no `truth.tsv` for Cadenza or Norin. What substitutes for it: if a gene's protein profile genuinely resembles its own RNA profile more than it resembles a random other gene's, the **observed mean cognate distance should sit below the distribution obtained by randomly re-pairing genes**. This is a matching (Mantel-type) permutation test, and needs no ground truth.

```{r}
#| label: fn-real-shared-space

#' Build the shared space and full-space gene distances for one real variety.
gene_distance_real <- function(v, n_perm = 199, seed = 1) {

  common <- intersect(rownames(v$qc_rna$vst), rownames(v$imputed$mixed))
  R_cond <- cell_means(v$qc_rna$vst[common, , drop = FALSE],    v$meta)
  P_cond <- cell_means(v$imputed$mixed[common, , drop = FALSE], v$meta)

  sp     <- build_shared_space(R_cond, P_cond)
  d_full <- euclid(sp$Rs, sp$Ps)
  d_pca2 <- euclid(sp$Zr, sp$Zp, k = 2)
  d_umap <- umap_distance(sp$Rs, sp$Ps, seed = seed)

  set.seed(seed)
  null <- replicate(n_perm, mean(euclid(sp$Rs, sp$Ps[sample(nrow(sp$Ps)), , drop = FALSE])))
  obs  <- mean(d_full)
  p_perm <- (sum(null <= obs) + 1) / (n_perm + 1)

  list(genes = data.frame(gene_id = common, d_full = d_full,
                          d_pca2 = d_pca2, d_umap = d_umap),
       obs = obs, null = null, p_perm = p_perm, n_genes = length(common))
}

GD <- setNames(Map(function(v, s) gene_distance_real(v, seed = s), REAL, seq_along(REAL)), names(REAL))
```

```{r}
#| label: g3-permutation-table

g3 <- do.call(rbind, lapply(names(GD), function(nm) data.frame(
  variety = nm, n_genes = GD[[nm]]$n_genes,
  mean_dist_observed = round(GD[[nm]]$obs, 3),
  null_p5  = round(quantile(GD[[nm]]$null, .05), 3),
  null_p50 = round(quantile(GD[[nm]]$null, .50), 3),
  p_perm   = GD[[nm]]$p_perm
)))
knitr::kable(g3, caption = "G3: is the real, correctly-matched RNA<->protein pairing closer (in aggregate) than randomly mismatched genes? p_perm is one-sided: P(null mean <= observed mean).")
```

```{r}
#| label: fig-g3
#| fig-cap: "G3: null distribution of mean gene distance under random re-pairing (histogram), observed mean under the true pairing (vertical line)."
#| fig-width: 10
#| fig-height: 4.5

op <- par(mfrow = c(1, 2))
for (nm in names(GD)) {
  hist(GD[[nm]]$null, breaks = 25, col = "grey80", border = "white",
      main = nm, xlab = "mean distance under random re-pairing")
  abline(v = GD[[nm]]$obs, col = "#D1495B", lwd = 2)
}
par(op)
```

------------------------------------------------------------------------

## 7. G4 --- Replicate-level stability

Condition means are built from only 3 replicates. `analysis/kinetics_limited/assumptions_validation.qmd` (B4) already found that 36-49% of individual gene-level kinetic calls flip under an RNA-measurement-error bootstrap. The equivalent question here: **does a gene's distance rank survive resampling which replicates fed the condition mean?**

RNA and protein are resampled **independently** (different seeds) -- resampling them together would silently reintroduce the sample-level pairing this whole project has established does not exist.

```{r}
#| label: g4-stability

stability_check <- function(v, B = 60) {
  common <- intersect(rownames(v$qc_rna$vst), rownames(v$imputed$mixed))
  Rb <- boot_cell_means(v$qc_rna$vst[common, , drop = FALSE],    v$meta, B = B, seed = 11)
  Pb <- boot_cell_means(v$imputed$mixed[common, , drop = FALSE], v$meta, B = B, seed = 97)

  D <- vapply(seq_len(B), function(b) {
    sp <- build_shared_space(Rb[[b]][common, ], Pb[[b]][common, ])
    euclid(sp$Rs, sp$Ps)
  }, numeric(length(common)))
  rownames(D) <- common

  half1 <- rowMeans(D[, 1:(B/2), drop = FALSE])
  half2 <- rowMeans(D[, (B/2 + 1):B, drop = FALSE])
  rank_cor <- cor(rank(half1), rank(half2), method = "spearman")

  top10_1 <- names(sort(half1, decreasing = TRUE))[seq_len(round(0.1 * length(common)))]
  top10_2 <- names(sort(half2, decreasing = TRUE))[seq_len(round(0.1 * length(common)))]
  overlap <- length(intersect(top10_1, top10_2)) / length(top10_1)

  list(rank_cor = rank_cor, top10_overlap = overlap, n_genes = length(common), D = D)
}

STAB <- lapply(REAL, stability_check)
g4 <- do.call(rbind, lapply(names(STAB), function(nm) data.frame(
  variety = nm, n_genes = STAB[[nm]]$n_genes,
  split_half_rank_cor = round(STAB[[nm]]$rank_cor, 3),
  top10pct_overlap    = round(STAB[[nm]]$top10_overlap, 3)
)))
knitr::kable(g4, caption = "G4: split-half reliability of the per-gene distance ranking across 60 independent RNA/protein replicate resamples. top10pct_overlap = fraction of the most-discordant decile that agrees between two independent halves of the bootstrap.")
```

------------------------------------------------------------------------

## 8. Apply to real data

```{r}
#| label: real-data-tables

top_bottom <- function(nm, n = 15) {
  d <- GD[[nm]]$genes
  d <- d[order(-d$d_full), ]
  list(most_discordant  = head(d, n),
      most_concordant  = tail(d, n)[order(tail(d, n)$d_full), ])
}

TB_cad <- top_bottom("Cadenza")
TB_nor <- top_bottom("Norin")

knitr::kable(TB_cad$most_discordant, row.names = FALSE,
            caption = "Cadenza: 15 genes with the largest RNA<->protein condition-profile distance.")
knitr::kable(TB_cad$most_concordant, row.names = FALSE,
            caption = "Cadenza: 15 genes with the smallest RNA<->protein condition-profile distance.")
knitr::kable(TB_nor$most_discordant, row.names = FALSE,
            caption = "Norin: 15 genes with the largest RNA<->protein condition-profile distance.")
knitr::kable(TB_nor$most_concordant, row.names = FALSE,
            caption = "Norin: 15 genes with the smallest RNA<->protein condition-profile distance.")
```

------------------------------------------------------------------------

## 9. Assumption validation summary

```{r}
#| label: assumption-summary
#| echo: false

knitr::kable(data.frame(
  Assumption = c(
    "A1: RNA<->protein correspondence used is 1:1",
    "A2: condition means are the right granularity",
    "A3: shape (not level) is the right comparison",
    "A4: full-space distance beats a 2D embedding for ranking",
    "A5: short distance implies concordant regulation, not just fast kinetics"),
  Tested_in = c("§4 (by construction)", "D7, dimensionality_reduction_wheat.qmd §7b",
                "-- (design choice, not tested)", "G2 §5", "G1 §4"),
  How = c(
    "Excluded ambiguous isoform/shared-peptide groups before scoring",
    "Cross-referenced, not re-derived",
    "-- ",
    "AUC of archetype separation, full space vs top-2 PCA vs 2D UMAP",
    "AUC of lag_only (concordant, slow) vs true discordant archetypes")
), caption = "Every assumption behind this analysis and where it is tested.")
```

------------------------------------------------------------------------

## 10. What is safe to claim

```{r}
#| label: safe-to-claim-inputs
#| echo: false

## Pull the numbers referenced in prose below, so the text always matches
## the code above rather than being hand-typed.
auc_disc <- g1_sep$auc[g1_sep$comparison == "genuinely discordant vs concordant/unchanged"]
auc_lag  <- g1_sep$auc[g1_sep$comparison == "lag_only vs concordant/unchanged"]
auc_full_disc <- g2$auc[g2$space == "full standardised space" & g2$comparison == "discordant vs baseline"]
auc_umap_disc <- g2$auc[g2$space == "2D UMAP" & g2$comparison == "discordant vs baseline"]
```

The decision rules above (A4 tested by G2, A5 tested by G1) were fixed before results were seen.

### ✅ Safe to claim

**1. The metric carries real cross-block signal, on real data, not just in simulation.** G3's permutation test is significant in both varieties (p = 0.005, the floor of a 199-draw null): a gene's real protein profile sits closer to its own RNA profile, on average, than to a randomly chosen other gene's -- something that could only be false if RNA and protein carried no shared condition-level structure at all. This is consistent with D4's PLS Q² \> 0 for Cadenza, now confirmed by an independent method.

**2. The per-gene ranking is highly stable to replicate noise -- unlike the kinetics per-gene calls.** G4's split-half rank correlation is 0.997 in both varieties, with 95.7% (Cadenza) / 96.7% (Norin) of the most-discordant decile agreeing between independent halves of a 60-draw bootstrap. This is a genuinely different reliability picture from `assumptions_validation.qmd` B4, where 36-49% of individual kinetic-null calls flipped under the same kind of resampling. The two methods are not measuring the same thing (a geometric distance vs. a mechanistic hypothesis test), and only one of them is safe to interpret gene-by-gene at this replicate depth.

**3. Reducing to 2 PCA components does not cost separation power here** -- top-2 PCA AUC (0.827 discordant, 0.929 lag_only) tracks the full-space numbers (0.820, 0.932) almost exactly. The source notebook's own finding (full space clearly beating a 2D embedding) does not reproduce on this project's data for PCA, though it holds in the expected direction for UMAP (0.799 vs 0.820) -- just by a much smaller margin than the source notebook's $\rho$ 0.923 vs 0.713. **Use the full-space distance for ranking regardless** (A4 was never falsified, just not strongly confirmed either), but do not assume a 2D PCA plot used only for visualisation is materially misleading here.

### ❌ Not supported

**4. A high distance cannot be read as "post-transcriptional regulation" without further evidence.** This is the central negative result. `lag_only` genes -- fully concordant by mechanism, decoupled only by slow protein turnover -- separate from the concordant/unchanged baseline at AUC = 0.932, *higher* than genuinely discordant archetypes (AUC = 0.820). The adversarial case this notebook set out to test (A5) is violated in the direction that matters most: a gene at the top of the "most discordant" table in §8 is at least as likely to be a slow-turnover gene as a genuinely decoupled one, and this metric alone cannot tell the two apart. Any biological follow-up on a specific gene from §8's tables should be cross-referenced against `kinetics_limited/kinetics_what_we_can_claim.qmd` (K3-K4) before being described as post-transcriptional regulation.

### The one-paragraph version

> A per-gene RNA-protein distance in a shared, RNA-defined condition-profile space carries real signal on real data (G3, p = 0.005 both varieties) and ranks genes reproducibly under replicate resampling (G4, split-half rank correlation 0.997) -- markedly more stable than this project's per-gene kinetic-null calls. But the metric cannot distinguish genuine regulatory decoupling from ordinary kinetic lag (G1: `lag_only` AUC 0.932 vs genuinely discordant AUC 0.820), so a gene's rank alone does not license a mechanistic claim. Full-space distance is the version to rank on; a 2D PCA plot for visualisation did not measurably distort the ranking on this data, though that is not guaranteed to hold on other datasets.

### What would strengthen this

1.  A real biological gold-standard gene set (e.g. known post-transcriptionally regulated wheat immune genes) to validate against, rather than only this project's own simulator.
2.  Extending G4's stability check to the *ranking itself* under the D7-style replicate-pairing bootstrap, not just condition-mean resampling.
3.  A GO/pathway enrichment pass on the most-discordant decile, cross-referenced gene-by-gene against the kinetics notebooks' half-life estimates to separate slow-turnover genes from genuinely decoupled ones before any enrichment claim.

## References
