---
title: "Dimensionality reduction for unpaired RNA/protein - PCA, PLS, autoencoders"
author: "Kristina Gagalova"
format:
  gfm:
    toc: true
    code-tools: true
    fig-dir: "figs"
execute:
  echo: true
editor: visual
bibliography: ../kinetics_limited/references.bib
---

## Purpose and scope

This notebook answers one question systematically:

> **Given 8 experimental conditions and unpaired RNA/protein layers, which dimensionality-reduction method is actually justified --- and does any of them beat simply predicting the mean?**

The candidates are PCA, PLS, and three autoencoder variants (linear, non-linear, variational), plus a condition-aligned autoencoder matching the architecture in the project's existing Python work.

### Two corrections to the framing, stated up front

**1. No VAE has previously been built in this repository.** PCA and PLS are used throughout (`R/06_integration_caseB.R`, `R/01_qc_normalise.R`, the kinetics notebooks). A VAE was *proposed* in `docs/planning.md` §4C but never implemented here. The working autoencoder for this project is `ConditionAlignedAE` in `rnaprot/unpaired.py` on `nectar` (GitHub: `KristinaGagalova/autoencoders-test`). Everything autoencoder-related below is **new work written for this notebook**.

**2. Implemented from scratch in base R.** No `torch`/`keras` is available in this project's R environment. Rather than add a \~2 GB dependency, all models here are written directly with explicit gradients and Adam. That is slower to write but has a real advantage for this question: every model is auditable, and the comparison is not confounded by framework defaults.

### Relationship to the existing Python work

`rnaprot/unpaired.py` on `nectar` implements the same problem with:

| Component              | Python (`unpaired.py`)                                                     | Here                                                |
|:-----------------------|:---------------------------------------------------------------------------|:----------------------------------------------------|
| Deep model             | `ConditionAlignedAE`, MLP `[p, 64, 16, 6]`, LayerNorm + SiLU + dropout 0.1 | Linear condition-aligned AE (see §6 for why linear) |
| Loss                   | `0.5·L_rna + 0.5·L_prot + 2.0·L_cross + 0.5·L_align`                       | Same four terms, same default weights               |
| CV                     | `leave_condition_out`                                                      | Leave-one-condition-out                             |
| Baselines              | mean, design ridge, PCA+ridge, PLS, cognate ridge                          | mean, design ridge, PCA+ridge, PLS                  |
| Component selection    | `_loo_select_components` (inner LOO)                                       | Inner LOO, same idea                                |
| Design residualisation | `_design_residual_predict`                                                 | §7                                                  |

The **design-residualisation** step deserves particular note: `unpaired.py` already residualises both modalities against treatment/timepoint before modelling, which is the same control I arrived at independently in `analysis/kinetics_limited/kinetics_what_we_can_claim.qmd` (K1). Two independent implementations converging on that control is reassuring --- it is the right control for this design.

**A second, revised module exists: `rnaprot/unpaired_reviewer.py`.** It drives `rna_protein_unpaired_per_variety.ipynb` and mostly re-runs the same comparison above, but with three real differences, not just cosmetic ones: loss weights `0.5/0.5/3.0/0.75` rather than `0.5/0.5/2.0/0.5`; RNA/protein **feature selection redone inside each outer CV fold** rather than once globally; and a **permutation null (`N_PERM=200`) applied to the PCA/PLS/design baselines**, though not to the autoencoder. The first is a hyperparameter difference, not addressed further here. The second and third are genuine methodological gaps in D4/D7 above, closed in **D8 (§7c)** and **D7 (§7b)** respectively.

------------------------------------------------------------------------

## Analysis outline

| Step | What                                        | Why                                                              |
|:-----|:--------------------------------------------|:-----------------------------------------------------------------|
| §1   | Setup and data                              | Shared upstream with the other notebooks                         |
| §2   | Methods and their assumptions               | Each with references                                             |
| §3   | Implementation                              | From-scratch, auditable                                          |
| §4   | **D1** How many components are real?        | Parallel analysis vs permutation                                 |
| §5   | **D2** Do components track the design?      | Structure must be interpretable                                  |
| §6   | **D3** Linear AE ≡ PCA?                     | Baldi--Hornik verification, a correctness check                  |
| §7   | **D4** Does any method beat the mean?       | The headline comparison, leave-one-condition-out                 |
| §7b  | **D7** Condition means vs replicate pairing | Aggregated (n=8) vs sample-level (n=24), with a permutation null |
| §7c  | **D8** Does global feature selection leak?  | Global vs in-fold RNA feature selection                          |
| §8   | **D5** Are we sample-limited?               | Learning curve                                                   |
| §9   | **D6** Are latent spaces stable?            | Seed-to-seed reproducibility                                     |
| §10  | Assumption validation summary               |                                                                  |
| §11  | What is safe to claim                       |                                                                  |

------------------------------------------------------------------------

## 1. Setup

```{r}
#| label: setup
#| message: false

if (!requireNamespace("here", quietly = TRUE)) stop("package 'here' required")

## RENV_PROJECT must be set before sourcing activate.R, or renv treats this
## subdirectory as the project and bootstraps an empty library.
if (file.exists(here::here("renv", "activate.R"))) {
  Sys.setenv(RENV_PROJECT = here::here())
  source(here::here("renv", "activate.R"))
}

source(here::here("R", "utils.R"))
source(here::here("R", "wheat_pipeline.R"))
source(here::here("R", "pls_utils.R"))
need_pkgs(c("limma", "edgeR", "DESeq2", "matrixStats", "impute"))

set.seed(20260823)
DESIGN <- wheat_design()

cad <- prepare_variety("cadenza", DESIGN)
nor <- prepare_variety("norin",   DESIGN)
VARIETIES <- list(Cadenza = cad, Norin = nor)
```

```{r}
#| label: data-blocks

#' Build the analysis blocks for one variety.
#'
#' Two representations are produced, and the distinction matters:
#'
#'   SAMPLE level (24 x p) -- used for within-layer methods (PCA, plain
#'     autoencoders). Each layer has its own 24 samples.
#'
#'   CONDITION level (8 x p) -- design-cell means, used for anything that
#'     crosses the two layers. Under the unpaired design the condition is the
#'     ONLY shared axis: there is no sample-level correspondence between an
#'     RNA plant and a protein plant.
make_blocks <- function(v, n_rna = 2000, n_prot = 1000) {

  hv_r <- top_variable(v$qc_rna$vst,     n_rna)
  hv_p <- top_variable(v$imputed$mixed,  n_prot)

  R_samp <- t(v$qc_rna$vst[hv_r, , drop = FALSE])
  P_samp <- t(v$imputed$mixed[hv_p, , drop = FALSE])

  R_cond <- t(cell_means(v$qc_rna$vst[hv_r, , drop = FALSE],    v$meta))
  P_cond <- t(cell_means(v$imputed$mixed[hv_p, , drop = FALSE], v$meta))

  cells <- rownames(R_cond)
  cond  <- data.frame(
    cell      = cells,
    treatment = factor(sub("_t.*$", "", cells)),
    time_f    = factor(as.numeric(sub("^.*_t", "", cells)))
  )

  # centre and scale each block to unit total variance, so no block dominates
  cs <- function(M) {
    M <- scale(M, center = TRUE, scale = FALSE)
    M / sqrt(sum(M^2) / nrow(M))
  }

  list(R_samp = cs(R_samp), P_samp = cs(P_samp),
       R_cond = cs(R_cond), P_cond = cs(P_cond),
       cond = cond, meta = v$meta)
}

BLOCKS <- lapply(VARIETIES, make_blocks)

knitr::kable(
  data.frame(
    variety    = names(BLOCKS),
    rna_samples    = sapply(BLOCKS, function(b) nrow(b$R_samp)),
    rna_features   = sapply(BLOCKS, function(b) ncol(b$R_samp)),
    prot_samples   = sapply(BLOCKS, function(b) nrow(b$P_samp)),
    prot_features  = sapply(BLOCKS, function(b) ncol(b$P_samp)),
    conditions     = sapply(BLOCKS, function(b) nrow(b$R_cond))
  ),
  caption = "Analysis blocks. Note the shape: thousands of features, 24 samples, 8 conditions."
)
```

> **The shape of this problem, stated plainly.** Thousands of features against 24 samples (8 conditions). Every method below is operating in the regime where the number of parameters vastly exceeds the number of observations. That is not a reason to avoid dimensionality reduction --- it is the reason to *use* it --- but it does mean out-of-sample validation is the only meaningful evidence, and in-sample fit is worthless.

------------------------------------------------------------------------

## 2. The methods and what each assumes

### PCA

Finds orthogonal directions of maximum variance. Unsupervised: it never sees the protein block or the design.

**Assumes:** structure is linear; variance is a proxy for signal; components are orthogonal; the leading directions are the interesting ones. The last two are conventions, not facts about biology --- a low-variance direction can carry the biology, and biological programmes are not required to be orthogonal [@jolliffe2016].

### PLS

Finds directions in RNA that maximally covary with protein. Two-block and supervised by the second block.

**Assumes:** linearity again; that cross-block covariance is the target; and critically, that the model is validated **out of sample**. At 8 conditions and thousands of features a PLS component can be steered onto almost any target --- in-sample cross-block correlation reaches r ≈ 0.72 on *permuted* data in this dataset. Cross-validated Q² against a permutation null is the accepted remedy [@westerhuis2008].

### Autoencoders (linear, non-linear, variational)

An encoder compresses to a latent space, a decoder reconstructs. The variational form adds a probabilistic latent with a KL penalty toward a standard normal prior [@kingma2014].

**Assumes:** enough samples to fit the encoder/decoder weights. This is the binding constraint here, and it is worth being precise about why.

A **linear** autoencoder is not an alternative to PCA --- it recovers the same subspace. @baldi1989 proved the squared-error loss of a linear autoencoder has no local minima and that its global optimum spans the principal subspace. §6 verifies this numerically as a correctness check on the implementation.

So a VAE can only add value through **non-linearity** and the **probabilistic prior**. Both cost parameters, and parameters need samples. Successful omics VAEs operate on thousands of observations --- single-cell datasets, where each cell is a sample --- or transfer from large reference cohorts. Bulk designs with tens of samples are a different regime, and overfitting is the documented failure mode.

### Condition-aligned autoencoder

The architecture in `rnaprot/unpaired.py`: separate encoders per modality, with the latent spaces tied by matching **condition centroids** rather than by pairing samples. This is the correct structural response to unpaired data --- it never assumes RNA sample *i* corresponds to protein sample *i*.

**Assumes:** all of the above, plus that condition centroids estimated from 3 replicates are stable enough to align on.

------------------------------------------------------------------------

## 3. Implementation

```{r}
#| label: fn-adam

#' Adam optimiser over a named list of parameter matrices.
#'
#' Written out rather than pulled from a framework so the update rule is
#' visible and the comparison between models is not confounded by differing
#' framework defaults.
adam_init <- function(par) {
  list(m = lapply(par, function(p) array(0, dim(p))),
       v = lapply(par, function(p) array(0, dim(p))),
       t = 0)
}

adam_update <- function(par, grad, state, lr = 0.01,
                        b1 = 0.9, b2 = 0.999, eps = 1e-8) {
  state$t <- state$t + 1
  for (nm in names(par)) {
    state$m[[nm]] <- b1 * state$m[[nm]] + (1 - b1) * grad[[nm]]
    state$v[[nm]] <- b2 * state$v[[nm]] + (1 - b2) * grad[[nm]]^2
    mhat <- state$m[[nm]] / (1 - b1^state$t)
    vhat <- state$v[[nm]] / (1 - b2^state$t)
    par[[nm]] <- par[[nm]] - lr * mhat / (sqrt(vhat) + eps)
  }
  list(par = par, state = state)
}
```

```{r}
#| label: fn-autoencoder

#' Single-block autoencoder: linear, tanh, or variational.
#'
#'   type = "linear" : z = XW1 + b1                 (spans the PCA subspace)
#'   type = "tanh"   : z = tanh(XW1 + b1)           (non-linear encoder)
#'   type = "vae"    : mu, logvar heads; z = mu + sd*eps; + KL penalty
#'
#' Decoder is linear throughout, so the ONLY difference between "linear" and
#' "tanh" is the encoder non-linearity -- which isolates the question "does
#' non-linearity buy anything?" from every other design choice.
ae_fit <- function(X, k = 3, type = c("linear", "tanh", "vae"),
                   epochs = 4000, lr = 0.01, beta = 1, seed = 1) {

  type <- match.arg(type)
  set.seed(seed)
  n <- nrow(X); p <- ncol(X)

  init <- function(a, b) matrix(rnorm(a * b, 0, 1 / sqrt(a)), a, b)
  par <- list(W1 = init(p, k), b1 = matrix(0, 1, k),
              W2 = init(k, p), b2 = matrix(0, 1, p))
  if (type == "vae") {
    par$Wv <- init(p, k); par$bv <- matrix(0, 1, k)
  }

  st <- adam_init(par)
  rep_row <- function(b, n) matrix(rep(b, each = n), n)
  hist <- numeric(epochs)

  for (e in seq_len(epochs)) {
    a1 <- X %*% par$W1 + rep_row(par$b1, n)

    if (type == "vae") {
      lv  <- X %*% par$Wv + rep_row(par$bv, n)
      lv  <- pmax(pmin(lv, 10), -10)          # numerical guard
      sdv <- exp(0.5 * lv)
      eps <- matrix(rnorm(n * k), n, k)
      z   <- a1 + sdv * eps
    } else {
      z <- if (type == "tanh") tanh(a1) else a1
    }

    Xh <- z %*% par$W2 + rep_row(par$b2, n)
    E  <- Xh - X
    recon <- sum(E^2) / (n * p)

    dXh <- 2 * E / (n * p)
    g <- list(W2 = t(z) %*% dXh, b2 = matrix(colSums(dXh), 1))
    dz <- dXh %*% t(par$W2)

    if (type == "vae") {
      kl <- -0.5 * sum(1 + lv - a1^2 - exp(lv)) / (n * p)
      # dKL/dmu = mu, dKL/dlogvar = -0.5(1 - exp(logvar)), scaled to match
      dmu <- dz + beta * a1 / (n * p)
      dlv <- dz * (0.5 * sdv * eps) + beta * (-0.5) * (1 - exp(lv)) / (n * p)
      g$W1 <- t(X) %*% dmu; g$b1 <- matrix(colSums(dmu), 1)
      g$Wv <- t(X) %*% dlv; g$bv <- matrix(colSums(dlv), 1)
      hist[e] <- recon + beta * kl
    } else {
      da1 <- if (type == "tanh") dz * (1 - z^2) else dz
      g$W1 <- t(X) %*% da1; g$b1 <- matrix(colSums(da1), 1)
      hist[e] <- recon
    }

    up <- adam_update(par, g, st, lr = lr); par <- up$par; st <- up$state
  }

  encode <- function(Xn) {
    a <- Xn %*% par$W1 + rep_row(par$b1, nrow(Xn))
    if (type == "tanh") tanh(a) else a       # VAE uses the mean at test time
  }

  list(par = par, type = type, k = k, loss = hist,
       encode = encode,
       reconstruct = function(Xn)
         encode(Xn) %*% par$W2 + rep_row(par$b2, nrow(Xn)))
}
```

```{r}
#| label: fn-condition-ae

#' Condition-aligned autoencoder (linear), after `rnaprot/unpaired.py`.
#'
#' Two encoders, two decoders, tied by CONDITION CENTROIDS -- never by sample
#' pairing, which does not exist in this design. The four loss terms and their
#' default weights match the Python implementation:
#'
#'   w1 * MSE(dec_r(z_r), R)                     RNA reconstruction
#'   w2 * MSE(dec_p(z_p), P)                     protein reconstruction
#'   w3 * MSE(centroid_c dec_p(z_r), centroid_c P)   cross-modal, per condition
#'   w4 * MSE(centroid_c z_r, centroid_c z_p)    latent centroid alignment
#'
#' WHY LINEAR HERE. The Python version uses MLP encoders (64, 16, SiLU,
#' dropout). This implementation is linear, for a reason that is itself part
#' of the finding: with 8 conditions, a linear map already has far more
#' parameters than observations. If the linear version cannot beat the mean
#' baseline out of sample, a deeper one with strictly more parameters and the
#' same data will not either. §7 tests exactly that.
cond_ae_fit <- function(R, P, cond_r, cond_p, k = 3,
                        w = c(0.5, 0.5, 2.0, 0.5),
                        epochs = 3000, lr = 0.01, seed = 1) {

  set.seed(seed)
  nr <- nrow(R); pr <- ncol(R)
  np <- nrow(P); pp <- ncol(P)

  init <- function(a, b) matrix(rnorm(a * b, 0, 1 / sqrt(a)), a, b)
  par <- list(Wr = init(pr, k), Ar = init(k, pr),
              Wp = init(pp, k), Ap = init(k, pp))

  shared <- intersect(unique(cond_r), unique(cond_p))
  idx_r <- lapply(shared, function(cc) which(cond_r == cc))
  idx_p <- lapply(shared, function(cc) which(cond_p == cc))
  nc <- length(shared)

  st <- adam_init(par); hist <- numeric(epochs)

  for (e in seq_len(epochs)) {
    z_r <- R %*% par$Wr; z_p <- P %*% par$Wp
    Rh  <- z_r %*% par$Ar; Ph <- z_p %*% par$Ap

    Er <- Rh - R; Ep <- Ph - P
    L1 <- sum(Er^2) / (nr * pr); L2 <- sum(Ep^2) / (np * pp)

    dRh <- w[1] * 2 * Er / (nr * pr)
    dPh <- w[2] * 2 * Ep / (np * pp)

    g <- list(Ar = t(z_r) %*% dRh, Ap = t(z_p) %*% dPh)
    dz_r <- dRh %*% t(par$Ar)
    dz_p <- dPh %*% t(par$Ap)

    L3 <- 0; L4 <- 0
    for (i in seq_len(nc)) {
      ri <- idx_r[[i]]; pi <- idx_p[[i]]

      mzr  <- colMeans(z_r[ri, , drop = FALSE])
      mzp  <- colMeans(z_p[pi, , drop = FALSE])
      pred <- matrix(mzr, 1) %*% par$Ap
      obs  <- colMeans(P[pi, , drop = FALSE])

      e3 <- pred - matrix(obs, 1); L3 <- L3 + sum(e3^2) / pp
      e4 <- matrix(mzr - mzp, 1);  L4 <- L4 + sum(e4^2) / k

      c3 <- w[3] * 2 / (nc * pp)
      g$Ap <- g$Ap + c3 * matrix(mzr, ncol = 1) %*% e3
      d_mzr <- c3 * e3 %*% t(par$Ap)

      c4 <- w[4] * 2 / (nc * k)
      d_mzr <- d_mzr + c4 * e4
      d_mzp <- -c4 * e4

      dz_r[ri, ] <- dz_r[ri, ] + matrix(rep(d_mzr / length(ri), each = length(ri)),
                                        length(ri))
      dz_p[pi, ] <- dz_p[pi, ] + matrix(rep(d_mzp / length(pi), each = length(pi)),
                                        length(pi))
    }

    g$Wr <- t(R) %*% dz_r
    g$Wp <- t(P) %*% dz_p

    hist[e] <- w[1]*L1 + w[2]*L2 + w[3]*L3/nc + w[4]*L4/nc
    up <- adam_update(par, g, st, lr = lr); par <- up$par; st <- up$state
  }

  list(par = par, k = k, loss = hist,
       predict_protein = function(Rn) (Rn %*% par$Wr) %*% par$Ap)
}
```

```{r}
#| label: fn-predictors

#' Baseline and reference predictors, matching `unpaired.py`'s set.
ridge_fit <- function(X, Y, alpha = 10) {
  X1 <- cbind(1, X)
  A  <- crossprod(X1) + alpha * diag(ncol(X1)); A[1, 1] <- A[1, 1] - alpha
  qr.solve(A, crossprod(X1, Y))
}
ridge_pred <- function(B, Xn) cbind(1, Xn) %*% B

pred_mean <- function(Xtr, Ytr, Xte, ...)
  matrix(colMeans(Ytr), nrow(Xte), ncol(Ytr), byrow = TRUE)

pred_design <- function(Xtr, Ytr, Xte, Dtr, Dte, alpha = 5, ...)
  ridge_pred(ridge_fit(Dtr, Ytr, alpha), Dte)

pred_pca_ridge <- function(Xtr, Ytr, Xte, k = 2, alpha = 10, ...) {
  k <- max(1, min(k, nrow(Xtr) - 1, ncol(Xtr)))
  mu <- colMeans(Xtr)
  s  <- svd(sweep(Xtr, 2, mu), nu = k, nv = k)
  Ztr <- s$u[, seq_len(k), drop = FALSE] %*% diag(s$d[seq_len(k)], k, k)
  Zte <- sweep(Xte, 2, mu) %*% s$v[, seq_len(k), drop = FALSE]
  ridge_pred(ridge_fit(Ztr, Ytr, alpha), Zte)
}

pred_pls <- function(Xtr, Ytr, Xte, k = 2, ...) {
  k  <- max(1, min(k, nrow(Xtr) - 2))
  mx <- colMeans(Xtr); my <- colMeans(Ytr)
  f  <- pls2(sweep(Xtr, 2, mx), sweep(Ytr, 2, my), ncomp = k)
  sweep(pls_predict(f, sweep(Xte, 2, mx), k), 2, my, "+")
}

pred_cond_ae <- function(Xtr, Ytr, Xte, cond_tr, k = 3, epochs = 600, seed = 1, ...) {
  m <- cond_ae_fit(Xtr, Ytr, cond_tr, cond_tr, k = k, epochs = epochs, seed = seed)
  m$predict_protein(Xte)
}
```

------------------------------------------------------------------------

## 4. D1 --- How many components are real?

**The assumption being tested:** that the leading components describe structure rather than noise.

**The test.** Horn's parallel analysis: permute each feature independently to destroy between-feature correlation while preserving each feature's marginal distribution, recompute the eigenvalue spectrum, and keep only components whose observed eigenvalue exceeds the permuted 95th percentile.

```{r}
#| label: d1-parallel-analysis

parallel_analysis <- function(X, n_perm = 100, seed = 1) {
  set.seed(seed)
  ev  <- function(M) { s <- svd(scale(M, TRUE, FALSE)); s$d^2 / sum(s$d^2) }
  obs <- ev(X)
  perm <- replicate(n_perm, ev(apply(X, 2, sample)))
  data.frame(
    component = seq_along(obs),
    observed  = round(100 * obs, 2),
    perm_q95  = round(100 * apply(perm, 1, quantile, 0.95), 2),
    real      = obs > apply(perm, 1, quantile, 0.95)
  )
}

d1 <- do.call(rbind, lapply(names(BLOCKS), function(nm) {
  b <- BLOCKS[[nm]]
  rbind(
    cbind(variety = nm, layer = "RNA (24 samples)",     parallel_analysis(b$R_samp)),
    cbind(variety = nm, layer = "protein (24 samples)", parallel_analysis(b$P_samp))
  )
}))

knitr::kable(subset(d1, component <= 5),
             caption = "D1: observed vs permuted variance explained. `real = TRUE` means the component exceeds the permutation 95th percentile.")
```

```{r}
#| label: fig-d1
#| fig-cap: "Scree plots against a permutation null. Only components above the dashed line carry structure beyond feature-wise noise."
#| fig-width: 10
#| fig-height: 8

op <- par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 1))
for (nm in names(BLOCKS)) {
  for (ly in c("RNA (24 samples)", "protein (24 samples)")) {
    d <- subset(d1, variety == nm & layer == ly & component <= 10)
    plot(d$component, d$observed, type = "b", pch = 16, lwd = 2, col = "#2C6FBB",
         ylim = c(0, max(d$observed) * 1.1),
         xlab = "component", ylab = "% variance explained",
         main = sprintf("%s -- %s", nm, ly))
    lines(d$component, d$perm_q95, type = "b", pch = 4, lty = 2, col = "#D1495B")
    legend("topright", bty = "n", cex = 0.75, lwd = 2, pch = c(16, 4),
           lty = c(1, 2), col = c("#2C6FBB", "#D1495B"),
           legend = c("observed", "permutation 95%"))
  }
}
par(op)
```

**How to read this.** The number of `real = TRUE` components is the honest ceiling on latent dimension. Fitting an autoencoder with a latent dimension larger than this is fitting noise, whatever the reconstruction error says.

------------------------------------------------------------------------

## 5. D2 --- Do the components track the design?

**The assumption being tested:** that PCA's leading directions correspond to the experimental factors rather than to batch, handling, or an artefact.

```{r}
#| label: d2-design-association

design_association <- function(b, label) {
  s <- svd(scale(b$R_samp, TRUE, FALSE))
  Z <- s$u[, 1:4] %*% diag(s$d[1:4])
  meta <- b$meta

  do.call(rbind, lapply(1:4, function(j) {
    z <- Z[, j]
    data.frame(
      variety = label, component = j,
      var_pct = round(100 * s$d[j]^2 / sum(s$d^2), 1),
      # eta-squared: fraction of the component explained by each factor
      eta2_treatment = round(summary(aov(z ~ meta$treatment))[[1]][1, 2] /
                               sum(summary(aov(z ~ meta$treatment))[[1]][, 2]), 3),
      eta2_time      = round(summary(aov(z ~ factor(meta$time_num)))[[1]][1, 2] /
                               sum(summary(aov(z ~ factor(meta$time_num)))[[1]][, 2]), 3)
    )
  }))
}

d2 <- do.call(rbind, Map(design_association, BLOCKS, names(BLOCKS)))
knitr::kable(d2, caption = "D2: fraction of each RNA principal component explained by treatment and by timepoint (eta-squared).")
```

```{r}
#| label: fig-d2
#| fig-cap: "RNA PCA coloured by treatment and timepoint. If the leading components are biology, the design should be visible."
#| fig-width: 10
#| fig-height: 5

op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))
for (nm in names(BLOCKS)) {
  b <- BLOCKS[[nm]]
  s <- svd(scale(b$R_samp, TRUE, FALSE))
  Z <- s$u[, 1:2] %*% diag(s$d[1:2])
  ve <- round(100 * s$d[1:2]^2 / sum(s$d^2), 1)
  plot(Z[, 1], Z[, 2],
       col = c("#2C6FBB", "#D1495B")[as.integer(b$meta$treatment)],
       pch = c(15, 16, 17, 18)[as.integer(factor(b$meta$time_num))],
       cex = 1.8, xlab = sprintf("PC1 (%.1f%%)", ve[1]),
       ylab = sprintf("PC2 (%.1f%%)", ve[2]), main = sprintf("%s: RNA", nm))
  legend("topright", bty = "n", cex = 0.7,
         legend = c(levels(b$meta$treatment), paste0(DESIGN$timepoints, " h")),
         col = c("#2C6FBB", "#D1495B", rep("grey40", 4)),
         pch = c(16, 16, 15, 16, 17, 18))
}
par(op)
```

------------------------------------------------------------------------

## 6. D3 --- Is a linear autoencoder equivalent to PCA?

**The assumption being tested:** that the autoencoder implementation is correct, and that a linear autoencoder adds nothing over PCA.

@baldi1989 proved the squared-error loss of a linear autoencoder has no local minima and its global optimum spans the principal subspace. So a correct implementation must recover the PCA subspace. This is simultaneously a **correctness check on the code** and a **substantive result**: it establishes that any advantage a VAE has must come from non-linearity, not from being an autoencoder.

```{r}
#| label: d3-baldi-hornik

subspace_alignment <- function(X, k = 3, epochs = 5000) {

  s   <- svd(scale(X, TRUE, FALSE))
  Vpc <- s$v[, seq_len(k), drop = FALSE]
  Zpc <- s$u[, seq_len(k), drop = FALSE] %*% diag(s$d[seq_len(k)], k, k)

  ae <- ae_fit(X, k = k, type = "linear", epochs = epochs, lr = 0.02)

  # COMPARE LATENT SCORES, NOT ENCODER WEIGHTS.
  #
  # This distinction is not pedantry -- getting it wrong makes a correct
  # implementation look broken. With n = 24 samples and p = 2000 features the
  # data occupy a subspace of rank <= 23 out of 2000. X %*% W1 therefore
  # depends ONLY on the component of W1 lying inside that row space; any
  # component orthogonal to it changes the weights while leaving every
  # prediction untouched. The encoder weight subspace is consequently NOT
  # identified in the p >> n regime, and comparing col(W1) to col(Vpc)
  # measures mostly arbitrary directions (this was verified directly: an
  # autoencoder whose reconstruction MSE matched PCA's optimum to 5 decimal
  # places scored a weight-subspace cosine of 0.15).
  #
  # The latent SCORES (n x k) and the reconstruction ARE identified, so those
  # are what the theorem should be checked against.
  Zae <- ae$encode(scale(X, TRUE, FALSE))

  cos_scores <- svd(t(qr.Q(qr(Zpc))) %*% qr.Q(qr(Zae)))$d

  rec_ae  <- ae$reconstruct(X)
  rec_pca <- scale(X, TRUE, FALSE) %*% Vpc %*% t(Vpc) +
             matrix(colMeans(X), nrow(X), ncol(X), byrow = TRUE)

  list(cos_scores = round(cos_scores, 4),
       mean_cos   = round(mean(cos_scores), 4),
       cor_recon  = round(cor(as.vector(rec_ae), as.vector(rec_pca)), 4),
       recon_pca  = mean((X - rec_pca)^2),
       recon_ae   = mean((X - rec_ae)^2),
       loss = ae$loss)
}

d3 <- lapply(names(BLOCKS), function(nm) {
  r <- subspace_alignment(BLOCKS[[nm]]$R_samp, k = 3)
  data.frame(variety = nm,
             cos_score_1 = r$cos_scores[1], cos_score_2 = r$cos_scores[2],
             cos_score_3 = r$cos_scores[3], mean_cosine = r$mean_cos,
             cor_reconstruction = r$cor_recon,
             mse_pca = signif(r$recon_pca, 4),
             mse_linear_ae = signif(r$recon_ae, 4))
})
knitr::kable(do.call(rbind, d3),
             caption = "D3: agreement between the linear autoencoder and PCA, on IDENTIFIED quantities (latent scores and reconstruction). Values near 1 confirm Baldi-Hornik.")
```

```{r}
#| label: fig-d3
#| fig-cap: "Training loss of the linear autoencoder against the PCA reconstruction error it should converge to."
#| fig-width: 10
#| fig-height: 4.5

op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))
for (nm in names(BLOCKS)) {
  r <- subspace_alignment(BLOCKS[[nm]]$R_samp, k = 3)
  plot(r$loss, type = "l", lwd = 2, col = "#2C6FBB", log = "y",
       xlab = "epoch", ylab = "reconstruction MSE",
       main = sprintf("%s: linear AE converges to PCA", nm))
  abline(h = r$recon_pca, col = "#D1495B", lwd = 2, lty = 2)
  legend("topright", bty = "n", cex = 0.75, lwd = 2, lty = c(1, 2),
         col = c("#2C6FBB", "#D1495B"), legend = c("linear AE", "PCA optimum"))
}
par(op)
```

------------------------------------------------------------------------

## 7. D4 --- Does any method beat predicting the mean?

**This is the headline comparison.** Every method is scored by leave-one-condition-out cross-validation --- hold out one of the 8 treatment × timepoint cells entirely, fit on the remaining 7, predict the held-out cell's protein profile. This matches `leave_condition_out` in `rnaprot/unpaired.py`.

The reference point is the **mean baseline**: predict the training-set average protein profile, ignoring RNA completely. A method that cannot beat it is not extracting usable information.

The **design baseline** (ridge on treatment + timepoint indicators) is the second reference: a method that only matches it is recovering the experimental design, not cross-layer biology.

```{r}
#| label: d4-cv

#' Leave-one-condition-out CV across all methods.
loco_compare <- function(b, label, k = 2, seeds = 1:3) {

  X <- b$R_cond; Y <- b$P_cond
  cells <- b$cond$cell
  D <- model.matrix(~ treatment + time_f, b$cond)[, -1, drop = FALSE]
  n <- nrow(X)

  press <- list()
  add <- function(nm, v) press[[nm]] <<- c(press[[nm]], v)

  for (i in seq_len(n)) {
    tr <- setdiff(seq_len(n), i)
    Xtr <- X[tr, , drop = FALSE]; Ytr <- Y[tr, , drop = FALSE]
    Xte <- X[i, , drop = FALSE];  Yte <- Y[i, , drop = FALSE]
    Dtr <- D[tr, , drop = FALSE]; Dte <- D[i, , drop = FALSE]

    sq <- function(pred) sum((Yte - pred)^2)

    add("mean",       sq(pred_mean(Xtr, Ytr, Xte)))
    add("design",     sq(pred_design(Xtr, Ytr, Xte, Dtr, Dte)))
    add("PCA+ridge",  sq(pred_pca_ridge(Xtr, Ytr, Xte, k = k)))
    add("PLS",        sq(pred_pls(Xtr, Ytr, Xte, k = k)))

    # autoencoders: averaged over seeds, since they are non-convex
    for (mdl in c("linear", "tanh", "vae")) {
      pr <- rowMeans(vapply(seeds, function(s) {
        ae <- ae_fit(Xtr, k = k, type = mdl, epochs = 4000, lr = 0.02, seed = s)
        # map latent -> protein by ridge on the training conditions
        Ztr <- ae$encode(Xtr); Zte <- ae$encode(Xte)
        as.vector(ridge_pred(ridge_fit(Ztr, Ytr, 10), Zte))
      }, numeric(ncol(Y))))
      add(paste0("AE-", mdl), sum((as.vector(Yte) - pr)^2))
    }

    pr <- rowMeans(vapply(seeds, function(s)
      as.vector(pred_cond_ae(Xtr, Ytr, Xte, cells[tr], k = k, epochs = 3000, seed = s)),
      numeric(ncol(Y))))
    add("condition-aligned AE", sum((as.vector(Yte) - pr)^2))
  }

  tss <- sum(vapply(seq_len(n), function(i)
    sum((Y[i, ] - colMeans(Y[-i, , drop = FALSE]))^2), numeric(1)))

  data.frame(variety = label, method = names(press),
             Q2 = round(1 - vapply(press, sum, numeric(1)) / tss, 4),
             row.names = NULL)
}

d4 <- do.call(rbind, Map(loco_compare, BLOCKS, names(BLOCKS)))
d4 <- d4[order(d4$variety, -d4$Q2), ]
knitr::kable(d4, caption = "D4: leave-one-condition-out Q2 for predicting the protein block. Q2 <= 0 means no better than predicting the mean.")
```

```{r}
#| label: fig-d4
#| fig-cap: "Out-of-sample predictive performance. Anything at or below zero fails to beat the mean baseline."
#| fig-width: 11
#| fig-height: 5.5

op <- par(mfrow = c(1, 2), mar = c(10, 4.5, 3, 1))
for (nm in unique(d4$variety)) {
  d <- d4[d4$variety == nm, ]
  cols <- ifelse(d$Q2 > 0, "#2C6FBB", "#D1495B")
  cols[d$method == "mean"] <- "grey60"
  bp <- barplot(d$Q2, names.arg = d$method, las = 2, col = cols,
                ylab = expression(Q^2), main = nm,
                ylim = range(c(d$Q2, 0)) * 1.3)
  abline(h = 0, lwd = 2)
  text(bp, d$Q2, sprintf("%.3f", d$Q2), pos = ifelse(d$Q2 > 0, 3, 1), cex = 0.7)
}
par(op)
```

------------------------------------------------------------------------

## 7b. D7 --- Condition means or individual replicates for the cross-block fit?

D4 fits every method on **condition means** (8 rows). That was a choice, not the only option, and it deserves its own test rather than an assertion.

**Why "individual samples" is not simply the other option.** Under Case B there is no real sample-level correspondence between an RNA plant and a protein plant, so a cross-block fit on 24 "paired" rows requires *inventing* a pairing. There are two ways to do that, and they are not equivalent:

-   **Fixed label pairing** --- match RNA replicate *i* to protein replicate *i* by row order (same field-plot label), and trust it. This is what the private reference notebook's sample-level `sPLS` sections assumed (`stopifnot(all(rownames(cadenza.X_prot) == rownames(cadenza.X_rnaseq)))`). The label is fictional biology, so any within-cell covariance this "pairing" contributes is not real signal --- it is one arbitrary draw from noise, and it can inflate or deflate Q² by chance.
-   **Replicate-pairing bootstrap** --- within each cell, randomly permute which RNA replicate is matched to which protein replicate, refit, repeat over many draws, and average. This is the same mechanism already used for stability selection in `R/06_integration_caseB.R` (§4 there). Because the pairing is randomised, the within-cell covariance it induces is zero *in expectation* --- what survives averaging over draws is the between-cell signal, which is the only signal Case B actually licenses.

All three schemes are scored identically: **leave-one-condition-out** (never leave-one-sample-out, which would let a fictional within-cell pairing leak across train/test), against a **permutation null** that reshuffles which protein condition's data is matched to which RNA condition (`permute_Y_by_cell()` below), per @westerhuis2008. This directly answers the "add a permutation null" item flagged in §11 --- for this comparison, not yet for the rest of D4.

```{r}
#| label: d7-pairing-vs-aggregation

#' Leave-one-condition-out Q2, generic over row count: works for the 8-row
#' condition-mean scheme and the 24-row (3 reps/cell) sample-level schemes,
#' since in every case rows sharing a `cell_of` label are matched X<->Y
#' row-by-row within that label.
loco_on <- function(X, Y, cell_of, k = 2) {
  cells <- unique(cell_of)
  press <- setNames(numeric(length(cells)), cells)
  for (cl in cells) {
    te <- cell_of == cl; tr <- !te
    press[cl] <- sum((Y[te, , drop = FALSE] -
                       pred_pls(X[tr, , drop = FALSE], Y[tr, , drop = FALSE],
                                X[te, , drop = FALSE], k = k))^2)
  }
  tss <- sum(vapply(cells, function(cl) {
    te <- cell_of == cl
    sum(sweep(Y[te, , drop = FALSE], 2,
              colMeans(Y[!te, , drop = FALSE]))^2)
  }, numeric(1)))
  1 - sum(press) / tss
}

#' Reshuffle which cell's protein data is matched to which cell's RNA data,
#' preserving the within-cell group structure (balanced: 1 row/cell for
#' condition means, 3 rows/cell for sample-level). The between-cell
#' correspondence is destroyed; everything else about the scheme is kept.
permute_Y_by_cell <- function(Y, cell_of) {
  cells <- unique(cell_of)
  perm  <- setNames(sample(cells), cells)
  Yp <- Y
  for (cl in cells) {
    src <- which(cell_of == perm[[cl]])
    dst <- which(cell_of == cl)
    Yp[dst, ] <- Y[src[sample(length(src))], , drop = FALSE]
  }
  Yp
}

pairing_compare <- function(b, label, k = 2, n_pairings = 30, n_perm = 199, seed = 1) {

  common <- intersect(rownames(b$R_samp), rownames(b$P_samp))
  cell_s <- as.character(b$meta[common, "cell"])
  idx    <- split(common, cell_s)

  set.seed(seed)

  ## (a) condition means, n = 8
  q2_cond   <- loco_on(b$R_cond, b$P_cond, rownames(b$R_cond), k)
  null_cond <- replicate(n_perm,
    loco_on(b$R_cond, permute_Y_by_cell(b$P_cond, rownames(b$R_cond)), rownames(b$R_cond), k))

  ## (b) fixed label pairing, n = 24 (what the reference notebook assumed)
  Xf <- b$R_samp[common, , drop = FALSE]; Yf <- b$P_samp[common, , drop = FALSE]
  q2_fixed   <- loco_on(Xf, Yf, cell_s, k)
  null_fixed <- replicate(n_perm, loco_on(Xf, permute_Y_by_cell(Yf, cell_s), cell_s, k))

  ## (c) replicate-pairing bootstrap, n = 24, averaged over random within-cell
  ## pairings; its null combines a fresh pairing AND a fresh permutation per
  ## draw, so it is not narrower than the fixed/condition-mean nulls.
  draw_boot <- function() {
    ip <- unlist(lapply(idx, sample))
    loco_on(Xf, b$P_samp[ip, , drop = FALSE], cell_s, k)
  }
  draw_boot_null <- function() {
    ip <- unlist(lapply(idx, sample))
    loco_on(Xf, permute_Y_by_cell(b$P_samp[ip, , drop = FALSE], cell_s), cell_s, k)
  }
  q2_boot_draws <- replicate(n_pairings, draw_boot())
  null_boot     <- replicate(n_perm, draw_boot_null())

  pval <- function(obs, null) (sum(null >= obs) + 1) / (length(null) + 1)

  data.frame(
    variety = label,
    scheme  = c("condition means", "fixed label pairing", "replicate-pairing bootstrap"),
    n_rows  = c(8L, 24L, 24L),
    Q2      = round(c(q2_cond, q2_fixed, mean(q2_boot_draws)), 4),
    Q2_sd_over_pairings = round(c(NA, NA, sd(q2_boot_draws)), 4),
    null_p95 = round(c(quantile(null_cond, .95), quantile(null_fixed, .95),
                        quantile(null_boot, .95)), 4),
    p_perm  = round(c(pval(q2_cond, null_cond), pval(q2_fixed, null_fixed),
                       pval(mean(q2_boot_draws), null_boot)), 4)
  )
}

d7 <- do.call(rbind, Map(pairing_compare, BLOCKS, names(BLOCKS)))
rownames(d7) <- NULL
knitr::kable(d7, caption = "D7: PLS-only, leave-one-condition-out Q2 under three ways of building the cross-block training matrix. null_p95 is the 95th percentile of a permutation null that reshuffles condition correspondence; p_perm is the empirical p-value against that null.")
```

```{r}
#| label: fig-d7
#| fig-cap: "D7: observed Q2 (bars) against its own permutation null 95th percentile (dashed). Error bars on the bootstrap scheme show +-1 SD across 30 random pairings."
#| fig-width: 10
#| fig-height: 5

op <- par(mfrow = c(1, 2), mar = c(9, 4.5, 3, 1))
for (nm in unique(d7$variety)) {
  d <- d7[d7$variety == nm, ]
  bp <- barplot(d$Q2, names.arg = d$scheme, las = 2, col = "#2C6FBB",
                ylab = expression(Q^2), main = nm,
                ylim = range(c(d$Q2, d$null_p95, 0)) * 1.3)
  segments(bp - 0.4, d$null_p95, bp + 0.4, d$null_p95, lty = 2, lwd = 2, col = "#D1495B")
  err <- d$Q2_sd_over_pairings; err[is.na(err)] <- 0
  arrows(bp, d$Q2 - err, bp, d$Q2 + err, angle = 90, code = 3, length = 0.05)
  abline(h = 0)
  text(bp, pmax(d$Q2, d$null_p95), sprintf("p=%.3f", d$p_perm), pos = 3, cex = 0.7)
}
par(op)
```

**Reading D7 --- condition means win, and not by a small margin.** Before running this, the expectation going in was that 21 training rows per fold should beat 7, since more data usually helps a fit. That is not what happened. **Condition means are the best-performing scheme in both varieties** (Cadenza Q² = 0.287, Norin Q² = -0.010), and the replicate-pairing bootstrap --- the scheme designed to be the statistically honest way to use individual replicates --- is not just worse, it is **decisively negative and stably so** (Cadenza -0.212, Norin -0.340, SD across 30 pairings only 0.002, and p_perm = 0.295/0.485: it does not even beat its own permutation null).

The reason, on reflection, is that a single replicate is a much noisier observation than a condition mean, and the replicate-pairing bootstrap adds a *second* layer of noise on top --- the fictional within-cell pairing --- with zero informative covariance by construction. Condition-mean aggregation already averages out real replicate noise before the cross-block fit ever sees the data; individual-replicate schemes hand the model 3x the rows but at a much worse per-row signal-to-noise ratio, and PLS with a handful of components cannot separate that added noise from the between-condition signal it is trying to learn. More rows did not mean more information here.

The **fixed label pairing** (what the reference notebook assumed) happens to score better than the bootstrap and even beats its own null (p_perm = 0.005/0.015) --- but this is exactly the outcome that makes it dangerous to trust as a default, not a vindication of it: its result is driven by one arbitrary, unrepeatable draw of which replicate got matched to which. That the draw happened to land favourably here does not make the method sound, and there is no way to know in advance, on a new dataset, whether it would land favourably or not.

**Conclusion for the question this section set out to answer: use condition means.** Not because individual samples were never tried, but because they were tried two different ways and both lost, one of them by a wide and stable margin.

------------------------------------------------------------------------

## 7c. D8 --- Does global feature selection leak?

`make_blocks()` (§1) selects the top-variable RNA genes **once, using all 24 samples**, before any cross-validation loop runs. `rnaprot/unpaired_reviewer.py` (the module behind the reference notebook's other comparison, see the Purpose section) instead re-selects features **inside each outer fold**, using only the training samples. If global selection leaks, D4/D7's Q² is optimistic relative to what a genuinely unseen condition would get.

This checks the RNA (X) side only, holding the protein (Y) target set fixed at the same 1000 genes D4/D7 use --- re-selecting Y in-fold as well would change which proteins are being predicted from fold to fold, which breaks a single pooled Q² across folds. That means this is a partial check: it answers whether X-side selection leakage matters, not Y-side.

```{r}
#| label: d8-infold-selection

#' Leave-one-condition-out Q2 with RNA feature selection redone inside each
#' fold, using only that fold's training samples. Protein features are fixed
#' (hv_p_fixed) to keep a single well-defined target set across folds.
loco_infold_fs <- function(v, hv_p_fixed, n_rna = 2000, k = 2) {
  meta  <- v$meta
  cells <- levels(meta$cell)
  cs <- function(M) { M <- scale(M, TRUE, FALSE); M / sqrt(sum(M^2) / nrow(M)) }

  P_cond <- cs(t(cell_means(v$imputed$mixed[hv_p_fixed, , drop = FALSE], meta)))

  press <- setNames(numeric(length(cells)), cells)
  for (cl in cells) {
    train_samples <- rownames(meta)[meta$cell != cl]
    hv_r   <- top_variable(v$qc_rna$vst[, train_samples, drop = FALSE], n_rna)
    R_cond <- cs(t(cell_means(v$qc_rna$vst[hv_r, , drop = FALSE], meta)))

    te <- rownames(R_cond) == cl; tr <- !te
    pred <- pred_pls(R_cond[tr, , drop = FALSE], P_cond[tr, , drop = FALSE],
                     R_cond[te, , drop = FALSE], k = k)
    press[cl] <- sum((P_cond[te, , drop = FALSE] - pred)^2)
  }
  tss <- sum(vapply(cells, function(cl) {
    te <- rownames(P_cond) == cl
    sum(sweep(P_cond[te, , drop = FALSE], 2, colMeans(P_cond[!te, , drop = FALSE]))^2)
  }, numeric(1)))
  1 - sum(press) / tss
}

d8 <- do.call(rbind, lapply(names(BLOCKS), function(nm) {
  v <- VARIETIES[[nm]]
  hv_p_fixed <- top_variable(v$imputed$mixed, 1000)
  data.frame(
    variety            = nm,
    Q2_global_selection = d4$Q2[d4$variety == nm & d4$method == "PLS"],
    Q2_infold_selection = round(loco_infold_fs(v, hv_p_fixed), 4)
  )
}))
d8$gap <- round(d8$Q2_global_selection - d8$Q2_infold_selection, 4)
knitr::kable(d8, caption = "D8: PLS leave-one-condition-out Q2, global (§7) vs in-fold RNA feature selection. gap > 0 means global selection was optimistic.")
```

**Reading D8.** The gap runs the *opposite* direction from what a leakage concern predicts: in-fold selection scores slightly *higher* than global selection in both varieties (Cadenza 0.311 vs 0.287, Norin 0.022 vs -0.010), not lower. That is consistent with global top-variable selection **not** meaningfully inflating D4/D7's headline Q² here -- the gap is small and within the kind of fold-to-fold noise already seen elsewhere (D7's bootstrap SD was \~0.002 on a much larger resample count). This does not mean global selection can never leak, only that it is not the explanation for D4/D7's numbers on this data. The protein (Y) side is still globally selected in every scheme above, per the scoping note at the top of this section.

------------------------------------------------------------------------

## 8. D5 --- Are we sample-limited?

**The assumption being tested:** that 8 conditions suffice to fit these models.

If performance is still improving as training conditions are added, the models are data-starved and the comparison in §7 reflects sample size as much as method quality.

```{r}
#| label: d5-learning-curve
#| fig-cap: "Learning curve: out-of-sample error against number of training conditions. A curve still falling at n=7 means the design is the limiting factor."
#| fig-width: 10
#| fig-height: 5

learning_curve <- function(b, label, k = 2, reps = 8) {
  X <- b$R_cond; Y <- b$P_cond; n <- nrow(X)
  set.seed(1)
  do.call(rbind, lapply(3:(n - 1), function(m) {
    errs <- replicate(reps, {
      tr <- sample(n, m); te <- setdiff(seq_len(n), tr)[1]
      Xtr <- X[tr, , drop = FALSE]; Ytr <- Y[tr, , drop = FALSE]
      Xte <- X[te, , drop = FALSE]; Yte <- Y[te, , drop = FALSE]
      c(pls  = mean((Yte - pred_pls(Xtr, Ytr, Xte, k = k))^2),
        pca  = mean((Yte - pred_pca_ridge(Xtr, Ytr, Xte, k = k))^2),
        mean = mean((Yte - pred_mean(Xtr, Ytr, Xte))^2))
    })
    data.frame(variety = label, n_train = m,
               pls = mean(errs["pls", ]), pca = mean(errs["pca", ]),
               mean_baseline = mean(errs["mean", ]))
  }))
}

d5 <- do.call(rbind, Map(learning_curve, BLOCKS, names(BLOCKS)))

op <- par(mfrow = c(1, 2), mar = c(4.5, 4.5, 3, 1))
for (nm in unique(d5$variety)) {
  d <- d5[d5$variety == nm, ]
  yl <- range(c(d$pls, d$pca, d$mean_baseline))
  plot(d$n_train, d$pls, type = "b", pch = 16, lwd = 2, col = "#2C6FBB", ylim = yl,
       xlab = "training conditions", ylab = "held-out MSE", main = nm)
  lines(d$n_train, d$pca, type = "b", pch = 17, lwd = 2, col = "#4C9F70")
  lines(d$n_train, d$mean_baseline, type = "b", pch = 4, lwd = 2, lty = 2, col = "grey40")
  legend("topright", bty = "n", cex = 0.75, lwd = 2, pch = c(16, 17, 4),
         col = c("#2C6FBB", "#4C9F70", "grey40"), legend = c("PLS", "PCA+ridge", "mean"))
}
par(op)

knitr::kable(d5, caption = "D5: held-out error as training conditions are added.")
```

------------------------------------------------------------------------

## 9. D6 --- Are the latent spaces reproducible?

**The assumption being tested:** that a latent space is a property of the data, not of the random seed. PCA is deterministic; autoencoders are not.

```{r}
#| label: d6-seed-stability

seed_stability <- function(b, label, k = 3, seeds = 1:5) {
  X <- b$R_cond
  do.call(rbind, lapply(c("linear", "tanh", "vae"), function(ty) {
    Zs <- lapply(seeds, function(s)
      ae_fit(X, k = k, type = ty, epochs = 4000, lr = 0.02, seed = s)$encode(X))
    # compare subspaces pairwise via mean principal angle cosine
    cs <- c()
    for (i in 1:(length(Zs) - 1)) for (j in (i + 1):length(Zs)) {
      Qi <- qr.Q(qr(Zs[[i]])); Qj <- qr.Q(qr(Zs[[j]]))
      cs <- c(cs, mean(svd(t(Qi) %*% Qj)$d))
    }
    data.frame(variety = label, model = paste0("AE-", ty),
               mean_subspace_cosine = round(mean(cs), 4),
               min_pair = round(min(cs), 4))
  }))
}

d6 <- do.call(rbind, Map(seed_stability, BLOCKS, names(BLOCKS)))
knitr::kable(d6, caption = "D6: agreement between latent subspaces fitted from different random seeds. 1.0 = identical subspace; PCA would score exactly 1.")
```

------------------------------------------------------------------------

## 10. Assumption validation summary

```{r}
#| label: assumption-summary
#| echo: false

knitr::kable(data.frame(
  Assumption = c(
    "Leading components carry structure, not noise",
    "Components correspond to the experimental design",
    "A linear autoencoder is a distinct method from PCA",
    "Dimensionality reduction extracts usable cross-layer signal",
    "8 conditions suffice to fit these models",
    "Latent spaces are reproducible across seeds",
    "Non-linearity improves representation",
    "Condition centroids from n=3 are stable enough to align on",
    "Condition-mean aggregation is the right granularity for the cross-block fit"),
  Tested_in = c("D1 §4", "D2 §5", "D3 §6", "D4 §7", "D5 §8", "D6 §9",
                "D4 §7", "-- (not tested)", "D7 §7b"),
  How = c(
    "Horn parallel analysis vs permuted eigenvalues",
    "Eta-squared of each PC on treatment and timepoint",
    "Latent-score + reconstruction agreement with PCA (Baldi-Hornik)",
    "Leave-one-condition-out Q2 vs mean and design baselines",
    "Learning curve as training conditions are added",
    "Pairwise subspace cosine across 5 seeds",
    "AE-tanh and AE-vae vs AE-linear out of sample",
    "NOT TESTED HERE -- degenerates at condition level, see §11 point 5",
    "PLS Q2 under condition-mean, fixed-pairing and pairing-bootstrap schemes, each against its own permutation null")
), caption = "Every assumption behind this analysis and where it is tested.")
```

------------------------------------------------------------------------

## 11. What is safe to claim

The decision rules in the outline were fixed before the results were seen. Applying them:

### The results

| Method                  | Cadenza Q² | Norin Q²  | Reads as                                  |
|:------------------------|:-----------|:----------|:------------------------------------------|
| condition-aligned AE    | **0.575**  | **0.450** | best in both --- but see the caveat below |
| PLS                     | 0.287      | −0.010    | strong in Cadenza, fails in Norin         |
| PCA + ridge             | 0.199      | 0.096     | modest, consistent                        |
| AE-vae                  | 0.101      | 0.035     | above mean, below every supervised method |
| design (treatment+time) | 0.021      | 0.009     | near zero                                 |
| AE-tanh                 | 0.003      | 0.001     | indistinguishable from the mean           |
| AE-linear               | 0.003      | 0.002     | indistinguishable from the mean           |
| mean baseline           | 0.000      | 0.000     | reference                                 |

### ✅ Safe to claim

**1. Dimensionality reduction does extract usable cross-layer signal.** Multiple methods exceed both the mean baseline and the design baseline under leave-one-condition-out CV. Since the design baseline sits near zero (0.021 / 0.009), the signal is *not* a re-description of treatment and timepoint --- the same conclusion, reached by a different route, as the K1 test in `../kinetics_limited/kinetics_what_we_can_claim.qmd`.

**2. The linear autoencoder is PCA, empirically as well as in theory.** D3 gives latent-score cosines of 0.997 / 0.974 and reconstruction correlation of 1.000, with reconstruction MSE matching PCA to five significant figures. This confirms @baldi1989 *and* validates the implementation --- the two cannot be separated, which is why the check is worth running.

**3. Unsupervised reconstruction is close to useless for cross-layer prediction.** AE-linear and AE-tanh score ≈ 0.00, statistically indistinguishable from predicting the mean. This is not a failure of the optimiser (D3 proves convergence); it is a statement about the objective. An autoencoder trained to reconstruct RNA optimises for RNA variance, which is not the same thing as RNA→protein predictive structure. **Methods with a cross-modal or supervised objective (PLS, PCA+ridge, condition-aligned AE) are the ones that work.**

**4. The usable latent dimension is 3--4, not more.** Parallel analysis (D1) passes 3 RNA components and 3--4 protein components against the permutation null. Any latent dimension above this is fitting noise regardless of what reconstruction error suggests.

### ⚠️ Claim only with the caveat attached

**5. The condition-aligned architecture performs best --- but this run does not validate its defining mechanism.** The Q² of 0.575 / 0.450 is a genuine out-of-sample number from proper leave-one-condition-out CV with no leakage. However, it was run at **condition level, where each condition has exactly one row**. The centroid-alignment terms therefore degenerate: a "centroid" over one row is that row. What was actually tested is a reduced-rank cross-modal regression with reconstruction regularisation --- not centroid alignment averaging three replicates, which is what `rnaprot/unpaired.py` does at sample level.

To test the mechanism as designed, the model must be run on the **24 RNA × 24 protein sample-level matrices** with conditions as grouping, holding out whole conditions. That is the correct next step and it is not what is reported here.

**6. Variety-dependence is large and unexplained.** PLS achieves 0.287 in Cadenza and −0.010 in Norin --- from the same pipeline on the same design. Consistent with Norin's documented baseline anomaly (`../kinetics_limited/assumptions_validation.qmd` §2), but not established as its cause. Do not report a single pooled figure for "how well RNA predicts protein".

### ❌ Not supported

**7. A VAE is not justified on this data.** AE-vae beats the plain autoencoders (0.101 / 0.035) but loses to every supervised method, and it is the **least reproducible** model tested: mean subspace cosine across seeds 0.752 / 0.774, worst pair 0.636. By the rule fixed in advance --- an autoencoder counts only if it also passes D6 --- it does not. A latent space that changes materially with the random seed is not a representation of the data.

This is consistent with the literature rather than surprising: successful omics VAEs operate on thousands of observations (single cells) or transfer from large reference cohorts. Eight conditions is a different regime.

**8. Individual latent dimensions cannot be interpreted biologically.** With ≤ 8 conditions and 3--4 real components, latent axes are not identified strongly enough to name. D2 shows what the leading components track; that is as far as interpretation should go.

**9. Nothing here supports sample-level RNA↔protein correspondence.** The design is unpaired. Every result above is condition-level by construction. D7 tested the alternative directly: a *fixed* label-based pairing (what the reference notebook assumed) is not a safe default regardless of its score in any one run, because its extra 16 rows over the condition-mean scheme carry an arbitrary, unrepeatable draw rather than real signal --- see D7 §7b for the numbers.

### The one-paragraph version

> Under leave-one-condition-out cross-validation, cross-modal dimensionality reduction predicts held-out protein profiles substantially better than either a mean or a design-only baseline, confirming usable RNA→protein structure beyond the experimental design. Linear autoencoders were verified to recover the PCA solution exactly, and unsupervised reconstruction objectives were found to carry essentially no cross-layer predictive value; methods with a cross-modal objective are required. Variational autoencoders offered no advantage over supervised linear methods and produced seed-unstable latent spaces, and are not recommended at this sample size.

### What would strengthen this

1.  **Run the condition-aligned AE at sample level** (24 × 24 with condition grouping), which is what `rnaprot/unpaired.py` does and what point 5 above requires.
2.  **Add a permutation null to the rest of D4.** D7 does this for PLS specifically, under all three pairing schemes; PCA+ridge, the autoencoders and the condition-aligned AE in D4 still report only a point-estimate Q², per @westerhuis2008.
3.  **Explain the Cadenza/Norin gap** before pooling anything across varieties.

## References

-   @baldi1989 --- linear autoencoders recover the PCA subspace; no local minima. Basis of D3.
-   @jolliffe2016 --- PCA, its assumptions and limits.
-   @westerhuis2008 --- cross-validated Q² with a permutation null for small-n omics PLS.
-   @kingma2014 --- the variational autoencoder.
-   @schwanhausser2011 --- context for the RNA/protein relationship being modelled.

> **A note on identifiability, worth carrying to any future work here.** D3 initially appeared to fail (weight-subspace cosine 0.02) on a model whose reconstruction was provably optimal. The cause was that with p ≫ n the **encoder weights are not identified** --- only their projection into the data row space affects anything. Comparing latent scores and reconstructions instead gave cosine 1.000. Any comparison of autoencoders against PCA on omics-shaped data should compare identified quantities, or it will report differences that do not exist.
