---
title: "Functional enrichment (fgsea) of RNA-protein integration statistics"
author: "Kristina Gagalova"
format:
  gfm:
    toc: true
    code-tools: true
    fig-dir: "figs"
execute:
  echo: true
editor: visual
---

## Purpose and scope

This notebook asks: **do the per-gene statistics already computed elsewhere in this project (discordance distance, kinetic amplitude, cross-variety comparison) concentrate in specific functional categories, more than chance would predict?**

Every other notebook in `analysis/` classifies or ranks genes individually and then states, explicitly, where individual-gene calls are unsafe -- `kinetics_limited` in `analysis/kinetics_limited`, "not supported at the individual-gene level" for K5 in `kinetics_what_we_can_claim.qmd`. Gene-set enrichment is the standard fix for exactly that gap: aggregate a noisy per-gene statistic over a functional category and ask whether the *category* moves, which needs far less per-gene precision than an individual call does.

The method is **not new** -- it is the same machinery already used in the companion project's variety-level enrichment write-up (`norinXcadenza-shared/analysis/rnaseq/wheat/Enrichment-varieties.md`): `fgsea::fgseaMultilevel()` on a ranked statistic, gene sets built with `check_id_overlap_annotations()` / `make_fgsea_annotation_table()` (`R/Enrichment.R`, ported verbatim). What is new here is *which* statistic gets ranked -- not DE `t`, but the gene-distance and kinetic statistics this project's own pipeline produces.

### What this pulls in from elsewhere

| Source                                                                | What it provides                                                                                                 |
|-----------------------------------------------------------------------|------------------------------------------------------------------------------------------------------------------|
| `data/annotations/*-CS-rbh-final-MAPMAN.tsv.gz`                       | MapMan/Mercator bins (`MapCave.bin.name`), wide format -- the primary vocabulary, matching the upstream notebook |
| `data/annotations/*-CS-rbh-final.tsv.gz`                              | GO / Pfam / InterPro, long format -- secondary vocabulary, see caveats below                                     |
| `results/gene-distance/{norin,cadenza}_gene_distances_8condition.csv` | `d_full`, `d_pca2`, `d_umap`, `correlation`, `amplitude` per gene (`analysis/gene-distance`)                     |
| `data/real/norin_cadenza_rbh-clean.csv`                               | Direct Norin\<-\>Cadenza reciprocal-best-hit map, for cross-variety comparison                                   |

### Annotation caveats, established before running anything

A profiling pass over these annotation files (done before writing this notebook) found:

1.  **The detected proteome is the binding constraint.** Norin: 5,879 quantified proteins, 4,561 annotated. Cadenza: 6,105 quantified, 5,548 annotated. Every gene set below is built **from the detected proteome only**, never the whole genome -- an enrichment background of "everything IWGSC ever annotated" would just recover "MS detects abundant, well-known proteins" (ribosome, Rubisco, glycolysis).
2.  **GO in these files is thin and unpropagated.** Within the Norin proteome universe there are only 48 GO:BP terms with \>= 10 genes (MF 88, CC 17), and terms are raw InterPro2GO leaves, not ancestor-propagated (`GO:0008152     metabolic process` sits alongside its own child terms as a direct annotation). GO is kept here as a secondary, coarser cross-check only.
3.  **MapMan/Mercator is the primary vocabulary** for exactly this reason -- it is purpose-built, hand-curated, and richer for wheat than raw InterPro2GO. This mirrors the choice already made in the companion project's own enrichment notebook.
4.  **RBH is many-to-many.** 408 of 4,561 Norin proteome genes map to more than one CS v2.1 gene (up to 6). `check_id_overlap_annotations()` already handles this (the `mixed_terms` status), so it is not re-solved here.

------------------------------------------------------------------------

## 1. Setup

```{r}
#| label: setup
#| message: false

if (!requireNamespace("here", quietly = TRUE)) stop("package 'here' required")

## RENV_PROJECT must be set before sourcing activate.R, or renv treats this
## subdirectory as the project and bootstraps an empty library.
if (file.exists(here::here("renv", "activate.R"))) {
  Sys.setenv(RENV_PROJECT = here::here())
  source(here::here("renv", "activate.R"))
}

source(here::here("R", "utils.R"))
need_pkgs(c("dplyr", "tidyr", "stringr", "tibble", "data.table", "readr",
            "fgsea", "ggplot2", "patchwork", "ggrepel", "BiocParallel"))

library(dplyr)
library(tidyr)
library(stringr)
library(tibble)
library(ggplot2)
library(fgsea)
library(patchwork)

source(here::here("R", "Enrichment.R"))

set.seed(20260825)

ANNO_DIR <- here::here("data", "annotations")
GD_DIR   <- here::here("results", "gene-distance")
REAL_DIR <- here::here("data", "real")

#' Canonical gene-ID normalisation for this notebook.
#'
#' Four ID conventions have to be reconciled here, and they disagree in both
#' CASE and SUFFIX:
#'   proteomics matrices  TraesNOR1A03G00000030            (upper, bare)
#'   MapMan (Norin)       TraesNOR3B03G01717390.1.CDS1     (upper, isoform+CDS)
#'   MapMan (Cadenza)     TraesCAD_scaffold_081796_01G000400.1  (upper, isoform)
#'   gene-distance table  TraesNOR1D03G00570400            (upper, bare)
#'
#' `check_id_overlap_annotations()` lowercases internally, so lowercase is the
#' project-wide join key. EVERY id that participates in a join must go through
#' this function -- normalising only one side is a silent zero-overlap join,
#' which is exactly what produced empty gene sets on the first run.
strip_isoform <- function(x) {
  x <- tolower(x)
  x <- gsub("\\.cds[0-9]+$", "", x)
  x <- gsub("\\.[0-9]+$", "", x)
  x
}

#' Guard against the silent-empty-join failure mode. fgsea returns 0 rows
#' rather than erroring when nothing overlaps, so the check has to be explicit.
assert_overlap <- function(a, b, what) {
  n <- length(intersect(a, b))
  if (n == 0) stop("zero overlap in ", what,
                   " -- check ID normalisation (case / isoform suffix)",
                   call. = FALSE)
  cat(sprintf("%-46s %6d ids overlap\n", what, n))
  invisible(n)
}
```

------------------------------------------------------------------------

## 2. Annotation universes

### 2.1 Detected proteome per variety

The enrichment background throughout this notebook -- gene sets are built *only* from genes actually quantified in the proteomics data, never the whole genome.

```{r}
#| label: proteome-universe

## strip_isoform() here is what makes the universe join-compatible with the
## annotation tables; without it every downstream %in% silently matches nothing.
read_id_col <- function(path) {
  x <- data.table::fread(path, select = 1L, header = TRUE)
  unique(strip_isoform(x[[1]]))
}

universe <- list(
  norin   = read_id_col(file.path(REAL_DIR, "norin-prot.csv")),
  cadenza = read_id_col(file.path(REAL_DIR, "cadenza-prot.csv"))
)

cat("Norin proteome:   ", length(universe$norin),   "quantified proteins\n")
cat("Cadenza proteome: ", length(universe$cadenza), "quantified proteins\n")
```

### 2.2 MapMan/Mercator bins (primary vocabulary)

Wide-format annotation, one row per gene, `MapCave.bin.name` carries the MapMan bin. Gene IDs here carry a `.<isoform>.CDS<n>` suffix that the upstream `check_id_overlap_annotations()` does not fully strip (it only removes the `.CDS<n>` part) -- so isoform suffixes are stripped explicitly here first, to match the bare gene IDs used everywhere else in this project.

```{r}
#| label: load-mapman
#| message: false

load_mapman <- function(variety) {
  path <- file.path(ANNO_DIR, sprintf("%s-CS-rbh-final-MAPMAN.tsv.gz",
                                       tools::toTitleCase(variety)))
  df <- readr::read_tsv(path, show_col_types = FALSE, progress = FALSE)
  df[[variety]] <- strip_isoform(df[[variety]])
  assert_overlap(df[[variety]], universe[[variety]],
                 sprintf("MapMan x %s proteome", variety))
  df[df[[variety]] %in% universe[[variety]], ]
}

mapman <- list(
  norin   = load_mapman("norin"),
  cadenza = load_mapman("cadenza")
)

cat("\nNorin MapMan rows restricted to proteome:  ", nrow(mapman$norin), "\n")
cat("Cadenza MapMan rows restricted to proteome:", nrow(mapman$cadenza), "\n")
```

Build the fgsea gene-set table exactly as the upstream notebook does -- `check_id_overlap_annotations()` classifies each gene as `specific` / `generic` / `mixed_generic` / `mixed_terms` / `non.annotated`, and `make_fgsea_annotation_table()` collapses that to one `term_for_fgsea` string per gene, ready to `split()` into pathways.

```{r}
#| label: mapman-fgsea-table

build_mapman_annotation_table <- function(variety) {
  genes_df <- tibble(gene_id = universe[[variety]])

  res <- check_id_overlap_annotations(
    ranked_df       = genes_df,
    annotation_df   = mapman[[variety]],
    rank_gene_col   = "gene_id",
    assign_gene_col = variety,
    term_col        = "MapCave.bin.name",
    not_assigned_terms = c("not assigned.unknown", "not assigned.no ontology")
  )

  make_fgsea_annotation_table(res$checked_df, id_col = "gene_id")
}

mapman_anno_table <- list(
  norin   = build_mapman_annotation_table("norin"),
  cadenza = build_mapman_annotation_table("cadenza")
)

lapply(mapman_anno_table, function(x) table(x$status))
```

### 2.3 GO / Pfam / InterPro (secondary vocabulary)

Long format: one row per (gene, annotation-type). Restricted to the detected proteome, split by `f.type` (and by `f.domain` for Gene Ontology), each category built into its own set of pathways.

```{r}
#| label: load-go-pfam-interpro
#| message: false

load_longfmt <- function(variety) {
  path <- file.path(ANNO_DIR, sprintf("%s-CS-rbh-final.tsv.gz",
                                       tools::toTitleCase(variety)))
  dt <- data.table::fread(path, sep = "\t", header = TRUE, quote = "")
  id_col <- variety
  dt[[id_col]] <- strip_isoform(dt[[id_col]])
  assert_overlap(dt[[id_col]], universe[[variety]],
                 sprintf("GO/Pfam/InterPro x %s proteome", variety))
  dt[dt[[id_col]] %in% universe[[variety]]]
}

longfmt <- list(
  norin   = load_longfmt("norin"),
  cadenza = load_longfmt("cadenza")
)

#' Build one fgsea pathway list from the long-format annotation.
#' @param f_type "Pfam", "InterPro", or "Gene Ontology"
#' @param go_domain restrict to a GO domain (BP/MF/CC); ignored otherwise
build_pathways_longfmt <- function(dt, id_col, f_type, go_domain = NULL,
                                    min_size = 5, max_size = 500) {
  sub <- dt[dt$f.type == f_type]
  if (!is.null(go_domain)) sub <- sub[sub$f.domain == go_domain]
  sub <- sub[!is.na(sub$f.name) & sub$f.name != ""]

  label <- ifelse(is.na(sub$f.label) | sub$f.label == "", sub$f.name, sub$f.label)
  prefix <- if (!is.null(go_domain)) sprintf("GO:%s", go_domain) else f_type
  sub$term <- sprintf("%s: %s (%s)", prefix, label, sub$f.name)

  sets <- split(sub[[id_col]], sub$term)
  sets <- lapply(sets, unique)
  sets[lengths(sets) >= min_size & lengths(sets) <= max_size]
}

build_all_longfmt_pathways <- function(variety) {
  dt <- longfmt[[variety]]
  c(
    build_pathways_longfmt(dt, variety, "Pfam"),
    build_pathways_longfmt(dt, variety, "InterPro"),
    build_pathways_longfmt(dt, variety, "Gene Ontology", "BP"),
    build_pathways_longfmt(dt, variety, "Gene Ontology", "MF"),
    build_pathways_longfmt(dt, variety, "Gene Ontology", "CC")
  )
}

longfmt_pathways <- list(
  norin   = build_all_longfmt_pathways("norin"),
  cadenza = build_all_longfmt_pathways("cadenza")
)

cat("Norin:   ", length(longfmt_pathways$norin),   "Pfam/InterPro/GO sets (proteome, size 5-500)\n")
cat("Cadenza: ", length(longfmt_pathways$cadenza), "Pfam/InterPro/GO sets (proteome, size 5-500)\n")
```

------------------------------------------------------------------------

## 3. Load the integration statistics to rank on

`analysis/gene-distance/gene_distance_shared_space.qmd` already produced, per gene, a shared-space discordance distance (`d_full`, full-dimensional; `d_pca2`/`d_umap`, low-dimensional), a total-least-squares `correlation` between the RNA and protein condition profiles, and an `amplitude` (protein range relative to RNA range). These are real-data outputs, already on disk -- nothing is recomputed here.

```{r}
#| label: load-gene-distance

gene_distance <- list(
  norin   = read.csv(file.path(GD_DIR, "norin_gene_distances_8condition.csv")),
  cadenza = read.csv(file.path(GD_DIR, "cadenza_gene_distances_8condition.csv"))
)

lapply(gene_distance, function(d) { d$gene_id <- strip_isoform(d$gene_id); nrow(d) })
gene_distance <- lapply(gene_distance, function(d) { d$gene_id <- strip_isoform(d$gene_id); d })

str(gene_distance$norin)
```

------------------------------------------------------------------------

## 4. E1 -- Discordance GSEA: which bins go out of RNA/protein sync?

Rank every gene by `d_full` (higher = more discordant) and ask which MapMan bins sit systematically toward one end of that ranking -- no arbitrary top-decile cutoff, the full ranking is used, exactly as `fgsea` is designed for.

```{r}
#| label: e1-run-fgsea
#| message: false

#' @param score_type passed to fgsea. "std" for a SIGNED statistic (enrichment
#'   can be at either tail); "pos" for a strictly NON-NEGATIVE one. Getting this
#'   wrong is not cosmetic: with an all-positive statistic under "std", fgsea's
#'   running sum can only ever walk one way, so the two-tailed null it compares
#'   against is the wrong null and the p-values are not interpretable. `d_full`
#'   is a Euclidean distance and so is strictly positive -- it needs "pos".
run_fgsea_on_stat <- function(df, stat_col, pathways, min_size = 5, max_size = 500,
                              score_type = c("std", "pos"), decreasing = TRUE) {
  score_type <- match.arg(score_type)

  ranks <- setNames(df[[stat_col]], df$gene_id)
  ranks <- ranks[is.finite(ranks)]
  ranks <- sort(ranks, decreasing = decreasing)

  assert_overlap(names(ranks), unlist(pathways, use.names = FALSE),
                 sprintf("ranked %s x gene sets", stat_col))

  fgseaMultilevel(
    pathways  = pathways,
    stats     = ranks,
    minSize   = min_size,
    maxSize   = max_size,
    scoreType = score_type,
    nproc     = 1
  )
}

e1_mapman_pathways <- lapply(mapman_anno_table, function(tab) {
  tab <- tab %>% filter(keep_for_fgsea, !is.na(term_for_fgsea))
  sets <- split(tab$gene_ids, tab$term_for_fgsea)
  sets <- lapply(sets, unique)
  sets[lengths(sets) >= 5 & lengths(sets) <= 500]
})

cat("MapMan gene sets (size 5-500) --",
    "Norin:", length(e1_mapman_pathways$norin),
    "| Cadenza:", length(e1_mapman_pathways$cadenza), "\n")

## d_full is a Euclidean distance -> strictly positive -> scoreType "pos".
e1_norin   <- run_fgsea_on_stat(gene_distance$norin,   "d_full", e1_mapman_pathways$norin,
                                score_type = "pos")
e1_cadenza <- run_fgsea_on_stat(gene_distance$cadenza, "d_full", e1_mapman_pathways$cadenza,
                                score_type = "pos")
```

```{r e1-bubble, fig.width=30, fig.height=15, fig.align='center'}
#| label: e1-bubble

e1_plots <- list(
  Norin   = plot_fgsea_bubble(e1_norin,   top_n = 25, plot_title = "Norin: d_full discordance"),
  Cadenza = plot_fgsea_bubble(e1_cadenza, top_n = 25, plot_title = "Cadenza: d_full discordance")
)
e1_plots <- Filter(Negate(is.null), e1_plots)
if (length(e1_plots) > 0) Reduce(`|`, e1_plots)
```

### Result (first successful render, 2026-08-25)

Of the 5,159 / 5,580 ranked genes, only **2,329 / 2,553** actually fall in a qualifying MapMan bin (5--500 genes) -- the rest are `non.annotated` or land in bins outside that size range, so that smaller number is the effective sample size for this test. **168 / 174** bins were tested; **13 / 12** survive `padj < 0.05`. That is a modest but real signal: about 7--8% of tested bins show significant discordance enrichment, well above the 5% expected under the null.

------------------------------------------------------------------------

## 5. E2 -- Three-metric decomposition: distance, correlation, amplitude

`d_full` alone conflates two different failure modes: a gene can be far in shared space because its RNA and protein *shapes* differ (low correlation -- lag/timing mismatch) or because the shapes agree but the *scale* differs (high correlation, extreme amplitude -- buffering/amplification). Running `fgsea` on each statistic separately is the direct way to ask which bins drive which failure mode.

Reading the three together is the point: a bin that is high on `d_full` **and** low on `correlation` is going out of sync in *timing*; a bin high on `d_full` but with correlation intact and `log2_amplitude` displaced is going out of sync in *magnitude* -- which is the buffered/amplified axis the kinetic model in `analysis/kinetics_limited` estimates per gene, reached here at category level where the per-gene call is not safe.

```{r}
#| label: e2-run-fgsea
#| message: false

## `amplitude` is a RATIO of protein to RNA response range: 1 = matched, <1
## buffered, >1 amplified. On the raw scale it is positive and multiplicatively
## symmetric around 1, so a rank on it is dominated by the amplified tail and
## compresses the buffered half into (0, 1). log2 puts buffering and
## amplification on symmetric, additive footing around 0 -- the same axis the
## kinetic model's `a` uses -- and makes scoreType "std" the correct choice.
gene_distance <- lapply(gene_distance, function(d) {
  d$log2_amplitude <- log2(d$amplitude)
  d
})

## Signed statistics take "std"; the strictly-positive distance takes "pos".
e2_spec <- list(
  d_full         = "pos",
  correlation    = "std",
  log2_amplitude = "std"
)

e2_results <- lapply(names(gene_distance), function(v) {
  lapply(names(e2_spec), function(s) {
    run_fgsea_on_stat(gene_distance[[v]], s, e1_mapman_pathways[[v]],
                      score_type = e2_spec[[s]])
  }) |> setNames(names(e2_spec))
})
names(e2_results) <- names(gene_distance)
e2_stats <- names(e2_spec)
```

```{r e2-bubble, fig.width=30, fig.height=15, fig.align='center'}
#| label: e2-bubble

e2_plots <- lapply(names(e2_results), function(v) {
  lapply(e2_stats, function(s) {
    plot_fgsea_bubble(e2_results[[v]][[s]], top_n = 15,
                      plot_title = sprintf("%s: %s", tools::toTitleCase(v), s))
  })
})
names(e2_plots) <- names(e2_results)

e2_plots_flat <- Filter(Negate(is.null), unlist(e2_plots, recursive = FALSE))
if (length(e2_plots_flat) > 0) Reduce(`|`, e2_plots_flat)
```

### Result (first successful render, 2026-08-25) -- and a bug in the `log2_amplitude` step above

`correlation` GSEA is clean: 2 / 5 bins pass `padj < 0.05` (Norin / Cadenza), on the full 2,329 / 2,553-gene ranking.

`log2_amplitude` is **not** clean, and the cause is a bug in the chunk above, not a biological null result. `analysis/gene-distance/gene_distance_shared_space.qmd` (its own line 657) documents that the `amplitude` column saved to `results/gene-distance/*_gene_distances_8condition.csv` is **already** `log2(protein response magnitude / RNA response magnitude)` -- signed, centred at 0. The comment above this chunk describes it instead as a raw positive ratio needing a first `log2()`, and applies one. In the actual data, `amplitude` is negative for roughly 60% of genes (3,115 / 5,159 Norin, 3,169 / 5,580 Cadenza; min about --3.65 / --4.55) -- so `log2(d$amplitude)` takes the log of a negative number for most genes, which is `NaN` (visible as the two `NaNs produced` warnings on render), silently dropped by `ranks[is.finite(ranks)]` inside `run_fgsea_on_stat()`.

Net effect: the `log2_amplitude` GSEA ran on only **904 / 1,118** genes instead of the \~2,300--2,500 the other two statistics use, and **0 bins** survive `padj < 0.05` in either variety. That zero is an artefact of the halved, biased sample (whichever genes happened to have positive `amplitude` survive; the rest are discarded), not evidence that amplitude-driven discordance has no functional structure.

**Not fixed in this pass, since a fix changes results and a re-render was asked to wait.** The fix, when applied: drop the second `log2()` in the chunk above and rank on `amplitude` directly (still `scoreType = "std"`, since it is already signed) -- e.g. replace `d$log2_amplitude <- log2(d$amplitude)` with `d$log2_amplitude <- d$amplitude` (kept the same column name so the rest of the chunk/plots need no other change), or rename throughout for clarity.

------------------------------------------------------------------------

## 6. E3 -- Cross-variety differential enrichment

Join the two varieties' `d_full` rankings on the direct Norin\<-\>Cadenza reciprocal-best-hit map (`data/real/norin_cadenza_rbh-clean.csv`) and compare NES for each MapMan bin between varieties -- which functional categories are discordant in one variety but not the other.

```{r}
#| label: e3-cross-variety
#| message: false

rbh_nc <- read.csv(file.path(REAL_DIR, "norin_cadenza_rbh-clean.csv"))
rbh_nc$query  <- strip_isoform(rbh_nc$query)
rbh_nc$target <- strip_isoform(rbh_nc$target)

cat("Norin<->Cadenza 1:1 RBH pairs:", nrow(distinct(rbh_nc, query, target)), "\n")

e3_table <- function(res, label) {
  as.data.frame(res) %>%
    transmute(pathway, NES, padj, size) %>%
    rename_with(~ paste0(., "_", label), c(NES, padj, size))
}

e3_compare <- full_join(
  e3_table(e1_norin,   "norin"),
  e3_table(e1_cadenza, "cadenza"),
  by = "pathway"
) %>%
  mutate(delta_NES = NES_norin - NES_cadenza) %>%
  arrange(desc(abs(delta_NES)))

head(e3_compare, 20)
```

```{r e3-plot, fig.width=15, fig.height=15, fig.align='center'}
#| label: e3-plot


e3_top <- e3_compare %>%
  filter(!is.na(delta_NES)) %>%
  slice_max(abs(delta_NES), n = 20)

if (nrow(e3_top) > 0) {
  ggplot(e3_top, aes(x = NES_norin, y = NES_cadenza)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
    geom_point(aes(size = abs(delta_NES)), alpha = 0.75) +
    ggrepel::geom_text_repel(aes(label = pathway), size = 3, max.overlaps = 15) +
    theme_bw(base_size = 13) +
    labs(title = "Discordance NES: Norin vs Cadenza",
         subtitle = "Off-diagonal = variety-specific discordance in this MapMan bin",
         x = "NES (Norin, d_full)", y = "NES (Cadenza, d_full)")
}
```

### Result (first successful render, 2026-08-25) -- read the top-20 table with padj in view

The join covers 105,305 Norin\<-\>Cadenza RBH pairs. The top 20 bins by `|delta_NES|` (`TCA / org transformation...malic`, `lipid metabolism...beta-oxidation`, `RNA.processing.RNA helicase`, etc.) are ranked on the raw NES gap **without conditioning on significance in either variety** -- most rows in that table have `padj_norin` or `padj_cadenza` at or near 1.0 (e.g. row 1: `padj_norin = 1.00`, `padj_cadenza = 0.58`; row 5: both `padj = 1.00`). So this table currently answers "biggest raw NES gap," not "biggest *credible* gap" -- a bin can top it purely from sampling noise in one variety's ranking. Tightening this to `filter(padj_norin < 0.05 | padj_cadenza < 0.05)` before `slice_max()` would restrict the comparison to bins that are actually significant in at least one variety, which is the comparison the section's framing ("which functional categories are discordant in one variety but not the other") implies. Not applied here for the same reason as the E2 note above -- flagged, not fixed, pending a deliberate re-render.

------------------------------------------------------------------------

## 6b. E4 -- An ML-derived per-gene ranking: PLS component-1 loading (Cadenza only)

Everything above ranks genes by kinetics or shared-space distance. Whether an ML method (PLS/PCA/autoencoder) could *also* contribute an individual-gene ranking was checked directly rather than assumed, because the obvious way to get one from these methods -- a point-estimate PLS/AE loading -- is exactly the failure mode already found in the reference PLS notebook (`norinXcadenza-shared/analysis/integrated/latent-space-projection/PLS-integration.qmd`): it fabricates a 1:1 RNA\<-\>protein sample correspondence via `sample_cond` label matching, which `dimensionality_reduction_wheat.qmd` §7b (D7) already quantified as an unrepeatable, apparently favourable draw (`fixed_pairing_pctile_of_random_pairings = 1.00` in **both** varieties) rather than real signal.

D7 also identified the scheme that *is* trustworthy: condition-mean PLS, with individual replicates entering only through **randomised within-cell resampling** (the same mechanism `R/06_integration_caseB.R` §2 already uses for component-trajectory bootstrap CIs). What was missing is a per-**gene** version of that same resampling -- 06's bootstrap tracks score stability, not loading stability. `pls_boot_loadings()` (added to `R/pls_utils.R`) closes that gap: it refits the condition-mean PLS over many within-cell replicate resamples and reports, per gene, the fraction of resamples whose loading sign agrees with the point estimate -- a stability score in the same spirit as `pairing_stability_selection.tsv`, but built on the resampling scheme D7 validated rather than the one it refuted.

Two more restrictions follow directly from the last conversation's audit, not from this section's own results:

-   **Cadenza only.** D7 found no validated aggregate coupling for Norin (condition-mean PLS `p_Q2 = 0.060`, missing even a lenient threshold) -- there is no validated axis to rank Norin genes against.
-   **Component 1 only**, decided by `q2_perm_pvalue()` below *before* looking at any gene-level output, per the pre-registration discipline the rest of this project follows.

```{r}
#| label: e4-setup
#| message: false

need_pkgs(c("edgeR", "DESeq2", "matrixStats", "impute", "limma"))
source(here::here("R", "wheat_pipeline.R"))
source(here::here("R", "pls_utils.R"))

DESIGN <- wheat_design()
cad_prep <- prepare_variety("cadenza", DESIGN)

cd_r_e4 <- cad_prep$meta[colnames(cad_prep$qc_rna$vst), , drop = FALSE]
cd_p_e4 <- cad_prep$meta[colnames(cad_prep$imputed$mixed), , drop = FALSE]
cells_e4 <- levels(cad_prep$meta$cell)

cat("RNA genes after QC:            ", nrow(cad_prep$qc_rna$vst), "\n")
cat("Protein genes after QC+impute: ", nrow(cad_prep$imputed$mixed), "\n")
```

```{r}
#| label: e4-component-selection
#| message: false

n_hvg_e4 <- min(2000, nrow(cad_prep$qc_rna$vst))
hv_r_e4  <- top_variable(cad_prep$qc_rna$vst, n_hvg_e4)
hv_p_e4  <- rownames(cad_prep$imputed$mixed)   # already small (~5-6k) -- use all

Xc_e4 <- prep_block(cell_means(cad_prep$qc_rna$vst[hv_r_e4, ],   cd_r_e4))
Yc_e4 <- prep_block(cell_means(cad_prep$imputed$mixed[hv_p_e4, ], cd_p_e4))
stopifnot(identical(rownames(Xc_e4), rownames(Yc_e4)))

## Which components carry validated, out-of-sample predictive signal, BEFORE
## any gene loading is examined -- the same discipline as D7/06's own p_Q2.
q2p_e4 <- q2_perm_pvalue(Xc_e4, Yc_e4, ncomp = 3, nperm = 300, seed = 20260825)
q2p_e4
```

**Reading the component table.** Component 1 alone is both the strongest (Q² = 0.246 vs. a permutation-null 95th percentile of 0.061) and, notably, cumulative Q² *falls* as components 2 and 3 are added (0.175, then 0.118) even though each cumulative Q² still nominally clears its own (increasingly negative) null. That pattern -- more components making held-out prediction worse -- is the signature of components 2/3 adding noise rather than additional shared signal, not evidence they should be included. **Component 1 only is used below.**

```{r}
#| label: e4-boot-loadings
#| message: false

e4_boot <- pls_boot_loadings(cad_prep$qc_rna$vst, cad_prep$imputed$mixed,
                             cd_r_e4, cd_p_e4, cells_e4,
                             hv_r_e4, hv_p_e4, ncomp = 1, B = 200, seed = 20260825)

e4_prot <- e4_boot$protein
e4_prot$gene_id <- strip_isoform(e4_prot$feature)

cat("Loading summary (component 1):\n"); print(summary(e4_prot$loading))
cat("Stability summary (fraction of 200 bootstrap draws agreeing in sign):\n")
print(summary(e4_prot$stability))
cat("Top-100 |loading| genes with stability >= 0.9:",
   sum(rank(-abs(e4_prot$loading)) <= 100 & e4_prot$stability >= 0.9), "/ 100\n")
cat("Top-100 |loading| genes with stability <  0.5 (would-be false leads):",
   sum(rank(-abs(e4_prot$loading)) <= 100 & e4_prot$stability <  0.5), "/ 100\n")
```

```{r}
#| label: e4-run-fgsea
#| message: false

## Reuses the SAME Cadenza MapMan gene sets as E1/E2 -- no new pathway-
## building code, and it makes the E1/E4 hit lists directly comparable.
## No Norin arm: D7 found no validated component to rank Norin genes against.
e4_prot$loading_c1 <- e4_prot$loading
e4_cadenza <- run_fgsea_on_stat(e4_prot, "loading_c1", e1_mapman_pathways$cadenza,
                                score_type = "std")

cat("MapMan bins tested:", nrow(e4_cadenza), "| padj < 0.05:",
   sum(e4_cadenza$padj < 0.05, na.rm = TRUE), "\n")
```

```{r e4-bubble, fig.width=14, fig.height=8, fig.align='center'}
#| label: e4-bubble

plot_fgsea_bubble(e4_cadenza, top_n = 20,
                  plot_title = "Cadenza: PLS component-1 loading (validated axis)")
```

### Result -- a clean, biologically coherent signal, and it is not just re-deriving E1

19--20 / 180 testable bins pass `padj < 0.05` (\~11%, versus E1's 12/174 ≈ 7%) -- `fgseaMultilevel`'s Monte Carlo p-value estimate moves one borderline bin across the 0.05 line from one run to the next (seeded, but not bit-identical across R sessions), so treat the count as \~19--20, not a fixed number. The top hit by a wide margin, stable across runs, is **`stress.biotic`** (NES = 2.15, padj ≈ 5e-11, size = 58) -- the dominant validated shared RNA/protein axis over this infection time-course is, unsurprisingly and reassuringly, the biotic-stress response. The full positive side reads as a coherent defence program: peroxidases, beta-1,3-glucanases (classic PR-protein defence enzymes), phenylpropanoid/lignin biosynthesis (cell-wall reinforcement), glutathione S-transferases, cytochrome P450s and UDP-glucosyl/glucoronyl transferases (xenobiotic/defence detoxification), ABC transporters. The negative side is the mirror image -- photosynthesis and light signalling turned down: PSII polypeptide subunits, NADH DH, `signalling.light`, PIP aquaporins, magnesium chelatase (chlorophyll synthesis). That defence-up / photosynthesis-down pattern is exactly what a validated shared axis over an infection course should look like, which is a meaningful sanity check on the whole E4 approach, independent of any annotation-level argument.

**Comparison with E1:** only 6 of these bins (`stress.biotic`, `stress.abiotic`, `misc.peroxidases`, the beta-1,3-glucanase bin, the PAL lignin bin, and the UDP-glucosyl/glucoronyl transferase bin) were also significant in E1's `d_full`-ranked discordance test -- **the other \~13 are new**, not a re-statement of the distance-based result. That is the concrete answer to "what more can we use `dimensionality_reduction`'s ML approach for": a validated PLS component contributes gene-set-level information that shape/scale discordance (E1/E2) does not surface on its own, precisely because it is picking up the dominant *direction* of coordinated cross-omics variation (infection response) rather than per-gene disagreement between the two layers.

**Why this is trustworthy where the reference notebook's version was not:** every ingredient here was checked before being used, not assumed -- `p_Q2 = 0.013` for the component itself (not raw Q², which D7/06 showed is gameable at n = 8), Cadenza-only (Norin has no validated axis), component 1 only (2/3 add noise, not signal), and the loading itself is bootstrap-stabilised under the resampling scheme D7 validated (100/100 of the top-100 \|loading\| genes have stability \>= 0.9; zero have stability \< 0.5). None of that machinery existed before this conversation; all of it is now in `R/pls_utils.R` (`q2_perm_pvalue()`, `pls_boot_loadings()`) for reuse.

```{r}
#| label: e4-outputs

## save_fgsea() is defined in §8 below, not yet in scope here -- inlined
## rather than forward-referenced, since chunks execute top to bottom.
save_fgsea_e4 <- function(res, path) {
  df <- as.data.frame(res)
  df[] <- lapply(df, function(x) if (is.list(x)) sapply(x, paste, collapse = ";") else x)
  write.table(df, path, sep = "\t", row.names = FALSE, quote = FALSE)
}

int_outdir <- ensure_dir(here::here("results", "integration"))
write.table(q2p_e4, file.path(int_outdir, "cadenza_real_joint_components_q2perm.tsv"),
           sep = "\t", row.names = FALSE, quote = FALSE)
write.table(e4_boot$protein, file.path(int_outdir, "cadenza_real_loadings_protein_boot.tsv"),
           sep = "\t", row.names = FALSE, quote = FALSE)
write.table(e4_boot$rna, file.path(int_outdir, "cadenza_real_loadings_rna_boot.tsv"),
           sep = "\t", row.names = FALSE, quote = FALSE)
save_fgsea_e4(e4_cadenza, here::here("results", "enrichment", "E4_pls_component1_mapman_cadenza.tsv"))
```

------------------------------------------------------------------------

## 6c. E5 -- A second ML-derived ranking: the condition-aligned autoencoder (both varieties)

E4 used PLS because D7 validated it and the reference notebook's PLS-based sample-matching was the specific method under audit. But `dimensionality_reduction_wheat.qmd`'s own D4 permutation-null table (already computed, on the real wheat data, before this section existed) shows something E4 could not: at `k = 2`, the **condition-aligned autoencoder** gives **Norin** a validated cross-omics signal (Q² = 0.1474, permutation-null 95th pctile = -0.0088, **p_perm = 0.02**) where plain PLS found nothing (D7: p = 0.06). Cadenza validates too, and more strongly (Q² = 0.3205, p_perm = 0.02). This is a direct, concrete answer to "any plans to use the autoencoder model": yes, specifically this one, and specifically because it is the one AE variant that already cleared both bars that disqualified the others (D6: seed-to-seed reproducibility 0.9992, versus 0.75-0.84 for AE-vae/AE-tanh).

Two caveats carried over from that notebook, stated before running anything new here:

-   **This is the condition-mean-level use of the architecture.** D9 (`dimensionality_reduction_wheat.qmd` Sec.7d) found that fitting the model at the replicate level it was actually designed for (real n=3 centroid alignment, not a degenerate n=1-per-cell alignment) gives a *lower* Q² (Cadenza 0.2619, Norin 0.0894) than the condition-mean fit used for the p_perm numbers above. Used here anyway, matching D4's own validated scheme exactly -- the alternative would require building a fresh permutation null for the replicate-level fit, not yet done anywhere.
-   **k = 1 is used below, not the validated k = 2.** At k \>= 2 the autoencoder's latent dimensions carry no ordering or identity constraint the way PLS's sequentially-deflated components do, so which resampled bootstrap dimension corresponds to which point-estimate dimension is not well defined -- a real obstacle to the same loading-stability check E4 relies on. k = 1 sidesteps it (only a sign ambiguity remains, and `align_sign()` already handles that generically), at the cost of needing its own, fresh Q²/permutation-null validation -- the k = 2 result above does not carry over automatically. That validation is what the next chunk does, before any gene loading is examined, in the same order as E4.

```{r}
#| label: e5-setup
#| message: false

source(here::here("R", "ae_utils.R"))

## Cadenza's prepared data already exists from E4 (cad_prep) -- reused, not
## recomputed. Norin needs its own prepare_variety() call.
nor_prep <- prepare_variety("norin", DESIGN)
cd_r_nor <- nor_prep$meta[colnames(nor_prep$qc_rna$vst), , drop = FALSE]
cd_p_nor <- nor_prep$meta[colnames(nor_prep$imputed$mixed), , drop = FALSE]
cells_e5 <- levels(nor_prep$meta$cell)

VAR_E5 <- list(
  cadenza = list(prep = cad_prep, cd_r = cd_r_e4, cd_p = cd_p_e4),
  norin   = list(prep = nor_prep, cd_r = cd_r_nor, cd_p = cd_p_nor)
)

hv_e5 <- lapply(VAR_E5, function(v) {
  n_hvg <- min(2000, nrow(v$prep$qc_rna$vst))
  list(hv_r = top_variable(v$prep$qc_rna$vst, n_hvg),
      hv_p = rownames(v$prep$imputed$mixed))
})
```

```{r}
#| label: e5-component-validation
#| message: false

## Fresh Q2 + permutation null at k = 1, condition-mean level -- the k = 2
## result quoted above does not validate k = 1. Same LOCO + permute_Y_by_cell
## machinery dimensionality_reduction_wheat.qmd's D4/D7 use, extracted
## verbatim into R/ae_utils.R. nperm = 29 matches that notebook's own stated
## setting (its D4 permutation-null chunk's code currently runs 49; 29 is
## used here for tractable runtime and is the value that notebook's own
## table caption describes).
q2p_e5 <- do.call(rbind, lapply(names(VAR_E5), function(nm) {
  v  <- VAR_E5[[nm]]
  Xc <- t(cell_means(v$prep$qc_rna$vst[hv_e5[[nm]]$hv_r, ],   v$cd_r))
  Yc <- t(cell_means(v$prep$imputed$mixed[hv_e5[[nm]]$hv_p, ], v$cd_p))
  res <- cond_ae_q2_perm(Xc, Yc, k = 1, seeds = 1:3, epochs = 3000,
                        nperm = 29, perm_epochs = 600, seed_base = 5000)
  cbind(variety = nm, res)
}))
q2p_e5
```

**Reading the table.** Both varieties validate at k = 1: **Cadenza** Q² = 0.2082 (null 95th pctile -0.3059, **p = 0.033**), **Norin** Q² = 0.0889 (null 95th pctile -0.1753, **p = 0.033**). Neither result is contingent on the k = 2 number quoted above -- this is a fresh LOCO + permutation test at k = 1 specifically, and it independently confirms the same qualitative finding: the condition-aligned AE finds real cross-omics signal in Norin where PLS (D7, p = 0.06) does not.

```{r}
#| label: e5-boot-loadings
#| message: false

## Restricted to the variety/varieties that validated at k = 1 above -- see
## the reading immediately above this chunk for which that is.
e5_varieties <- q2p_e5$variety[q2p_e5$p_perm < 0.05]
cat("Varieties validated at k = 1:", paste(e5_varieties, collapse = ", "), "\n")

e5_boot <- setNames(lapply(e5_varieties, function(nm) {
  v <- VAR_E5[[nm]]
  cond_ae_boot_loadings(v$prep$qc_rna$vst, v$prep$imputed$mixed,
                        v$cd_r, v$cd_p, cells_e5,
                        hv_e5[[nm]]$hv_r, hv_e5[[nm]]$hv_p,
                        k = 1, B = 100, epochs = 1000, seed = 20260825)
}), e5_varieties)

for (nm in e5_varieties) {
  cat("\n---", nm, "loading/stability summary ---\n")
  print(summary(e5_boot[[nm]]$protein$loading))
  print(summary(e5_boot[[nm]]$protein$stability))
}
```

```{r}
#| label: e5-run-fgsea
#| message: false

e5_results <- setNames(lapply(e5_varieties, function(nm) {
  prot <- e5_boot[[nm]]$protein
  prot$gene_id <- strip_isoform(prot$feature)
  run_fgsea_on_stat(prot, "loading", e1_mapman_pathways[[nm]], score_type = "std")
}), e5_varieties)

for (nm in e5_varieties) {
  cat(nm, "-- MapMan bins tested:", nrow(e5_results[[nm]]), "| padj < 0.05:",
     sum(e5_results[[nm]]$padj < 0.05, na.rm = TRUE), "\n")
}
```

```{r e5-bubble, fig.width=14, fig.height=8, fig.align='center'}
#| label: e5-bubble

e5_plots <- Filter(Negate(is.null), lapply(e5_varieties, function(nm)
  plot_fgsea_bubble(e5_results[[nm]], top_n = 20,
                    plot_title = sprintf("%s: condition-aligned AE (k=1) loading",
                                        tools::toTitleCase(nm)))))
if (length(e5_plots) > 0) Reduce(`|`, e5_plots)
```

### Result -- a genuine negative finding: real aggregate signal, unstable per-gene loadings

Both varieties give the **same result, independently**: mean bootstrap-loading stability of 0.502 (Cadenza) and 0.503 (Norin) -- indistinguishable from a coin flip -- versus PLS's 0.92 for the same top-100-gene check in E4. Zero of the top-100 \|loading\| genes reach stability \>= 0.9 in either variety (E4's PLS component had 100/100). Consequently `fgsea` finds **0 significant MapMan bins in either variety** (0/180 Cadenza, 0/177 Norin) -- not because there is no shared signal (the Q²/permutation test above says there is), but because the per-gene loadings that would have to carry a gene-set signal do not reproduce across resamples.

**Why PLS's loadings are stable and this model's are not, despite both resting on the same validated resampling scheme.** PLS's NIPALS iteration provably converges to the dominant direction of the cross-block covariance, so resampled fits land on essentially the same direction (up to the sign `align_sign()` corrects for). `cond_ae_fit()`, even at k = 1 and even though the resulting *map* is linear, is trained by Adam gradient descent on a **non-convex joint objective** over four separately-initialised weight matrices (`Wr, Ar, Wp, Ap`) from a **different random initialisation on every bootstrap draw** (`seed = seed + b`). Nothing in that objective anchors which of the (infinitely many, loss-equivalent) `(Wr, Ap)` factorisations a given run converges to beyond a global sign, which `align_sign()` does correct -- the near-exactly-0.5 mean in *both* independent variety runs is the signature of individual genes' relative signs within the vector varying freely across runs, not merely a global sign ambiguity slipping past the alignment step. This was not chased further with longer bootstrap-draw training (1000 epochs here, versus the 3000 used to validate Q² above) because two independent varieties landing on almost the identical \~0.50 value is itself evidence of a structural property of this optimiser, not a convergence budget that a few more epochs would fix -- worth confirming if this model is revisited, but not assumed here.

**The honest bottom line, and the direct answer to "any plans to use the autoencoder model":** yes, and it was worth doing -- it is a real, validated way to detect *that* cross-omics signal exists (and, for Norin, the only method in this project that finds any), but **not**, at least with this training procedure, a source of individual-gene or gene-set-level information the way PLS's component is. That asymmetry -- a method can validate at the aggregate level and still fail the per-gene stability bar E4 introduced -- is itself the finding, and it is exactly the kind of check this project's own discipline (validate before using, at every level you intend to make a claim at) is built to catch.

```{r}
#| label: e5-outputs

write.table(q2p_e5, file.path(int_outdir, "cond_ae_k1_q2perm.tsv"),
           sep = "\t", row.names = FALSE, quote = FALSE)
for (nm in e5_varieties) {
  write.table(e5_boot[[nm]]$protein,
             file.path(int_outdir, sprintf("%s_real_loadings_protein_ae_k1_boot.tsv", nm)),
             sep = "\t", row.names = FALSE, quote = FALSE)
  save_fgsea_e4(e5_results[[nm]],
               here::here("results", "enrichment",
                         sprintf("E5_condAE_k1_mapman_%s.tsv", nm)))
}
```

------------------------------------------------------------------------

## 7. Not yet wired up here

Two items from the enrichment-extension catalogue are deliberately **not** implemented in this pass, because the inputs they need do not exist on disk yet -- rather than fake them, they are listed here as the next step.

-   **Kinetic archetype / amplitude enrichment** (rank by `log2(a)` or `t_half`, or ORA on archetype membership). `analysis/kinetics_limited/de_proteomics_wheat.qmd` computes archetype classification and per-gene amplitude `a_hat` for the real wheat data (482/431 amplified, 1278/978 concordant, etc. -- Cadenza/Norin), but classifies **inline in that notebook's own session** and never writes an `archetypes.tsv` for real data (unlike the simulated pipeline's `R/05_concordance_archetypes.R`, which does). Before A1/A4 from the catalogue can run here, that classification block needs factoring out into a function that writes `results/kinetics/archetypes_{norin,cadenza}.tsv`.
-   **Per-condition discordance heatmap** (8 separate rankings, one per treatment x timepoint cell, MapMan bins x cells). The saved `*_gene_distances_8condition.csv` files carry only the *aggregate* `d_full`/`correlation`/`amplitude` per gene, not the per-condition contribution to that distance -- that decomposition exists only inside `gene_distance_shared_space.qmd` §8b/V3c (in-session, not saved). Needs a `results/gene-distance/*_per_condition_contrib.csv` export from that notebook first.

------------------------------------------------------------------------

## 8. Outputs

```{r}
#| label: write-outputs

outdir <- here::here("results", "enrichment")
ensure_dir(outdir)

save_fgsea <- function(res, path) {
  df <- as.data.frame(res)
  df[] <- lapply(df, function(x) if (is.list(x)) sapply(x, paste, collapse = ";") else x)
  write.table(df, path, sep = "\t", row.names = FALSE, quote = FALSE)
}

save_fgsea(e1_norin,   file.path(outdir, "E1_discordance_mapman_norin.tsv"))
save_fgsea(e1_cadenza, file.path(outdir, "E1_discordance_mapman_cadenza.tsv"))
write.table(e3_compare, file.path(outdir, "E3_cross_variety_discordance.tsv"),
            sep = "\t", row.names = FALSE, quote = FALSE)

list.files(outdir)
```

------------------------------------------------------------------------

## 9. Summary and known issues (first successful render, 2026-08-25)

**Annotation coverage.** MapMan reaches 4,643 / 4,882 of the 5,879 / 6,105 detected proteins (Norin/Cadenza); GO/Pfam/InterPro reaches more (5,242 / 5,548) but is thinner per-term (see the caveats in the purpose section). 181 / 185 MapMan bins and 1,461 / 1,517 GO/Pfam/InterPro terms qualify for GSEA (size 5--500).

**Results.**

| Statistic                               | Norin: bins tested / padj\<0.05 | Cadenza: bins tested / padj\<0.05 |
|-----------------------------------------|---------------------------------|-----------------------------------|
| `d_full` (E1)                           | 168 / **13**                    | 174 / **12**                      |
| `correlation` (E2)                      | 168 / **2**                     | 174 / **5**                       |
| `log2_amplitude` (E2)                   | 64 / **0**\*\*                  | 79 / **0**\*\*                    |
| PLS component-1 loading (E4)            | not run -- no validated axis    | 180 / **\~19**                    |
| Condition-aligned AE (k=1) loading (E5) | 177 / **0**\*\*\*               | 180 / **0**\*\*\*                 |

\*\* Not a null result -- see known issue 1 below. \*\*\* Also not a null result of "no signal" -- the *aggregate* Q² validates in both varieties (Cadenza p=0.033, Norin p=0.033); the 0 reflects unstable per-gene loadings, not absence of shared signal. See known issue 5.

**Bottom line:** real, modest discordance signal (E1: \~7--8% of tested bins significant vs. 5% expected by chance) and a clean, if smaller, `correlation` signal (E2). **E4 remains the strongest hit rate (\~11%) and the most biologically legible** -- its top bin, `stress.biotic`, is exactly what a validated shared RNA/protein axis over an infection time-course should surface, and only 6 of its \~19 significant bins overlap E1's. **E5 (added 2026-08-25) is a genuine negative result worth keeping, not a bug**: the condition-aligned autoencoder is the one AE variant in `dimensionality_reduction_wheat.qmd` that already cleared both the Q²-validation and seed-reproducibility bars that disqualified AE-tanh/AE-vae, and it gives **Norin its first validated ML cross-omics signal anywhere in this project** (PLS found none, D7 p=0.06) -- but its per-gene loadings do not survive the same replicate-resampling stability check E4's PLS component passed easily (mean stability \~0.50, coin-flip, in both varieties independently, versus PLS's 0.92), so no gene-set claim is drawn from it. The `log2_amplitude` arm of E2 and the raw `delta_NES` ranking in E3 are still not trustworthy as reported -- both remain flagged inline with the specific fix, neither has been applied.

**Known issues, not fixed in this pass (flagged only, per instruction to hold off on re-rendering):**

1.  **`log2_amplitude` double-logs an already-log2 column** (§5) -- `amplitude` is signed and already log-scale; taking `log2()` of it produces `NaN` for \~60% of genes, which are silently dropped, leaving only \~40% of genes in that one GSEA and 0 significant bins in either variety. Fix: rank on `amplitude` directly, no second `log2()`.
2.  **`longfmt_pathways` (GO/Pfam/InterPro, §2.3) is built but never used** -- no `run_fgsea_on_stat()` call in E1/E2/E3/E4/E5 is passed these gene sets; only the MapMan sets are actually tested. The "secondary vocabulary" described in the purpose section is not yet wired into any analysis.
3.  **E3's top-20 table is not filtered on significance** (§6) -- it ranks by raw `|delta_NES|`, and most of the current top rows have `padj` near 1 in one or both varieties, so the table currently surfaces noise as often as signal. Fix: filter to `padj_norin < 0.05 | padj_cadenza < 0.05` before `slice_max()`.
4.  **E4's bin count is a range, not a fixed number** (§6b) -- `fgseaMultilevel`'s Monte Carlo p-value estimate moves one borderline bin across `padj = 0.05` between otherwise-identical runs (confirmed: a standalone run gave 20, the full in-notebook run gave 19, same top hits). Report as "\~19--20," not either number precisely, unless `nperm`/the multilevel algorithm's internal sampling is increased to tighten the borderline estimate.
5.  **E5's per-gene loadings are unstable, likely a property of the optimiser, not (yet fully ruled out as) a training-budget artefact** (§6c) -- bootstrap draws used 1000 epochs (versus 3000 for the Q²-validated point estimate); two independent varieties landing on the same \~0.50 stability is evidence of a structural issue (non-convex joint optimisation from random per-draw initialisation, unlike PLS's NIPALS convergence guarantee) rather than under-training, but a longer-epoch bootstrap run would confirm this rather than assume it, and has not been done.

**Deferred from the start** (§7, unchanged): kinetic archetype / amplitude enrichment (needs an `archetypes.tsv` this project does not yet write for real data) and the per-condition discordance heatmap (needs a per-condition distance export `gene_distance_shared_space.qmd` does not yet produce).
